---
title: "prompt/completion pairs chapter 3"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: ML/Stat | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a data frame of 100 rows and 6 columns with the following structure called 'data'. 

```{r}
    [,1]     [,2]     [,3]     [,4]     [,5]     [,6]
[1,] 295.0368 198.0631 306.7758 259.3416 265.7973 131.3946
[2,] 117.8228 256.6800 325.0575 111.3225 271.2894 207.0569
[3,] 307.3622 220.4294 162.1763 227.7250 202.3732 114.8394
```
  
How could I calculate the means and variances of the rows of the following data set, and plot the distributions of means
and variances using `hist()` and `boxplot()` functions?
How can I calculate the standard deviation of the distribution of the means using the `sd()` function?
How can I compare that to the expected standard error obtained from the central limit theorem keeping in mind the
population parameters were `σ = 70` and `n = 6`? How does the standard deviation of the means of the rows change when we
simulate more data with different number of samples and/or different number of observations?

completion:
Assuming that your data frame is consisted of numerical values we can calculate the means and variances of the rows
using `rowMeans()` and `rowVars()` and plot the distributions with `hist()` and `boxplot()`. 
To calculate the standard deviation of the distribution of the means, we will use the `sd()` function, and compare
that to the expected standard error calculated with the formula `SE = σ / √n`. 
Then we will test the changes in the estimated standard error in case of 6000 observations with the sample size of
6 (data1), 6000 observations with the sample size of 10 (data2), and 600 observations with the sample size of 10
(data3).

```{r,eval=FALSE}

# load the library matrixStats
require(matrixStats)

# Calculate the means of the rows
means <- rowMeans(data)
# Calculate the variances of the rows
vars <- rowVars(data)

# Plot the means and variances with histogram
hist(means)
hist(vars) 

# Plot the means and variances with boxplot
boxplot(means)
boxplot(vars)

# Calculate the standard deviation of the distribution of the means
samples <- sd(means) # 28.079

# Expected standard error from population parameters: SE = σ / √n
clt.se <- 70/sqrt(6) # 28.577


# Generate data with same mean and sd, same number of samples, but 10 times of observation.
data1 <- matrix(rnorm(6000,mean=200,sd=70),ncol=6)
# Calculate the standard deviation of the means of the rows
sd(rowMeans(data1)) # 27.813

# Generate data with same mean and sd, number of samples = 10, 10 times of observation.
data2 <- matrix(rnorm(6000,mean=200,sd=70),ncol=10)
# Calculate the standard deviation of the means of the rows
sd(rowMeans(data2)) # 21.358

# Generate data with same mean and sd, number of samples = 10, same number of observation.
data3 <- matrix(rnorm(600,mean=200,sd=70),ncol=10)
# Calculate the standard deviation of the means of the rows
sd(rowMeans(data3)) # 22.055

# We can see that the standard deviation of the distribution of the means are very close to the expected standard error.
# The estimate of the standard error does not change much when the number of observation increases,
# it will only decrease if we sample more data points per set (increase number of samples).

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stat | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a vector of 30 numerical values, which represent the observation for 30 variables, called 'pois1'.
I also have a vector with 1000 numerical values, called 'sample_means' which is the means of the observed values of these 30 variables.
I want to plot the sampling distributions of the means using a histogram and get the 2.5th and 97.5th percentiles
of the distribution. How can I complete these task?
I also want to calculate confidence intervals of the mean of "pois1" with `t.test()` function.
Then I want to use the bootstrap confidence interval for the mean on "pois1" and compare the interval with the
theoretical confidence interval of the mean from the `t.test()`.

completion:
To calculate the 2.5th and 97.5th percentiles of the distribution, we will first load the package `mosaic`
and use function `quantile()` for `sample_mean` with 2.5th and 97.5th percentages.
We will also plot the distribution using `hist()` on `sample_mean`.
We will use `t.test()` on `pois1` to find out the confidence interval.
Then we will use the bootstrap confidence interval for the mean on `pois1` and compare those values with the
theoretical values from the `t.test()`.


```{r,eval=FALSE} 

# Load the mosaic package
require(mosaic)

# Calculate the 2.5th and 97.5th percentiles of the distribution
quantile(sample_means,probs=c(0.025,0.975))

# Plot the histogram for sample_means
hist(sample_means)

# Do t-test on pois1 and calculate confidence intervals
t.test(pois1)

# Use the bootstrap confidence interval for the mean on `pois1`.
quantile((do(1000)*mean(sample(pois1,30,replace=T)))[,1],probs=c(0.025,0.975))
 

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stat | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 vectors called `gene1` and `gene2`.
Each vector has 30 numeric values, which represent gene expression values obtained under different conditions.
How can I test the difference of means of the `gene1` and `gene2` using the randomization, `t.test()` and
`wilcox.test()` functions? I also want to plot the distribution with randomization using histogram.

completion:
We will do the randomization with the following process. First, we will calculate the differences in the means and
do the randomization procedure to get a null distribution when we assume there is no difference between samples,H0.
We then calculate how often we would get the original difference under the assumption that H0 is true.
Then we can do the t-test and wilcox-test easily by using `t.test()` and `wilcox.test()` functions.
Lastly, we will plot the distribution of the means by randomization using histogram.


```{r,eval=FALSE} 

# Randomization-based testing
# Calculate the original differences in the means
org.diff <- mean(gene1)-mean(gene2)

# Make gene1 and gene2 into one data frame, with gene1 marked as test and gene2 marked as control group.
gene.df <- data.frame(exp=c(gene1,gene2),
                  group=c( rep("test",30),rep("control",30) ) )

# Do the randomization process to get a null distribution when we assume there is no difference between samples (H0).
exp.null <- do(1000) * diff(mosaic::mean(exp ~ shuffle(group), data=gene.df))

# Calculate how often we would get the original difference under the assumption that H0 is true
sum(exp.null>org.diff)/1000

# Do t-test on gene1 and gene2
t.test(gene1,gene2)

# Do wilcox-test on gene1 and gene2
wilcox.test(gene1,gene2)

# Plot the histogram with the null distribution
hist(exp.null[,1], xlab = "null distribution | No difference in samples",
     main = expression(paste(H[0], " :no difference in means")),
     xlim = c(-2,2), col = "lightpink", border = "white")

# Draw a line at the 95th percentile of null distribution.
abline(v=quantile(exp.null[,1],0.95),col="red" )
text(x=quantile(exp.null[,1],0.95),y=250,"0.05",adj=c(1,0),col="red")

# Draw a line at the point of original difference of the means
abline(v=org.diff,col="blue" )
text(x=org.diff,y=200,"org. diff.",adj=c(1,0),col="blue")

```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: ML/Stat | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 50 numerical values in a vector called 'x', which represents the histone modification score and is between 1 and 100. 
I also have a vector 'y' which has 50 numerical values and represent gene expression values.
How can I fit a line to predict Y based on X? I also want to plot the scatter plot and the fitted line.
How can I calculate the correlation and R^2 and run the `summary()` function and extract P-values from there?
I want to plot the residuals vs. the fitted values plot, how do I do that?

completion:
To fit a line to predict y, we can simply use `lm()` function. To create a scatterplot and add the fitted line,
we can first use `plot()` which plots the points, and add `abline()` which draws a line of a given linear model.
To calculate the correlation we will use `cor()` which gives pearson correlation, and R^2 can be easily calculated
by doing `cor()^2`. We can use `summary()` function on the model we created by `lm()` and extract the p values by
pointing to the column that gives out p-values.
To plot the residuals vs. the fitted values, we can call the `plot()` function with the model returned by `lm()` as
the first argument, and `which=1` as the second argument. 


```{r,eval=FALSE}  

# Use lm function to fit a line
mod <- lm(y~x)

# Plot the data points
plot(x,y)
# Draw the fitted line of the model
abline(mod)

# Calculate the correlation between x and y with pearson method (default)
cor(x,y)
# R squared
cor(x,y)^2 

# Use the summary function
lm_summary <- summary(mod)
lm_summary

# Extract the p-value by pointing at the second row of "Coefficients:", column of "Pr(>|t|)"
lm_summary$coefficients[2, "Pr(>|t|)"]

# Call the plot() function with the model as the first argument, and which=1 as the second argument
plot(mod,which=1)
 
```
--- TASK END

