---
title: "prompt/completion pairs chapter 3"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a gene expression values stored in a vector 'x'.
I want to calculate the mean and median of this vector. How can I accomplish this task?

completion: 
To calculate the mean and median of the vector 'x', you can use the following code.

```{r,eval=FALSE}

# Calculate mean
mean(x)

# Calculate median
median(x)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats \| visualization , numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset with 20 numerical values, called 'x', which are observations from a set of variables.
I would like to calculate the variance and standard deviation of this dataset.
How can I calculate these values?
I also want to calculate IQR(interquartile range), obtain quantiles, and plot the boxplot.

completion:
First, you can use `var()` and `sd()` to calculate the variance and standard deviation of the vector.
You can also calculate IQR by `IQR()` function.
To obtain quantiles of 0,25,50,75,100%, use `quantile()`, and plot the boxplot using `boxplot()`.

```{r,eval=FALSE}

# variance of the dataset
var(x)
# standard deviation of the dataset
sd(x)
# Calculate interquartile range (IQR)
IQR(x)
# Obtain quantiles
quantile(x)
# Create a box plot
boxplot(x, horizontal = T)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset with continuous numerical values, called 'data'.
I already calculated the mean and standard deviation of the 'data', which is 0 and 2, respectively.
I want to perform various calculations using a normal distribution with `mean=0` and `sd=2`.
Specifically, I would like to calculate the probability density function at `x=-2`.
I also want to find probabilities for `P(X =< -2)` and `P(X > -2)`.
Lastly, I want to determine the y value that corresponds to `P(X > y) = 0.15`.
How can I perform these calculations?

completion:
In R, there is a family of `*norm` functions `rnorm`,`dnorm`,`qnorm` and `pnorm`.
These can be used to operate with the normal distribution.
First the function `dnorm()` will be used to calculate the probablity density function.
Then the `pnorm()` function can calculate probability of `P(X =< -2)`. 
The probablity of `P(X > -2)` can be calculated by setting the `lower.tail = FALSE`.
The y value corresponding to `P(X > y) = 0.15` can be calculated using `qnorm()` function.

```{r,eval=FALSE}

# get the value of probability density function when X= -2, where mean=0 and sd=2
dnorm(-2, mean=0, sd=2)

# get the probability of P(X =< -2) where mean=0 and sd=2
pnorm(-2, mean=0, sd=2)

# get the probability of P(X > -2) where mean=0 and sd=2
pnorm(-2, mean=0, sd=2,lower.tail = FALSE)

# get y value corresponding to P(X > y) = 0.15 with  mean=0 and sd=2
qnorm(0.15, mean=0 , sd=2)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats \| visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I took 50 samples randomly from a population, which is called 'sample1'.
Since I do not have access to the population, I would like to build confidence intervals using bootstrap resampling
to estimate the precision of population parameter. How do I do bootstrap resampling with 1000 resamples?
How can I visualize the distribution of sample means using histogram and show 95% confidence interval?
Additionally, I would like to use the central limit theorem (CLT) to calculate 95% confidence interval and compare
it to the confidence interval calculated with bootstrap. How can I do that?

completion:
We will do bootstrap resampling for the 'sample1' to create 1000 resamples by randomly selecting values with
replacement from the initial sample. These resamples will allow us to construct a distribution of sample means.
Visualizations will play a crucial role in depicting this distribution; a histogram will provide a clear view,
and confidence intervals will be visually represented by red vertical lines at the 2.5th and 97.5th percentiles.
Additionally, the Central Limit Theorem will be used to calculate a confidence interval for the mean of the
original dataset. This interval's validity will be compared against the bootstrap-based interval.

```{r,eval=FALSE}
library(mosaic)

# do bootstrap resampling, sampling with replacement
boot.means <- do(1000) * mean(resample(sample1))

# get percentiles from the bootstrap means
q <- quantile(boot.means[,1],p=c(0.025,0.975))

# plot the histogram
hist(boot.means[,1],col="cornflowerblue",border="white",
                    xlab="sample means")
abline(v=c(q[1], q[2] ),col="red")
text(x=q[1],y=200,round(q[1],3),adj=c(1,0))
text(x=q[2],y=200,round(q[2],3),adj=c(0,0))

# Calculating confidence intervals using CLT (Central Limit Theorem)
alpha <- 0.05
sd <- 5
n <- 50
mean(sample1)+qnorm(c(alpha/2,1-alpha/2))*sd/sqrt(n)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats \| visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 vectors called `gene1` and `gene2`.
Each vector has 30 numeric values, which represent gene expression values obtained under different conditions.
How can I test the difference of means of the `gene1` and `gene2` using the randomization?
How can I plot the distribution using histogram?
I also want to do a Welch's t-test and t-test with equal variance assumption. How can I do that?

completion:
We will do the randomization with the following process.
First, we will calculate the differences in the means and do the randomization procedure to get a null distribution
when we assume there is no difference between samples, H0. We then calculate how often we would get the original
difference we calculated under the assumption that H0 is true.
Then, we will plot the null distribution for differences of means obtained via randomization using histograms.
The original difference will be marked via the blue line, and the value that corresponds to P-value of 0.05 will be
marked via the red line.
Then we can do the Welch's t-test by simply using `stats::t.test()` and t-test with equal variance assumption by
indicating `var.equal = TRUE` in the `t.test()`.

```{r,eval=FALSE}

# Randomization-based testing

# Calculate the original differences in the means
org.diff <- mean(gene1)-mean(gene2)

# Make gene1 and gene2 into one data frame, with gene1 marked as test and gene2 marked as control group.
gene.df <- data.frame(exp=c(gene1,gene2),
                  group=c( rep("test",30),rep("control",30) ) )

# Do the randomization process to get a null distribution when we assume there is no difference between samples (H0).
exp.null <- do(1000) * diff(mosaic::mean(exp ~ shuffle(group), data=gene.df))

# Plot the histogram with the null distribution
hist(exp.null[,1],xlab="null distribution | no difference in samples",
     main=expression(paste(H[0]," :no difference in means") ),
     xlim=c(-2,2),col="cornflowerblue",border="white")

# Draw a line at the 95th percentile of null distribution.
abline(v=quantile(exp.null[,1],0.95),col="red" )
abline(v=org.diff,col="blue" )

# Draw a line at the point of original difference of the means
text(x=quantile(exp.null[,1],0.95),y=200,"0.05",adj=c(1,0),col="red")
text(x=org.diff,y=200,"org. diff.",adj=c(1,0),col="blue")

# Calculation of p-value
p.val <- sum(exp.null[,1]>org.diff)/length(exp.null[,1])
p.val

# Welch's t-test
stats::t.test(gene1,gene2)

# t-test with equal variance assumption
stats::t.test(gene1,gene2,var.equal=TRUE)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats \| visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a data from the cancer gene expression study comparing two mutation types of genes, called 'hedenfalk'.
I would like to perform a multiple test correction on the p-values of this data. How can I do that?
How can I perform a Bonferroni correction, FDR, and q-value approach?
I would like to make a plot to compare the three methods. How can I do that?

completion:
The `qvalue` package from Bioconductor will be used.
In the data set, p-value will be specified with `hedenfalk$p`, since we will only use the p-value for our task.
First the three methods will be implemented. The q-value method will be implemented by `qvalue()` function,
the Bonferroni correction will be implemented by `p.adjust()` with `method = "bonferroni"` and the FDR approach
will be implemented by `p.adjust()` with `method = "fdr"`.
Lastly, the p-values adjusted by the three methods will be plotted to show the difference.

```{r,eval=FALSE}

# For q-value adjustment, we need library 'qvalue'
library(qvalue)

# q-value adjustment method
qvalues <- qvalue(hedenfalk$p)$q

# Bonferroni adjustment method
bonf.pval <- p.adjust(hedenfalk$p,method ="bonferroni")

# FDR-adjustment method
fdr.adj.pval <- p.adjust(hedenfalk$p,method ="fdr")

# Plot the adjusted p-values. First plot p-values adjusted with q-value adjustment.
plot(hedenfalk$p,qvalues,pch=19,ylim=c(0,1),
     xlab="raw P-values",ylab="adjusted P-values")

# Then add points for FDR and Bonferroni, distinguished with colors.
points(hedenfalk$p,bonf.pval,pch=19,col="red")
points(hedenfalk$p,fdr.adj.pval,pch=19,col="blue")

# Write the legend for the three adjustments.
legend("bottomright",legend=c("q-value","FDR (BH)","Bonferroni"),
       fill=c("black","blue","red"))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a gene expression data frame that has 6 columns, called 'data'. The first 3 columns are the test samples,
and the last 3 are the control samples. Each row is a gene, and there are 500 rows total. I would like to calculate
moderated and unmoderated t-test statistics and plot the p-value distributions for tests using histograms.
How can I complete these tasks? Also, how can I calculate the number of significant tests with 0.05 threshold?

completion:
Assuming 'data' data frame has numerical values for gene expression, we will calculate moderated and unmoderated
t-test statistics. We will first estimate t-statistic without moderation by calculating the difference in the means
(calculated row by row) and the standard error for the difference in means of two groups. We will also calculate
the p-value of rejecting null.
Then we will estimate t-statistic with moderation. That can be done by using same method for t-statistic without
moderation - we just have to modify the standard error by shrinking it towards median.
Lastly we will then make 2 histograms, one for unmoderated p and one for moderated p.
The number of significant tests with the 0.05 threshold will be calculated and written on the histograms.

```{r,eval=FALSE}

# set test/control groups
group1 <- 1:3
group2 <- 4:6

# Set n1 and n2, which is the number of the test / control samples.
n1 <- 3
n2 <- 3

# Calculate the mean of each row (test group and control group) and take the difference
dx <- rowMeans(data[,group1])-rowMeans(data[,group2])

# load package 'matrixStats'
require(matrixStats)

# get the esimate of pooled variance 
stderr <- sqrt( (rowVars(data[,group1])*(n1-1) + 
       rowVars(data[,group2])*(n2-1)) / (n1+n2-2) * ( 1/n1 + 1/n2 ))

# do the shrinking towards median
mod.stderr <- (stderr + median(stderr)) / 2 # moderation in variation

# esimate t statistic with moderated variance
t.mod <- dx / mod.stderr

# calculate P-value of rejecting null 
p.mod <- 2*pt( -abs(t.mod), n1+n2-2 )

# esimate t statistic without moderated variance
t <- dx / stderr

# calculate P-value of rejecting null 
p <- 2*pt( -abs(t), n1+n2-2 )

# This command to present 2 histograms horizontally
par(mfrow=c(1,2))

# First make a histogram for unmoderated p values.
# You can change the settings such as colors, title, x-label, etc.
hist(p,col="cornflowerblue",border="white",main="",xlab="P-values t-test")

# Write the number of the significant tests with the 0.05 threshold
mtext(paste("signifcant tests:",sum(p<0.05))  )

# Do the same for the moderated p values.
hist(p.mod,col="cornflowerblue",border="white",main="",
     xlab="P-values mod. t-test")
mtext(paste("signifcant tests:",sum(p.mod<0.05))  )

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats \| visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 50 numerical values in a vector called 'x', which represents the histone modification score and is between 1 and 100. 
I also have a vector 'y' which has 50 numerical values and represent gene expression values.
How can I fit a line to predict Y based on X? I also want to plot the scatter plot and the fitted line.
Using the summary function I would like to test the coefficients for the null hypothesis.
I also want to get the confidence intervals and pull out the estimated coefficients from the model.
How can I accomplish these tasks?

completion:
To fit a line to predict y, we can simply use `lm()` function. To create a scatterplot and add the fitted line, we 
can first use `plot()` which plots the points, and add `abline()` which draws a line of a given linear model.
We can use `summary()` function on the model we created by `lm()` and it will show the results of tested
coefficients along with other information. Lastly, we can use the `confint()` function to get the confidence
intervals on the coefficients and the `coef()` function to pull out the estimated coefficients from the model.

```{r,eval=FALSE}

# Use lm function to fit a line
mod1 <- lm(y~x)

# plot the data points
plot(x,y,pch=20,
     ylab="Gene Expression",xlab="Histone modification score")
# plot the linear fit
abline(mod1,col="blue")

# summary function to test all the coefficients for the null hypothesis βi=0.
summary(mod1)

# Get confidence intervals on the coefficients
confint(mod1)

# Pull out the estimated coefficients from the model
coef(mod1)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats \| visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 vectors called `gene1` and `gene2`. Each vector has 30 numeric values, which represent gene expression
values obtained under different conditions. I would like to encode the two differential gene expression as
categorical variables in 0 and 1. How can I make a linear regression model with the data so that the two groups
are encoded in 0 and 1? What are the estimated coefficients? How can I plot the model?
I would also like to try comparing three levels with 'A','B','C' and see how the estimated coefficients look like.

completion:
In differential gene expression, we usually test the difference between different conditions, which can be encoded
as categorical variables in a linear regression. We will first make the model with categorical variables including
two levels that will be encoded in 0 and 1. Then, we will run the `summary()` function.
To visualize the model, we will simply load the library `mosaic()` and use `plotModel()` function to plot.
Lastly, we can try making model with categorical variables including three levels.
However, since there are only two groups (gene1 and gene2), we will make gene1 as 'A', gene2 as 'B' and 'C'.
We will then run the `summary()` function.
Since groupB and C had same data set of `gene2`, we can see that the coefficients for the two group are identical.

```{r,eval=FALSE}

# Make gene1 and gene2 into one data frame, with gene1 encoded as 1 and gene2 encoded as 0.
gene.df <- data.frame(exp=c(gene1,gene2),
                  group=c( rep(1,30),rep(0,30) ) )

# Linear regression model with gene.df data, which shows the relationship between exp and group
mod2 <- lm(exp~group,data=gene.df)
# Run the summary function
summary(mod2)

# Using mosaic library, plot the model
require(mosaic)
plotModel(mod2)

# Data table for three categorial variables
gene.df <- data.frame(exp=c(gene1,gene2,gene2),
                  group=c( rep("A",30),rep("B",30),rep("C",30) ) 
                  )
# linear regression model with three categorial variables
mod3 <- lm(exp~group,data=gene.df)
# Run the summary function
summary(mod3)

```

--- TASK END
