---
title: "prompt/completion pairs chapter 3"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a gene expression data set named "geneExpMat.txt". The data has 100 differentially expressed genes.
The first 3 columns are the test samples, and the last 3 are the control samples.
How can I do a t-test for each gene (each row is a gene), and record the p-values?
Also, how can I do a moderated t-test and record the p-values?
Using the p-values, I want to make a p-value histogram and compare two approaches in terms of the number of
significant tests with the 0.05 threshold. I also want to use FDR (BH), Bonferroni and q-value adjustment methods
on the unmoderated p-values and plot them to compare.
Calculate how many adjusted p-values are below 0.05 for each approach.

Completion:
The table must be read in first.
We can use `data.table::fread()` function for this purpose as it can read files faster.
Since the test and control samples are divided by the column number, we can compare the column 1-3 and column 4-6.
We will first estimate t-statistic without moderation by calculating the difference in the means
(calculated row by row) and the standard error for the difference in means of two groups.
We will also calculate the p-value of rejecting null.
Then we will estimate t-statistic with moderation. That can be done by using same method for t-statistic without
moderation - we just have to modify the standard error by shrinking it towards median.
We will then make 2 histograms, one for unmoderated p and one for moderated p.
The number of significant tests with the 0.05 threshold will be calculated and written on the histograms.
Then we will use FDR (BH), Bonferroni and q-value adjustment methods on the p-values and calculate how many
adjusted p-values are below 0.05 for each.


```{r,eval=FALSE}  

library(data.table)
# read gene expression data set
geneExpMat <- data.table::fread("geneExpMat.txt", header = TRUE, 
                              data.table=FALSE)

# Later the `rowVars` function requires matrix, so change the data frame into matrix.
geneExpMat <- as.matrix(geneExpMat)

# load package 'matrixStats'
library(matrixStats)

# Set n1 and n2, which is the number of the test / control samples.
n1 <- 3
n2 <- 3

# Calculate the mean of each row (test group and control group) and take the difference
dx <- rowMeans(geneExpMat[,1:3]) - rowMeans(geneExpMat[,4:6])

# Get the estimate of pooled variance
stderr <- sqrt(((n1-1)*rowVars(geneExpMat[,1:3])+(n2-1)*rowVars(geneExpMat[,4:6])) / (n1+n2-2)*(1/n1 + 1/n2) )

# Estimate t-statistic without moderated variance
t <- dx / stderr

# Calculate P-value of rejecting null
p <- 2*pt (-abs(t), n1+n2-2)

# Do the shrinking towards median for moderation in variance
mod.stderr <- (stderr + median(stderr)) / 2

# Estimate t-statistic with moderated variance and calculate P-value of rejecting null 
t.mod <- dx / mod.stderr
p.mod <- 2*pt (-abs(t.mod), n1+n2-2)

# This command to present 2 histograms horizontally
par(mfrow=c(1,2))

# First make a histogram for unmoderated p values.
# You can change the settings such as colors, title, x-label, etc.
hist(p,col="cornflowerblue",border="white",main="",xlab="P-values t-test")

# Write the number of the significant tests with the 0.05 threshold
mtext(paste("signifcant tests:",sum(p<0.05))  )

# Do the same for the moderated p values.
hist(p.mod,col="cornflowerblue",border="white",main="",
     xlab="P-values mod. t-test")
mtext(paste("signifcant tests:",sum(p.mod<0.05))  )

# Use FDR (BH), Bonferroni and q-value adjustment methods on the p-values

# For q-value adjustment, we need library 'qvalue'
library(qvalue)

# q-value adjustment method
qvalues <- qvalue(p)$qvalues

# FDR-adjustment method
fdr.p <- p.adjust(p, method = "fdr")

# Bonferroni adjustment method
bonf.p <- p.adjust(p, method = "bonferroni")

# Plot the adjusted p-values. First plot p-values adjusted with q-value adjustment.
plot(p,qvalues,pch=19,ylim=c(0,1),xlab="raw P-values",ylab="adjusted P-value")

# Then add points for FDR and Bonferroni, distinguished with colors.
points(p,bonf.p,pch=19,col="red")
points(p,fdr.p,pch=19,col="blue")

# Write the legend for the three adjustments.
legend("bottomright",legend=c("q-value","FDR (BH)","Bonferroni"), fill=c("black","blue","red"))

# Calculate how many adjusted p-values are below 0.05 for each approach
sum(qvalues<0.05)
sum(fdr.p<0.05)
sum(bonf.p<0.05)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization , numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a gene expression data set.
The file name is "geneExpMat.txt". The data has 100 differentially expressed genes.
The first 3 columns are the test samples, and the last 3 are the control samples.
How can I do a t-test for each gene (each row is a gene), and record the p-values?
I also want to use FDR (BH), Bonferroni and q-value adjustment methods on the p-values and plot them to compare.
Calculate how many adjusted p-values are below 0.05 for each approach.

completion:
The tables must be read in first. We can use `data.table::fread()` function for this purpose as it can read files faster. 
Since the test and control samples are divided by the column number, we can compare the column 1-3 and column 4-6.
We will first estimate t-statistic without moderation by calculating the difference in the means
(calculated row by row) and the standard error for the difference in means of two groups.
We will also calculate the p-value of rejecting null.
Then we will use FDR (BH), Bonferroni and q-value adjustment methods on the p-values and calculate how many
adjusted p-values are below 0.05 for each.


```{r,eval=FALSE}  

library(data.table)
# read gene expression data set
geneExpMat <- data.table::fread("geneExpMat.txt", header = TRUE, 
                              data.table=FALSE)

# Later the `rowVars` function requires matrix, so change the data frame into matrix.
geneExpMat <- as.matrix(geneExpMat)

# load package 'matrixStats'
library(matrixStats)

# Set n1 and n2, which is the number of the test / control samples.
n1 <- 3
n2 <- 3

# Calculate the mean of each row (test group and control group) and take the difference
dx <- rowMeans(geneExpMat[,1:3]) - rowMeans(geneExpMat[,4:6])

# Get the estimate of pooled variance
stderr <- sqrt(((n1-1)*rowVars(geneExpMat[,1:3])+(n2-1)*rowVars(geneExpMat[,4:6])) / (n1+n2-2)*(1/n1 + 1/n2) )

# Estimate t-statistic without moderated variance
t <- dx / stderr

# Calculate P-value of rejecting null
p <- 2*pt (-abs(t), n1+n2-2)

# Use FDR (BH), Bonferroni and q-value adjustment methods on the p-values

# For q-value adjustment, we need library 'qvalue'
library(qvalue)

# q-value adjustment method
qvalues <- qvalue(p)$qvalues

# FDR-adjustment method
fdr.p <- p.adjust(p, method = "fdr")

# Bonferroni adjustment method
bonf.p <- p.adjust(p, method = "bonferroni")

# Plot the adjusted p-values. First plot p-values adjusted with q-value adjustment.
plot(p,qvalues,pch=19,ylim=c(0,1),xlab="raw P-values",ylab="adjusted P-value")

# Then add points for FDR and Bonferroni, distinguished with colors.
points(p,bonf.p,pch=19,col="red")
points(p,fdr.p,pch=19,col="blue")

# Write the legend for the three adjustments.
legend("bottomright",legend=c("q-value","FDR (BH)","Bonferroni"), fill=c("black","blue","red"))

# Calculate how many adjusted p-values are below 0.05 for each approach
sum(qvalues<0.05)
sum(fdr.p<0.05)
sum(bonf.p<0.05)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a gene expression data set. The file name is "geneExpMat.txt". The data has 100 differentially expressed genes.
The first 3 columns are the test samples, and the last 3 are the control samples.
How can I do a t-test for each gene (each row is a gene), and record the p-values?
Also, how can I do a moderated t-test and record the p-values?
Using the p-values, I want to make a p-value histogram and compare two approaches in terms of the number of
significant tests with the 0.05 threshold.

completion:
The tables must be read in first. We can use `data.table::fread()` function for this purpose as it can read files faster. 
Since the test and control samples are divided by the column number, we can compare the column 1-3 and column 4-6.
We will first estimate t-statistic without moderation by calculating the difference in the means (calculated row by row)
and the standard error for the difference in means of two groups. We will also calculate the p-value of rejecting null.
Then we will estimate t-statistic with moderation. That can be done by using same method for t-statistic without 
moderation - we just have to modify the standard error by shrinking it towards median.
We will then make 2 histograms, one for unmoderated p and one for moderated p.
The number of significant tests with the 0.05 threshold will be calculated and written on the histograms.


```{r,eval=FALSE}  

library(data.table)
# read gene expression data set
geneExpMat <- data.table::fread("geneExpMat.txt", header = TRUE, 
                              data.table=FALSE)

# Later the `rowVars` function requires matrix, so change the data frame into matrix.
geneExpMat <- as.matrix(geneExpMat)

# load package 'matrixStats'
library(matrixStats)

# Set n1 and n2, which is the number of the test / control samples.
n1 <- 3
n2 <- 3

# Calculate the mean of each row (test group and control group) and take the difference
dx <- rowMeans(geneExpMat[,1:3]) - rowMeans(geneExpMat[,4:6])

# Get the estimate of pooled variance
stderr <- sqrt(((n1-1)*rowVars(geneExpMat[,1:3])+(n2-1)*rowVars(geneExpMat[,4:6])) / (n1+n2-2)*(1/n1 + 1/n2) )

# Estimate t-statistic without moderated variance
t <- dx / stderr

# Calculate P-value of rejecting null
p <- 2*pt (-abs(t), n1+n2-2)

# Do the shrinking towards median for moderation in variance
mod.stderr <- (stderr + median(stderr)) / 2

# Estimate t-statistic with moderated variance and calculate P-value of rejecting null 
t.mod <- dx / mod.stderr
p.mod <- 2*pt (-abs(t.mod), n1+n2-2)

# This command to present 2 histograms horizontally
par(mfrow=c(1,2))

# First make a histogram for unmoderated p values.
# You can change the settings such as colors, title, x-label, etc.
hist(p,col="cornflowerblue",border="white",main="",xlab="P-values t-test")

# Write the number of the significant tests with the 0.05 threshold
mtext(paste("signifcant tests:",sum(p<0.05))  )

# Do the same for the moderated p values.
hist(p.mod,col="cornflowerblue",border="white",main="",
     xlab="P-values mod. t-test")
mtext(paste("signifcant tests:",sum(p.mod<0.05))  )

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a histone modification data set. The file name is "HistoneModeVSgeneExp.txt".
There are 3 columns in the dataset. These are measured levels of H3K4me3, H3K27me3 and gene expression per gene 
(named 'measured_log2'). How can I plot the scatterplot for H3K4me3 vs.expression?
How can I fit the model for prediction of expression data using 1) only H3K4me3 as explanatory variable
2) only H3K27me3 as explanatory variable 3) both H3K4me3 and H3K27me3 as explanatory variable? 
How would I know which terms are significant?
How can I plot H3k4me3 vs. H3k27me3 and inspect the points that do not follow a linear trend? 

completion:
The tables must be read in first.
We can use `data.table::fread()` function for this purpose as it can read files faster.
Then we can plot H3K27me3 vs.expression using `plot()`.
Then we can create linear models for each cases with `lm()` (using H3k4me3, H3k27me3, H3k4me3+H3k27me3). 
After that we can inspect the significance of three models using `summary()` function.
To inspect the points that do not follow a linear trend, we can use the `smoothScatter()` function, which produces
a smoothed color density representation of a scatterplot.

```{r,eval=FALSE}  
library(data.table)

# read methylation age data
hmodFile <- data.table::fread("HistoneModeVSgeneExp.txt", header = TRUE, 
                              data.table=FALSE)

# Plot H3K27me3 vs.expression
plot(hmodFile$H3k4me3,hmodFile$measured_log2)

# Create three linear models for each case.
mod1<-lm(measured_log2~H3k4me3,data=hmodFile)
mod2<-lm(measured_log2~H3k27me3,data=hmodFile)
mod3<-lm(measured_log2~H3k4me3+H3k27me3,data=hmodFile)

# Use summary function for each model
summary(mod1)
summary(mod2)
summary(mod3)

# smoothScatter function
smoothScatter(hmodFile$H3k27me3,hmodFile$H3k4me3)
```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a histone modification data set. The file name is "HistoneModeVSgeneExp.txt". There are 3 columns in the dataset.
These are measured levels of H3K4me3, H3K27me3 and gene expression per gene(named 'measured_log2').
How can I plot the scatterplot for H3K4me3 vs.expression?

completion:
The tables must be read in first. We can use `data.table::fread()` function for this purpose as it can read files faster.
Then we can plot H3K27me3 vs.expression using `plot()`.

```{r,eval=FALSE}  
library(data.table)

# read methylation age data
hmodFile <- data.table::fread("HistoneModeVSgeneExp.txt", header = TRUE, 
                              data.table=FALSE)

# Plot H3K27me3 vs.expression
plot(hmodFile$H3k4me3,hmodFile$measured_log2)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats , numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a histone modification data set. The file name is "HistoneModeVSgeneExp.txt". There are 3 columns in the dataset.
These are measured levels of H3K4me3, H3K27me3 and gene expression per gene(named 'measured_log2'). 
How can I fit the model for prediction of expression data using 1) only H3K4me3 as explanatory variable 2) only H3K27me3
as explanatory variable 3) both H3K4me3 and H3K27me3 as explanatory variable? How would I know which terms are
significant?

completion:
The tables must be read in first. We can use `data.table::fread()` function for this purpose as it can read files faster.
Then we can plot H3K27me3 vs.expression using `plot()`.
Then we can create linear models for each cases with `lm()` (using H3k4me3, H3k27me3, H3k4me3+H3k27me3).
After that we can inspect the significance of three models using `summary()` function.
To inspect the points that do not follow a linear trend, we can use the `smoothScatter()` function, which produces a
smoothed color density representation of a scatterplot.

```{r,eval=FALSE}  
library(data.table)

# read methylation age data
hmodFile <- data.table::fread("HistoneModeVSgeneExp.txt", header = TRUE, 
                              data.table=FALSE)

# Create three linear models for each case.
mod1<-lm(measured_log2~H3k4me3,data=hmodFile)
mod2<-lm(measured_log2~H3k27me3,data=hmodFile)
mod3<-lm(measured_log2~H3k4me3+H3k27me3,data=hmodFile)

# Use summary function for each model
summary(mod1)
summary(mod2)
summary(mod3)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a histone modification data set. The file name is "HistoneModeVSgeneExp.txt". There are 3 columns in the dataset.
These are measured levels of H3K4me3, H3K27me3 and gene expression per gene(named 'measured_log2'). 
How can I plot H3k4me3 vs. H3k27me3 and inspect the points that do not follow a linear trend? 

completion:
The tables must be read in first.
We can use `data.table::fread()` function for this purpose as it can read files faster.
To inspect the points that do not follow a linear trend, we can use the `smoothScatter()` function, which produces
a smoothed color density representation of a scatterplot.

```{r,eval=FALSE}  
library(data.table)

# read methylation age data
hmodFile <- data.table::fread("HistoneModeVSgeneExp.txt", header = TRUE, 
                              data.table=FALSE)

# smoothScatter function
smoothScatter(hmodFile$H3k27me3,hmodFile$H3k4me3)
```
--- TASK END