---
title: "prompt/completion pairs chapter 4"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set with gene expression values from 60 bone marrow samples of patients with one of the four main
types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls, trimmed down to the top 1000 most variable genes.
The file is called "leukemiaExpressionSubset.txt". The matrix that contains gene expressions has the genes in the
rows and the patients in the columns. How can I use this data set to cluster the patients and display the values as a heatmap? 

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. We will now use this data set to cluster the patients and display the values as a heatmap and a dendrogram.
The heatmap shows the expression values of genes across patients in a color coded manner. The heatmap function,
`pheatmap()`, that we will use performs the clustering as well. The matrix that contains gene expressions has the genes in
the rows and the patients in the columns. Therefore, we will also use a column-side color code to mark the patients based
on their leukemia type. For this, we can set the annotation for each sample by taking the first three letters of the
column names as substring. For the hierarchical clustering, we will use Ward's method designated by the
`clustering_method` argument to the `pheatmap()` function. 

```{r,eval=FALSE}  

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Load needed library
library(pheatmap)

# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(mat),1,3))
# Set the row names of our annotation_col as column names of the matrix
rownames(annotation_col) <- colnames(mat)

# Use pheatmap() to create heatmap. Use "ward.D2" as clustering method as asked.
pheatmap(mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale="none", clustering_method = "ward.D2",
         clustering_distance_cols="euclidean",
         )

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set with gene expression values from 60 bone marrow samples of patients with one of the four main
types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls, trimmed down to the top 1000 most variable genes.
The file is called "leukemiaExpressionSubset.txt". The matrix that contains gene expressions has the genes in the
rows and the patients in the columns. How can I see the relationship of the clusters using hierarchical clustering
with the "complete linkage" and plotting the dendrogram? I would also like to visualize the cluster I will get if 
I cut the tree at `height = 80`. What is the number of samples for each cluster when I cut the tree at `height = 80`? 
How many samples are there in each cluster when I cut so that there are 5 clusters?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then, we will do the hierarchical clustering using `hclust()` with Euclidian distance and the default "complete
linkage". We want to calculate the distance between columns, so we transposed the matrix. We will also plot the cluster
dendrogram and add red rectangle to the plot at the `height = 80`. This visualizes the clusters we would get if we cut the 
tree at that height. After we will cut the tree using `cutree()` function at the height 80, and show the number of samples
for each cluster using `table()`. We will also cut the tree so that there are 5 clusters by specifying `k = 5` in
the `cutree()` function and show the number of samples for each cluster.


```{r,eval=FALSE} 

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Hierarchical clustering using hclust() with Euclidian distance. We want to calculate the distance between columns, so we transposed the matrix. 
hcl <- hclust(dist(t(mat)))

# Plot the cluster dendrogram with no labels for the leaves. hang = -1 to control how far the labels are from the branches.
plot(hcl,labels = FALSE, hang= -1)

# Add red rectangle to the plot at the height = 80
rect.hclust(hcl, h = 80, border = "red")

# cut tree so that there are 5 clusters
clu.k5 <- cutree(hcl,k=5) 

# cut tree/dendrogram from height 80
clu.h80 <- cutree(hcl,h=80) 

# number of samples for each cluster when we cut the tree so that there are 5 clusters
table(clu.k5) 

# number of samples for each cluster when we cut the tree from height 80
table(clu.h80)


```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set with gene expression values from 60 bone marrow samples of patients with one of the four main types of
leukemia (ALL, AML, CLL, CML) or no-leukemia controls, trimmed down to the top 1000 most variable genes.
The file is called "leukemiaExpressionSubset.txt". The matrix that contains gene expressions has the genes in the rows
and the patients in the columns. How do I use the k-means clustering method on this gene expression data set and check
the percentage of each leukemia type in each cluster? Also, how can I use the k-medoids clustering method on the gene
expression set and check the percentage of each leukemia type in each cluster? I also want to visualize the clustering
from partitioning using "multi-dimensional scaling" and plot the patients in a 2D plot color coded by their cluster assignments.

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. After setting a seed for reproducibility, we will run the K-means clustering algorithm using the function
`kmeans()`. Since our original matrix has the genes in the rows and the patients in the columns, we have to transpose the
matrix so that we calculate distances between patients. We will initialize the algorithm with `centers = 5`. We can check
the percentage of each leukemia type in each cluster after doing the K-means clustering.
For that we will make a data frame with rows as the Leukemia type and columns as our clusters to see if each of the 5
clusters predominantly represents one of the 4 leukemia types or the control.
Then, we can use the k-medoids clustering method by using the k-medoids clustering function `pam()` from the `cluster`
package. We can check the percentage of each leukemia type in each cluster just as we did it for the K-means clustering.
For visualization of the clustering from partitioning methods we can compress the between-patient-distances to a
2-dimensional plot.  For this we will first calculate euclidean distances between patients with `dist()` function,
and calculate the "multi-dimensional scaling" with `cmdscale()` function. Then, we will plot the patients in the 2D space
with `plot()` and add legend for cluster colors.

```{r,eval=FALSE}  

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# K-means clustering
# Setting seed for reproducibility
set.seed(101)
# we have to transpose the matrix with t() so that we calculate distances between patients
kclu <- kmeans(t(mat),centers=5)  

# number of data points in each cluster
table(kclu$cluster)

# Check the percentage of each leukemia type in each cluster
type2kclu <- data.frame(
                    LeukemiaType =substr(colnames(mat),1,3),
                    cluster=kclu$cluster)
# Visualize it as a table
table(type2kclu)


# K-medoids Clustering
# Use the k-medoids clustering function `pam()` from the `cluster` package.
kmclu <- cluster::pam(t(mat),k=5) 

# make a data frame with Leukemia type and cluster id
type2kmclu <- data.frame(
                    LeukemiaType =substr(colnames(mat),1,3),
                    cluster=kmclu$cluster)

table(type2kmclu)

# Visualization of the clustering from partitioning methods
# Calculate distances
dists <- dist(t(mat))

# calculate MDS
mds <- cmdscale(dists)

# plot the patients in the 2D space
plot(mds,pch=19,col=rainbow(5)[kclu$cluster])

# set the legend for cluster colors
legend("bottomright",
       legend=paste("clu",unique(kclu$cluster)),
       fill=rainbow(5)[unique(kclu$cluster)],
       border=NA,box.col=NA)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set with gene expression values from 60 bone marrow samples of patients with one of the four main types of
leukemia (ALL, AML, CLL, CML) or no-leukemia controls, trimmed down to the top 1000 most variable genes.
The file is called "leukemiaExpressionSubset.txt". The matrix that contains gene expressions has the genes in the rows
and the patients in the columns. How can I determine the optimal number of clusters using silhouette values for k-medoids?
I want to see the silhouette value in plot for `k = 5`. Then I want to calculate the average silhouette value across k
values between 2 and 7 and plot them. I also want to calculate the gap statistic for my patient gene expression dataset
with k-medoids and plot them across k values 1 to 8. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will first calculate silhouette values for k-medoids clustering with the `pam()` function with `k = 5` and
plot the resulting plot. With this we can see the silhouette values for each cluster at `k = 5`. We will then calculate the
average silhouette value for different k values and compare. We will use `sapply()` function to get average silhouette
values across k values between 2 and 7. Within `sapply()` there is an anonymous function that that does the clustering and
calculates average silhouette values for each k. We will plot the average silhouette values across k values between 2 and
7 to visualize. For the gap statistic, we can easily use the `cluster::clusGap()` function. We will define the clustering
function and put it in the `clusGap()` with maximum value of k as 8 and number of bootstrap samples as 50.
We will then plot the Gap statistic value across k values between 1 and 8.

```{r,eval=FALSE}  

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Silhouette

# Load the library
library(cluster)
# Set the seed
set.seed(101)

# Calculate the sihouette values for k-medoids clustering with the `pam()` function with `k = 5`
pamclu <- cluster::pam(t(mat),k=5)
# Plot the resulting plot
plot(silhouette(pamclu),main=NULL)

# Get average silhouette values across k values between 2 and 7
Ks <- sapply(2:7,
    function(i) 
      summary(silhouette(pam(t(mat),k=i)))$avg.width)
# Plot the Ks
plot(2:7,Ks,xlab="k",ylab="av. silhouette",type="b",
     pch=19)

# Gap Statistic

set.seed(101)
# define the clustering function
pam1 <- function(x,k) 
  list(cluster = pam(x,k, cluster.only=TRUE))

# calculate the gap statistic with maximum number of k as 8.
pam.gap <- clusGap(t(mat), FUN = pam1, K.max = 8,B=50)

# plot the gap statistic accross k values
plot(pam.gap, main = "Gap statistic for the 'Leukemia' data")

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set with gene expression values from 60 bone marrow samples of patients with one of the four main types of
leukemia (ALL, AML, CLL, CML) or no-leukemia controls, trimmed down to the top 1000 most variable genes. 
The file is called "leukemiaExpressionSubset.txt". The matrix that contains gene expressions has the genes in the rows 
and the patients in the columns. I want to represent the patients with expression profiles of just two genes, CD33
(ENSG00000105383) and PYGL (ENSG00000100504). I want to do the principal component analysis, in which I draw the
principal components over the original scatter plot and show the new coordinate system based on the principal components
in the next plot. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. We can create the subset of the data with two genes only by extracting information for those whose rownames are one
of the two gene codes. For the sub_matrix we transpose the matrix so that the samples are on the columns, and use `scale()`
function to center the data. We then will first plot our genes of interest as scatter plot. To color the points based on
Leukemia types, we set the leukemia type annotation for each sample. We also added the legend for the leukemia types. To
plot the eigenvectors, we will calculate the PCA with the `princomp()` function - this function returns the new coordinate
s as well. Then we will plot the direction of eigenvectors which were returned by `princomp()`. Lastly we will plot the
samples in the new coordinate system and plot the new coordinate basis vectors.


```{r,eval=FALSE}  

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

par(mfrow=c(1,2))

# create the subset of the data with two genes only
# notice that we transpose the matrix so samples are on the columns
sub.mat <- scale(t(mat[rownames(mat) %in% c("ENSG00000100504","ENSG00000105383"),]))

# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(mat),1,3))

# ploting our genes of interest as scatter plots
plot(sub.mat[colnames(sub.mat)=="ENSG00000100504",],
     sub.mat[colnames(sub.mat)=="ENSG00000105383",],
     pch=19,
     ylab="CD33 (ENSG00000105383)",
     xlab="PYGL (ENSG00000100504)",
     col=as.factor(annotation_col$LeukemiaType),
     xlim=c(-2,2),ylim=c(-2,2))

# create the legend for the Leukemia types
legend("bottomright",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# calculate the PCA only for our genes and all the samples
pr <- princomp(sub.mat)

# plot the direction of eigenvectors
# pr$loadings returned by princomp has the eigenvectors
arrows(x0=0, y0=0, x1 = pr$loadings[1,1], 
       y1 = pr$loadings[2,1],col="pink",lwd=3)
arrows(x0=0, y0=0, x1 = pr$loadings[1,2], 
       y1 = pr$loadings[2,2],col="gray",lwd=3)

# plot the samples in the new coordinate system
plot(-pr$scores,pch=19,
     col=as.factor(annotation_col$LeukemiaType),
     ylim=c(-2,2),xlim=c(-4,4))

# plot the new coordinate basis vectors
arrows(x0=0, y0=0, x1 =-2, 
       y1 = 0,col="pink",lwd=3)
arrows(x0=0, y0=0, x1 = 0, 
       y1 = -1,col="gray",lwd=3)


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set with gene expression values from 60 bone marrow samples of patients with one of the four main types of
leukemia (ALL, AML, CLL, CML) or no-leukemia controls, trimmed down to the top 1000 most variable genes.
The file is called "leukemiaExpressionSubset.txt". The matrix that contains gene expressions has the genes in the rows
and the patients in the columns. I first want to do a singular value decomposition (SVD) on the matrix and use that
to plot samples on the reduced dimensions. I also want to perfom the PCA on transposed matrix and plot the sampes on the
reduced coordinates. I also want to do a independent component analysis (ICA) to extract 2 components and plot the
rows of matrix S. Lastly, I want to use non-negative matrix factorization algorithms to decompose the matrix and plot the
values of metagenes (rows of H). How can I complete these tasks? 

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. To color the points based on Leukemia types, we set the leukemia type annotation for each sample.
Then, we will start with the singular value decomposition (SVD). We will apply the function `svd()` on the scaled matrix.
Then we can get projection on eigenassays by doing U^T * X operation, which is same as SV^T.
We will project the columns in our original expression data on eigenarrays and use the first two dimensions in the 
scatter plot. We will also perform the PCA this time with the `prcomp()` function on the transposed genes-by-samples 
matrix to get similar information, and plot the samples on the reduced coordinates. For the independent component analysis 
(ICA), we will load the library `fastICA`. We will apply `fastICA()` function on the transposed matrix and specify that
`n.comp = 2` to extract 2 components. We will plot the rows of matrix S which represents metagenes. For the non-negative
matrix factorization we will load the library `NMF`. We will apply `nmf()` function on the matrix with `rank = 3` for
specification of factorization rank of 3, and `seed = "nndsvd"` to indicate the non-negative double singular value
decomposition. With `basis()` we get the matrix W, and with `coef()` we get the matrix H. We will plot the values of
metagenes (rows of H) for components 1 and 3.

```{r,eval=FALSE}  

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(mat),1,3))

# 1. Singular value decomposition (SVD): X = USV^T

par(mfrow=c(1,2))

# apply SVD on the scaled matrix
d <- svd(scale(mat)) 

# projection on eigenassays : U^T * X = SV^T 
assays <- t(d$u) %*% scale(mat) 

# Plot the eigenassays
plot(assays[1,],assays[2,],pch=19,
     col=as.factor(annotation_col$LeukemiaType))

# apply PCA on transposed matrix
pr <- prcomp(t(mat),center=TRUE,scale=TRUE) 

# plot new coordinates from PCA, projections on eigenvectors
plot(pr$x[,1],pr$x[,2],col=as.factor(annotation_col$LeukemiaType))

# 2. Independent component analysis (ICA): X = SA

# Load the fastICA library
library(fastICA)
# apply ICA
ica.res <- fastICA(t(mat),n.comp=2) 

# plot reduced dimensions
plot(ica.res$S[,1],ica.res$S[,2],col=as.factor(annotation_col$LeukemiaType))

# 3. Non-negative matrix factorization (NMF): X = WH

# Load the NMF library
library(NMF)

# NMF with 3 components/factors
res <- NMF::nmf(mat,rank=3,seed="nndsvd") 
w <- basis(res) # get W
h <- coef(res)  # get H

# plot 1st factor against 3rd factor
plot(h[1,],h[3,],col=as.factor(annotation_col$LeukemiaType),pch=19)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set with gene expression values from 60 bone marrow samples of patients with one of the four main types of
leukemia (ALL, AML, CLL, CML) or no-leukemia controls, trimmed down to the top 1000 most variable genes.
The file is called "leukemiaExpressionSubset.txt". The matrix that contains gene expressions has the genes in the rows
and the patients in the columns. I want to do a classical multi-dimensional scaling and isotonic multi-dimensional scaling 
on the matrix and plot them in a scatterplot.I also want to use t-Distributed Stochastic Neighbor Embedding (t-SNE) and
plot them. How can I do that? 

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. To color the points based on Leukemia types, we set the leukemia type annotation for each sample.
Then we can run the classical MDS using `cmdscale()` function on the distance between samples of the transposed matrix.
For isotonic MDS, we can use `MASS::isoMDS()` function. Then we can plot the Leukemia gene expression values per patient
on reduced dimensions by classical MDS and isometric MDS.
For the t-SNE we will first load the library `"Rtsne"`. Then we will set the random seed because the t-SNE optimization
algorithm has random starting points and this might create non-identical results in every run. After calculating the t-SNE
lower dimension embeddings we will plot the points in a 2D scatter plot and create the legend for the leukemia types.

```{r,eval=FALSE}  

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(mat),1,3))

# 1. Multi-dimensional scaling

# Classical MDS
mds <- cmdscale(dist(t(mat)))
# Isotonic MDS
isomds <- MASS::isoMDS(dist(t(mat)))

# plot the patients in the 2D space
par(mfrow=c(1,2))
# Classical MDS
plot(mds,pch=19,col=as.factor(annotation_col$LeukemiaType),
     main="classical MDS")
# Isotonic MDS
plot(isomds$points,pch=19,col=as.factor(annotation_col$LeukemiaType),
     main="isotonic MDS")

# 2. t-Distributed Stochastic Neighbor Embedding (t-SNE)

# Load the library
library("Rtsne")

# Set a seed for reproducible results
set.seed(42) 

# Run TSNE
tsne_out <- Rtsne(t(mat),perplexity = 10) 
#image(t(as.matrix(dist(tsne_out$Y))))

# Show the objects in the 2D tsne representation
plot(tsne_out$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19)

# create the legend for the Leukemia types
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

```
--- TASK END

