---
title: "prompt/completion pairs chapter 4"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:

I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
I would first like to observe the effect of data transformation with scaling. How can I take the logarithm of the data
with the `log2()` function and then scale the expression matrix using `scale()` function? After that, I would like to
also make boxplots of the unscaled and scaled data sets using the `boxplot()` function.

With the same data set, use the ward.d distance in hierarchical clustering and plot multiple heatmaps.
How can I plot a heatmap with a dendrogram? I want to create the heatmap for both scaled and unscaled matrix.

For the transformed and untransformed data sets, I want to use the silhouette for deciding number of clusters using
hierarchical clustering. I also want to plot them. How can I do that?

Lastly, I want to use the Gap Statistic for deciding the number of clusters in hierarchical clustering.
Is it the same number of clusters identified by two methods?

completion:

The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will first start with scaling the data by using `log2()` and `scale()` function.
We can make boxplots using `boxplots()` to compare the unscaled and scaled expression matrices.

Then we will load the `pheatmap` library to create heatmaps using ward.d distance. To create the heatmap, we will first
set the leukemia type annotation for each sample by taking the first three letters of the column names as substring:
for example, ALL_GSM330151.CEL -> ALL. Then we will set the row names of our annotation_col as our column names of the
matrix. Then we can use the function `pheatmap()`, using "ward.D2" as clustering method as asked.
We will do the process twice for unscaled and scaled matrices.

Then we will use the silhouette for deciding number of clusters using hierarchical clustering. First we will calculate
the silhouette values for k-medoids clustering with the pam() function with k = 5 and plot the values. Then we will
determine best k values by testing k=2 to k=7 using sapply and plot the results. We will do the process twice for
unscaled and scaled matrices.

Lastly we will use Gap Statistic to decide the number of clusters for unscaled and scaled data. First defining a
clustering function is needed, then we can use the `clusGap()` function to determine the Gap Statistic.
Here we set the maximum number of clusters to be 8 and the number of bootstrap samples to 50.
We will find the Gap statistics for unscaled and scaled data and plot them.


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# 1. Observe the effect of data transformation with scaling

# Take Logarithm and then Scale.
scaled_mat <- scale(log2(mat))

# Make boxplots for unscaled and scaled expression matrix
par(mfrow=c(1,2))
boxplot(mat, xlab="Gene",ylab="Expression",col="lightblue",main="Unscaled Expression Matrix")
boxplot(scaled_mat, xlab="Gene",ylab="Expression",col="lightpink",main="Scaled Expression Matrix")

# 2. Use ward.d distance in hierarchical clustering and plot multiple heatmaps

# Load needed library
library(pheatmap)

# Heatmap for unscaled expression matrix
# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(mat),1,3))
# Set the row names of our annotation_col as column names of the matrix
rownames(annotation_col) <- colnames(mat)

# Use pheatmap() to create heatmap. Use "ward.D2" as clustering method as asked.
pheatmap(mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,scale="none",
         clustering_method = "ward.D2",clustering_distance_cols="euclidean",
         main = "Unscaled expression matrix")

# Heatmap for scaled expression matrix
# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(scaled_mat),1,3))
# Set the row names of our annotation_col as column names of the matrix
rownames(annotation_col) <- colnames(scaled_mat)

# Use pheatmap() to create heatmap. Use "ward.D2" as clustering method as asked.
pheatmap(scaled_mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,scale="none",
         clustering_method = "ward.D2",clustering_distance_cols="euclidean",
         main = "Scaled expression matrix")

# 3. Use the silhouette for deciding number of clusters using hierarchical clustering

# Load the needed library
library(cluster)

# Set random number seed, so that the random numbers from the text is the same when you run the code.
set.seed(101)

# For unscaled data set
# Calculate the silhouette values for k-medoids clustering with the pam() function with k = 5
pamclu <- cluster::pam(t(mat),k=5)
# Plot the silhouette values
plot(silhouette(pamclu),main="Unscaled data")

# Determine best k values by testing k=2 to k=7 using sapply
Ks <- sapply(2:7,
            function(i)
              summary(silhouette(pam(t(mat),k=i)))$avg.width)
# Plot the k values in accordance with average silhouette values
plot(2:7,Ks,xlab="k",ylab="avg.silhouette",type="b",pch=19)

# For scaled data set
# Calculate the silhouette values for k-medoids clustering with the pam() function with k = 5
pamclu <- cluster::pam(t(scaled_mat),k=5)
# Plot the silhouette values
plot(silhouette(pamclu),main="Scaled data")

# Determine best k values by testing k=2 to k=7 using sapply
Ks <- sapply(2:7,
            function(i)
              summary(silhouette(pam(t(scaled_mat),k=i)))$avg.width)
# Plot the k values in accordance with average silhouette values
plot(2:7,Ks,xlab="k",ylab="avg.silhouette",type="b",pch=19)


# 4 Gap Statistic to decide the number of clusters
set.seed(101)

# Clustering Function
pam1 <- function(x,k)
  list(cluster = pam(x,k,cluster.only=TRUE))

# Gap Statistic for unscaled data
# K.max: max number of clusters to consider(at least 2)
# FUN: Clustering Function
# B: Number of Bootstrap samples
pam.gap <- clusGap(t(mat), FUN = pam1, K.max=8, B=50)
plot(pam.gap, main="Gap Statistic(unscaled)")

# Gap Statistic for scaled data
pam.gap1 <- clusGap(t(scaled_mat), FUN = pam1, K.max=8, B=50)
plot(pam.gap1, main="Gap Statistic(scaled)")
```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:

I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
I would like to observe the effect of data transformation with scaling. How can I take the logarithm of the data with the
`log2()` function and then scale the expression matrix using `scale()` function? After that, I would like to also make
boxplots of the unscaled and scaled data sets using the `boxplot()` function.


completion:

The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will first start with scaling the data by using `log2()` and `scale()` function.
We can make boxplots using `boxplots()` to compare the unscaled and scaled expression matrices.


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Observe the effect of data transformation with scaling

# Take Logarithm and then Scale.
scaled_mat <- scale(log2(mat))

# Make boxplots for unscaled and scaled expression matrix
par(mfrow=c(1,2))
boxplot(mat, xlab="Gene",ylab="Expression",col="lightblue",main="Unscaled Expression Matrix")
boxplot(scaled_mat, xlab="Gene",ylab="Expression",col="lightpink",main="Scaled Expression Matrix")


```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:

I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
I would like to observe the effect of data transformation with scaling.
How can I take the logarithm of the data with the `log2()` function and then scale the expression matrix using `scale()` function?
I would like to use the ward.d distance in hierarchical clustering and plot multiple heatmaps.
How can I plot a heatmap with a dendrogram? I want to create the heatmap for both scaled and unscaled matrix.

completion:

The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will first start with scaling the data by using `log2()` and `scale()` function. 
Then we will load the `pheatmap` library to create heatmaps using ward.d distance. To create the heatmap, we will first
set the leukemia type annotation for each sample by taking the first three letters of the column names as substring:
for example, ALL_GSM330151.CEL -> ALL. Then we will set the row names of our annotation_col as our column names of the
matrix. Then we can use the function `pheatmap()`, using "ward.D2" as clustering method as asked. We will do the process
twice for unscaled and scaled matrices.


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Take Logarithm and then Scale.
scaled_mat <- scale(log2(mat))

# Use ward.d distance in hierarchical clustering and plot multiple heatmaps

# Load needed library
library(pheatmap)

# Heatmap for unscaled expression matrix
# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(mat),1,3))
# Set the row names of our annotation_col as column names of the matrix
rownames(annotation_col) <- colnames(mat)

# Use pheatmap() to create heatmap. Use "ward.D2" as clustering method as asked.
pheatmap(mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,scale="none",
         clustering_method = "ward.D2",clustering_distance_cols="euclidean",
         main = "Unscaled expression matrix")

# Heatmap for scaled expression matrix
# Set the leukemia type annotation for each sample by taking the first three letters of the column names as substring
annotation_col <- data.frame(LeukemiaType = substr(colnames(scaled_mat),1,3))
# Set the row names of our annotation_col as column names of the matrix
rownames(annotation_col) <- colnames(scaled_mat)

# Use pheatmap() to create heatmap. Use "ward.D2" as clustering method as asked.
pheatmap(scaled_mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,scale="none",
         clustering_method = "ward.D2",clustering_distance_cols="euclidean",
         main = "Scaled expression matrix")


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:

I have this dataset "leukemiaExpressionSubset.rds" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
I would first like to observe the effect of data transformation with scaling. How can I take the logarithm of the data
with the `log2()` function and then scale the expression matrix using `scale()` function?
For the transformed and untransformed data sets, I want to use the silhouette for deciding number of clusters using
hierarchical clustering. I also want to plot them. How can I do that?

completion:

The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will first start with scaling the data by using `log2()` and `scale()` function.
Then we will use the silhouette for deciding number of clusters using hierarchical clustering. First we will calculate
the silhouette values for k-medoids clustering with the pam() function with k = 5 and plot the values. Then we will
determine best k values by testing k=2 to k=7 using sapply and plot the results. We will do the process twice for
unscaled and scaled matrices.


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Observe the effect of data transformation with scaling

# Take Logarithm and then Scale.
scaled_mat <- scale(log2(mat))

# Use the silhouette for deciding number of clusters using hierarchical clustering

# Load the needed library
library(cluster)

# Set random number seed, so that the random numbers from the text is the same when you run the code.
set.seed(101)

# For unscaled data set
# Calculate the silhouette values for k-medoids clustering with the pam() function with k = 5
pamclu <- cluster::pam(t(mat),k=5)
# Plot the silhouette values
plot(silhouette(pamclu),main="Unscaled data")

# Determine best k values by testing k=2 to k=7 using sapply
Ks <- sapply(2:7,
            function(i)
              summary(silhouette(pam(t(mat),k=i)))$avg.width)
# Plot the k values in accordance with average silhouette values
plot(2:7,Ks,xlab="k",ylab="avg.silhouette",type="b",pch=19)

# For scaled data set
# Calculate the silhouette values for k-medoids clustering with the pam() function with k = 5
pamclu <- cluster::pam(t(scaled_mat),k=5)
# Plot the silhouette values
plot(silhouette(pamclu),main="Scaled data")

# Determine best k values by testing k=2 to k=7 using sapply
Ks <- sapply(2:7,
            function(i)
              summary(silhouette(pam(t(scaled_mat),k=i)))$avg.width)
# Plot the k values in accordance with average silhouette values
plot(2:7,Ks,xlab="k",ylab="avg.silhouette",type="b",pch=19)


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:

I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
I would first like to observe the effect of data transformation with scaling. How can I take the logarithm of the data
with the `log2()` function and then scale the expression matrix using `scale()` function?
Then I want to use the Gap Statistic for deciding the number of clusters in hierarchical clustering. 

completion:

The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will first start with scaling the data by using `log2()` and `scale()` function.
After that we will use Gap Statistic to decide the number of clusters for unscaled and scaled data. First defining a
clustering function is needed, then we can use the `clusGap()` function to determine the Gap Statistic. Here we set the
maximum number of clusters to be 8 and the number of bootstrap samples to 50. We will find the Gap statistics for
unscaled and scaled data and plot them.


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# The effect of data transformation with scaling

# Take Logarithm and then Scale.
scaled_mat <- scale(log2(mat))

# Gap Statistic to decide the number of clusters
set.seed(101)

# Clustering Function
pam1 <- function(x,k)
  list(cluster = pam(x,k,cluster.only=TRUE))

# Gap Statistic for unscaled data
# K.max: max number of clusters to consider(at least 2)
# FUN: Clustering Function
# B: Number of Bootstrap samples
pam.gap <- clusGap(t(mat), FUN = pam1, K.max=8, B=50)
plot(pam.gap, main="Gap Statistic(unscaled)")

# Gap Statistic for scaled data
pam.gap1 <- clusGap(t(scaled_mat), FUN = pam1, K.max=8, B=50)
plot(pam.gap1, main="Gap Statistic(scaled)")
```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.

I want to do PCA(Principal component analysis) on the expression matrix using the `princomp()` function and then use the
screeplot() function to visualize the explained variation by eigenvectors. How do I determine how many top components
explain 95% of the variation?

Then I want to remove eigenvectors that make up 95% of variation and reconstruct the matrix using SVD, then calculate the
reconstruction error as the difference between original and reconstructed matrix.

How do I produce a 10-component ICA(Independent component analysis) from the expression data set? How do I remove each
component and measure the reconstruction error without that component? I want to rank the components by decreasing
reconstruction-error.

Lastly I want to use the Rtsne() function on the leukemia expression data set and try to increase and decrease perplexity
t-sne, and describe the observed changes in 2D plots.


completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then first we will do PCA on the expression matrix using `princomp()` function and use `screenplot()` to
visualize. To calculate how many top components explain 95% of the variation, we will first calculate the cumulative
variance and find out at which point the cumulative variance is over 95.

For the next part, we will first use `svd()` to create a svd matrix. Then we will equalize eigenvalue for the top
components that make up 95% of variation to 0 and then reconstruct the matrix. We can calculate the reconstruction error
by getting the difference of the original and reconstructed matrix.

To do the ICA, we will first load the `library(fastICA)` and perform ICA with 10 components. To measure the
reconstruction error without the component for the 10-component, we will first initialize a list to store the errors.
Then we can create a for-loop from 1 to 10 that remove the i-th component and measure reconstruction error each time.
Using the list of reconstruction errors, we can rank the components by decreasing reconstruction error.

For the last part, we will load the `library("Rtsne")`. We will run TSNE using `Rtsne()` function create a plot and
legend to visualize. We will do it for 4 different perplexity - 4,8,12,16 - but you can also change the perplexity as you
want. However, it should remain so that 3 * perplexity < nrow(X) - 1. 


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# 1. Do PCA on the expression matrix and visualize
pr <- princomp(mat, scale=TRUE)
screeplot(pr, type = "line")

# How many top components explain 95% of variation? 24
cumulative_variance <- cumsum(pr$sdev^2) / sum(pr$sdev^2)
top_components <- sum(cumulative_variance <= 0.95) + 1

# 2. X = USV^T
# Use SVD
svd_mat <- svd(mat)

# Equalize eigenvalue to 0 for top components for variation
svd_mat$d[1:24] <- 0

# reconstruct the matrix
re_mat <- svd_mat$u %*% diag(svd_mat$d) %*% t(svd_mat$v)

# Calculate the reconstruction error as the difference between original and reconstructed matrix
reconst_error <- mat - re_mat

# 3. X = SA
library(fastICA)

# Perform ICA with 10 components
ica_res <- fastICA(t(mat), n.comp = 10)

# Initialize a list to store the reconstruction errors
recon_err <- vector("list", 10)

# Remove each component and measure reconstruction error
for (i in 1:10) {
  # Remove the i-th component from the ICA result
  mod_ica_res <- ica_res
  mod_ica_res$S[i, ] <- 0  # Setting the i-th component to 0
  
  # Reconstruct the data without the i-th component
  recon_mat <- (mod_ica_res$S) %*% (mod_ica_res$A)
  
  # Calculate the reconstruction error
  recon_err[[i]] <- t(mat) - recon_mat
}

# Rank the components by decreasing reconstruction error
sorted_components <- order(sapply(recon_err, function(error) sum(error^2)), decreasing = TRUE)

# 4
# Load the library
library("Rtsne")

# Set random number seed, so that the random numbers from the text is the same when you run the code.
set.seed(42)

# To put 4 graphs into columns and rows of 2.
par(mfrow=c(2,2))

# Run TSNE
tsne_out4 <- Rtsne(t(mat),perplexity = 4)

# Show the objects in the 2D tsne representation
plot(tsne_out4$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 4")
# create the legend for the Leukemia types
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# Repeat for perplexity = 8
tsne_out8 <- Rtsne(t(mat),perplexity = 8)
plot(tsne_out8$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 8")
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# Repeat for perplexity = 12
tsne_out12 <- Rtsne(t(mat),perplexity = 12)
plot(tsne_out12$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 12")
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# Repeat for perplexity = 16
tsne_out16 <- Rtsne(t(mat),perplexity = 16)
plot(tsne_out16$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 16")
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)
```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.

I want to do PCA(Principal component analysis) on the expression matrix using the `princomp()` function and then use the
screeplot() function to visualize the explained variation by eigenvectors. How do I determine how many top components
explain 95% of the variation?


completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then first we will do PCA on the expression matrix using `princomp()` function and use `screenplot()` to
visualize. To calculate how many top components explain 95% of the variation, we will first calculate the cumulative
variance and find out at which point the cumulative variance is over 95.


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Do PCA on the expression matrix and visualize
pr <- princomp(mat, scale=TRUE)
screeplot(pr, type = "line")

# How many top components explain 95% of variation? 24
cumulative_variance <- cumsum(pr$sdev^2) / sum(pr$sdev^2)
top_components <- sum(cumulative_variance <= 0.95) + 1

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling , numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
I want to remove eigenvectors that make up 95% of variation - the top 24 - and reconstruct the matrix using SVD, then
calculate the reconstruction error as the difference between original and reconstructed matrix.


completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. We will first use `svd()` to create a svd matrix. Then we will equalize eigenvalue for the top components that
make up 95% of variation to 0 and then reconstruct the matrix. We can calculate the reconstruction error by getting the
difference of the original and reconstructed matrix.


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# X = USV^T

# Use SVD
svd_mat <- svd(mat)

# Equalize eigenvalue to 0 for top components for variation
svd_mat$d[1:24] <- 0

# reconstruct the matrix
re_mat <- svd_mat$u %*% diag(svd_mat$d) %*% t(svd_mat$v)

# Calculate the reconstruction error as the difference between original and reconstructed matrix
reconst_error <- mat - re_mat

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
How do I produce a 10-component ICA(Independent component analysis) from the expression data set? How do I remove each
component and measure the reconstruction error without that component? I want to rank the components by decreasing
reconstruction-error.


completion:
The data set must be read first.  We will use `data.table::fread()` function to read in the text file as it can read
files faster. To do the ICA, we will first load the `library(fastICA)` and perform ICA with 10 components. To measure the
reconstruction error without the component for the 10-component, we will first initialize a list to store the errors.
Then we can create a for-loop from 1 to 10 that remove the i-th component and measure reconstruction error each time.
Using the list of reconstruction errors, we can rank the components by decreasing reconstruction error.



```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# X = SA
library(fastICA)

# Perform ICA with 10 components
ica_res <- fastICA(t(mat), n.comp = 10)

# Initialize a list to store the reconstruction errors
recon_err <- vector("list", 10)

# Remove each component and measure reconstruction error
for (i in 1:10) {
  # Remove the i-th component from the ICA result
  mod_ica_res <- ica_res
  mod_ica_res$S[i, ] <- 0  # Setting the i-th component to 0
  
  # Reconstruct the data without the i-th component
  recon_mat <- (mod_ica_res$S) %*% (mod_ica_res$A)
  
  # Calculate the reconstruction error
  recon_err[[i]] <- t(mat) - recon_mat
}

# Rank the components by decreasing reconstruction error
sorted_components <- order(sapply(recon_err, function(error) sum(error^2)), decreasing = TRUE)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have this dataset "leukemiaExpressionSubset.txt" which is consisted of gene expression values from 60 bone marrow
samples of patients with one of the four main types of leukemia (ALL, AML, CLL, CML) or no-leukemia controls.
I want to use the Rtsne() function on the leukemia expression data set and try to increase and decrease perplexity t-sne,
and describe the observed changes in 2D plots.

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. We will load the `library("Rtsne")`. We will run TSNE using `Rtsne()` function create a plot and legend to
visualize. We will do it for 4 different perplexity - 4,8,12,16 - but you can also change the perplexity as you want.
However, it should remain so that 3 * perplexity < nrow(X) - 1. 


```{r,eval=FALSE}

# read leukemia expression data set
mat <- data.table::fread("leukemiaExpressionSubset.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(mat) <- mat[,1]
mat <- mat[,-1]

# Load the library
library("Rtsne")

# Set random number seed, so that the random numbers from the text is the same when you run the code.
set.seed(42)

# To put 4 graphs into columns and rows of 2.
par(mfrow=c(2,2))

# Run TSNE
tsne_out4 <- Rtsne(t(mat),perplexity = 4)

# Show the objects in the 2D tsne representation
plot(tsne_out4$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 4")
# create the legend for the Leukemia types
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# Repeat for perplexity = 8
tsne_out8 <- Rtsne(t(mat),perplexity = 8)
plot(tsne_out8$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 8")
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# Repeat for perplexity = 12
tsne_out12 <- Rtsne(t(mat),perplexity = 12)
plot(tsne_out12$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 12")
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# Repeat for perplexity = 16
tsne_out16 <- Rtsne(t(mat),perplexity = 16)
plot(tsne_out16$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19, main="Perplexity 16")
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)
```
--- TASK END
