---
title: "prompt/completion pairs chapter 4"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a data table named `df` (shown below). I would like to calculate the distance metric between gene expression
vectors between the patient pairs using "Manhattan distance" , "Euclidean Distance" and the "correlation distance".
How can I do that? Also, how can I scale the gene expression values? I want to get a tree structure or a dendrogram that
shows the relationship between the individual data points and clusters. How do I do a hierarchical clustering on Euclidean
distances between patients and plot the resulting clustering tree?

```{r expTable,echo=FALSE}
df <- data.frame(
    IRX4=c(11,13,2,1),
    OCT4=c(10,13,4,3 ),
    PAX6=c(1 ,3 ,10,9),
    row.names=c("patient1","patient2","patient3","patient4")
  )
```

completion:
We will first the given data table `df` to be able to use for the tasks. Then we will use the `dist()` function for
Manhattan and Euclidean distances. To calculate the correlation distance, we will use the formula of $d_{AB}=1-\rho$,
where $\rho$ is the Pearson correlation coefficient between two vectors. For the standardization we can use `scale()`
function. After that, we can do a hierarchical clusting on Euclidean distance. We will store the Euclidean distance in a
variable, and use `hclust()` function on that. We will use `method = "complete"` for complete linkage. Lastly, we can plot
the clustering tree by simply using `plot()` on the hierarchical clusting.

```{r,eval=FALSE}  

# Copy the data frame
df <- data.frame(
    IRX4=c(11,13,2,1),
    OCT4=c(10,13,4,3 ),
    PAX6=c(1 ,3 ,10,9),
    row.names=c("patient1","patient2","patient3","patient4")
  )

# Calculate the distance metric using Manhattan method
dist(df,method="manhattan")

# Calculate the distance metric using Euclidean method
dist(df,method="euclidean")

# Calculate the distance metric using correlation distance
as.dist(1-cor(t(df))) 

# Scale the data frame
scale(df)

# When not specified, default is Euclidean method
d <- dist(df)

# Use the base dunction in R to do hierarchical clusting
hc=hclust(d,method="complete")

# Plot the hierarchical clusting
plot(hc)

```
--- TASK END
