---
title: "prompt/completion pairs chapter 5"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.rds", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I first want to preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero
variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors, center the
data and remove higly correlated variables with cutoff = 0.9. Also, how do I merge these two files to create a table?

I want to run the k-NN classifier on the data. What is the effect on classification accuracy for k-NN predicting the CIMP
and noCIMP status of the patient? Then I want to use bootstrap resampling with k-NN for the prediction accuracy. How
different is it from cross-validation for different ks? There are a number of ways to get variable importance for a
classification problem. I want to use random forests on the classification problem. How do I plot the top 10 important
variables based on permutation-based method for the random forest classification? How do I normalize the importance
scores obtained by the random forest method?

completion:
The data set must be read first. We will use `readRDS()` function to read the rds file. Then we will start the
preprocessing of the gene expression data with taking a log to tame the extreme values and add 1 pseudocount to deal with
0 values. Then, before doing other tasks we need to transpose our gene expression data, since our gene expression data
are on the row side, and we need to put it on the column side to use machine learning algorithms. Then we will load the
`caret` library to do the filtering. First we will create a filter with `preProcess()` function and apply it to remove
near zero variation for the columns at least 85% of the values are the same. Then we will choose to take the top 1000
variable predictors, which have the biggest variances. Then we will center the data, and create a filter to remove higly
correlated variables. Lastly we can merge the two data set by row names, which are now for both data set the patient-IDs.
After the merge we have the sample-Ids on the first column, and we need to put it back on the row names. For that we will
set the row names as the first column of the current data set and delete the first column.

Now to go through knn-classifier, bootstrap resampling and random forest method, we will first split the data set with
`caret::createPartition()` function. We will get indices for the 70 % of the data set and separate test and training sets
using the indices (70 % for training and 30 % for test).

To run the k-NN algorithm, we will use the `caret::knn3()` function with k = 5. To test the training set accuracy, we
will get the predictions on the training set using `predict()` and compare the predicted to real by
`caret::confusionMatrix()`. For the test set accuracy we will do the same for the testing set.

Next we will run the bootstrap resampling by using `trainControl()` function with method as "boot". Then we will train
the k-NN model with `train()` function with method as "knn". To compare that with cross validation we will repeat the
process, this time `trainControl()` function with method as "cv". 

For the random forest model, we will train it with `train()` function with method as "ranger" and indicate the importance
as "permutation" to calculate the importance. Variable importance can be calculated by `caret::varImp()` function. We can
plot the top 10 important variables using `plot()`. We can normalize the importance scores by applying a function that
does (x - min(x)) / (max(x) - min(x)) to all importance scores.

```{r,eval=FALSE}  

# read gene expression data set from glioblastoma patients
gexp <- readRDS("LGGrnaseq.rds")

# read patient annotation data set from glioblastoma patients
patient <- readRDS("patient2LGGsubtypes.rds")

# 1. Preprocessing----------------------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)
# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
nzv_tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
data <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(data) <- data[,1]
data <- data[,-1]

# 2. k-NN classifier on the data ----------

# get indices for 70% of the data set
intrain <- createDataPartition(y = data[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- data[intrain,]
testing <- data[-intrain,]

set.seed(123)
# Run k-NN classifier
knnFit <- knn3(x=training[,-1], # training set
              y=as.factor(training[,1]), # training set class labels
              k=5)

# predictions on the training set
trainPred <- predict(knnFit,training[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(training[,1]),reference=trainPred)

# predictions on the test set, return class labels
testPred <- predict(knnFit,testing[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(testing[,1]),reference=testPred)

# 3. Bootstrap resampling with k-NN for the prediction accuracy----------------

set.seed(123)

# train control with bootstrap
trctrl <- trainControl(method = "boot",number=20, returnResamp = "all")

# Train k-NN model
knn_fit <- train(subtype ~., data = training, 
                 method = "knn",
                 trControl=trctrl,
                 tuneGrid = data.frame(k=1:12)
                 )

# best k value
knn_fit$bestTune 

# plot k vs prediction error
plot(x=1:12,1-knn_fit$results[,2],pch=19,
     ylab="prediction error",xlab="k")
lines(loess.smooth(x=1:12,1-knn_fit$results[,2],degree=2),
      col="#CC0000")

# How different is it from cross-validation for different ks?

set.seed(123)

# Cross validation method
trctrl_cv <- trainControl(method = "cv",number=10)

# we will now train k-NN model
knn_fit_cv <- train(subtype~., data = training, 
                 method = "knn",
                 trControl=trctrl_cv,
                 tuneGrid = data.frame(k=1:12))

# best k value by cross-validation accuracy
knn_fit_cv$bestTune

# plot k vs prediction error
plot(x=1:12,1-knn_fit_cv$results[,2],pch=19,
     ylab="prediction error",xlab="k")
lines(loess.smooth(x=1:12,1-knn_fit_cv$results[,2],degree=2),
      col="green")


# 4. Run random forests ---------------------

set.seed(123)

# we will now train random forest model
rfFit <- train(subtype~., 
               data = data, 
               method = "ranger",
               trControl= trainControl(method = "none"),
               importance="permutation", # calculate importance
               tuneGrid = data.frame(mtry=100,
                                     min.node.size = 1,
                                     splitrule="gini")
)
# print OOB error
rfFit$finalModel$prediction.error

# Variable Importance
varImp(rfFit)
plot(varImp(rfFit),top=10)

# Extract the top 10 variables from both results
top_10_rfFit <- row.names(varImp(rfFit))[1:10]


# Normalize the importance scores from random forests
# Compute importance scores using Random Forests (rfFit)
importance_rf <- varImp(rfFit)

# Normalizing function
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
# Normalize the importance scores to a range between 0 and 1
normalized_importance_rf <- normalize(importance_rf$importance)



```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling , numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.rds", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I first want to preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero
variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors, center the
data and remove higly correlated variables with cutoff = 0.9. Also, how do I merge these two files to create a table?

completion:
The data set must be read first. We will use `readRDS()` function to read the rds file. Then we will start the
preprocessing of the gene expression data with taking a log to tame the extreme values and add 1 pseudocount to deal with
0 values. Then, before doing other tasks we need to transpose our gene expression data, since our gene expression data
are on the row side, and we need to put it on the column side to use machine learning algorithms. Then we will load the
`caret` library to do the filtering. First we will create a filter with `preProcess()` function and apply it to remove
near zero variation for the columns at least 85% of the values are the same. Then we will choose to take the top 1000
variable predictors, which have the biggest variances. Then we will center the data, and create a filter to remove higly
correlated variables. Lastly we can merge the two data set by row names, which are now for both data set the patient-IDs.
After the merge we have the sample-Ids on the first column, and we need to put it back on the row names. For that we will
set the row names as the first column of the current data set and delete the first column.

```{r,eval=FALSE}  

# read gene expression data set from glioblastoma patients
gexp <- readRDS("LGGrnaseq.rds")

# read patient annotation data set from glioblastoma patients
patient <- readRDS("patient2LGGsubtypes.rds")

# Preprocessing----------------------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)
# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
nzv_tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
data <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(data) <- data[,1]
data <- data[,-1]

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling , numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I first want to preprocess the data by taking the log and filtering the data. By filtering, I want to take the top 1000
variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. Also, how do I merge these
two files to create a table?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme values and
add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene expression data,
since our gene expression data are on the row side, and we need to put it on the column side to use machine learning
algorithms. Then we will load the `caret` library to do the filtering. We will choose to take the top 1000 variable
predictors, which have the biggest variances. Then we will center the data, and create a filter to remove higly
correlated variables. Lastly we can merge the two data set by row names, which are now for both data set the patient-IDs.
After the merge we have the sample-Ids on the first column, and we need to put it back on the row names. For that we will
set the row names as the first column of the current data set and delete the first column.

```{r,eval=FALSE}  

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocessing----------------------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
data <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(data) <- data[,1]
data <- data[,-1]

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I first want to preprocess the data by taking the log and filtering the data. By filtering, I want to take the top 1000
variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. Also, how do I merge these
two files to create a table?

I want to run the k-NN classifier on the data. What is the effect on classification accuracy for k-NN predicting the CIMP
and noCIMP status of the patient?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme values and
add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene expression data,
since our gene expression data are on the row side, and we need to put it on the column side to use machine learning
algorithms. Then we will load the `caret` library to do the filtering. We will choose to take the top 1000 variable
predictors, which have the biggest variances. Then we will center the data, and create a filter to remove higly
correlated variables. Lastly we can merge the two data set by row names, which are now for both data set the patient-IDs.
After the merge we have the sample-Ids on the first column, and we need to put it back on the row names. For that we will
set the row names as the first column of the current data set and delete the first column.

Now to go through knn-classifier, we will first split the data set with `caret::createPartition()` function. We will get
indices for the 70 % of the data set and separate test and training sets using the indices (70 % for training and 30 %
for test).

To run the k-NN algorithm, we will use the `caret::knn3()` function with k = 5. To test the training set accuracy, we
will get the predictions on the training set using `predict()` and compare the predicted to real by
`caret::confusionMatrix()`. For the test set accuracy we will do the same for the testing set.

```{r,eval=FALSE}  

# 0. Data Read-------------------------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# 1. Preprocessing----------------------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
data <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(data) <- data[,1]
data <- data[,-1]

# 2. k-NN classifier on the data ----------

# get indices for 70% of the data set
intrain <- createDataPartition(y = data[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- data[intrain,]
testing <- data[-intrain,]

set.seed(123)
# Run k-NN classifier
knnFit <- knn3(x=training[,-1], # training set
              y=as.factor(training[,1]), # training set class labels
              k=5)

# predictions on the training set
trainPred <- predict(knnFit,training[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(training[,1]),reference=trainPred)

# predictions on the test set, return class labels
testPred <- predict(knnFit,testing[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(testing[,1]),reference=testPred)

```

--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I first want to preprocess the data by taking the log and filtering the data. By filtering, I want to take the top 1000
variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. Also, how do I merge these
two files to create a table? I want to use bootstrap resampling with k-NN for the prediction accuracy. How different is
it from cross-validation for different ks? 

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme values and
add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene expression data,
since our gene expression data are on the row side, and we need to put it on the column side to use machine learning
algorithms. Then we will load the `caret` library to do the filtering. We will choose to take the top 1000 variable
predictors, which have the biggest variances. Then we will center the data, and create a filter to remove higly
correlated variables. Lastly we can merge the two data set by row names, which are now for both data set the patient-IDs.
After the merge we have the sample-Ids on the first column, and we need to put it back on the row names. For that we will
set the row names as the first column of the current data set and delete the first column.

We will first split the data set with `caret::createPartition()` function. We will get indices for the 70 % of the data
set and separate test and training sets using the indices (70 % for training and 30 % for test). Next we will run the
bootstrap resampling by using `trainControl()` function with method as "boot". Then we will train the k-NN model with
`train()` function with method as "knn". To compare that with cross validation we will repeat the process, this time
`trainControl()` function with method as "cv". 


```{r,eval=FALSE}  

# 0. Data Read-------------------------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# 1. Preprocessing----------------------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
data <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(data) <- data[,1]
data <- data[,-1]

# get indices for 70% of the data set
intrain <- createDataPartition(y = data[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- data[intrain,]
testing <- data[-intrain,]


# 2. Bootstrap resampling with k-NN for the prediction accuracy----------------

set.seed(123)

# train control with bootstrap
trctrl <- trainControl(method = "boot",number=20, returnResamp = "all")

# Train k-NN model
knn_fit <- train(subtype ~., data = training, 
                 method = "knn",
                 trControl=trctrl,
                 tuneGrid = data.frame(k=1:12)
                 )

# best k value
knn_fit$bestTune 

# plot k vs prediction error
plot(x=1:12,1-knn_fit$results[,2],pch=19,
     ylab="prediction error",xlab="k")
lines(loess.smooth(x=1:12,1-knn_fit$results[,2],degree=2),
      col="#CC0000")

# How different is it from cross-validation for different ks?

set.seed(123)

# Cross validation method
trctrl_cv <- trainControl(method = "cv",number=10)

# we will now train k-NN model
knn_fit_cv <- train(subtype~., data = training, 
                 method = "knn",
                 trControl=trctrl_cv,
                 tuneGrid = data.frame(k=1:12))

# best k value by cross-validation accuracy
knn_fit_cv$bestTune

# plot k vs prediction error
plot(x=1:12,1-knn_fit_cv$results[,2],pch=19,
     ylab="prediction error",xlab="k")
lines(loess.smooth(x=1:12,1-knn_fit_cv$results[,2],degree=2),
      col="green")


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I first want to preprocess the data by taking the log and filtering the data. By filtering, I want to take the top 1000
variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. Also, how do I merge these
two files to create a table?

There are a number of ways to get variable importance for a classification problem. I want to use random forests on the
classification problem. How do I plot the top 10 important variables based on permutation-based method for the random
forest classification? How do I normalize the importance scores obtained by the random forest method?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme values and
add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene expression data,
since our gene expression data are on the row side, and we need to put it on the column side to use machine learning
algorithms. Then we will load the `caret` library to do the filtering. We will choose to take the top 1000 variable
predictors, which have the biggest variances. Then we will center the data, and create a filter to remove higly
correlated variables. Lastly we can merge the two data set by row names, which are now for both data set the patient-IDs.
After the merge we have the sample-Ids on the first column, and we need to put it back on the row names. For that we will
set the row names as the first column of the current data set and delete the first column.

Now we will first split the data set with `caret::createPartition()` function. We will get indices for the 70 % of the
data set and separate test and training sets using the indices (70 % for training and 30 % for test). For the random
forest model, we will train it with `train()` function with method as "ranger" and indicate the importanc as
"permutation" to calculate the importance. Variable importance can be calculated by `caret::varImp()` function. We can
plot the top 10 important variables using `plot()`. We can normalize the importance scores by applying a function that
does (x - min(x)) / (max(x) - min(x)) to all importance scores.

```{r,eval=FALSE}  

# 0. Data Read-------------------------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# 1. Preprocessing----------------------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
data <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(data) <- data[,1]
data <- data[,-1]

# get indices for 70% of the data set
intrain <- createDataPartition(y = data[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- data[intrain,]
testing <- data[-intrain,]

# 2. Run random forests ---------------------

set.seed(123)

# we will now train random forest model
rfFit <- train(subtype~., 
               data = data, 
               method = "ranger",
               trControl= trainControl(method = "none"),
               importance="permutation", # calculate importance
               tuneGrid = data.frame(mtry=100,
                                     min.node.size = 1,
                                     splitrule="gini")
)
# print OOB error
rfFit$finalModel$prediction.error

# Variable Importance
varImp(rfFit)
plot(varImp(rfFit),top=10)

# Extract the top 10 variables from both results
top_10_rfFit <- row.names(varImp(rfFit))[1:10]


# Normalize the importance scores from random forests
# Compute importance scores using Random Forests (rfFit)
importance_rf <- varImp(rfFit)

# Normalizing function
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
# Normalize the importance scores to a range between 0 and 1
normalized_importance_rf <- normalize(importance_rf$importance)


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set about the DNA methylation levels, called "CpGmeth2Age.rds". The data set has 108 samples and each row
is a different sample. In the first column the age of the sample is indicated, followed by the methylation values in the
other columns. I would like to first remove the ones that have less than 0.1 standard deviation across the samples. Then
I would like to run the random forest regression and plot the importance metrics.
I would also like to split 20 % of the methylation-age data as test data and run elastic net regression on the training
portion to tune parameters and test it on the test portion. How can I do that?
Lastly, I want to run an ensemble model for regression using the caretEnsemble or mlr package and compare the results
with the elastic net and random forest model. Did the test accuracy increase?

completion:
The data set must be read first. We will use `readRDS()` function to read the rds file. Then we will remove the columns
that have less than 0.1 standard deviation across the samples. Then we will train the random forest model by `train()`
with `method = "ranger"`. For the train control we will use cross validation method. We will calculate importance by
setting `importance="permutation"` inside the function `train()` and using function `varImp()`. To plot the importance
metrics, we will plot the top 10 variable importances using `plot()`.

To split 20 % of the data as test data, we will get indices for 80% of the data set and use the indices to separate test
and training data sets. To run elastic net regression on the training portion to tune parameters we need the
`library(glmnet)`. We will set up 10-fold cross validation as the train control, and train the elastic net model with
`train()` function, `method = "glmnet"`. We will test the alpha and lambda values, and the best alpha and lambda values
by cross-validation accuracy will be determined. After the training, we will test the accuracy of trained set. We will
use `predict()` function to predict the age by running our model on the methylation values. Then we will use
`confusionMatrix()` to compare the predicted value(age) with actual age on the first column.

For the last part, we will use the mlr package. We'll start by loading the necessary packages glmnet and mlr. Creating a
task and learner for Elastic Net regression follows, where "Age" becomes the target variable. The "regr.glmnet" learner
sets up the Elastic Net model framework. Then, we'll define a parameter set for Elastic Net hyperparameters and set up
tuning control. We'll tune the Elastic Net model to optimize its performance.

Next, a task for Random Forest regression is created with "Age" as the target. Hyperparameters for Random Forest are
defined, and the model is tuned. We'll manually build a stack learner using Elastic Net and Random Forest models. This
stack learner predicts responses using the "average" method. After training the stack learner and generating predictions,
we'll evaluate its performance using measures like RMSE and R-squared.


```{r,eval=FALSE}  

library(data.table)
library(matrixStats)
library(caret)

# Reading the DNA methylation level data set.
ameth <- readRDS("CpGmeth2Age.rds")

# Remove the variables that have low variation (sd <= 0.1) across samples
ameth <- ameth[,c(TRUE, matrixStats::colSds(as.matrix(ameth[,-1]))>0.1)]
dim(ameth)

# Use random forest regression to predict the age from methylation values

set.seed(18)

# Use cross validation method
trctrl <- trainControl(method = "cv",number=10)

# we will now train random forest model
rfregFit <- train(Age~.,
               data = ameth, 
               method = "ranger",
               trControl=trctrl,
               importance="permutation", 
               tuneGrid = data.frame(mtry=50,
                                     min.node.size = 5,
                                     splitrule="variance")
               )


# Calculate variable importance and plot
varImp(rfregFit)
plot(varImp(rfregFit),top=10)

# 2-----------------------------------------------------------------------------
# Split 20% of the methylation-age data as test data

# get indices for 80% of the data set
ind_train <- createDataPartition(y = ameth[,1], p= 0.8)[[1]]

# seperate test and training sets
training <- ameth[ind_train,]
testing <- ameth[-ind_train,]

# Run elastic net regression on the training portion to tune parameters
set.seed(19)
library(glmnet)

# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# Train elastic net model
enetFit <- train(Age~.,
                 data = training, 
                 method = "glmnet",
                 trControl=trctrl,
                 tuneGrid = data.frame(alpha=0.5,
                                       lambda=seq(0.1,0.7,0.05)))

# best alpha and lambda values by cross-validation accuracy
enetFit$bestTune

# test it on the test portion
class.res <- predict(enetFit,testing[,-1])
class.res_factor <- factor(class.res, levels = levels(factor(testing[,1])))
confusionMatrix(factor(testing[,1]),class.res_factor)$overall[1]

# 3.----------------------------------------------------------------------------

# Run an ensemble model for regression using the caretEnsemble or mlr package
# Compare the results with the elastic net and random forest model
# Did the test accuracy increase?

# Load the required libraries
library(mlr)
library(glmnet)

# 1. Create the task and learner for Elastic Net
task_enet <- makeRegrTask(data = training, target = "Age")
learner_enet <- makeLearner("regr.glmnet")

# 2. Set hyperparameters for Elastic Net learner
param_set_enet <- makeParamSet(
  makeDiscreteParam("alpha", values = 0.5),
  makeNumericParam("lambda", lower = 0.1, upper = 0.7)
)

# 3. Create the tuning control for Elastic Net
ctrl_enet <- makeTuneControlRandom(maxit = 10)

# 4. Tune the Elastic Net model
resampling_enet <- makeResampleDesc("CV", iters = 10)
tune_result_enet <- tuneParams(learner_enet, task_enet, resampling_enet, par.set = param_set_enet, control = ctrl_enet)

# 5. Create the task and learner for Random Forest
task_rf <- makeRegrTask(data = ameth, target = "Age")
learner_rf <- makeLearner("regr.ranger")

# 6. Set hyperparameters for Random Forest learner
param_set_rf <- makeParamSet(
  makeIntegerParam("mtry", lower = 50, upper = 50),
  makeIntegerParam("min.node.size", lower = 5, upper = 5)
)

# 7. Create the tuning control for Random Forest
ctrl_rf <- makeTuneControlRandom(maxit = 10)

# 8. Tune the Random Forest model
resampling_rf <- makeResampleDesc("Bootstrap", iters = 1)
tune_result_rf <- tuneParams(learner_rf, task_rf, resampling_rf, par.set = param_set_rf, control = ctrl_rf)

# 9. Create a random forest model using the best hyperparameters
rf_model <- setHyperPars(makeLearner("regr.ranger"), par.vals = tune_result_rf$x)

# 10. Create a stack learner manually
stack_learner <- makeStackedLearner(
  base.learners = list(learner_enet, rf_model),
  predict.type = "response",
  method = "average"
)

# 11. Train the stack learner
stack_model <- train(stack_learner, task_enet)

# 12. Make predictions using the stack model
predictions <- predict(stack_model, task_enet)

# 13. Evaluate the stack model's performance
perf <- performance(predictions, measures = list(rmse, rsq))
print(perf)

```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set about the DNA methylation levels, called "CpGmeth2Age.txt". The data set has 108 samples and each row
is a different sample. In the first column the age of the sample is indicated, followed by the methylation values in the
other columns. I would like to first remove the ones that have less than 0.1 standard deviation across the samples. Then
I would like to run the random forest regression and plot the importance metrics. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will remove the columns that have less than 0.1 standard deviation across the samples. Then we will train
the random forest model by `train()` with `method = "ranger"`. For the train control we will use cross validation method.
We will calculate importance by setting `importance="permutation"` inside the function `train()` and using function
`varImp()`. To plot the importance metrics, we will plot the top 10 variable importances using `plot()`.


```{r,eval=FALSE}  

library(data.table)
library(matrixStats)
library(caret)

# Reading the DNA methylation level data set.
ameth <- data.table::fread("CpGmeth2Age.txt", data.table = FALSE)

# Remove the variables that have low variation (sd <= 0.1) across samples
ameth <- ameth[,c(TRUE, matrixStats::colSds(as.matrix(ameth[,-1]))>0.1)]
dim(ameth)

# Use random forest regression to predict the age from methylation values

set.seed(18)

# Use cross validation method
trctrl <- trainControl(method = "cv",number=10)

# we will now train random forest model
rfregFit <- train(Age~.,
               data = ameth, 
               method = "ranger",
               trControl=trctrl,
               importance="permutation", 
               tuneGrid = data.frame(mtry=50,
                                     min.node.size = 5,
                                     splitrule="variance")
               )


# Calculate variable importance and plot
varImp(rfregFit)
plot(varImp(rfregFit),top=10)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set about the DNA methylation levels, called "CpGmeth2Age.txt". The data set has 108 samples and each row
is a different sample. In the first column the age of the sample is indicated, followed by the methylation values in the
other columns. I would like to first remove the ones that have less than 0.1 standard deviation across the samples. Then
I would like to split 20 % of the methylation-age data as test data and run elastic net regression on the training
portion to tune parameters and test it on the test portion. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will remove the columns that have less than 0.1 standard deviation across the samples.
To split 20 % of the data as test data, we will get indices for 80% of the data set and use the indices to separate test
and training data sets. To run elastic net regression on the training portion to tune parameters we need the
`library(glmnet)`. We will set up 10-fold cross validation as the train control, and train the elastic net model with
`train()` function, `method = "glmnet"`. We will test the alpha and lambda values, and the best alpha and lambda values
by cross-validation accuracy will be determined. After the training, we will test the accuracy of trained set. We will
use `predict()` function to predict the age by running our model on the methylation values. Then we will use
`confusionMatrix()` to compare the predicted value(age) with actual age on the first column.


```{r,eval=FALSE}  

library(data.table)
library(matrixStats)
library(caret)

# Reading the DNA methylation level data set.
ameth <- data.table::fread("CpGmeth2Age.txt", data.table = FALSE)

# Remove the variables that have low variation (sd <= 0.1) across samples
ameth <- ameth[,c(TRUE, matrixStats::colSds(as.matrix(ameth[,-1]))>0.1)]
dim(ameth)


# -----------------------------------------------------------------------------
# Split 20% of the methylation-age data as test data

# get indices for 80% of the data set
ind_train <- createDataPartition(y = ameth[,1], p= 0.8)[[1]]

# seperate test and training sets
training <- ameth[ind_train,]
testing <- ameth[-ind_train,]

# Run elastic net regression on the training portion to tune parameters
set.seed(19)
library(glmnet)

# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# Train elastic net model
enetFit <- train(Age~.,
                 data = training, 
                 method = "glmnet",
                 trControl=trctrl,
                 tuneGrid = data.frame(alpha=0.5,
                                       lambda=seq(0.1,0.7,0.05)))

# best alpha and lambda values by cross-validation accuracy
enetFit$bestTune

# test it on the test portion
class.res <- predict(enetFit,testing[,-1])
class.res_factor <- factor(class.res, levels = levels(factor(testing[,1])))
confusionMatrix(factor(testing[,1]),class.res_factor)$overall[1]


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set about the DNA methylation levels, called "CpGmeth2Age.txt". The data set has 108 samples and each row
is a different sample. In the first column the age of the sample is indicated, followed by the methylation values in the
other columns. I would like to first remove the ones that have less than 0.1 standard deviation across the samples. 
Then I would like to split 20 % of the methylation-age data as test data and run elastic net regression on the training
portion to tune parameters and test it on the test portion. How can I do that?
Lastly, I want to run an ensemble model for regression using the caretEnsemble or mlr package and compare the results
with the elastic net and random forest model. Did the test accuracy increase?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read files
faster. Then we will remove the columns that have less than 0.1 standard deviation across the samples.
To split 20 % of the data as test data, we will get indices for 80% of the data set and use the indices to separate test
and training data sets. To run elastic net regression on the training portion to tune parameters we need the
`library(glmnet)`. We will set up 10-fold cross validation as the train control, and train the elastic net model with
`train()` function, `method = "glmnet"`. We will test the alpha and lambda values, and the best alpha and lambda values
by cross-validation accuracy will be determined. After the training, we will test the accuracy of trained set. We will
use `predict()` function to predict the age by running our model on the methylation values. Then we will use
`confusionMatrix()` to compare the predicted value(age) with actual age on the first column.

For the last part, we will use the mlr package. We'll start by loading the necessary packages glmnet and mlr.
Creating a task and learner for Elastic Net regression follows, where "Age" becomes the target variable.
The "regr.glmnet" learner sets up the Elastic Net model framework. Then, we'll define a parameter set for Elastic Net
hyperparameters and set up tuning control. We'll tune the Elastic Net model to optimize its performance.
Next, a task for Random Forest regression is created with "Age" as the target. Hyperparameters for Random Forest are
defined, and the model is tuned. We'll manually build a stack learner using Elastic Net and Random Forest models. This
stack learner predicts responses using the "average" method. After training the stack learner and generating predictions,
we'll evaluate its performance using measures like RMSE and R-squared.


```{r,eval=FALSE}  

library(data.table)
library(matrixStats)
library(caret)

# Reading the DNA methylation level data set.
ameth <- data.table::fread("CpGmeth2Age.txt", data.table = FALSE)

# Remove the variables that have low variation (sd <= 0.1) across samples
ameth <- ameth[,c(TRUE, matrixStats::colSds(as.matrix(ameth[,-1]))>0.1)]
dim(ameth)


# 1.-----------------------------------------------------------------------------
# Split 20% of the methylation-age data as test data

# get indices for 80% of the data set
ind_train <- createDataPartition(y = ameth[,1], p= 0.8)[[1]]

# seperate test and training sets
training <- ameth[ind_train,]
testing <- ameth[-ind_train,]

# Run elastic net regression on the training portion to tune parameters
set.seed(19)
library(glmnet)

# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# Train elastic net model
enetFit <- train(Age~.,
                 data = training, 
                 method = "glmnet",
                 trControl=trctrl,
                 tuneGrid = data.frame(alpha=0.5,
                                       lambda=seq(0.1,0.7,0.05)))

# best alpha and lambda values by cross-validation accuracy
enetFit$bestTune

# test it on the test portion
class.res <- predict(enetFit,testing[,-1])
class.res_factor <- factor(class.res, levels = levels(factor(testing[,1])))
confusionMatrix(factor(testing[,1]),class.res_factor)$overall[1]

# 2.----------------------------------------------------------------------------

# Run an ensemble model for regression using the caretEnsemble or mlr package
# Compare the results with the elastic net and random forest model
# Did the test accuracy increase?

# Load the required libraries
library(mlr)
library(glmnet)

# 1. Create the task and learner for Elastic Net
task_enet <- makeRegrTask(data = training, target = "Age")
learner_enet <- makeLearner("regr.glmnet")

# 2. Set hyperparameters for Elastic Net learner
param_set_enet <- makeParamSet(
  makeDiscreteParam("alpha", values = 0.5),
  makeNumericParam("lambda", lower = 0.1, upper = 0.7)
)

# 3. Create the tuning control for Elastic Net
ctrl_enet <- makeTuneControlRandom(maxit = 10)

# 4. Tune the Elastic Net model
resampling_enet <- makeResampleDesc("CV", iters = 10)
tune_result_enet <- tuneParams(learner_enet, task_enet, resampling_enet, par.set = param_set_enet, control = ctrl_enet)

# 5. Create the task and learner for Random Forest
task_rf <- makeRegrTask(data = ameth, target = "Age")
learner_rf <- makeLearner("regr.ranger")

# 6. Set hyperparameters for Random Forest learner
param_set_rf <- makeParamSet(
  makeIntegerParam("mtry", lower = 50, upper = 50),
  makeIntegerParam("min.node.size", lower = 5, upper = 5)
)

# 7. Create the tuning control for Random Forest
ctrl_rf <- makeTuneControlRandom(maxit = 10)

# 8. Tune the Random Forest model
resampling_rf <- makeResampleDesc("Bootstrap", iters = 1)
tune_result_rf <- tuneParams(learner_rf, task_rf, resampling_rf, par.set = param_set_rf, control = ctrl_rf)

# 9. Create a random forest model using the best hyperparameters
rf_model <- setHyperPars(makeLearner("regr.ranger"), par.vals = tune_result_rf$x)

# 10. Create a stack learner manually
stack_learner <- makeStackedLearner(
  base.learners = list(learner_enet, rf_model),
  predict.type = "response",
  method = "average"
)

# 11. Train the stack learner
stack_model <- train(stack_learner, task_enet)

# 12. Make predictions using the stack model
predictions <- predict(stack_model, task_enet)

# 13. Evaluate the stack model's performance
perf <- performance(predictions, measures = list(rmse, rsq))
print(perf)


```
--- TASK END
