---
title: "prompt/completion pairs chapter 5"
author: "authors"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.
I want to read both tables and see the head of the tables. For the gene expression data set show only the head from
column 1 to 5. I also want to know the dimensions of both tables.

Then I want to preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero
variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors, center
the data and remove higly correlated variables with cutoff = 0.9. How can I accomplisch these tasks?

Furthermore, I want to take care of the missing values by removing the missings, and imputing with median imputation and
knn-imputation. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables.

We can check if there are missings with `anyNA()` function. To remove the missings, we will run the code that checks
which values are NA in the matrix, then runs a column sum and keeps everything where the column sum is equal to 0.
The column sums where there are NA values will be higher than 0 depending on how many NA values there are in a column.
To impute the missings, we will first do the median imputation using `caret::preProcess()` with method specified as
`method = "medianImpute"`. Then we can also do the knn-imputation using the same function with `method = "knnImpute"`.


```{r,eval=FALSE}  

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# Show the head of the gene expression data table, from column 1 to 5
head(gexp[,1:5])
# Show the dimension of the gene expression data table
dim(gexp)

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Show the head of the patient data table
head(patient)
# Show the dimension of the patient data table
dim(patient)

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)


# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)


# check if there are NA values
missing_tgexp <- tgexp
anyNA(missing_tgexp)

# Remove missing genes
gexpnoNA <- missing_tgexp[ , colSums(is.na(missing_tgexp)) == 0]

# Median Imputation
mImpute <- preProcess(missing_tgexp,method="medianImpute")
imputedGexp <- predict(mImpute,missing_tgexp)

# Imputing the missing values based on the nearest neighbors of the samples
library(RANN)
knnImpute <- preProcess(missing_tgexp,method="knnImpute")
knnimputedGexp <- predict(knnImpute,missing_tgexp)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I want to preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero 
variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. How can I accomplisch these tasks?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables.


```{r,eval=FALSE}  

# Data Read ----------------------------------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Filtering ---------------------------------------------------------------------------

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I want to preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero
variation for the columns at least 85% of the values are the same. How can I accomplisch these tasks?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. We will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same.


```{r,eval=FALSE}  

# Data Read ------------------------------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Non-Zero Filtering ----------------------------------------------------------

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.

I want to preprocess the data by taking the log and filtering the data. By filtering, I want to take the top 1000
variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. How can I accomplisch these
tasks?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will choose to take the
top 1000 variable predictors, which have the biggest variances. Then we will center the data, and create a filter to
remove higly correlated variables.


```{r,eval=FALSE}  

# Data Read ----------------------------------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Filtering ---------------------------------------------------------------------------

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.
I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero
variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors, center
the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to merge
those two tables, so that the rows represent individual patients and the first column states CIMP subtype, followed by
the gene expressions. How can I accomplish these tasks?

Then I would like to split 30% of the data as the test set, and rest of it as training set. With the training set, I
would like to get k-NN prediction on the training data with `k = 5` and calculate the accuracy metrics by comparing the
predicted labels to real labels. Then I want to see the test set accuracy on the predicted and real classes. Lastly, I
would like to create a ROC curve for the k-NN probabilities on the test set, plot the curve and return the area under the
curve. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column.

Now that the response variable is merged with our dataset, we can split it into test and training sets with the
`caret::createPartition()` function. We will first get the indices for 70% of the data set, and use that indices to
indicate the training set. Rest will be our testing set. We will then run the k-NN algorithm with the `caret::knn3()`
function with `k = 5`, setting the x (training set) as the training matrix without the first column (subtypes) and y
(class labels) as the first column (subtypes). To calculate the training set accuracy we will first get the predictions
by using the `predict()` function on our training set. Then we will use the function `caret::confusionMatrix()` and
compare the predicted labels to real labels. For the test set accuracy we will repeat the process with the testing set.

For the ROC curve, we will first get the k-NN class prediction probabilities on the test set with the `predict()`
function. Then we will use the information on the `pROC::roc()` function with specified response and predictor.
Currently, our levels for the first column of testing set is "CIMP" and "noCIMP". Since the function`roc()` assumes that
the second class is the class of interest, so we will reverse it for the function argument `levels` so that the levels
would be "noCIMP" and "CIMP". Then we can plot the ROC curve with `plot()` and return the area under the curve with
`pROC::auc()` function.

```{r,eval=FALSE}  

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Training with knn----------------------------------------

# set the random number seed for reproducibility 
set.seed(3031) 

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# get k-NN prediction on the training data itself, with k=5
knnFit <- knn3(x=as.matrix(training[,-1]), # training set
            y=as.factor(training[,1]), # training set class labels
            k = 5)

# predictions on the training set
trainPred <- predict(knnFit,training[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(training[,1]),reference=trainPred)

# predictions on the test set, return class labels
testPred <- predict(knnFit,testing[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(testing[,1]),reference=testPred)

library(pROC)

# get k-NN class probabilities
# prediction probabilities on the test set
testProbs <- predict(knnFit,testing[,-1])

# get the roc curve
rocCurve <- pROC::roc(response = testing[,1],
                      predictor = testProbs[,1],
                      ## This function assumes that the second
                      ## class is the class of interest, so we
                      ## reverse the labels.
                      levels = rev(levels(as.factor(testing[,1]))))
# plot the curve
plot(rocCurve, legacy.axes = TRUE)

# return area under the curve
pROC::auc(rocCurve)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.
I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero
variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors, center
the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to merge
those two tables, so that the rows represent individual patients and the first column states CIMP subtype, followed by
the gene expressions. How can I accomplish these tasks?

Then I would like to split 30% of the data as the test set, and rest of it as training set. With the training set, I
would like to get k-NN prediction on the training data with `k = 5` and calculate the accuracy metrics by comparing the
predicted labels to real labels. Then I want to see the test set accuracy on the predicted and real classes. 

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column.

Now that the response variable is merged with our dataset, we can split it into test and training sets with the
`caret::createPartition()` function. We will first get the indices for 70% of the data set, and use that indices to
indicate the training set. Rest will be our testing set. We will then run the k-NN algorithm with the `caret::knn3()`
function with `k = 5`, setting the x (training set) as the training matrix without the first column (subtypes) and y
(class labels) as the first column (subtypes). To calculate the training set accuracy we will first get the predictions
by using the `predict()` function on our training set. Then we will use the function `caret::confusionMatrix()` and
compare the predicted labels to real labels. For the test set accuracy we will repeat the process with the testing set.


```{r,eval=FALSE}  

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Training with knn----------------------------------------

# set the random number seed for reproducibility 
set.seed(3031) 

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# get k-NN prediction on the training data itself, with k=5
knnFit <- knn3(x=as.matrix(training[,-1]), # training set
            y=as.factor(training[,1]), # training set class labels
            k = 5)

# predictions on the training set
trainPred <- predict(knnFit,training[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(training[,1]),reference=trainPred)

# predictions on the test set, return class labels
testPred <- predict(knnFit,testing[,-1],type="class")

# compare the predicted labels to real labels
# get different performance metrics
confusionMatrix(data=as.factor(testing[,1]),reference=testPred)


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP existence.
I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove near zero
variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors, center
the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to merge
those two tables, so that the rows represent individual patients and the first column states CIMP subtype, followed by
the gene expressions. How can I accomplish these tasks?

Then I would like to split 30% of the data as the test set, and rest of it as training set. With the training set, I
would like to get k-NN prediction on the training data with `k = 5`. Then I would like to create a ROC curve for the k-NN
probabilities on the test set, plot the curve and return the area under the curve. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column.

Now that the response variable is merged with our dataset, we can split it into test and training sets with the
`caret::createPartition()` function. We will first get the indices for 70% of the data set, and use that indices to
indicate the training set. Rest will be our testing set. We will then run the k-NN algorithm with the `caret::knn3()`
function with `k = 5`, setting the x (training set) as the training matrix without the first column (subtypes) and y
(class labels) as the first column (subtypes). 

For the ROC curve, we will first get the k-NN class prediction probabilities on the test set with the `predict()`
function. Then we will use the information on the `pROC::roc()` function with specified response and predictor.
Currently, our levels for the first column of testing set is "CIMP" and "noCIMP". Since the function`roc()` assumes that
the second class is the class of interest, so we will reverse it for the function argument `levels` so that the levels
would be "noCIMP" and "CIMP". Then we can plot the ROC curve with `plot()` and return the area under the curve with
`pROC::auc()` function.


```{r,eval=FALSE}  

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Training with knn----------------------------------------

# set the random number seed for reproducibility 
set.seed(3031) 

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# get k-NN prediction on the training data itself, with k=5
knnFit <- knn3(x=as.matrix(training[,-1]), # training set
            y=as.factor(training[,1]), # training set class labels
            k = 5)

# ROC Curve---------------------------------------------------------

library(pROC)

# get k-NN class probabilities
# prediction probabilities on the test set
testProbs <- predict(knnFit,testing[,-1])

# get the roc curve
rocCurve <- pROC::roc(response = testing[,1],
                      predictor = testProbs[,1],
                      ## This function assumes that the second
                      ## class is the class of interest, so we
                      ## reverse the labels.
                      levels = rev(levels(as.factor(testing[,1]))))
# plot the curve
plot(rocCurve, legacy.axes = TRUE)

# return area under the curve
pROC::auc(rocCurve)

```
--- TASK END



### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

I want to run k-NN algorithm. However, I don't know which k to pick for the best result. I would like to try k = 1 to 12
and find out which k value I should use. How can I calculate training error and test error for each k values and plot
them with a smooth line for the trend? I also want to do a cross-validation on the data. How do I run a k-NN model with
cross validation? How do I use cross-validation to train models for different k values? I also want to try bootstrap
resampling and check the prediction error that way. How can I do that?

I want to run the random forest algorithm using `caret()` package interface on the data set. How do I do that? How can I
get the prediction error and plot the variable importance for the top 10 variables?

I also want to use one predictor variable - the expression of one gene - to classify tumor samples to "CIMP" and
"noCIMP" subtypes. This time I want to use PDPN gene expression to model a binary response variable by logistic
regression. How can I do that? I also want to create a data frame to be plotted in a sigmoid curve together with the
points from original data, to see how good was our logistic regression model to predict the subtype. In addition, I want
to assess the performance of the model with the test and training set. Lastly, what happens when I use all genes as
predictor variables in the logistic regression model? What are the training and test accuracy?

To apply regularization, I want to use elastic net method on the tumor subtype prediction problem. How can I train the
elastic net model with `alpha = 0.5`? Let cross-validation select the best lambda. What are the training and test
accuracy? Using the fitted elastic net model, plot the variable importance of top 10 variables.

Lastly, I want to do ensemble learning by combining k-NN, random forest, and elastic net models and making class
predictions based on the votes from those different models. How can I get predictions for three different models on the
test set, use majority voting to decide on the class label, and then check performance?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

To figure the appropriate k-value out, we will first set k as integer from 1 to 12 and set two empty vectors to each
store the training and testing errors. Then we will make a loop, so that for i in k, we will calculate predictions and
store errors on the training and testing set for each k. After running through the loop we will then have two vectors
with training and test errors, so we can simply plot them with scatterplot and smooth lines.

To run a k-NN model with cross-validation, we will use `caret::train()` function. This function will use
cross-validation to train models for different k values. To use this function, we need a control method for the
training. For this, we will use `trainControl` function. This time we will set up a 10-fold cross validation. Then we
will use the `train()` function and every k value will be trained and tested with cross-validation to estimate
prediction performance for each k. We will then plot the cross-validation error. To use bootstrap resampling we will use
the same `caret::train()` function, the only difference would be the control method. We will set up the control method
with `trainControl()` function with method set to bootstrap resampling.

We will use the `caret` package interface to the `ranger` random forest package. This is a fast implementation of the
original random forest algorithm. For the control method, we will do no resampling based prediction error and leave the
`method = "none"`. For simplicity we will just set `mtry = 100` and `min.node.size = 1`, but you can also find out the
best `mtry` value based on the cross-validation performance and tune the mininum node size of terminal nodes to control
the depth of the trees grown. We will then train random forest model with `train()` function with `method = "ranger"`. 
To print out the prediction error, we can simply point to the prediction error of the final trained model. Since random
forests come with built-in variable importance metrics, we can use `caret::varImp()` function to access the importance
values and plot them using the `plot()` function.

To fit a logistic regression model, we will use the formula interface in `caret`, where we will supply the names of the
response and predictor variables in a formula. In this case, we will be using a core R function, `glm()`, from the
`stats` package. In the function `train()`, we will specify that the subtype is in relation to PDPN, and define the
logistic regression by setting `method = "glm"` and `family = "binomial"`. After training the model, we will create a
data to plot the sigmoid curve and predict the probabilities for the simulated data based on the model. We will also
plot the curve with the simulated data and points from the original data. To assess the performance of our model with
the test and training set, we will use `caret::predict()` and `caret::confusionMatrix()` functions. Lastly, we will try
fitting the model with all genes as predictor variables. For this, we will specify that `subtype ~ .` in the train
function. We will again use the `predict()` and `confusionMatrix()` to find out the training and test accuracy.

To do a elastic net regularization, we will first load the package `glmnet`. Then we will set the train control method
with 10-fold cross validation. Now we can use `caret::train()` function in which `method = "glmnet"` and `alpha = 0.5`.
For lambda parameters we will give it three values to try. After training the model, we can point to the `bestTune` to
see the best alpha and lambda values by cross-validation accuracy. Then we will check the training and test accuracy
with `caret::predict()` and `caret::confusionMatrix()` functions. The `caret::varImp()` function uses the coefficients
to rank the variables from the elastic net model and we are going to plot the top 10 important variables which are
normalized to the importance of the most important variable.

To do the ensemble learning with combination of k-NN, random forest and elastic net models, we will first get the
predictions for each model using the fitted model we obtained. Then we will do voting for class labels, which will find
out the most frequent class label per row. Then we will check performance by using `caret::confusionMatrix()` function
on our test set, with reference being our voted class labels.

```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Model Tuning-------------------------------------------------------------

set.seed(101)
# set k values
k <- 1:12
# set vector for training errors
trainErr <- c()

for( i in k){
  
  knnFit <- knn3(x=training[,-1], # training set
              y=as.factor(training[,1]), # training set class labels
              k=i)
  
  # predictions on the training set
  class.res=predict(knnFit,training[,-1],type="class")
  # training error
  trainErr[i] <- 1-confusionMatrix(as.factor(training[,1]),
                                   class.res)$overall[1]
}

set.seed(31)
k <- 1:12
# set vector for testing errors
testErr <- c()
for( i in k){
  
  knnFit <- knn3(x=training[,-1], # training set
              y=as.factor(training[,1]), # training set class labels
              k=i)
  
  # predictions on the testing set
  class.res <- predict(knnFit,testing[,-1],type="class")
  # testing error
  testErr[i] <- 1-confusionMatrix(as.factor(testing[,1]),
                               class.res)$overall[1]
}
# plot training error
plot(k,trainErr,type="p",col="#CC0000",
     ylim=c(0.000,0.08),
     ylab="prediction error (1-accuracy)",pch=19)
# add a smooth line for the trend
lines(loess.smooth(x=k, trainErr,degree=2), col="#CC0000")

# plot test error
points(k,testErr,col="#00CC66",pch=19) 
lines(loess.smooth(x=k,testErr,degree=2), col="#00CC66")
# add legend
legend("bottomright",fill=c("#CC0000","#00CC66"),
       legend=c("training","test"),bty="n")

# Cross validation (knn)--------------------------------------------------------

set.seed(17)
# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# we will now train k-NN model
knn_fit <- train(subtype~., data = training, 
                 method = "knn",
                 trControl=trctrl,
                 tuneGrid = data.frame(k=1:12))

# best k value by cross-validation accuracy
knn_fit$bestTune

# plot k vs prediction error
plot(x=1:12,1-knn_fit$results[,2],pch=19,
     ylab="prediction error",xlab="k")
lines(loess.smooth(x=1:12,1-knn_fit$results[,2],degree=2),
      col="#CC0000")

# Bootstrapping (knn)-------------------------------------------------

set.seed(17)
# this method controls everything about training
# we will just set up 100 bootstrap samples and for each 
# bootstrap OOB samples to test the error
trctrl <- trainControl(method = "boot",number=20,
                       returnResamp="all")

# we will now train k-NN model
knn_fit_b <- train(subtype~., data = training, 
                 method = "knn",
                 trControl=trctrl,
                 tuneGrid = data.frame(k=1:12))

# Trees and Forests--------------------------------------------------------

set.seed(17)

# we will do no resampling based prediction error
# although it is advised to do so even for random forests
trctrl <- trainControl(method = "none")

# we will now train random forest model
rfFit <- train(subtype~., 
               data = training, 
               method = "ranger",
               trControl=trctrl,
               importance="permutation", # calculate importance
               tuneGrid = data.frame(mtry=100,
                                     min.node.size = 1,
                                     splitrule="gini")
)
# print OOB error
rfFit$finalModel$prediction.error

plot(varImp(rfFit),top=10)

# Logistic regression and regularization-----------------------------------

# fit logistic regression model
# method and family defines the type of regression
# in this case these arguments mean that we are doing logistic regression
lrFit <- train(subtype ~ PDPN,  
              data=training, trControl=trainControl("none"),
              method="glm", family="binomial")

# create data to plot the sigmoid curve
newdat <- data.frame(PDPN=seq(min(training$PDPN),
                              max(training$PDPN),len=100))

# predict probabilities for the simulated data
newdat$subtype <- predict(lrFit, newdata=newdat, type="prob")[,1]

# plot the sigmoid curve and the training data
plot(ifelse(subtype=="CIMP",1,0) ~ PDPN, 
     data=training, col="red4",
     ylab="subtype as 0 or 1", xlab="PDPN expression")
lines(subtype ~ PDPN, newdat, col="green4", lwd=2)

# training accuracy 
class.res <- predict(lrFit,training[,-1])
confusionMatrix(as.factor(training[,1]),class.res)$overall[1]

# test accuracy 
class.res <- predict(lrFit,testing[,-1])
confusionMatrix(as.factor(testing[,1]),class.res)$overall[1]


lrFit2 <- train(subtype ~ .,  
               data=training, 
               # no model tuning with sampling
               trControl=trainControl("none"),
               method="glm", family="binomial")

# training accuracy 
class.res <- predict(lrFit2,training[,-1])
confusionMatrix(as.factor(training[,1]),class.res)$overall[1]

# test accuracy 
class.res <- predict(lrFit2,testing[,-1])
confusionMatrix(as.factor(testing[,1]),class.res)$overall[1]

# Regularization (elastic net model) ----------------------------------------------------------

set.seed(17)
library(glmnet)

# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# we will now train elastic net model
enetFit <- train(subtype~., data = training, 
                 method = "glmnet",
                 trControl=trctrl,
                 # alpha and lambda paramters to try
                 tuneGrid = data.frame(alpha=0.5,
                                       lambda=seq(0.1,0.7,0.05)))

# best alpha and lambda values by cross-validation accuracy
enetFit$bestTune

# training accuracy 
class.res <- predict(enetFit,training[,-1])
confusionMatrix(as.factor(training[,1]),class.res)$overall[1]

# test accuracy 
class.res <- predict(enetFit,testing[,-1])
confusionMatrix(as.factor(testing[,1]),class.res)$overall[1]

# Variable importance
plot(varImp(enetFit),top=10)

# Ensemble learning---------------------------------------------------------

# predict with k-NN model
knnPred <- as.character(predict(knnFit,testing[,-1],type="class"))
# predict with elastic Net model
enetPred <- as.character(predict(enetFit,testing[,-1]))
# predict with random forest model
rfPred <- as.character(predict(rfFit,testing[,-1]))

# do voting for class labels
# code finds the most frequent class label per row
votingPred <- apply(cbind(knnPred,enetPred,rfPred),1,
                 function(x) names(which.max(table(x))))

# check accuracy
confusionMatrix(data=as.factor(testing[,1]),
                reference=as.factor(votingPred))$overall[1]

```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

I want to run k-NN algorithm. However, I don't know which k to pick for the best result. I would like to try k = 1 to 12
and find out which k value I should use. How can I calculate training error and test error for each k values and plot
them with a smooth line for the trend? 

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

To figure the appropriate k-value out, we will first set k as integer from 1 to 12 and set two empty vectors to each
store the training and testing errors. Then we will make a loop, so that for i in k, we will calculate predictions and
store errors on the training and testing set for each k. After running through the loop we will then have two vectors
with training and test errors, so we can simply plot them with scatterplot and smooth lines.


```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Model Tuning-------------------------------------------------------------

set.seed(101)

# set k values
k <- 1:12

# set vector for training errors
trainErr <- c()

for( i in k){
  
  knnFit <- knn3(x=training[,-1], # training set
              y=as.factor(training[,1]), # training set class labels
              k=i)
  
  # predictions on the training set
  class.res=predict(knnFit,training[,-1],type="class")
  # training error
  trainErr[i] <- 1-confusionMatrix(as.factor(training[,1]),
                                   class.res)$overall[1]
}

set.seed(31)
k <- 1:12
# set vector for testing errors
testErr <- c()
for( i in k){
  
  knnFit <- knn3(x=training[,-1], # training set
              y=as.factor(training[,1]), # training set class labels
              k=i)
  
  # predictions on the testing set
  class.res <- predict(knnFit,testing[,-1],type="class")
  # testing error
  testErr[i] <- 1-confusionMatrix(as.factor(testing[,1]),
                               class.res)$overall[1]
}

# plot training error
plot(k,trainErr,type="p",col="#CC0000",
     ylim=c(0.000,0.08),
     ylab="prediction error (1-accuracy)",pch=19)
# add a smooth line for the trend
lines(loess.smooth(x=k, trainErr,degree=2), col="#CC0000")

# plot test error
points(k,testErr,col="#00CC66",pch=19) 
lines(loess.smooth(x=k,testErr,degree=2), col="#00CC66")
# add legend
legend("bottomright",fill=c("#CC0000","#00CC66"),
       legend=c("training","test"),bty="n")


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

I want to run a k-NN model with cross-validation on the data. How do I run a k-NN model with cross validation?
How do I use cross-validation to train models for different k values?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

To run a k-NN model with cross-validation, we will use `caret::train()` function. This function will use
cross-validation to train models for different k values. To use this function, we need a control method for the
training. For this, we will use `trainControl` function. This time we will set up a 10-fold cross validation. Then we
will use the `train()` function and every k value will be trained and tested with cross-validation to estimate
prediction performance for each k. We will then plot the cross-validation error. 


```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Cross validation (knn)--------------------------------------------------------

set.seed(17)
# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# we will now train k-NN model
knn_fit <- train(subtype~., data = training, 
                 method = "knn",
                 trControl=trctrl,
                 tuneGrid = data.frame(k=1:12))

# best k value by cross-validation accuracy
knn_fit$bestTune

# plot k vs prediction error
plot(x=1:12,1-knn_fit$results[,2],pch=19,
     ylab="prediction error",xlab="k")
lines(loess.smooth(x=1:12,1-knn_fit$results[,2],degree=2),
      col="#CC0000")


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

I want to run k-NN algorithm with bootstrap resampling and check the prediction error that way. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

To use k-NN algorithm with bootstrap resampling we will use the `caret::train()` function. We will set up the control method with `trainControl()` function with method set to bootstrap resampling.


```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Bootstrapping (knn)-------------------------------------------------

set.seed(17)
# this method controls everything about training
# we will just set up 20 bootstrap samples and for each 
# bootstrap OOB samples to test the error
trctrl <- trainControl(method = "boot",number=20,
                       returnResamp="all")

# we will now train k-NN model
knn_fit_b <- train(subtype~., data = training, 
                 method = "knn",
                 trControl=trctrl,
                 tuneGrid = data.frame(k=1:12))


```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

I want to run the random forest algorithm using `caret()` package interface on the data set. How do I do that? How can I
get the prediction error and plot the variable importance for the top 10 variables?


completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

We will use the `caret` package interface to the `ranger` random forest package. This is a fast implementation of the
original random forest algorithm. For the control method, we will do no resampling based prediction error and leave the
`method = "none"`. For simplicity we will just set `mtry = 100` and `min.node.size = 1`, but you can also find out the
best `mtry` value based on the cross-validation performance and tune the mininum node size of terminal nodes to control
the depth of the trees grown. We will then train random forest model with `train()` function with `method = "ranger"`. 
To print out the prediction error, we can simply point to the prediction error of the final trained model. Since random
forests come with built-in variable importance metrics, we can use `caret::varImp()` function to access the importance
values and plot them using the `plot()` function.


```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]


# Trees and Forests--------------------------------------------------------

set.seed(17)

# we will do no resampling based prediction error
# although it is advised to do so even for random forests
trctrl <- trainControl(method = "none")

# we will now train random forest model
rfFit <- train(subtype~., 
               data = training, 
               method = "ranger",
               trControl=trctrl,
               importance="permutation", # calculate importance
               tuneGrid = data.frame(mtry=100,
                                     min.node.size = 1,
                                     splitrule="gini")
)
# print OOB error
rfFit$finalModel$prediction.error

plot(varImp(rfFit),top=10)


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

I want to use one predictor variable - the expression of one gene - to classify tumor samples to "CIMP" and
"noCIMP" subtypes. This time I want to use PDPN gene expression to model a binary response variable by logistic
regression. How can I do that? I also want to create a data frame to be plotted in a sigmoid curve together with the
points from original data, to see how good was our logistic regression model to predict the subtype. In addition, I want
to assess the performance of the model with the test and training set. Lastly, what happens when I use all genes as
predictor variables in the logistic regression model? What are the training and test accuracy?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

To fit a logistic regression model, we will use the formula interface in `caret`, where we will supply the names of the
response and predictor variables in a formula. In this case, we will be using a core R function, `glm()`, from the
`stats` package. In the function `train()`, we will specify that the subtype is in relation to PDPN, and define the
logistic regression by setting `method = "glm"` and `family = "binomial"`. After training the model, we will create a
data to plot the sigmoid curve and predict the probabilities for the simulated data based on the model. We will also
plot the curve with the simulated data and points from the original data. To assess the performance of our model with
the test and training set, we will use `caret::predict()` and `caret::confusionMatrix()` functions. Lastly, we will try
fitting the model with all genes as predictor variables. For this, we will specify that `subtype ~ .` in the train
function. We will again use the `predict()` and `confusionMatrix()` to find out the training and test accuracy.


```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Logistic regression and regularization-----------------------------------

# fit logistic regression model
# method and family defines the type of regression
# in this case these arguments mean that we are doing logistic regression
lrFit <- train(subtype ~ PDPN,  
              data=training, trControl=trainControl("none"),
              method="glm", family="binomial")

# create data to plot the sigmoid curve
newdat <- data.frame(PDPN=seq(min(training$PDPN),
                              max(training$PDPN),len=100))

# predict probabilities for the simulated data
newdat$subtype <- predict(lrFit, newdata=newdat, type="prob")[,1]

# plot the sigmoid curve and the training data
plot(ifelse(subtype=="CIMP",1,0) ~ PDPN, 
     data=training, col="red4",
     ylab="subtype as 0 or 1", xlab="PDPN expression")
lines(subtype ~ PDPN, newdat, col="green4", lwd=2)

# training accuracy 
class.res <- predict(lrFit,training[,-1])
confusionMatrix(as.factor(training[,1]),class.res)$overall[1]

# test accuracy 
class.res <- predict(lrFit,testing[,-1])
confusionMatrix(as.factor(testing[,1]),class.res)$overall[1]


lrFit2 <- train(subtype ~ .,  
               data=training, 
               # no model tuning with sampling
               trControl=trainControl("none"),
               method="glm", family="binomial")

# training accuracy 
class.res <- predict(lrFit2,training[,-1])
confusionMatrix(as.factor(training[,1]),class.res)$overall[1]

# test accuracy 
class.res <- predict(lrFit2,testing[,-1])
confusionMatrix(as.factor(testing[,1]),class.res)$overall[1]


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

To apply regularization, I want to use elastic net method on the tumor subtype prediction problem. How can I train the
elastic net model with `alpha = 0.5`? Let cross-validation select the best lambda. What are the training and test
accuracy? Using the fitted elastic net model, plot the variable importance of top 10 variables.

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

To do a elastic net regularization, we will first load the package `glmnet`. Then we will set the train control method
with 10-fold cross validation. Now we can use `caret::train()` function in which `method = "glmnet"` and `alpha = 0.5`.
For lambda parameters we will give it three values to try. After training the model, we can point to the `bestTune` to
see the best alpha and lambda values by cross-validation accuracy. Then we will check the training and test accuracy
with `caret::predict()` and `caret::confusionMatrix()` functions. The `caret::varImp()` function uses the coefficients
to rank the variables from the elastic net model and we are going to plot the top 10 important variables which are
normalized to the importance of the most important variable.


```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Regularization (elastic net model) ----------------------------------------------------------

set.seed(17)
library(glmnet)

# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# we will now train elastic net model
enetFit <- train(subtype~., data = training, 
                 method = "glmnet",
                 trControl=trctrl,
                 # alpha and lambda paramters to try
                 tuneGrid = data.frame(alpha=0.5,
                                       lambda=seq(0.1,0.7,0.05)))

# best alpha and lambda values by cross-validation accuracy
enetFit$bestTune

# training accuracy 
class.res <- predict(enetFit,training[,-1])
confusionMatrix(as.factor(training[,1]),class.res)$overall[1]

# test accuracy 
class.res <- predict(enetFit,testing[,-1])
confusionMatrix(as.factor(testing[,1]),class.res)$overall[1]

# Variable importance
plot(varImp(enetFit),top=10)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.rds", which indicates whether the patient has CIMP(CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to remove
near zero variation for the columns at least 85% of the values are the same, then take the top 1000 variable predictors,
center the data and remove higly correlated variables with cutoff = 0.9. After preprocessing the data, I would like to
merge those two tables, so that the rows represent individual patients and the first column states CIMP subtype,
followed by the gene expressions. Then I would like to split 30% of the data as the test set, and rest of it as training
set. How can I accomplish these tasks?

I want to run k-NN algorithm. However, I don't know which k to pick for the best result. I would like to try k = 1 to 12
and find out which k value I should use. How can I calculate training error and test error for each k values and plot
them with a smooth line for the trend? I also want to do a cross-validation on the data. How do I run a k-NN model with
cross validation? How do I use cross-validation to train models for different k values? I also want to try bootstrap
resampling and check the prediction error that way. How can I do that?

I want to run the random forest algorithm using `caret()` package interface on the data set. How do I do that? How can I
get the prediction error and plot the variable importance for the top 10 variables?

I also want to use one predictor variable - the expression of one gene - to classify tumor samples to "CIMP" and
"noCIMP" subtypes. This time I want to use PDPN gene expression to model a binary response variable by logistic
regression. How can I do that? I also want to create a data frame to be plotted in a sigmoid curve together with the
points from original data, to see how good was our logistic regression model to predict the subtype. In addition, I want
to assess the performance of the model with the test and training set. Lastly, what happens when I use all genes as
predictor variables in the logistic regression model? What are the training and test accuracy?

To apply regularization, I want to use elastic net method on the tumor subtype prediction problem. How can I train the
elastic net model with `alpha = 0.5`? Let cross-validation select the best lambda. What are the training and test
accuracy? Using the fitted elastic net model, plot the variable importance of top 10 variables.

Lastly, I want to do ensemble learning by combining k-NN, random forest, and elastic net models and making class
predictions based on the votes from those different models. How can I get predictions for three different models on the
test set, use majority voting to decide on the class label, and then check performance?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will create a filter
with `preProcess()` function and apply it to remove near zero variation for the columns at least 85% of the values are
the same. Then we will choose to take the top 1000 variable predictors, which have the biggest variances. Then we will
center the data, and create a filter to remove higly correlated variables. Since we already transposed the gene
expression data, we can simply merge the two tables by row names, which are sample-IDs by the base function `merge()`.
Using the `merge()` function in accordance with row names causes that an extra column called "Row.names" is added at the
left, so we will push the sample ids back to the row names and delete the first column. We can split the data set into
test and training sets with the `caret::createPartition()` function. We will first get the indices for 70% of the data
set, and use that indices to indicate the training set. Rest will be our testing set.

To do the ensemble learning with combination of k-NN, random forest and elastic net models, we will first get run each
algorithm to get the fitted model. For the k-NN model we will use `caret::knn3()` function and set `k = 5`.

For the random forest model we will use the `caret` package interface to the `ranger` random forest package. This is a
fast implementation of the original random forest algorithm. For the control method, we will do no resampling based
prediction error and leave the `method = "none"`. For simplicity we will just set `mtry = 100` and `min.node.size = 1`,
but you can also find out the best `mtry` value based on the cross-validation performance and tune the mininum node size
of terminal nodes to control the depth of the trees grown. We will then train random forest model with `train()` function
with `method = "ranger"`. To print out the prediction error, we can simply point to the prediction error of the final
trained model.

To do a elastic net regularization, we will first load the package `glmnet`. Then we will set the train control method
with 10-fold cross validation. Now we can use `caret::train()` function in which `method = "glmnet"` and `alpha = 0.5`.
For lambda parameters we will give it three values to try.

To do the ensemble learning with combination of k-NN, random forest and elastic net models, we will first get the
predictions for each model using the fitted model we obtained. Then we will do voting for class labels, which will find
out the most frequent class label per row. Then we will check performance by using `caret::confusionMatrix()` function
on our test set, with reference being our voted class labels.

```{r,eval=FALSE} 

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms 
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Create a filter to remove near zero variation for the columns at least 85% of the values are the same
nzv <- preProcess(tgexp,method="nzv",uniqueCut = 15)

# apply the filter using "predict" function
# return the filtered dataset and assign it to nzv_tgexp
tgexp <- predict(nzv,tgexp)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# k-NN -------------------------------------------------------------

library(caret)

knnFit=knn3(x=training[,-1], # training set
            y=as.factor(training[,1]), # training set class labels
            k=5)

# Trees and Forests--------------------------------------------------------

set.seed(17)

# we will do no resampling based prediction error
# although it is advised to do so even for random forests
trctrl <- trainControl(method = "none")

# we will now train random forest model
rfFit <- train(subtype~., 
               data = training, 
               method = "ranger",
               trControl=trctrl,
               importance="permutation", # calculate importance
               tuneGrid = data.frame(mtry=100,
                                     min.node.size = 1,
                                     splitrule="gini")
)

# Regularization (elastic net model) ----------------------------------------------------------

set.seed(17)
library(glmnet)

# this method controls everything about training
# we will just set up 10-fold cross validation
trctrl <- trainControl(method = "cv",number=10)

# we will now train elastic net model
enetFit <- train(subtype~., data = training, 
                 method = "glmnet",
                 trControl=trctrl,
                 # alpha and lambda paramters to try
                 tuneGrid = data.frame(alpha=0.5,
                                       lambda=seq(0.1,0.7,0.05)))

# Ensemble learning---------------------------------------------------------

# predict with k-NN model
knnPred <- as.character(predict(knnFit,testing[,-1],type="class"))
# predict with elastic Net model
enetPred <- as.character(predict(enetFit,testing[,-1]))
# predict with random forest model
rfPred <- as.character(predict(rfFit,testing[,-1]))

# do voting for class labels
# code finds the most frequent class label per row
votingPred <- apply(cbind(knnPred,enetPred,rfPred),1,
                 function(x) names(which.max(table(x))))

# check accuracy
confusionMatrix(data=as.factor(testing[,1]),
                reference=as.factor(votingPred))$overall[1]

```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to take the
top 1000 variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. After
preprocessing the data, I would like to merge those two tables, so that the rows represent individual patients and the
first column states CIMP subtype, followed by the gene expressions. Then I would like to split 30% of the data as the
test set, and rest of it as training set. How can I accomplish these tasks?

After preprocessing the data, I want to use various supervised machine learning algorithms on this cancer subtype
classification data. First I want to do gradient boosting and try different learning rate parameters. In this instance,
I also want to subsample 50% of the dataset before training each tree. How can I do that? How can I know the best
parameters by cross-validation accuracy? Then, I would like to try the support vector machines (SVM) method to train the
model. Lastly, I want to train the neural network on the data set. For the neural network I want to try 1 or 2 nodes in
the hidden layer without doing any regularization. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will choose to take the
top 1000 variable predictors, which have the biggest variances. Then we will center the data, and create a filter to
remove higly correlated variables. Since we already transposed the gene expression data, we can simply merge the two
tables by row names, which are sample-IDs by the base function `merge()`. Using the `merge()` function in accordance with
row names causes that an extra column called "Row.names" is added at the left, so we will push the sample ids back to the
row names and delete the first column. We can split the data set into test and training sets with the
`caret::createPartition()` function. We will first get the indices for 70% of the data set, and use that indices to
indicate the training set. Rest will be our testing set.

To do gradient boosting, we will use the `XGboost` package. For the control method, we will set up a 5-fold cross
validation. Then we will use `caret::train()` with `method = "xgbTree"` to train our model. The most inmportant
parameters are number of trees (`nrounds`), tree depth (`max_depth`), and learning rate or shrinkage (`eta`). We will
set them to 200 trees, tree depth of 4 and will try three values for learning rate. We will sample 50% of the data and
use only that portion to train the tree by setting `subsample = 0.5`. After training the model, we can find out the best
parameters by cross-validation accuracy by pointing to `$bestTune` of the model.

For the support vector machines (SVM) we will use the `kernlab` package. For the control method, we will set up a 5-fold
cross validation. Then we will use `caret::train()` with `method = "svmRadial"` to train our model. Here the important
parameter is $C$, which controls the margin. Since it is recommended that $C$ is in the form of $2^k$ where $k$ is in
the range of -5 and 15, we will try 0.25, 0.5 and 1. 

To train a neural network on our cancer data set, we will use the `nnet` package. For the control method, we will set up
a 5-fold cross validation. Then we will use `caret::train()` with `method = "nnet"` to train our model. We will try 1 or
2 nodes in the hidden layer by setting `size = 1:2`. In addition, we will set `decay = 0`, which will correspond to not
doing any regularization. 

```{r,eval=FALSE}  

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Gradient Boosting----------------------------------------------

library(xgboost)
set.seed(17)

# we will just set up 5-fold cross validation
trctrl <- trainControl(method = "cv",number=5)

# we will now train gradient boosting model
gbFit <- train(subtype~., data = training, 
               method = "xgbTree",
               trControl=trctrl,
               # paramters to try
               tuneGrid = data.frame(nrounds=200,
                                     eta=c(0.05,0.1,0.3),
                                     max_depth=4,
                                     gamma=0,
                                     colsample_bytree=1,
                                     subsample=0.5,
                                     min_child_weight=1))

# best parameters by cross-validation accuracy
gbFit$bestTune

# Support Vector Machines-----------------------------------------------

library(kernlab)
set.seed(17)

# we will just set up 5-fold cross validation
trctrl <- trainControl(method = "cv",number=5)

# we will now train support vector machines model
svmFit <- train(subtype~., data = training, 
                # this SVM used radial basis function
                method = "svmRadial", 
                trControl=trctrl,
                tuneGrid=data.frame(C=c(0.25,0.5,1),
                                    sigma=1))

# Neural networks---------------------------------------------------------

#svm code here
library(nnet)
set.seed(17)

# we will just set up 5-fold cross validation
trctrl <- trainControl(method = "cv",number=5)

# we will now train neural networks model
# it will try
nnetFit <- train(subtype~., data = training, 
                 method = "nnet",
                 trControl=trctrl,
                 tuneGrid=data.frame(size=1:2,decay=0
                                      ),
                 # this is maximum number of weights
                 # needed for the nnet method
                 MaxNWts=2000) 

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to take the
top 1000 variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. After
preprocessing the data, I would like to merge those two tables, so that the rows represent individual patients and the
first column states CIMP subtype, followed by the gene expressions. Then I would like to split 30% of the data as the
test set, and rest of it as training set. How can I accomplish these tasks?

After preprocessing the data, I want to do gradient boosting and try different learning rate parameters. In this
instance, I also want to subsample 50% of the dataset before training each tree. How can I do that? How can I know the
best parameters by cross-validation accuracy? 


completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will choose to take the
top 1000 variable predictors, which have the biggest variances. Then we will center the data, and create a filter to
remove higly correlated variables. Since we already transposed the gene expression data, we can simply merge the two
tables by row names, which are sample-IDs by the base function `merge()`. Using the `merge()` function in accordance with
row names causes that an extra column called "Row.names" is added at the left, so we will push the sample ids back to the
row names and delete the first column. We can split the data set into test and training sets with the
`caret::createPartition()` function. We will first get the indices for 70% of the data set, and use that indices to
indicate the training set. Rest will be our testing set.

To do gradient boosting, we will use the `XGboost` package. For the control method, we will set up a 5-fold cross
validation. Then we will use `caret::train()` with `method = "xgbTree"` to train our model. The most inmportant
parameters are number of trees (`nrounds`), tree depth (`max_depth`), and learning rate or shrinkage (`eta`). We will
set them to 200 trees, tree depth of 4 and will try three values for learning rate. We will sample 50% of the data and
use only that portion to train the tree by setting `subsample = 0.5`. After training the model, we can find out the best
parameters by cross-validation accuracy by pointing to `$bestTune` of the model.


```{r,eval=FALSE}  

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Gradient Boosting----------------------------------------------

library(xgboost)
set.seed(17)

# we will just set up 5-fold cross validation
trctrl <- trainControl(method = "cv",number=5)

# we will now train gradient boosting model
gbFit <- train(subtype~., data = training, 
               method = "xgbTree",
               trControl=trctrl,
               # paramters to try
               tuneGrid = data.frame(nrounds=200,
                                     eta=c(0.05,0.1,0.3),
                                     max_depth=4,
                                     gamma=0,
                                     colsample_bytree=1,
                                     subsample=0.5,
                                     min_child_weight=1))

# best parameters by cross-validation accuracy
gbFit$bestTune


```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to take the
top 1000 variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. After
preprocessing the data, I would like to merge those two tables, so that the rows represent individual patients and the
first column states CIMP subtype, followed by the gene expressions. Then I would like to split 30% of the data as the
test set, and rest of it as training set. How can I accomplish these tasks?

After preprocessing the data, I would like to try the support vector machines (SVM) method to train the model.
How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will choose to take the
top 1000 variable predictors, which have the biggest variances. Then we will center the data, and create a filter to
remove higly correlated variables. Since we already transposed the gene expression data, we can simply merge the two
tables by row names, which are sample-IDs by the base function `merge()`. Using the `merge()` function in accordance with
row names causes that an extra column called "Row.names" is added at the left, so we will push the sample ids back to the
row names and delete the first column. We can split the data set into test and training sets with the
`caret::createPartition()` function. We will first get the indices for 70% of the data set, and use that indices to
indicate the training set. Rest will be our testing set.

For the support vector machines (SVM) we will use the `kernlab` package. For the control method, we will set up a 5-fold
cross validation. Then we will use `caret::train()` with `method = "svmRadial"` to train our model. Here the important
parameter is $C$, which controls the margin. Since it is recommended that $C$ is in the form of $2^k$ where $k$ is in
the range of -5 and 15, we will try 0.25, 0.5 and 1. 


```{r,eval=FALSE}  

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Support Vector Machines-----------------------------------------------

library(kernlab)
set.seed(17)

# we will just set up 5-fold cross validation
trctrl <- trainControl(method = "cv",number=5)

# we will now train support vector machines model
svmFit <- train(subtype~., data = training, 
                # this SVM used radial basis function
                method = "svmRadial", 
                trControl=trctrl,
                tuneGrid=data.frame(C=c(0.25,0.5,1),
                                    sigma=1))



```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 data sets. One is the gene expression data of glioblastoma tumor samples, called "LGGrnaseq.txt", in which the
columns are individual patients and the rows are the gene expressions. The other set is the patient annotation data from
the glioblastoma patient, called "patient2LGGsubtypes.txt", which indicates whether the patient has CIMP (CpG island
methylator phenotype) or noCIMP subtype. Here is the row individual patients and the column indicates the CIMP
existence. I want to first preprocess the data by taking the log and filtering the data. By filtering, I want to take the
top 1000 variable predictors, center the data and remove higly correlated variables with cutoff = 0.9. After
preprocessing the data, I would like to merge those two tables, so that the rows represent individual patients and the
first column states CIMP subtype, followed by the gene expressions. Then I would like to split 30% of the data as the
test set, and rest of it as training set. How can I accomplish these tasks?

After preprocessing the data, I want to train the neural network on the data set. For the neural network I want to try 1
or 2 nodes in the hidden layer without doing any regularization. How can I do that?

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. Then we will start the preprocessing of the gene expression data with taking a log to tame the extreme
values and add 1 pseudocount to deal with 0 values. Then, before doing other tasks we need to transpose our gene
expression data, since our gene expression data are on the row side, and we need to put it on the column side to use
machine learning algorithms. Then we will load the `caret` library to do the filtering. First we will choose to take the
top 1000 variable predictors, which have the biggest variances. Then we will center the data, and create a filter to
remove higly correlated variables. Since we already transposed the gene expression data, we can simply merge the two
tables by row names, which are sample-IDs by the base function `merge()`. Using the `merge()` function in accordance with
row names causes that an extra column called "Row.names" is added at the left, so we will push the sample ids back to the
row names and delete the first column. We can split the data set into test and training sets with the
`caret::createPartition()` function. We will first get the indices for 70% of the data set, and use that indices to
indicate the training set. Rest will be our testing set.

To train a neural network on our cancer data set, we will use the `nnet` package. For the control method, we will set up
a 5-fold cross validation. Then we will use `caret::train()` with `method = "nnet"` to train our model. We will try 1 or
2 nodes in the hidden layer by setting `size = 1:2`. In addition, we will set `decay = 0`, which will correspond to not
doing any regularization. 

```{r,eval=FALSE}  

# Data Read------------------------------------------------

# read gene expression data set from glioblastoma patients
gexp <- data.table::fread("LGGrnaseq.txt", data.table = FALSE)

# Push the gene ID from the first column to the row names and delete the first column
rownames(gexp) <- gexp[,1]
gexp <- gexp[,-1]

# read patient annotation data set from glioblastoma patients
patient <- data.table::fread("patient2LGGsubtypes.txt", data.table = FALSE)

# Push the patient ID from the first column to the row names and delete the first column
rownames(patient) <- patient[,1]
patient <- subset(patient, select = -1)

# Preprocess------------------------------------------------

# Taking a log to tame the extreme values and add 1 pseudocount to deal with 0 values
gexp <- log10(gexp+1)

# We need to transpose our data, since the predictor variables are gene expression and the machine learning algorithms
# require that the predictor variables are on the column side.
tgexp <- t(gexp)

# Load the needed library
library(caret)

# Choose to take the top 1000 variable predictors
SDs <- apply(tgexp,2,sd)
topPreds <- order(SDs,decreasing = TRUE)[1:1000]
tgexp <- tgexp[,topPreds]

# Centering the data using `preProcess()` function
processCenter <- preProcess(tgexp, method = c("center"))
tgexp <- predict(processCenter,tgexp)

# create a filter for removing higly correlated variables
# if two variables are highly correlated only one of them is removed
corrFilt <- preProcess(tgexp, method = "corr",cutoff = 0.9)
tgexp <- predict(corrFilt,tgexp)

# merge patient with transposed gexp
tgexp <- merge(patient,tgexp,by="row.names")

# push sample ids back to the row names
rownames(tgexp) <- tgexp[,1]
tgexp <- tgexp[,-1]

# Splitting the data------------------------------------------------------

# get indices for 70% of the data set
intrain <- createDataPartition(y = tgexp[,1], p= 0.7)[[1]]

# seperate test and training sets
training <- tgexp[intrain,]
testing <- tgexp[-intrain,]

# Neural networks---------------------------------------------------------

#svm code here
library(nnet)
set.seed(17)

# we will just set up 5-fold cross validation
trctrl <- trainControl(method = "cv",number=5)

# we will now train neural networks model
# it will try
nnetFit <- train(subtype~., data = training, 
                 method = "nnet",
                 trControl=trctrl,
                 tuneGrid=data.frame(size=1:2,decay=0
                                      ),
                 # this is maximum number of weights
                 # needed for the nnet method
                 MaxNWts=2000) 

```
--- TASK END


### Task `r i`,`r i<-i+1` taskType: read | data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a data set about the DNA methylation levels, called "CpGmeth2Age.txt". In the first column the age of the sample
is indicated, followed by the methylation values in the other columns. First I want to see the dimension of the data
set. Then, I would like to run the summary for the first three columns and plot histogram of methylation values. How can
I do that? I also would like to remove the ones that have less than 0.1 standard deviation across the samples. Then I
would like to run the random forest regression without any cross-validation and plot the two plots, one for observed
vs.predicted age, another for residuals. 

completion:
The data set must be read first. We will use `data.table::fread()` function to read in the text file as it can read
files faster. We can see the dimension of the data set by `dim()` function. Then we will run the `summary()` for the
first three columns and plot histogram of methylation values. For the histogram we will exclude the first column, and
use only the methylation values. Then we will check the standard deviation for each column, and remove the ones that
have less than 0.1 variation. Now we can use random forest regression to predict the age from methylation values. For
the control method, we will leave the method as "none" since we are not doing any cross-validation. Then we will use
`caret::train()` with `method = "ranger"` to train our model. For our first plot, we will plot predicted values from the
model in accordance with the observed value. We will use scatterplot form with x being observed age and y being
predicted age. In addition, the R-squared value will be indicated on the plot. For our second plot, we will plot the
residuals calculated by "predicted - observed" in accordance with the observed age.


```{r,eval=FALSE} 

# Predicting continuous variables: Regression with machine learning--------

# read methylation-age table
ameth <- data.table::fread("CpGmeth2Age.txt", data.table = FALSE)

# Dimension of the original data set
dim(ameth)

# Summary
summary(ameth[,1:3])

# plot histogram of methylation values. ameth[,-1]: excluding first column
hist(unlist(ameth[,-1]),border="white",
     col="cornflowerblue",main="",xlab="methylation values")


# remove CpGs that have less than 0.1 standard deviation
ameth <- ameth[,c(TRUE,matrixStats::colSds(as.matrix(ameth[,-1]))>0.1)]
dim(ameth)

## Use random forest regression to predict the age from methylation values----

set.seed(18)
library(caret)
par(mfrow=c(1,2))

# we are not going to do any cross-validatin and rely on OOB error
trctrl <- trainControl(method = "none")

# we will now train random forest model
rfregFit <- train(Age~., 
                  data = ameth, 
                  method = "ranger",
                  trControl=trctrl,
                  # calculate importance
                  importance="permutation", 
                  tuneGrid = data.frame(mtry=50,
                                        min.node.size = 5,
                                        splitrule="variance")
)

# plot Observed vs OOB predicted values from the model
plot(ameth$Age,rfregFit$finalModel$predictions,
     pch=19,xlab="observed Age",
     ylab="OOB predicted Age")
mtext(paste("R-squared",
            format(rfregFit$finalModel$r.squared,digits=2)))

# plot residuals
plot(ameth$Age,(rfregFit$finalModel$predictions-ameth$Age),
     pch=18,ylab="residuals (predicted-observed)",
     xlab="observed Age",col="blue3")
abline(h=0,col="red4",lty=2)


```

--- TASK END