---
title: "prompt/completion pairs chapter 6"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a BED file called "cpgi.hg19.chr21.bed". 
I want to read in this file and convert it to 
a GRanges object. How do I do this?

completion:
To read in a BED file and convert it to 
a GRanges object, you can make use of the
`rtracklayer` library in R. The function 
`import.bed` reads in BED files and automatically
converts them to a GRanges object.     

```{r,eval=FALSE}
require(rtracklayer)

# read in data
data <- import.bed ("cpgi.hg19.chr21.bed")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have an Encode formatted broad peak file called
"wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz"
which holds information from my ChIP-seq experiment. 
I want to read this file in and convert it to a GRanges
object. How do I do this? 

completion:
To read an Encode formatted broad peak file and turn 
it into a GRanges object, you can make use of the 
`readBroadPeak` function from the genomation library.
Here is how you do it:

```{r,eval=FALSE}
library(genomation)

# read in file
bp <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have an Encode formatted broad peak file called
"wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz"
which holds information from my ChIP-seq experiment. 
I have another file called "cpgi.hg19.chr21.bed". This
is a BED file, and holds information about the locations 
of CpG islands. I want to see if any peaks in my ChIP-seq 
experiment are located in CpG islands. How do I do this?

completion:
To determine if any peaks from your ChIP-seq experiment are located 
in CpG islands in R, you can use the GenomicRanges package. To read in 
BED files, you can make use of the import.bed function from the 
rtracklayer library. To read in Encode formatted broad peak files
you can make use of the `readBroadPeak` function from the genomation
library. After reading in the files, you can find overlaps by using the
`subsetByOverlaps` function from the GenomicRanges library. 

```{r,eval=FALSE}
require(genomation)
require(GenomicRanges)
require(rtracklayer)

# read in files
bp <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")
cpg <- import.bed ("cpgi.hg19.chr21.bed")

# find overlaps
res <- subsetByOverlaps(bp,cpg)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have an Encode formatted broad peak file called
"wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz"
which holds information from my ChIP-seq experiment. 
I have another file called "cpgi.hg19.chr21.bed". This
is a BED file, and holds information about the locations 
of CpG islands. I want to count the number of peaks in my ChIP-seq 
experiment that overlap with certain CpG islands. How do I do this?

completion:
To count the number of peaks from your ChIP-seq experiment 
that overlap with the CpG islands, you can make use of the 
GenomicRanges library in R. To read in BED files, you can 
make use of the import.bed function from the rtracklayer 
library. To read in Encode formatted broad peak files
you can make use of the `readBroadPeak` function from the 
genomation library. After reading in the files, you can use
the function `countOverlaps` to count the amount
of peaks that overlap with the CpG islands. 

```{r,eval=FALSE}
require(genomation)
require(GenomicRanges)
require(rtracklayer)

# read in files
bp <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")
cpg <- import.bed ("cpgi.hg19.chr21.bed")

# count overlaps
counts <- countOverlaps(cpg, bp)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have an Encode formatted broad peak file called
"wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz"
which holds information from peaks in my ChIP-seq experiment. 
I have another file called "cpgi.hg19.chr21.bed". This
is a BED file, and holds information about the locations 
of CpG islands. I want to find the nearest CpG island
for each peak in my ChIP-seq experiment. How do I do this?

completion:
To find the nearest CpG island for each peak in your ChIP-seq experiment,
you can use the GenomicRanges package. To read in BED files, you can 
make use of the import.bed function from the rtracklayer 
library. To read in Encode formatted broad peak files
you can make use of the `readBroadPeak` function from the 
genomation library. After reading in the files, you can
find the nearest CpG island for each ChIP-seq peak by 
using the `nearest` function. Here's how you can do it:

```{r,eval=FALSE}
require(genomation)
require(GenomicRanges)
require(rtracklayer)

# read in files
bp <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")
cpg <- import.bed ("cpgi.hg19.chr21.bed")

# find nearest CpG island for each
# peak in ChIP-seq experiment
res <- nearest(bp,cpg)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have an Encode formatted broad peak file called
"wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz"
which holds information from peaks in my ChIP-seq experiment. 
I have another file called "cpgi.hg19.chr21.bed". This
is a BED file, and holds information about the locations 
of CpG islands. I want to find the distane to the 
nearest CpG island for each peak in my ChIP-seq experiment. 
How do I do this?

completion:
To find the distane to the nearest CpG island for each peak in 
your ChIP-seq experiment,you can use the GenomicRanges package. 
To read in BED files, you can make use of the import.bed function 
from the rtracklayer library. To read in Encode formatted broad 
peak files you can make use of the `readBroadPeak` function 
from the genomation library. After reading in the files, you can
find the nearest CpG island for each ChIP-seq peak by 
using the `nearest` function. Here's how you can do it:


```{r,eval=FALSE}
require(genomation)
require(GenomicRanges)
require(rtracklayer)

# read in files
bp <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")
cpg <- import.bed ("cpgi.hg19.chr21.bed")

# find distance to nearest CpG island
dists <- distanceToNearest(bp,cpg,select="arbitrary")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "wgEncodeHaibTfbsGm12878Sp1Pcr1xAlnRep1.chr21.bam"
which is in BAM format. I want to extract ranges between 9411240 and 
9412324 of chromosome 21. How do I do this?

completion:
To read in information between a certain range from a BAM file, 
you can make use of the Rsamtools and GenomicAlignments libraries in R.
The range of interest can be defined in a GRanges object. This object 
is then used to create a ScanBamParam object, which can be used to 
scan the BAM file for the specified region. Here is how you do it:

```{r,eval=FALSE}
require(GenomicAlignments)
require(Rsamtools)

# create GRanges object
gr <- GRanges (seqnames = "chr21", 
               ranges = IRanges (start =9411240,end = 9412324 ))

# create ScanBamParam object
region <- ScanBamParam(which=gr)

# read in bam file with specified ranges
alns <- readGAlignments("wgEncodeHaibTfbsGm12878Sp1Pcr1xAlnRep1.chr21.bam",
                        param = region)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "wgEncodeHaibTfbsGm12878Sp1Pcr1xAlnRep1.chr21.bam"
which is in BAM format. I want to get the coverage of this file. 
How do I do this?

completion:
To get the coverage vector of information in a BAM file,
you can make use of the `coverage` function from the 
GenomicAlignments library. Here is how you do it:

```{r,eval=FALSE}
require (GenomicAlignments)

covs <- coverage("wgEncodeHaibTfbsGm12878Sp1Pcr1xAlnRep1.chr21.bam")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "wgEncodeHaibTfbsA549.chr21.bw" which is 
in BigWig format. I want to read in the information of 
chromosome 21 in the range between 9415815 and 
9417514. How do I do this?

completion:
To read in information between a certain range from a BigWig
file, you can make use of the `import` function. To import
a certain range, you can make a GRanges object with the region 
of interest. This can then be passed to the `which` argument
from the `import` function to exclusively import information
in the range of interest. 

```{r,eval=FALSE}
require(GenomicAlignments)

# create GRanges object
gr <- GRanges (seqnames = "chr21", 
               ranges = IRanges (start =9415815,end = 9417514 ))

# read file
bw_res <- import("wgEncodeHaibTfbsA549.chr21.bw", which = gr)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "wgEncodeHaibTfbsA549.chr21.bw" which is 
in BigWig format. I want to read in the information of 
chromosome 21 in the range between 9415815 and 
9417514 and get the coverage. How do I do this?

completion:
To read in information between a certain range from a BigWig
file, you can make use of the `import` function. To import
a certain range, you can make a GRanges object with the region 
of interest. This can then be passed to the `which` argument
from the `import` function to exclusively import information
in the range of interest. To get the coverage vector, you can
make use of the `coverage` function from the GenomicAlignments
library. 

```{r,eval=FALSE}
require(GenomicAlignments)

# create GRanges object
gr <- GRanges (seqnames = "chr21", 
               ranges = IRanges (start =9415815,end = 9417514 ))

# read file
bw_res <- import("wgEncodeHaibTfbsA549.chr21.bw", which = gr)

# get coverage vector
cov_res <- coverage(bw_res)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "cpgi.hg19.chr21.bed"
which is a BED file holding information about
the position of CpG islands on chromosome 21. 
I want to visualize the positions of 
CpG islands on chromosome 21 in a plot. 
How do I do this?

completion:
To visualize the positions of CpG islands
on chromosome 21 you can make use of the Gviz
library in R. Here is how you can do it:

```{r,eval=FALSE}
require(Gviz)
require(rtracklayer)

# read in file
cpg <- import.bed ("cpgi.hg19.chr21.bed")

# creata data track
cpg_track <- AnnotationTrack(cpg, name="CpG")

# create genome track
gtrack <- GenomeAxisTrack()

# create ideogram track
itrack <- IdeogramTrack(genome = "hg19", chromosome = "chr21")

# plot tracks
plotTracks(list(itrack, gtrack, cpg_track))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a RefSeq file called "refseq.hg19.chr20.bed"
which is in BED format. This holds information
of ranges on chromosome 20 which are associated with
promoters, exons, introns and TSSes. I also have ChIP-seq
data in a file called "H1.ESC.H3K4me3.chr20.bw" which is
in BigWig format. I want to create a matrix of ChIP-seq scores 
for the promotor regions. I then want to plot the average
enrichment values around promoters. How do I do this?

completion:
To create a matrix of ChIP-seq scores for promoter 
regions and plot the average enrichment values around 
promoters, you can make use of the genomation library
to analyze your data. You can read in the refseq data using
the readTranscriptFeatures function. To create
a score matrix you can make use of the `ScoreMatrix`
function. To then plot the results, you can use the function
plotMeta or heatMatrix. Here is how you do it:

```{r,eval=FALSE}
require(genomation)

# reading in file
# You may want to adjust the promoter region size according to your analysis requirements:
rs_file <- readTranscriptFeatures("refseq.hg19.chr20.bed", 
                                  up.flank = 500, down.flank=500)

# create score matrix
sm <- ScoreMatrix("H1.ESC.H3K4me3.chr20.bw",rs_file$promoters,
               type="bigWig",strand.aware = TRUE)

# plot average enrichment
plotMeta(sm, xcoords = c(-500,500),
         ylab="Enrichment",dispersion = "se",
         xlab="bases around promoter")

heatMatrix(sm,order=TRUE,xcoords = c(-500,500),
           xlab="bases around promoter")         
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a RefSeq file called "refseq.hg19.chr20.bed"
which is in BED format. This holds information
of ranges on chromosome 20 which are associated with
promoters, exons, introns and TSSes. I also have ChIP-seq
data in 2 files from 2 experiments called 
"H1.ESC.H3K4me3.chr20.bw" and "H1.ESC.dnase.chr20.bw" which are
in BigWig format. I want to create a matrix of ChIP-seq scores 
for the promotor regions. I then want to plot the average
enrichment values around promoters. How do I do this?

completion:
To create a matrix of ChIP-seq scores for promoter 
regions and plot the average enrichment values around 
promoters, you can make use of the genomation library
to analyze your data. You can read in the refseq data using
the readTranscriptFeatures function. To create
a score matrix you can make use of the `ScoreMatrixList`
function. To then plot the results, you can use the function
plotMeta or multiHeatMatrix. Here is how you do it:

```{r,eval=FALSE}
require(genomation)

# reading in file
# You may want to adjust the promoter region size according to your analysis requirements:
rs_file <- readTranscriptFeatures("refseq.hg19.chr20.bed", 
                                  up.flank = 500, down.flank=500)


# create score matrix
sm <- ScoreMatrixList(c(exp1="H1.ESC.H3K4me3.chr20.bw",
                        exp2="H1.ESC.dnase.chr20.bw"),
                      rs_file$promoters,
                      type="bigWig",strand.aware = TRUE)

# create plots
plotMeta(sm)

multiHeatMatrix(sm,order=TRUE,xcoords = c(-500,500),
                xlab="bases around promoter",winsorize = c(0,95),
                matrix.main = c("exp1","exp2"),
                column.scale=TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a BED file called "cpgi.hg19.chr21.bed".
This holds information about the location of CpG 
islands on chromosome 21. I want to plot the location
of these CpG islands in a karyogram. I have another
file called "wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz". 
This is an Encode formatted broad peak file which holds 
the information of where peaks occurred in my ChIPseq 
experiment. In the extra data there is a column score. 
I would like to also plot the positions of the beginning of
the peaks of the ChIPseq experiment on the karyogram, 
and have a way of visualizing the score as well. How do I do this?

completion:
To plot the locations of CpG islands and ChIP-seq peaks on 
chromosome 21, and visualize the scores of ChIP-seq peaks, you can use the 
ggbio library. The data is read in as a GRanges object, and only
information of chromosome 21 is retained. To then plot a karyogram
you can make use of the ideoCyto data from the biovizBase package. 
This allows you to plot an empty karyogram, on which you can plot
the position of CpG islands, and the position and score of your
ChIP-seq experiment. Below you can find an example on how to do this:

```{r,eval=FALSE}
require(rtracklayer)
require(ggbio)
require(BSgenome.Hsapiens.UCSC.hg19)
require(genomation)

# read in data
cpg <- import.bed ("cpgi.hg19.chr21.bed")
bp <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")

# get info of chr21
bp <- bp[seqnames(bp)=="chr21"]
seqlevels(bp) <- "chr21"

# set seqlevels to length of chromosome 21
seqlengths(bp) <- seqlengths(Hsapiens)["chr21"]

# set seqlengths of cpg of chromosome 21
seqlengths(cpg) <-  seqlengths(Hsapiens)["chr21"]

# create object for plotting:
data(ideoCyto, package = "biovizBase")
region <- ideoCyto$hg19[seqnames(ideoCyto$hg19)=="chr21"]
seqlevels(region) <- "chr21"
p <- autoplot(seqinfo(region), layout = "karyogram")

# plot cpg info and ChIPseq info
p + layout_karyogram (cpg) + 
layout_karyogram(bp,aes(x=start,y=score),
                geom="point",color="red")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file named "wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz," which 
is an Encode formatted broad peak file. This file provides 
information regarding the locations of peaks in my ChIP-seq 
experiment. My goal is to create a circos plot to 
visualize the position of the peaks. 
There is a column for 
scores in the extra data of this file. I want to
to include  a way to visualize the scores 
effectively. The ChIPseq data is data from the whole human
genome, except for chromosome Y. 
How can I do this?

completion:
To create a Circos plot to visualize the positions of ChIP-seq peaks, 
including an effective way to visualize the scores, you can make use
of the ggbio library. You can make use of the ideoCyto data from the biovizBase 
package, to get the general information for the whole genome, removing chromosome Y,
so that you can set an empty circle with chromosome names for further plotting. 
Then, you can plot your ChIP-seq data in this. Below is an example on 
how to do it:

```{r,eval=FALSE}
require(rtracklayer)
require(ggbio)
require(genomation)

# read in data
bp <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")

# set the chromsome in a circle, excluding chromosome Y
data(ideoCyto, package = "biovizBase")
chr_info <- ideoCyto$hg19[seqnames(ideoCyto$hg19)!="chrY"]
seqlevels(chr_info,pruning.mode="coarse") <- seqlevels(chr_info)[seqlevels(chr_info)!="chrY"]
seqlengths(chr_info) <- seqlengths(chr_info)[names(seqlengths(chr_info))!="chrY"]


# prepare object for final plot
# set the chromsome in a circle
p <- ggplot() + layout_circle(chr_info, geom = "ideo", fill = "white",
                              colour="black",
                              radius = 39, trackWidth = 0.1)

# set chromosome names
p <- p + layout_circle(as(seqinfo(chr_info),"GRanges"), 
                       geom = "text", aes(label = seqnames), 
                      vjust = 0, radius = 40, trackWidth = 1,
                      size=3)

# plot ChIP data
p <- p + layout_circle(bp, 
                       geom = "point",aes(x=start,y=score), radius = 30,
                       trackWidth=8,
                       color="red")

# show plot
p
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have transcription factor binding sites (ChIP-seq peaks) data determined by 
ChIP-seq experiments in a file called 
"wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz". 
I would like to find the distances to the nearest TSS from CpG islands using 
CpG island data stored in "cpgi.hg19.chr21.bed" and plot them.
How can I do this?

completion:
To find the distance to the nearest TSS from CpG islands in your ChIP-seq 
experiment data, you can proceed as follows. 
Read your ChIP.seq data file to get the peaks using readBroadPeak() 
function from genomation package, read in your CpG island data using
import.bed() function from rtracklayer package.
Then to get the distance to the nearest TSS use distanceToNearest() function 
from GenomicRanges package, and then plot the distances using hist() function. 

```{r,eval=FALSE}

require(genomation)
require(GenomicRanges)
require(rtracklayer)

# read the peaks from the bed file
pks <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")
# read CpGi data set
cpgi <- import.bed("cpgi.hg19.chr21.bed")

# find distance to nearest TSS
dists <- distanceToNearest(pks,cpgi,select="arbitrary")

# histogram of the distances to nearest TSS
dist2plot=mcols(dists)[,1]
hist(log10(dist2plot),xlab="log10(dist to nearest TSS)",
     main="Distances")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have peaks data from two replicates in files
"wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz" and "wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep2.broadPeak.gz".
I would like to get the peaks that overlap (canonical peaks) on chr21.
How do I do it?

completion:
To get the canonical peaks you can proceed as follows. 
Read the peaks data file using readBroadPeak() function from genomation package.
Then to get the overlapping peaks use findOverlaps() or subsetByOverlaps() 
function for genomation package. After this subset the data for chr21. 

```{r,eval=FALSE}

require(genomation)

# read the peaks from the bed file
pks1 <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep1.broadPeak.gz")
pks2 <- readBroadPeak("wgEncodeHaibTfbsGm12878Sp1Pcr1xPkRep2.broadPeak.gz")

# get the peaks that overlap 
overlap_all <- subsetByOverlaps(pks1,pks2)

# get the peaks overlap for chr21
overlap_ch21 <- GRanges(Rle("chr21", length(overlap_all)), 
                        ranges(overlap_all, use.mcols=TRUE), 
                        strand=strand(overlap_all))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have the reference transcript file "refseq.hg19.chr20.bed", and
ChIP-seq enrichment data in files,
"H1.ESC.H3K4me3.chr20.bw" and
"H1.ESC.H3K27ac.chr20.bw".
I would like to extract -500,+500 bp regions around the TSSes on chr20 and 
then use the ChIP-seq data to create and visualize the average signal profiles
for regions around TSSes as a line plot.
How do I do it?

completion:
You can make use of genonation package for this as follows.
Extract features from the reference sequence file using 
readTranscriptFeatures() function, then get promotors from the the extracted 
features. Create average signal profiles using ScoreMatrixList() function. 
Then plot it using plotMeta() function to visualize the profiles.

```{r,eval=FALSE}

require(genomation)

# get transcription start sites on chr20
feat=readTranscriptFeatures("refseq.hg19.chr20.bed",
                            remove.unusual = TRUE,
                            up.flank = 500, down.flank = 500)

# get promoters from the features
prom=feat$promoters 

#require(GenomicAlignments)
# read file
#h3k4 <- import("H1.ESC.H3K4me3.chr20.bw")
#hek27 <- import("H1.ESC.H3K27ac.chr20.bw")

# create scoring matrix list
sml=ScoreMatrixList(c(H3K4me3="H1.ESC.H3K4me3.chr20.bw",
                      H3K27ac="H1.ESC.H3K27ac.chr20.bw"),prom,
                      type="bigWig",strand.aware = TRUE)

# plot scoring matrix list
plotMeta(sml)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have the reference transcript file "refseq.hg19.chr20.bed", and
ChIP-seq enrichment data in files,
"H1.ESC.H3K4me3.chr20.bw" and
"H1.ESC.H3K27ac.chr20.bw".
I would like to extract -500,+500 bp regions around the TSSes on chr20 and 
then use the ChIP-seq data to create average signal profiles for regions 
around TSSes and visualize it side by side as heatmaps for easy comparison. 
How do I do it?

completion:
You can make use of genonation package for this as follows.
Extract features from the reference sequence file using 
readTranscriptFeatures() function, then get promotors from the the extracted 
features. Create average signal profiles using ScoreMatrixList() function. 
Then plot it using multiHeatMatrix() function to visualize the profiles 
side by side.

```{r,eval=FALSE}

require(genomation)

# get transcription start sites on chr20
feat=readTranscriptFeatures("refseq.hg19.chr20.bed",
                            remove.unusual = TRUE,
                            up.flank = 500, down.flank = 500)

# get promoters from the features
prom=feat$promoters 

# create scoring matrix list
sml=ScoreMatrixList(c(H3K4me3="H1.ESC.H3K4me3.chr20.bw",
                      H3K27ac="H1.ESC.H3K27ac.chr20.bw"),prom,
                      type="bigWig",strand.aware = TRUE)

# set seed
set.seed(1029)
# plot heat maps side by side
multiHeatMatrix(sml,order=TRUE,xcoords = c(-500,500),
                xlab="bases around TSS",winsorize = c(0,95),
                matrix.main = c("H3K4me3","H3K27ac"),
                column.scale=TRUE,
                clustfun=function(x) kmeans(x, centers=3)$cluster)

```

--- TASK END