---
title: "prompt/completion pairs chapter 6"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I want to make a GRanges object
with 3 ranges in it. They are all
chr21, and have the ranges 0-100,100-200
and 200-300. They are from strand -,- and + 
respectively. How do I do this?

completion:
To create a GRanges object with three ranges on 
chromosome "chr21" with the specified coordinates and 
strands, you can use the GenomicRanges package in R. 
Here's how you can create the GRanges object with the 
provided information:


```{r,eval=FALSE}
require(GenomicRanges)
require(IRanges)

res <- GRanges(seqnames = c("chr21", "chr21", "chr21"),
               ranges = IRanges(start = c(0, 100, 200),
                                end = c(100, 200, 300)),
               strand = c("-","-","+")
              )

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a Granges object called res containing 3 ranges
that I want to add a metadata column to 
with the information sampleID : 1,2,3. 
How do I do this?

completion:
To add a metadata column with the information 
"sampleID: 1, 2, 3" to a GRanges object called res 
containing three ranges, you can use the mcols function 
from the GenomicRanges package. Here's how you can do it:

```{r,eval=FALSE}
require(GenomicRanges)

# add metadata column to GRanges object
mcols(res)$sampleID <- c(1, 2, 3)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a GRanges object called res containing 3 ranges.
I want to extract the metadata from this GRanges object. 
How do I do this?

completion:
To extract metadata from a GRanges object you 
can make use of the `values` function in R. 
Here is how you do it: 

```{r,eval=FALSE}

# extract metadata
met_dat <- values(res)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a GRanges object called res containing 3 ranges.
This object has 1 metadata column called sampleID that 
I wish to remove. How do I do this?

completion:
To remove a metadata column named "sampleID" from a GRanges object called res, 
you can use the $ operator to manipulate the metadata columns. Here is how
you can do it:

```{r,eval=FALSE}
# remove metadata column sampleID
res$sampleID <- NULL
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called data with 
4 columns which holds information about 
sequence reads.The first column holds
sequence names, the second column holds the
start position, the third column holds the end
position and the fourth column holds information
on sampleID. I want to convert this dataframe to 
a GRanges object. How do I do this?

completion:
To convert your dataframe `data` to a GRanges object
we can set the column names for your dataframe to the
conventional names for a GRanges object, and then make
use of the function `makeGRangesFromDataFrame`. 
Setting the argument `keep.extra.columns = TRUE` ensures
you keep the metadata from your dataframe as well. 

```{r,eval=FALSE}
# set columnnames to conventional names
colnames (data) <- c("seqnames", "start", "end", "sampleID")

# make GRanges object
gr_data <- makeGRangesFromDataFrame(data, keep.extra.columns = TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 GRanges objects called bp and cpg.
I want to see if there are any ranges in 
bp which overlap with cpg. How do I do this?

completion:
To determine if any ranges overlap between 
your two GRanges objects, you can use the GenomicRanges 
package. Using the function `findOverlaps` will find
one-to-one overlaps between your two objects. 

```{r,eval=FALSE}
require(GenomicRanges)

# find overlaps
res <- findOverlaps(bp,cpg)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 GRanges objects called bp and cpg.
I want to count how many ranges in bp
overlap with ranges in cpg. How do I 
do this?

completion:
To count the amount of ranges from bp
that overlap with the individual ranges 
in cpg, you can make use of the `countOverlaps`
function from the GenomicRanges library. 

```{r,eval=FALSE}
require(GenomicRanges)

# count overlaps
counts <- countOverlaps(cpg, bp)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 GRanges objects called bp and cpg.
I want to find the ranges in cpg which are
nearest to each of the ranges in bp. 
How do I do this?

completion:
To find the ranges in the cpg GRanges object that 
are nearest to each of the ranges in the bp GRanges 
object you can use the GenomicRanges package nearest function. 
Here's how you can do it:

```{r,eval=FALSE}
require(GenomicRanges)

# find closest ranges
res <- nearest (bp,cpg)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 GRanges objects called bp and cpg.
I want to find the distance for all bp ranges to 
the nearest cpg ranges. How do I do this?

completion:
To find the distance from each range in the bp GRanges object 
to the nearest range in the cpg GRanges object you can use the 
GenomicRanges package distanceToNearest function. 
Here is how you do it:

```{r,eval=FALSE}
require(GenomicRanges)

# find distances
dists <- distanceToNearest(bp, cpg, select = "arbitrary")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a SummarizedExperiment object called `se`
from which I want to extract the column associated
tables. How do I do this?

completion:
In a SummarizedExperiment object, the column-associated 
tables are typically stored in the colData slot. To extract 
the column-associated tables from a SummarizedExperiment object 
called se, you can access the colData slot using the `colData`
function. Here is how you do it:

```{r,eval=FALSE}
# extract column associated tables
col_dat <- colData(se)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a SummarizedExperiment object called `se`
from which I want to extract the row associated
tables. How do I do this?

completion:
In a SummarizedExperiment object, the row-associated 
tables are typically stored in the rowData slot. 
To extract the row-associated tables from a SummarizedExperiment 
object called se, you can access the rowData slot as follows:

```{r,eval=FALSE}
# extract row associated tables
row_dat <- rowData(se)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a SummarizedExperiment object called `se`
from which I want to extract the main table. 
How do I do this?

completion:
In a SummarizedExperiment object, the main data table is 
typically stored in the assay slot. To extract the main table 
from a SummarizedExperiment object called se, you can access 
the assay slot as follows:

```{r,eval=FALSE}

# extract main table(s):
main <- assays(se)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a SummarizedExperiment object called `se`. 
I want to extract a subsection based on 
a colData called timepoint. I only want
items for which timepoint is equal to 1.
How do I do this?

completion:
To extract a subsection of a SummarizedExperiment 
object based on the values in the timepoint column 
of the colData, you can use subsetting. 
Here's how you can do it:

```{r,eval=FALSE}
se_subset <- se[, se$timepoint==1]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a SummarizedExperiment object called `se`. 
I want to extract a subsection for rows which are in 
chr1:100,000-1,100,000. How do I do this?

completion:
To extract a subset of rows from a SummarizedExperiment 
object that fall within the region "chr1:100,000-1,100,000," 
you can use the subsetByOverlaps function from the GenomicRanges 
package. Here's how you can do it:


```{r,eval=FALSE}
require(GenomicRanges)

# get subset
roi <- GRanges(seqnames="chr1", ranges=100000:1100000)
se_subset <- subsetByOverlaps(se, roi)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a ScoreMatrix object called sm of 
which I want to plot enrichment values. 
How do I do this?

completion:
To plot enrichment values from a ScoreMatrix
object, you can make use of the genomation library
function `plotMeta`. Here is how you do it:

```{r,eval=FALSE}
require(genomation)

# plot enrichment
plotMeta(sm)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a ScoreMatrixList object called sm of 
which I want to plot enrichment values. 
How do I do this?

completion:
To plot enrichment values from a ScoreMatrixList
object, you can make use of the genomation library
function `plotMeta`. Here is how you do it:

```{r,eval=FALSE}
require(genomation)

# plot enrichment
plotMeta(sm)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a ScoreMatrix object called sm of 
which I want to plot enrichment values in a heatmap. 
How do I do this?

completion:
To plot enrichment values from a ScoreMatrix
object in a heatmap, you can make use of the genomation library
function `heatMatrix`. Here is how you do it:

```{r,eval=FALSE}
require(genomation)

# plot heatMatrix
heatMatrix(sm,order=TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a ScoreMatrixList object called sm of 
which I want to plot enrichment values in a heatmap. 
How do I do this?

completion:
To plot enrichment values from a ScoreMatrixList
object in a heatmap, you can make use of the genomation library
function `multiHeatMatrix`. Here is how you do it:

```{r,eval=FALSE}
require(genomation)

# plot heatMatrix
multiHeatMatrix(sm, order=TRUE)
```

--- TASK END
