---
title: "prompt/completion pairs chapter 7"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to plot the sequence quality per cycle. 
How do I do this?

completion:
To plot the sequence quality per cycle, you first need to read
in the data. this can be done using the `rqc` function. Then,
you can make use of the`rqcCycleQualityBoxPlot` function from the `Rqc` 
library. Here is how you do it:

```{r,eval=FALSE}
require(Rqc)

# read in data
qcRes <- rqc(pattern = "ERR127302_1_subset.fastq.gz", openBrowser = FALSE)

# plot quality per cycle. 
rqcCycleQualityBoxPlot(qcRes)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to visualize the nucleotide proportions for each position.
How do I do this?

completion:
To plot the nucleotide proportions for each position,
you first need to read in the data. This can be done using
the `rqc` function from the `Rqc` library. Then, the function
`rqcCycleBaseCallsLinePlot` can be used to plot the 
nucleotide proportion per position. 

```{r,eval=FALSE}
require(Rqc)

# read in data
qcRes <- rqc(pattern = "ERR127302_1_subset.fastq.gz", openBrowser = FALSE)

# plot nucleotide proportion per position
rqcCycleBaseCallsLinePlot(qcRes)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz" and
another called "ERR127302_2_subset.fastq.gz". 
which hold sequencing reads information and are in fastq format.
I want to plot the read frequency plot. How do I do this?

completion:
To plot the read frequency plot you first need to read in the data. 
This can be done using the `rqc` function from the `Rqc` library.
Since both files contain "ERR127302_" we can use this as the pattern
for reading in both files simultaneously. To then plot the read 
frequency, you can make use of the `rqcReadFrequencyPlot`
function. 


```{r,eval=FALSE}

# read in data
qcRes <- rqc(pattern = "ERR127302_", openBrowser = FALSE)

# plot read frequency plot
rqcReadFrequencyPlot(qcRes)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz" and
another called "ERR127302_2_subset.fastq.gz". 
which hold sequencing reads information and are in fastq format.
I want some general QC reports to check the quality of my 
reads. I want to save these statistics in a new directory called 
"fastqc_results". I then also want to look at a preview of this. 
How do I do this?

completion:
The most popular tool for sequencing quality control is the fastQC tool, 
which is written in Java. It produces several quality plots and statistics.
The R package fastqcr can run this Java tool and produce R-based plots and reports. 
This package simply calls the Java tool and parses its results. 
Below, we show how to do that.

```{r,eval=FALSE}
require(fastqcr)

# install the FASTQC java tool
fastqc_install()

# call FASTQC on all files in directory 
# and record the resulting statistics
# in fastqc_results folder
fastqc(fq.dir = ".",qc.dir = "fastqc_results")

# view the report rendered by R functions
qc_report(qc.path="fastqc_results", 
          result.file="reportFile", preview = TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to remove reads which have more then 1 unidentified base, and 
save the results to a new file called "processed.fastq" also in 
fastq format. How do I do this?

completion:
To remove reads which have more then 1 base in a fastq file,
you can make use of the `QuasR` library in R. 
Using the function `preprocessReads` and setting the
argument nBases=1, all reads which have more then 1 unidentified 
base will be filtered out. 

```{r,eval=FALSE}
require(QuasR)

preprocessReads("ERR127302_1_subset.fastq.gz", "processed.fastq", 
                nBases=1)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to remove 3 bases from each read at the end and 
save the results in a fastq file called "processed.fastq". 
How do I do this?

completion:
To remove 3 bases at the end from each read, you can make use of the
`QuasR` library in R. Using the function `preprocessReads` and
setting the argument `truncateEndBases=3` will ensure
3 bases are trimmed of the end of each read. 

```{r,eval=FALSE}
require(QuasR)

preprocessReads("ERR127302_1_subset.fastq.gz", "processed.fastq", 
                truncateEndBases=3)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to remove reads which are shorter then 40 basepairs and 
save the results in a fastq file called "processed.fastq". 
How do I do this?

completion:
To remove reads which are shorter then 40 basepairs, and 
save the results in a file called "processed.fastq", you can
make use of the `QuasR` library in R. Using the function `preprocessReads` 
and setting the argument `minLength=40` will ensure that reads shorter then
40 basepairs will be removed. 

```{r,eval=FALSE}
require(QuasR)

preprocessReads("ERR127302_1_subset.fastq.gz", "processed.fastq", 
                minLength=40)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to remove the ACCCGGGA patern if it occurs at the start
of a read. How do I do this?

completion:
To remove a specific pattern, such as "ACCCGGGA", if it occurs at the start of
each read in a FASTQ file you can make use of the `QuasR` library in R. 
Using the function `preprocessReads` and setting the argument `Lpattern="ACCCGGGA"`
will remove this pattern at the start of each read. 

```{r,eval=FALSE}
preprocessReads("ERR127302_1_subset.fastq.gz", "processed.fastq", 
                Lpattern="ACCCGGGA")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to remove the ACCCGGGA patern if it occurs at the end
of a read. How do I do this?

completion:
To remove a specific pattern, such as "ACCCGGGA", if it occurs at the end of
each read in a FASTQ file you can make use of the `QuasR` library in R. 
Using the function `preprocessReads` and setting the argument `Rpattern="ACCCGGGA"`
will remove this pattern at the end of each read. 

```{r,eval=FALSE}
preprocessReads("ERR127302_1_subset.fastq.gz", "processed.fastq", 
                Rpattern="ACCCGGGA")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to remove reads which have a quality score below 20, and
save the results in a fastq file called "Qfiltered.fastq".
How do I do this?

completion:
To filter and remove reads with a quality score below 20 from a 
FASTQ file and save the filtered reads in a new FASTQ file, you can use
the `ShortRead` library in R. reading in the FASTQ file is done using the
`readFastq` function. To get the quality scores, you can make use of the 
`quality`function. You can then filter for bases which have reads with 
a quality score below 20, and write the results to a new FASTQ file 
using the `writeFastq` function. 

```{r,eval=FALSE}
require(ShortRead)

# read fastq file
fq <- readFastq("ERR127302_1_subset.fastq.gz")

# get quality scores per base as a matrix
qPerBase <- as(quality(fq), "matrix")

# get counts of bases per read that have quality score below 20
qcount <- rowSums( qPerBase < 20) 

# Reads where all scores >= 20
reads_filtered <- fq[qcount == 0]

# write to fastq file
writeFastq(reads_filtered, 
           "Qfiltered.fastq")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called "ERR127302_1_subset.fastq.gz"
which holds sequencing reads information and is in fastq format.
I want to align these reads to the human genome. How do I do this?

completion:
To align reads in a FASTQ file to the human genome, you can 
make use of the QuasR library. The function qAlign can be
used to make an alignment. 
This function is designed to be easy. For example, it creates a 
genome index automatically if it does not exist, and will look for 
existing indices before it creates one. We provided only two arguments, 
a text file containing sample names and fastq file paths and a reference 
genome file. In fact, this function also has many knobs and you can change 
its behavior by supplying different arguments in order to affect the behavior 
of Bowtie. For example, you can supply parameters to Bowtie using the 
alignmentParameter argument. However the qAlign() function is optimized for 
different types of alignment problems and selects alignment parameters 
automatically. It is designed to work with alignment and quantification 
tasks for RNA-seq, ChIP-seq, small-RNA sequencing, 
Bisulfite sequencing (DNA methylation) and allele-specific analysis. 
If you want to change the default bowtie parameters, only do it for 
simple alignment problems such as ChIP-seq and RNA-seq.

```{r,eval=FALSE}
library(ShortRead)
library(BSgenome.Hsapiens.UCSC.hg19)

# prepare txt file for alignment function
tbl <- data.frame (FileName = "ERR127302_1_subset.fastq.gz", SampleName = "sample1")
write.table(tbl, "inputs.txt",sep="\t")

# get alignment
alignment_res <- qAlign("inputs.txt", "BSgenome.Hsapiens.UCSC.hg19")

```

--- TASK END
