---
title: "prompt/completion pairs chapter 7"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a variable called `qcRes` which is a
RqcResultSet object which holds information
from my sequence reads. I want to plot 
the sequence quality per cycle. How do I do 
this?

completion:
To plot the sequence quality per cycle for a RqcResultSet object, 
you can make use of the`rqcCycleQualityBoxPlot` function from the 
`Rqc` library. Here is how you do it:

```{r,eval=FALSE}
require(Rqc)

# plot quality per cycle. 
rqcCycleQualityBoxPlot(qcRes)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a variable called `qcRes` which is a
RqcResultSet object which holds information
from my sequence reads.
I want to visualize the nucleotide proportions for each position.
How do I do this?

completion:
To plot the nucleotide proportion for each position 
for a RqcResultSet object, you can make use of the
`rqcCycleBaseCallsLinePlot` function from the 
`Rqc` library. Here is how you do it:

```{r,eval=FALSE}
require(Rqc)

# plot nucleotide proportion per position
rqcCycleBaseCallsLinePlot(qcRes)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a variable called `qcRes` which is a
RqcResultSet object which holds information
from my sequence reads. I want to plot 
the read frequency. How do I do this?

completion:
to plot the read frequency, you can make
use of the `rqcReadFrequencyPlot` function 
from the `Rqc` library. Here is how you do it:

```{r,eval=FALSE}
require(Rqc)

# plot read frequency plot
rqcReadFrequencyPlot(qcRes)

```

--- TASK END
