---
title: "prompt/completion pairs chapter 8"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `raw_counts`. 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. 
I want to calculate the CPM for every column except for the width column. 
How do I do this?

completion:
To calculate the Counts Per Million (CPM) for the columns in 
your `raw_counts` data. you remove the "width" column, calculate 
the sum of counts for each sample, and then calculate the CPM 
by dividing each count by the corresponding sample's total count 
and multiplying by 1 million. Finally, we save the CPM values
in a new variable. 

```{r,eval=FALSE}
# calculate cpm values
cpm <- apply(subset(raw_counts, select = c(-width)), 2,
             function(x) x/sum(as.numeric(x)) * 10^6)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `raw_counts`. 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. I want to compute the RPKM. 
The width column should be used for the length of the genes.
How do I do this?

completion:
To compute the Reads Per Kilobase Million (RPKM) using a 
DataFrame called raw_counts where the "width" column contains 
the length of the corresponding genes, you multiply the raw reads by 10^9. 
Then divide by the corresponding gene lengths (in kilobases) and divide the above 
result by the sum of all counts for that sample. This gives you the RPKM. 

```{r,eval=FALSE}

# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [, "width"])


# compute rpkm
rpkm <- apply(X = subset(raw_counts, select = c(-width)),
              MARGIN = 2, 
              FUN = function(x) {
                10^9 * x / gene_lengths / sum(as.numeric(x))
               })
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `raw_counts`. 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. I want to compute the TPM. 
The width column should be used for the length of the genes.
How do I do this?

completion:
To compute the Transcripts Per Million (TPM) values for your 
gene expression data using the provided raw_counts DataFrame 
and the gene length information from the "width" column, you can 
first calculate the gene length normalized values by dividing 
the counts by the corresponding gene lengths divided by 1000. 
then, divide the above result by the sum of all counts for that sample 
times 1 million. 

```{r,eval=FALSE}

# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [, "width"])

# compute TPM
#find gene length normalized values 
rpk <- apply( subset(raw_counts, select = c(-width)), 2, 
              function(x) x/(gene_lengths/1000))
#normalize by the sample size using rpk values
tpm <- apply(rpk, 2, function(x) x / sum(as.numeric(x)) * 10^6)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `counts` which holds
RNA-seq count information. 
The rows correspond to the genes and the columns 
represent different samples, with corresponding
row- and columnnames. I have another dataframe called `col_data`. 
Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. I expect the normal samples
to cluster together, and the cancer samples to cluster together. 
I want to perform PCA (The width column can be discarded 
for this analysis) to see if this is really the case, and plot
the results in a scatterplot. Annotation of the scatterplot 
can be done using the `group` column. How do I do this?

completion:
To perform PCA on your RNA-seq data and create a scatterplot to visualize 
the clustering of normal and cancer samples based on the "group" information,
You can perform PCA with help of the
`prcomp` function. To then visualize the results in a scatterplot, 
you can make use of the `ggplot2` library in R. 
Alternatively, you could also make use of the 
autoplot function. Here is how you do it:


```{r,eval=FALSE}
# prepare annotation vector
annotation <- as.factor(col_data[, "group"])
names(annotation) <- rownames(col_data)

# perform PCA 
t_counts <- t(counts)
res <- prcomp(t_counts)

# ensure same ordering in annotation vector
annotation <- annotation[rownames(res$x)]

# plot first 2 PC's in scatterplot and 
# annotate 
ggplot2::ggplot (data = res$x,
                 aes(x = PC1,
                     y = PC2)) +
        geom_point(aes(colour = annotation)) +
        geom_text(hjust=-0.1, vjust=0,label=annotation)+
        ggtitle ("PCA of RNA seq samples")

# using the autoplot function
# scales data automatically.
library(ggfortify)
autoplot(res, data = col_data, colour = "group")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `counts` which holds
RNA-seq normalized count information. 
The rows correspond to the genes and the columns 
represent different samples, with corresponding
row- and columnnames. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. I want to select the top 100 genes 
with the highest variance. How do I do this?

completion:
To select the top 100 genes with the highest variance in an 
R DataFrame called counts, you can compute the variance of 
each gene across all samples. The result of which we can 
order, and extract the names of the 100 genes which showed 
the highest variance. 

```{r,eval=FALSE}
require(pheatmap)

# compute the variance of each gene across samples
var_count <- apply(counts,1,var)

# order in decreasing value
var_count <- var_count[order(var_count, decreasing=TRUE),drop=FALSE]

# get names of first 100 genes
selectedGenes <- names(var_count)[1:100]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `counts` which holds
RNA-seq normalized count information. 
The rows correspond to the genes and the columns 
represent different samples, with corresponding
row- and columnnames. I have another dataframe called `col_data`. 
Here, the samples are the rows with corresponding rownames and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. I expect the normal samples
to cluster together, and the cancer samples to cluster together. For
this purpose I want to create a heatmap. For this heatmap, the column
`width` can be discarded. I want to create this heatmap only with 
the top 100 genes with the highest variance. I want to annotate 
this heatmap using information from the group` column in the `col_data` 
dataframe. How do I do this?

completion:
To select the top 100 genes with the highest variance in an 
R DataFrame called counts, and plot a heatmap of the 
counts of these genes, you can compute the variance of 
each gene across all samples. The result of which we can 
order, and extract the names of the 100 genes which showed 
the highest variance. These you can then use to plot a heatmap using
the `pheatmap` library. 


```{r,eval=FALSE}
require (pheatmap)

# prepare annotation dataframe
annotation <- col_data[, "group",drop=FALSE]

# compute the variance of each gene across samples
var_count <- apply(counts,1,var)

# order in decreasing value
var_count <- var_count[order(var_count, decreasing=TRUE),drop=FALSE]

# get names of first 100 genes
selectedGenes <- names(var_count)[1:100]

# ensure same ordering for annotation
annotation <- annotation [colnames(counts), ,drop=FALSE]

# plot pheatmap
pheatmap::pheatmap (counts[selectedGenes, ],
                    annotation_col = annotation,
                    scale = 'row')
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `counts` which holds
RNA-seq normalized count information. 
The rows correspond to the genes and the columns 
represent different samples, with corresponding
row- and columnnames. I have another dataframe called `col_data`. 
Here, the samples are the rows with corresponding rownames and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from control individuals. I want to perform PCA 
to see if control individuals and cancer individuals cluster together.
I want to visualize the results in a scatterplot. I want to annotate
the points in the scatterplot using the `group` column from the `col_data`
dataframe. How do I do this?

completion:
To perform PCA and visualize whether control and cancer individuals 
cluster together using the counts DataFrame and annotate the points 
using the group column from the col_data DataFrame, you can use the 
prcomp function to perform PCA on the transposed counts DataFrame. 
You can choose to scale the data using the scale parameter if necessary.

```{r,eval=FALSE}
require (ggplot2)

# prepare annotation
annotation <- as.factor(col_data[, "group"])
names(annotation) <- rownames(col_data)

# transpose dataframe for PCA
t_counts <- t(counts)

# perform PCA
res <- prcomp(t_counts)$x 

# ensure same ordering
annotation <- annotation[rownames(res)]

# plot first 2 PC's in scatterplot and 
# annotate 
ggplot2::ggplot (data = res,
                 aes(x = PC1,
                     y = PC2)) +
        geom_point(aes(colour = annotation)) +
        geom_text(hjust=-0.1, vjust=0,label=annotation)+
        ggtitle ("PCA of RNA seq samples")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `counts` which holds
RNA-seq normalized count information. 
The rows correspond to the genes and the columns 
represent different samples, with corresponding
row- and columnnames. I want to calculate the 
correlation between samples. How do I do this?

completion:
To calculate the correlation between samples (columns) i
n an RNA-seq normalized count DataFrame like counts in R, 
you can use the cor() function. y default, cor() computes Pearson 
correlation coefficients, which are suitable for most applications. 
If you want to calculate other types of correlation, you can specify 
the method using the method argument (e.g., method = "spearman" 
for Spearman's rank correlation).

```{r,eval=FALSE}

# caluclate correlations
correlation_matrix <- cor(counts)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `counts` which holds
RNA-seq normalized count information. 
The rows correspond to the genes and the columns 
represent different samples, with corresponding
row- and columnnames. The first 5 columns are
from case samples. The next 5 are from control 
samples. I want to calculate the 
correlation between samples. I then want to 
visualize the results in a heatmap and annotate
it with case and control. How do I do this?

completion:
To calculate the correlation between samples (columns) i
n an RNA-seq normalized count DataFrame like counts in R, 
you can use the cor() function. y default, cor() computes Pearson 
correlation coefficients, which are suitable for most applications. 
If you want to calculate other types of correlation, you can specify 
the method using the method argument (e.g., method = "spearman" 
for Spearman's rank correlation). You can then use the `pheatmap` function
in R to plot a heatmap. For the annotation, you can make a dataframe holding
the sample names as rows, with a column for the different annotations.
This you can use to annotate the heatmap. 

```{r,eval=FALSE}
# caluclate correlations
correlation_matrix <- cor(counts)

# create dataframe for annotation
annotation <- data.frame (anno = c(rep("CASE", 5), rep("Control", 5)),
                          row.names = colnames(counts))

# plot heatmap
pheatmap(correlation_matrix,
         annotation_col = annotation)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DESeqDataSet object called `dds`
which holds information from my RNA-seq experiments. 
Genes are in rows, and samples are in columns. 
I want to remove genes which have no count 
across all samples. How do I do this?

completion:
To remove genes that have no counts across all samples in a
DESeqDataSet object in R, you can use the rowSums() function 
to calculate the sum of counts for each gene and then filter 
out the genes with zero counts. 

```{r,eval=FALSE}
require (DESeq2)

# select entries which have a rowsum of 1 or higher
dds <- dds[rowSums(DESeq2::counts(dds)) >= 1, ]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DESeqDataSet object called dds
which holds information from my RNA-seq experiments. 
It holds information of CASE and CTRL groups the annotation
of which is in a column called `groups`. I want to calculate the 
differential expression status of case samples compared 
to the control samples. I then want to order the results 
based on p-value. How do I do this?

completion:
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable. 
Now, we can compare and contrast the samples based on different variables of 
interest. In this case, the group variable that determines if a sample belongs 
to the CASE group or the CTRL group.

```{r,eval=FALSE}
require (DESeq2)

# run DESeq
dds <- DESeq(dds)

#compute the contrast for the 'group' variable where 'CTRL' 
#samples are used as the control group. 
DEresults <- results(dds, contrast = c("group", 'CASE', 'CTRL'))

#sort results by increasing p-value
DEresults <- DEresults[order(DEresults$pvalue), ]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DESeqDataSet object called dds
which holds information from my RNA-seq experiments. 
It holds information of CASE and CTRL groups the annotation
of which is in a column called `groups`. I want to calculate the 
differential expression status of case samples compared 
to the control samples. I then want to order the results 
based on p-value. After this, I want to plot an MA plot. How do 
I do this?

completion:
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable. 
Now, we can compare and contrast the samples based on different variables of 
interest. In this case, the group variable that determines if a sample belongs 
to the CASE group or the CTRL group.
To then plot an MA plot, you can make use of the `DESeq2::plotMA` function. 

```{r,eval=FALSE}
require (DESeq2)

# run DESeq
dds <- DESeq(dds)

# compute the contrast for the 'group' variable where 'CTRL' 
# samples are used as the control group. 
DEresults <- results(dds, contrast = c("group", 'CASE', 'CTRL'))

# sort results by increasing p-value
DEresults <- DEresults[order(DEresults$pvalue), ]

# plot MA plot
plotMA(object = dds, ylim = c(-5, 5))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DESeqDataSet object called dds
which holds information from my RNA-seq experiments. 
It holds information of CASE and CTRL groups the annotation
of which is in a column called `groups`. I want to calculate the 
differential expression status of case samples compared 
to the control samples. I then want to order the results 
based on p-value. After this, I want to plot the P-value distribution. 
How do I do this?

completion:
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable. 
Now, we can compare and contrast the samples based on different variables of 
interest. In this case, the group variable that determines if a sample belongs 
to the CASE group or the CTRL group. You can then plot the P-value 
distribution using the `ggplot` function. 

```{r,eval=FALSE}
require(DESeq2)
require (ggplot2)

# run DESeq
dds <- DESeq(dds)

# compute the contrast for the 'group' variable where 'CTRL' 
# samples are used as the control group. 
DEresults <- results(dds, contrast = c("group", 'CASE', 'CTRL'))

# sort results by increasing p-value
DEresults <- DEresults[order(DEresults$pvalue), ]

# plot P-value distribution
ggplot(data = as.data.frame(DEresults), aes(x = pvalue)) + 
  geom_histogram(bins = 100)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DESeqDataSet object called dds
which holds information from my RNA-seq experiments. 
I want to calculate the differential expression status 
of case samples compared to the control samples. Then 
I want to extract the normalized counts from the resulting 
DESeqDataSet object. How do I do this?

completion:
To calculate the differential expression status of case samples 
compared to control samples using a DESeqDataSet object in R, 
you can use the DESeq2 package, which is designed for differential gene 
expression analysis in RNA-seq data. Performing differential expression analysis
can be done using the `DESeq` function. Extracting the 
normalized counts from the DESeqDataSet object can be done using the 
`DESeq2::counts` function. 

```{r,eval=FALSE}
require(DESeq2)

# run DESeq
dds <- DESeq(dds)

# extract normalized counts from the DESeqDataSet object
countsNormalized <- DESeq2::counts(dds, normalized = TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DESeqDataSet object called dds
which holds information from my RNA-seq experiments. 
I want to calculate the differential expression status 
of case samples compared to the control samples. Then 
I want to extract the normalized counts from the resulting 
DESeqDataSet object. I then want to select the top 500
genes which are most variable across samples using the 
normalized counts, and plot a heatmap of this. 
How do I do this?

completion:
To calculate the differential expression status of case samples 
compared to control samples using a DESeqDataSet object in R, 
you can use the DESeq2 package, which is designed for differential gene 
expression analysis in RNA-seq data. Performing differential expression analysis
can be done using the `DESeq` function. Extracting the 
normalized counts from the DESeqDataSet object can be done using the 
`DESeq2::counts` function.  To select the top 500 genes which are 
most variable across samples we make use of the `var` function. 
You can then plot a heatmap using the `pheatmap` function in R. 

```{r,eval=FALSE}
require(DESeq2)
require (pheatmap)

# run DESeq
dds <- DESeq(dds)

# extract normalized counts from the DESeqDataSet object
countsNormalized <-DESeq2::counts(dds, normalized = TRUE)


# select most variable 500 genes across samples
selectedGenes <- names(sort(apply(countsNormalized, 1, var), 
                            decreasing = TRUE)[1:500])

# create new matrix holding only counts from these genes
selected_counts <- as.matrix(countsNormalized[selectedGenes, ])

# plot heatmap
pheatmap::pheatmap (selected_counts,scale = "row",
                    show_rownames=FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `raw_counts`. 
It holds the raw read counts of each gene for different 
samples. The rows correspond to the genes and the columns 
represent different samples. I want to make an RLE plot
of this data. How do I do this?

completion:
To plot an RLE plot of the read counts in `raw_counts`
you can make use of the `EDASeq` library in R. Here is how 
you do it:

```{r,eval=FALSE}
require(EDASeq)

raw_counts <- as.matrix(raw_counts)

plotRLE(raw_counts, outline=FALSE, ylim=c(-4, 4),
        main = 'Raw Counts',xaxt="n")
axis (las=2,side=1, labels = colnames(raw_counts), at=1:length(colnames(raw_counts)))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DESeqDataSet object called dds
which holds information from my RNA-seq experiments. 
It holds information of CASE and CTRL groups the annotation
of which is in a column called `groups`. I want to calculate the 
differential expression status of case samples compared 
to the control samples. I consider something differentially expressed 
when it has an adjusted p-value of <0.05.Then I want to extract differentially 
expressed genes, and perform GO term enrichment analysis on this set, 
and visualize the results. How do I do this?

completion:
o calculate the differential expression status of case samples 
compared to control samples using a DESeqDataSet object in R, 
you can use the DESeq2 package, which is designed for differential gene 
expression analysis in RNA-seq data. Performing differential expression analysis
can be done using the `DESeq` function. The results can then be extracted using the
`results` function. Performing GO term enrichment can be done using the 
`grpofiler2` library. To visualize the results you can make use of the
`gostplot` function. Here is how you do it:

```{r,eval=FALSE}
require(DESeq2)
require(gprofiler2)

# run DESeq
dds <- DESeq(dds)

# extract results
DEresults <- results(dds, contrast = c('group', 'CASE', 'CTRL'))

# remove results with pvalue of NA
DEresults <- DEresults[!is.na(DEresults[,"padj"]),]

# get differentially expressed genes:
diff_exp <- rownames(DEresults[DEresults[,"padj"]<0.05, ])

# calculate enriched GO terms
goResults <- gost(query = diff_exp, 
                  organism = "hsapiens",
                  user_threshold = 0.05)

gostplot(goResults, capped = TRUE, interactive = TRUE)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a SeqExpressionSet object called `set_s`
from which I removed unwanted variation using the 
`RUVSeq` library. Now I want to extract the normalized
counts from this object. How do I do this?

completion:
To extract the normalized counts from a SeqExpressionSet object
you can make use of the `normCounts` function from the `EDASeq`
library. Here is how you do it:

```{r,eval=FALSE}
require(EDASeq)

# extract normalized counts that are cleared from unwanted variation using RUVs
normCountData <- normCounts(set_s)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a SeqExpressionSet object called `set_s`
from which I removed unwanted variation using the 
`RUVSeq` library. Now I want to extract the normalized
counts from this object, and plot a heatmap of the results using
only the top 500 genes which are most variable across samples, annotating
it with information I stored in a dataframe called `col_data`. This has
samples as rows and 2 columns storing annotation information. 
How do I do this?

completion:
To extract the normalized counts from a SeqExpressionSet object and create a 
heatmap of the top 500 most variable genes, annotated with information from a 
separate dataframe (col_data), you can first extract the normalized counts
using the `normCounts` function from the EDASeq library. You can 
then use the var function to extract the top 500 most variable genes
across all samples, and finally plot a heatmap using the `pheatmap` function
from the pheatmap library. 

```{r,eval=FALSE}
require(EDASeq)
require(pheatmap)

# extract normalized counts that are cleared from unwanted variation using RUVs
normCountData <- normCounts(set_s)

# select top 500 genes based on variance
selectedGenes <- names(sort(apply(normCountData, 1, var), 
                            decreasing = TRUE))[1:500]

# plot annotated heatmap
pheatmap(normCountData[selectedGenes,], 
         annotation_col = col_data, 
         show_rownames = FALSE, 
         scale = "row")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:  data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `counts` which holds
RNA-seq count information. 
The rows correspond to the genes and the columns 
represent different samples, with corresponding
row- and columnnames. I have another dataframe called `col_data`. 
Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. The `group` column holds 
values "CTRL" for control samples and "CASE" for case samples.
This is actually the criteria on which I want to split the samples. 
Furthermore, I have a SeqExpressionSet object called `set_s`
from which I removed unwanted variation in my RNA-seq experiment 
using the `RUVSeq` library using a K=2. Now I want to integrate the sources
of variation to do differential expresison analysis on the original
count data which is in a dataframe called `counts`, based on the
division of the `group` column. How do I do this?

completion:
To integrate the sources of variation found, you can make use of the 
pData function. This will extract all the covariate information from 
your analysis, and this you can use to form a DESeqDataSet object, adding
the newly found covariates to the design formula. This 
you can then use to perform differential expression analysis. 

```{r,eval=FALSE}

col_data <- pData(set_s)

dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = col_data, 
                              # since K=2, we add W_1 and W_2 to the design formula, 
                              # leaving the variable of interest for last.
                              design = ~ W_1 + W_2 + group)

# run differential expression analysis
dds <- DESeq(dds)

# extract results:
res <- results(dds, contrast = c("group", "CASE", "CTRL"))

```

--- TASK END
