---
title: "prompt/completion pairs chapter 8"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. 
I want to calculate the CPM for every column except for the width column. 
How do I do this?

completion:
To calculate the Counts Per Million (CPM) for the columns in 
your raw counts data, you first read in the data using the
`data.table::fread` function. To then calculate
the CPM, you remove the "width" column, calculate 
the sum of counts for each sample, and then calculate the CPM 
by dividing each count by the corresponding sample's total count 
and multiplying by 1 million. Finally, we save the CPM values
in a new variable. 


```{r,eval=FALSE}

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# calculate cpm values
cpm <- apply(subset(raw_counts, select = c(-width)), 2,
             function(x) x/sum(as.numeric(x)) * 10^6)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. I want to compute the RPKM. 
The width column should be used for the length of the genes. 

completion:
To calculate the Reads Per Kilobase Million (RPKM)  for the columns in 
your raw counts data, you first read in the data using the
`data.table::fread` function. To then compute the RPKM using the "width" column 
for the length of the corresponding genes, you multiply the raw reads by 10^9. 
Then divide by the corresponding gene lengths (in kilobases) and divide the above 
result by the sum of all counts for that sample. This gives you the RPKM. 

```{r,eval=FALSE}

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [,"width"])


# compute rpkm 
rpkm <- apply(X = subset(raw_counts, select = c(-width)),
              MARGIN = 2, 
              FUN = function(x) {
                10^9 * x / gene_lengths / sum(as.numeric(x))
               })
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. 
I want to compute the TPM. 
The width column should be used for the length of the genes.
How do I do this?


completion:
To compute the Transcripts Per Million (TPM) values for your 
gene expression data using the provided txt file,
and the gene length information from the "width" column, you
first read in the data using the `data.table::fread` function. Then,
you can calculate the gene length normalized values by dividing 
the counts by the corresponding gene lengths divided by 1000. 
then, divide the above result by the sum of all counts for that sample 
times 1 million. 

```{r,eval=FALSE}

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [,"width"])

# compute TPM
#find gene length normalized values 
rpk <- apply( subset(raw_counts, select = c(-width)), 2, 
              function(x) x/(gene_lengths/1000))
#normalize by the sample size using rpk values
tpm <- apply(rpk, 2, function(x) x / sum(as.numeric(x)) * 10^6)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. 
I also have another txt file called 
"SRP029880.colData.txt". Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. I expect the normal samples
to cluster together, and the cancer samples to cluster together. 
I want to perform PCA (The width column can be discarded 
for this analysis) to see if this is really the case, and plot
the results in a scatterplot. Annotation of the scatterplot 
can be done using the `group` column from the "SRP029880.colData.txt"
file. How do I do this ?

completion:
To perform PCA on your RNA-seq data and create a scatterplot to visualize 
the clustering of normal and cancer samples based on the "group" information 
from the "SRP029880.colData.txt" file, you first read in the data using the
`data.table::fread` function. You can perform PCA with help of the
`prcomp` function. To then visualize the results in a scatterplot, 
you can make use of the `ggplot2` library in R. 

```{r,eval=FALSE}
require(ggplot2)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

col_data <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

# prepare annotation vector
annotation <- as.factor(col_data[, "group"])
names(annotation) <- rownames(col_data)

# prepare data frame so that samples are in rows
# and features are in columns:
data <- t(subset(raw_counts, select = c(-width)))

# perform PCA excluding width column
res <- prcomp(data)$x
res <- as.data.frame(res)

# ensure same ordering in annotation vector
annotation <- annotation[rownames(res)]

# plot first 2 PC's in scatterplot and 
# annotate 
ggplot2::ggplot (data = res,
                 aes(x = PC1,
                     y = PC2)) +
        geom_point(aes(colour = annotation)) +
        geom_text(hjust=-0.1, vjust=0,label=annotation)+
        ggtitle ("PCA of RNA seq samples")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. I want to calculate the TPM and then
select the top 100 genes with the highest variance across
all samples. How do I do this?

completion:
To select the top 100 genes with the highest variance in from
the txt file called "SRP029880.raw_counts.txt", you can first
read in the data using the `data.table::fread` function. Then, 
the TPM is calculated. Next, you can compute the variance of 
each gene across all samples. The result of this analysis is 
ordered, and the names of the 100 genes which showed the 
highest variance can then be selected. 

```{r,eval=FALSE}

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [,"width"])

# compute TPM
#find gene length normalized values 
rpk <- apply( subset(raw_counts, select = c(-width)), 2, 
              function(x) x/(gene_lengths/1000))

#normalize by the sample size using rpk values
tpm <- apply(rpk, 2, function(x) x / sum(as.numeric(x)) * 10^6)

# compute the variance of each gene across samples
var_count <- apply(tpm,1,var)

# order in decreasing value
var_count <- var_count[order(var_count, decreasing=TRUE),drop=FALSE]

# get names of first 100 genes
selectedGenes <- names(var_count)[1:100]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. 
I also have another txt file called 
"SRP029880.colData.txt". Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. I expect the normal samples
to cluster together, and the cancer samples to cluster together.
To check this, I want to create a heatmap. I first want to normalize the 
counts by calculating the TPM. The width column should be used for the 
length of the genes. I then want to create a heatmap of the TPM counts, 
butI want to create this heatmap only with 
the top 100 genes with the highest variance. I want to annotate 
this heatmap using information from the `group` column in the
"SRP029880.colData.txt" txt file . How do I do this?

completion:
To create a heatmap of the 100 genes with the highest variance using the
TPM count, you can first read in the data using the `data.table::fread` 
function. Then, the TPM is calculated. Next, you can compute 
the variance of each gene across all samples. The result of 
this analysis is ordered, and the names of the 100 genes 
which showed the highest variance can then be selected. 
These you can then use to plot a heatmap using
the `pheatmap` library. 

```{r,eval=FALSE}
require(pheatmap)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

col_data <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

# prepare annotation vector
annotation <- col_data[, "group", drop=FALSE]

# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [,"width"])

# compute TPM
#find gene length normalized values 
rpk <- apply( subset(raw_counts, select = c(-width)), 2, 
              function(x) x/(gene_lengths/1000))

#normalize by the sample size using rpk values
tpm <- apply(rpk, 2, function(x) x / sum(as.numeric(x)) * 10^6)

# compute the variance of each gene across samples
var_count <- apply(tpm,1,var)

# order in decreasing value
var_count <- var_count[order(var_count, decreasing=TRUE),drop=FALSE]

# get names of first 100 genes
selectedGenes <- names(var_count)[1:100]

# plot pheatmap
pheatmap::pheatmap (tpm[selectedGenes, ],
                    annotation_col = annotation,
                    scale = "row")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 5 columns are
case samples, and the next 5 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs. I first want to calculate the
TPM. The width column should be used for the 
length of the genes.
Then I want to compute the correlation 
scores between each pair of samples using the TPM counts 
and draw a correlation plot. I also want to draw a
heatmap of the correlation and annotate with either
case or control. How do I do this?

completion:
To create a heatmap of the 100 genes with the highest variance using the
TPM count, you can first read in the data using the `data.table::fread` 
function. Then, the TPM is calculated. Next, you can compute 
the correlation between samples using the `cor` function. 
After this you can visualize the results using `corrplot` and/or 
`pheatmap` functions. 

```{r,eval=FALSE}
require(corrplot)
require(pheatmap)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [,"width"])

# compute TPM
# find gene length normalized values 
rpk <- apply( subset(raw_counts, select = c(-width)), 2, 
              function(x) x/(gene_lengths/1000))

# normalize by the sample size using rpk values
tpm <- apply(rpk, 2, function(x) x / sum(as.numeric(x)) * 10^6)

# create annotation dataframe for later annotation
annotation <- data.frame (anno = c(rep("CASE", 5), rep("Control", 5)),
                          row.names = colnames(tpm))

# caluclate correlations
correlation_matrix <- cor(tpm)

corrplot(correlation_matrix, order = 'hclust', 
         addrect = 2, addCoef.col = 'white', 
         number.cex = 0.7)

# split the clusters into two based on the clustering similarity 
pheatmap(correlation_matrix,
         cutree_cols = 2,
         annotation_col = annotation)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 5 columns are
case samples, and the next 5 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs which sould be discarded. 
I also have another txt file called 
"SRP029880.colData.txt". Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. I want to create a 
DESeqDataSet object of the data in these 2 txt files, which
would seperate based on the column `groups from the "SRP029880.colData.txt"
txt file. How do I do this? 

completion:
To create a DESeqDataSet object from the mentioned txt files, we first
read in the data using the `data.table::fread` function. We can then
create a DESeqDataSet object using the `DESeqDataSetFromMatrix` function.
The DESeqDataSet object contains all the information about the experimental 
setup, the read counts, and the design formulas. Certain functions can be used 
to access this information separately: rownames(dds) shows which features are 
used in the study (e.g. genes), colnames(dds) displays the studied samples, 
counts(dds) displays the count table, and colData(dds) displays the experimental setup.

```{r,eval=FALSE}
require(DESeq2)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- subset(raw_counts, select = c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

#define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData 
dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = col_data,
                              design = as.formula(designFormula))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 5 columns are
case samples, and the next 5 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs which sould be discarded. 
I also have another txt file called 
"SRP029880.colData.txt". Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. The group column holds 
the values `CTRL` or `CASE` based on if it is a control or case
sample. 
I want to create a DESeqDataSet object of the data in these 2 txt 
files, which would seperate based on the column `groups from 
the "SRP029880.colData.txt" txt file. Then, I want to remove 
genes which have no count across all samples. 
Finally, I want to calculate the 
differential expression status of CASE samples compared 
to the CTRL samples. I then want to order the results 
based on p-value. How do I do this?

completion:
To create a DESeqDataSet object from the mentioned txt files, we first
read in the data using the `data.table::fread` function. We can then
create a DESeqDataSet object using the `DESeqDataSetFromMatrix` function.
The DESeqDataSet object contains all the information about the experimental 
setup, the read counts, and the design formulas. Certain functions can be used 
to access this information separately: rownames(dds) shows which features are 
used in the study (e.g. genes), colnames(dds) displays the studied samples, 
counts(dds) displays the count table, and colData(dds) displays the experimental setup.
To remove genes that have no counts across all samples in a
DESeqDataSet object in R, you can use the rowSums() function 
to calculate the sum of counts for each gene and then filter 
out the genes with zero counts. 
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable. 
Now, we can compare and contrast the samples based on different variables of 
interest. In this case, the group variable that determines if a sample belongs 
to the CASE group or the CTRL group.

```{r,eval=FALSE}
require(DESeq2)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- subset(raw_counts, select = c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

#define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData 
dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = col_data,
                              design = as.formula(designFormula))

# select entries which have a rowsum of 1 or higher
dds <- dds[rowSums(DESeq2::counts(dds)) >= 1, ]

# run DESeq
dds <- DESeq(dds)

#compute the contrast for the 'group' variable where 'CTRL' 
#samples are used as the control group. 
DEresults <- results(dds, contrast = c("group", 'CASE', 'CTRL'))

#sort results by increasing p-value
DEresults <- DEresults[order(DEresults$pvalue), ]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 5 columns are
case samples, and the next 5 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs which sould be discarded. 
I also have another txt file called 
"SRP029880.colData.txt". Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. The group column holds 
the values `CTRL` or `CASE` based on if it is a control or case
sample. 
I want to create a DESeqDataSet object of the data in these 2 txt 
files, which would seperate based on the column `groups from 
the "SRP029880.colData.txt" txt file. Then, I want to remove 
genes which have no count across all samples. 
Finally, I want to calculate the 
differential expression status of CASE samples compared 
to the CTRL samples. I then want to order the results 
based on p-value, and plot the p-value distribution.
How do I do this?

completion:
To create a DESeqDataSet object from the mentioned txt files, we first
read in the data using the `data.table::fread` function. We can then
create a DESeqDataSet object using the `DESeqDataSetFromMatrix` function.
The DESeqDataSet object contains all the information about the experimental 
setup, the read counts, and the design formulas. Certain functions can be used 
to access this information separately: rownames(dds) shows which features are 
used in the study (e.g. genes), colnames(dds) displays the studied samples, 
counts(dds) displays the count table, and colData(dds) displays the experimental setup.
To remove genes that have no counts across all samples in a
DESeqDataSet object in R, you can use the rowSums() function 
to calculate the sum of counts for each gene and then filter 
out the genes with zero counts. 
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable. 
Now, we can compare and contrast the samples based on different variables of 
interest. In this case, the group variable that determines if a sample belongs 
to the CASE group or the CTRL group. You can then plot the P-value 
distribution using the `ggplot` function. 

```{r,eval=FALSE}
require(DESeq2)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- subset(raw_counts, select = c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

#define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData 
dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = col_data,
                              design = as.formula(designFormula))

# select entries which have a rowsum of 1 or higher
dds <- dds[rowSums(DESeq2::counts(dds)) >= 1, ]

# run DESeq
dds <- DESeq(dds)

#compute the contrast for the 'group' variable where 'CTRL' 
#samples are used as the control group. 
DEresults <- results(dds, contrast = c("group", 'CASE', 'CTRL'))

#sort results by increasing p-value
DEresults <- DEresults[order(DEresults$pvalue), ]

# plot P-value distribution
ggplot(data = as.data.frame(DEresults), aes(x = pvalue)) + 
  geom_histogram(bins = 100)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 5 columns are
case samples, and the next 5 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs which sould be discarded. 
I also have another txt file called 
"SRP029880.colData.txt". Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. The group column holds 
the values `CTRL` or `CASE` based on if it is a control or case
sample. 
I want to create a DESeqDataSet object of the data in these 2 txt 
files, which would seperate based on the column `groups from 
the "SRP029880.colData.txt" txt file. Then, I want to remove 
genes which have no count across all samples. 
Finally, I want to calculate the 
differential expression status of CASE samples compared 
to the CTRL samples. I then want to extract the
normalized counts. How do I do this?

completion:
To create a DESeqDataSet object from the mentioned txt files, we first
read in the data using the `data.table::fread` function. We can then
create a DESeqDataSet object using the `DESeqDataSetFromMatrix` function.
The DESeqDataSet object contains all the information about the experimental 
setup, the read counts, and the design formulas. Certain functions can be used 
to access this information separately: rownames(dds) shows which features are 
used in the study (e.g. genes), colnames(dds) displays the studied samples, 
counts(dds) displays the count table, and colData(dds) displays the experimental setup.
To remove genes that have no counts across all samples in a
DESeqDataSet object in R, you can use the rowSums() function 
to calculate the sum of counts for each gene and then filter 
out the genes with zero counts. 
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable. 
To then extract the normalized counts, you can make use of the `DESeq::counts`
function. 

```{r,eval=FALSE}
require(DESeq2)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- subset(raw_counts, select = c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

#define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData 
dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = col_data,
                              design = as.formula(designFormula))

# select entries which have a rowsum of 1 or higher
dds <- dds[rowSums(DESeq2::counts(dds)) >= 1, ]

# run DESeq
dds <- DESeq(dds)

# extract normalized counts from the DESeqDataSet object
countsNormalized <-DESeq2::counts(dds, normalized = TRUE)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP029880.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 5 columns are
case samples, and the next 5 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs which sould be discarded. 
I also have another txt file called 
"SRP029880.colData.txt". Here, the samples are the rows and the
file contains 2 columns called `source_name` and `group`. The 
first column describes if the sample is a type of cancer or if
the sample is from a normal individual. The group column holds 
the values `CTRL` or `CASE` based on if it is a control or case
sample. 
I want to create a DESeqDataSet object of the data in these 2 txt 
files, which would seperate based on the column `groups from 
the "SRP029880.colData.txt" txt file. Then, I want to calculate the 
differential expression status of CASE samples compared 
to the CTRL samples. I want to extract the gene names
of genes which are differentially expressed, with an adjusted p-value
of < 0.05. I then want to do GO enrichment analysis and visualize the results. 
How do I do this?

completion:
To create a DESeqDataSet object from the mentioned txt files, we first
read in the data using the `data.table::fread` function. We can then
create a DESeqDataSet object using the `DESeqDataSetFromMatrix` function.
The DESeqDataSet object contains all the information about the experimental 
setup, the read counts, and the design formulas. Certain functions can be used 
to access this information separately: rownames(dds) shows which features are 
used in the study (e.g. genes), colnames(dds) displays the studied samples, 
counts(dds) displays the count table, and colData(dds) displays the experimental setup.
To remove genes that have no counts across all samples in a
DESeqDataSet object in R, you can use the rowSums() function 
to calculate the sum of counts for each gene and then filter 
out the genes with zero counts. 
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable.
The results can be extracted using the `results` function. 
Performing GO term enrichment can be done using the 
`grpofiler2` library. To visualize the results you can make use of the
`gostplot` function. Here is how you do it:

```{r,eval=FALSE}
require(DESeq2)
require(gprofiler2)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- subset(raw_counts, select = c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

#define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData
dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = col_data,
                              design = as.formula(designFormula))

# run DESeq
dds <- DESeq(dds)

# extract results
DEresults <- results(dds, contrast = c("group", "CASE", "CTRL"))

# remove results with pvalue of NA
DEresults <- DEresults[!is.na(DEresults[,"padj"]),]

# get differentially expressed genes:
diff_exp <- rownames(DEresults[DEresults[,"padj"]<0.05, ])

# calculate enriched GO terms
goResults <- gost(query = diff_exp,
                  organism = "hsapiens",
                  user_threshold = 0.05)

gostplot(goResults, capped = TRUE, interactive = TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have raw counts from a RNA-seq experiment
in a txt file called "SRP021193.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 6 columns are
case samples, and the next 6 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs which sould be used for calculating the TPM. 
I also have another txt file called 
"SRP021193.colData.txt". Here, the samples are the rows and the
file contains 3 columns called `diagnosis`, `group` and `librarySelection`. 
The first column describes if the sample is a type of heart disease or if
the sample is from a normal individual. The group column holds 
the values `CTRL` or `CASE` based on if it is a control or case
sample. The `librarySelection` column holds information about how
the library was selected. I want to calculate the TPM counts for the
raw reads, and select the top 100 genes with the greatest variance. Then,
I want to visualize these counts in a heatmap, to see
if `CTRL` and `CASE` groups cluster together, or if for instance 
the difference in library selection causes samples to cluster together. 
How do I do this?

completion:
To compute the Transcripts Per Million (TPM) values for your 
gene expression data using the provided txt file,
and the gene length information from the "width" column, you
first read in the data using the `data.table::fread` function. Then,
you can calculate the gene length normalized values by dividing 
the counts by the corresponding gene lengths divided by 1000. 
then, divide the above result by the sum of all counts for that sample 
times 1 million. The `var` function can be used to select the top 100 genes
with the highest variance across samples. 
To then create a heatmap to see which factors potentially influence
the clustering of the different samples, we can make use of the `pheatmap` library. 
You can annotate the heatmap with the information in the `SRP021193.colData.txt` file. 

```{r,eval=FALSE}
require (DESeq2)
require(pheatmap)

# Reading in and preparing data
raw_counts <- as.data.frame(data.table::fread("SRP021193.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts [, -1]

col_data <- as.data.frame(data.table::fread("SRP021193.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

# calculate TPM counts
# create vector with gene lengths 
gene_lengths <- as.vector(raw_counts [,"width"])

# compute TPM
#find gene length normalized values 
rpk <- apply( subset(raw_counts, select = c(-width)), 2, 
              function(x) x/(gene_lengths/1000))
#normalize by the sample size using rpk values
tpm <- apply(rpk, 2, function(x) x / sum(as.numeric(x)) * 10^6)

# select top 100 genes with greatest variance
selectedGenes <- names(sort(apply(tpm, 1, var),
                            decreasing = T)[1:100])


pheatmap(tpm[selectedGenes,], 
             scale = 'row',
             annotation_col = col_data,
             show_rownames = FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have counts from a RNA-seq experiment
in a txt file called "SRP021193.raw_counts.txt". 
The rows correspond to the genes and the columns 
represent different samples. The first 6 columns are
case samples, and the next 6 are control samples.
There is an additional column 
width that contains the length of the corresponding gene in 
the unit of base pairs which should be discarded
I also have another txt file called 
"SRP021193.colData.txt". Here, the samples are the rows and the
file contains 3 columns called `diagnosis`, `group` and `librarySelection`. 
The first column describes if the sample is a type of heart disease or if
the sample is from a normal individual. The group column holds 
the values `CTRL` or `CASE` based on if it is a control or case
sample. The `librarySelection` column holds information about how
the library was selected. I know that the `librarySelection` is an 
unwanted covariate. I want to perform differential expression analysis
for `CTRL` vs `CASE` (so the `group` column) whilst taking the covariate
into account. How do I do this?

completion:
To perform a differential expression analysis for "CTRL" vs. "CASE" samples 
while taking the covariate "librarySelection" into account, 
you can use the DESeq2 package. DESeq2 is a popular package for analyzing 
RNA-seq data and can handle the incorporation 
of covariates into your analysis. When creating the DESeqDataSet object, 
you can specify the design formula that includes the covariate "librarySelection."
When constructing the design formula, it is very important to pay attention 
to the sequence of variables. We leave the variable of interest to the 
last and we can add as many covariates as 
we want to the beginning of the design formula.
To calculate the differential expression status of case 
samples compared to the control samples you can use the DESeq() 
function of DESeq2, which is a wrapper function that implements 
estimation of size factors to normalize the counts, estimation 
of dispersion values, and computing a GLM model based on the 
experimental design formula. This function returns a 
DESeqDataSet object, which is an updated version of the dds variable.
The results can be extracted using the `results` function. 

```{r,eval=FALSE}
require (DESeq2)

# Reading in and preparing data
counts <- as.data.frame(data.table::fread("SRP021193.raw_counts.txt", header=TRUE))
rownames (counts) <- counts[, 1]
counts <- subset(counts, select=c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP021193.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

# create DESeqDataset whilst taking covariate into account
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = col_data,
                              design = ~ LibrarySelection + group)
# run DESeq
dds <- DESeq(dds)

# extract results
DEresults <- results(dds, contrast = c("group", "CASE", "CTRL"))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have RNA-seq counts from a file named "SRP049988.raw_counts.txt." 
The rows represent genes, and the columns correspond to various samples. 
The initial 5 columns are for case samples, followed by 5 columns for control samples. 
There's an additional column, "width," which contains gene lengths in base pairs 
that should be omitted.I also possess another text file named "SRP049988.colData.txt." 
In this file, the rows represent samples, and it contains 2 columns: "source_name" 
and "group." The "source_name" column describes whether the sample is associated 
with lung carcinoma or a normal individual. The "group" column holds values "CTRL" 
for control samples and "CASE" for case samples. I have identified unwanted variation in the 
data that needs removal. To address this, I've collected a set of housekeeping genes, 
listed in a file called "HK_genes.txt." These genes should exhibit consistent expression 
levels across all samples. I want to utilize this information to eliminate the unwanted variation.
How do I do this?

completion:
To eliminate unwanted variation by using information from the house keeping genes, 
you can make use of the `RUVSeq` library in R. Since we have information of genes
which should not change their expression over the different samples, the `RUVg` function
can be used to remove unwanted variation, based on this information. 
You can run RUVg() with different numbers 
of factors of unwanted variation (k). Below is an example of how to run this function with 
k=1:4.  You can then plot the PCA after removing the unwanted variation. You should be able 
to see which k values (number of factors) produce better separation between sample groups.

```{r,eval=FALSE}
require(RUVSeq)
require(EDASeq)

# Reading in and preparing data
counts <- as.data.frame(data.table::fread("SRP049988.raw_counts.txt", header=TRUE))
rownames (counts) <- counts[, 1]
counts <- subset(counts, select = c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP049988.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

# read in housekeeping gene information
hk_genes <- data.table::fread("HK_genes.txt",header=F)

# convert counts to seqExpressionSet object
set <- newSeqExpressionSet(counts = as.matrix(counts),
                           phenoData = col_data)

# intersection of which housekeeping genes are available
house_keeping_genes <- intersect(rownames(set), hk_genes$V1)

# run RUVg with different k and plot PCA results
par(mfrow = c(2, 2))
for(k in 1:4) {
  set_g <- RUVg(x = set, cIdx = house_keeping_genes, k = k)
  plotPCA(set_g, col=as.numeric(col_data$group), cex = 0.9, adj = 0.5, 
          main = paste0('with RUVg, k = ',k), 
          ylim = c(-1, 1), xlim = c(-1, 1))
}
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization,  numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have RNA-seq counts from a file named "SRP049988.raw_counts.txt." 
The rows represent genes, and the columns correspond to various samples. 
The initial 5 columns are for case samples, followed by 5 columns for control samples. 
There's an additional column, "width," which contains gene lengths in base pairs 
that should be omitted.I also possess another text file named "SRP049988.colData.txt." 
In this file, the rows represent samples, and it contains 2 columns: "source_name" 
and "group." The "source_name" column describes whether the sample is associated 
with lung carcinoma or a normal individual. The "group" column holds values "CTRL" 
for control samples and "CASE" for case samples. I have identified unwanted variation in the 
data that needs removal. I however have no additional information I can use. The variation 
between samples from the 'CTRL' and 'CASE' should be minimized. How do I do this?

completion:
To eliminate unwanted variation by using information from the house keeping genes, 
you can make use of the `RUVSeq` library in R. Since the variation 
between samples from the 'CTRL' and 'CASE' should be minimized, but you have
no additional information to use for this, you can make use of the `RUVs` function. 
You can run RUVs() with different numbers 
of factors of unwanted variation (k). Below is an example of how to run this function with 
k=1:4. You can then plot the PCA after removing the unwanted variation. You should be able 
to see which k values (number of factors) produce better separation between sample groups.

```{r,eval=FALSE}
require(RUVSeq)
require(EDASeq)

# Reading in and preparing data
counts <- as.data.frame(data.table::fread("SRP049988.raw_counts.txt", header=TRUE))
rownames (counts) <- counts[, 1]
counts <- subset(counts, select = c(-1,-width))

col_data <- as.data.frame(data.table::fread("SRP049988.colData.txt", header=TRUE))
rownames(col_data) <- col_data [, 1]
col_data <- col_data [, -1]

# convert counts to seqExpressionSet object
set <- newSeqExpressionSet(counts = as.matrix(counts),
                           phenoData = col_data)

# make a table of sample groups from col_data 
differences <- makeGroups(col_data$group)

par(mfrow = c(2, 2))
for(k in 1:4) {
  set_s <- RUVs(set, unique(rownames(set)), 
                k=k, differences) #all genes
  plotPCA(set_s, col=as.numeric(col_data$group), 
          cex = 0.9, adj = 0.5, 
        main = paste0('with RUVs, k = ',k), 
        ylim = c(-1, 1), xlim = c(-0.6, 0.6))
}
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have RNA-seq raw count data in file "SRP029880.raw_counts.txt" and the 
respective experimental setup data in file "SRP029880.colData.txt".
I would like to perform differential expression analysis and visualize the 
results in a volcano plot.
How can I do this?

completion:
To perform and visualize differential expression analysis we will use 
DESeq2 and ggplot2 package and proceed as follow. 
We read in our count data and experimental setup data using
data.table::fread(), it reads large files faster. 
Next we create a matrix out of the count data without the "width" column, 
define the design formula as "group", to describe the variable of interest 
for our differential expression analysis, "CASE" & "CTRL", which are in "group".
Using the count matrix and the setup data we create a DESeq data set object
using DESeqDataSetFromMatrix() function. Remove genes that have almost 
no information in any of the given samples. Using DESeq() function we estimate 
the size factors to normalize the counts, estimate the dispersion values, 
and compute a GLM model based on the experimental design formula. 
Using results() function we compare and contrast the samples based on 
different variables of interest to perform differential expression 
analysis. We then sort the results by increasing p-value and use them to 
obtain a volcano plot using ggplot() function. 
 
```{r,eval=FALSE}

library(DESeq2)
library(ggplot2)

# read the count data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# create a matrix out of count data without the 'width' column
countData <- as.matrix(subset(raw_counts, select = c(-width)))

# read the experimental setup data
colData <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(colData) <- colData [, 1]
colData <- colData [, -1]

# define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData 
dds <- DESeqDataSetFromMatrix(countData = countData, 
                              colData = colData, 
                              design = as.formula(designFormula))

# for each gene, we count the total number of reads for that gene in all samples 
# -and remove those that don't have at least 1 read
dds <- dds[ rowSums(DESeq2::counts(dds)) > 1, ]

# estimate size factors, dispersion values, gene-wise dispersion,
# -mean-dispersion relationship and computer a GLM model based on 
# -experimental design formula
dds <- DESeq(dds)

# differential expression analysis
DEresults = results(dds, contrast = c("group", 'CASE', 'CTRL'))

# sort results by increasing p-value
DEresults <- DEresults[order(DEresults$pvalue),]

# volcano plot
volplot <- ggplot(data = as.data.frame(DEresults), aes(x = log2FoldChange, y = -log10(pvalue))) +
  # add threshold lines
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
  geom_point() 
  
volplot + theme(text = element_text(size = 10)) 

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have RNA-seq raw count data in file "SRP029880.raw_counts.txt" and the 
respective experimental setup data in file "SRP029880.colData.txt".
I would like to perform differential expression analysis and visualize the 
results in a volcano plot that also indicates which genes are UP- or -DOWN- 
regulated.
How can I do this?

completion:
To perform and visualize differential expression analysis we will use 
DESeq2 and ggplot2 package and proceed as follow. 
We read in our count data and experimental setup data using
data.table::fread(), it reads large files faster. 
Next we create a matrix out of the count data without the "width" column, 
define the design formula as "group", to describe the variable of interest 
for our differential expression analysis, "CASE" & "CTRL", which are in "group".
Using the count matrix and the setup data we create a DESeq data set object
using DESeqDataSetFromMatrix() function. Remove genes that have almost 
no information in any of the given samples. Using DESeq() function we estimate 
the size factors to normalize the counts, estimate the dispersion values, 
and compute a GLM model based on the experimental design formula. 
Using results() function we compare and contrast the samples based on 
different variables of interest to perform differential expression 
analysis. We then sort the results by increasing p-value. To indicate
which genes are UP- or -DOWN- regulated we add a column to the data frame
and specify if the genes are UP- or -DOWN- regulated based on the p-value
significance of their log2FoldChange being positive or negative respectively.
We then obtain a volcano plot using ggplot() function. 

```{r,eval=FALSE}

library(DESeq2)
library(ggplot2)

# read the count data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# create a matrix out of count data without the 'width' column
countData <- as.matrix(subset(raw_counts, select = c(-width)))

# read the experimental setup data
colData <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(colData) <- colData [, 1]
colData <- colData [, -1]

# define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData 
dds <- DESeqDataSetFromMatrix(countData = countData, 
                              colData = colData, 
                              design = as.formula(designFormula))

# for each gene, we count the total number of reads for that gene in all samples 
# -and remove those that don't have at least 1 read
dds <- dds[ rowSums(DESeq2::counts(dds)) > 1, ]

# estimate size factors, dispersion values, gene-wise dispersion,
# -mean-dispersion relationship and computer a GLM model based on 
# -experimental design formula
dds <- DESeq(dds)

# differential expression analysis
DEresults = results(dds, contrast = c("group", 'CASE', 'CTRL'))

# sort results by increasing p-value
DEresults <- DEresults[order(DEresults$pvalue),]

# convert to data frame
DEresults <- as.data.frame(DEresults)

# add a column to the data frame to specify if the genes are UP- or 
# -DOWN- regulated (log2fc respectively positive or negative)
DEresults$diffexpressed <- "NO" 

# if log2Foldchange > 0.0 and pvalue < 0.05, set as "UP"
DEresults$diffexpressed[DEresults$log2FoldChange > 0.0 & DEresults$pvalue < 0.05] <- "UP"

# if log2Foldchange < 0.0 and pvalue < 0.05, set as "DOWN"
DEresults$diffexpressed[DEresults$log2FoldChange < 0.0 & DEresults$pvalue < 0.05] <- "DOWN"

# volcano plot
volplot <- ggplot(data = DEresults, aes(x = log2FoldChange, y = -log10(pvalue), col = diffexpressed)) +
  # add threshold lines
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
  geom_point() + 
  # to set the colours of our variable
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"), 
                     # override and set the categories labels
                     labels = c("Downregulated", "Not significant", "Upregulated")) + 
  labs(color = "Diff Expression") +
  ggtitle('Differential gene expression in "control" vs "case" samples')

volplot + theme(text = element_text(size = 10)) 
  
```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have RNA-seq raw count data in file "SRP029880.raw_counts.txt" and the 
respective experimental setup data in file "SRP029880.colData.txt".
I would like to estimate per-gene dispersion estimate and visualize it 
together with the fitted mean-dispersion relationship.
How can I do this?

completion:
To perform and visualize differential expression analysis we will use 
DESeq2 package and proceed as follow. 
We read in our count data and experimental setup data using
data.table::fread(), it reads large files faster. 
Next we create a matrix out of the count data without the "width" column, 
define the design formula as "group", to describe the variable of interest 
for our differential expression analysis, "CASE" & "CTRL", which are in "group".
Using the count matrix and the setup data we create a DESeq data set object
using DESeqDataSetFromMatrix() function. Remove genes that have almost 
no information in any of the given samples. Using DESeq() function we estimate 
the size factors to normalize the counts, estimate the dispersion values, 
and compute a GLM model based on the experimental design formula. 
We the obtain a dispersion plot using plotDispEsts() function.

```{r,eval=FALSE}

library(DESeq2)

# read the count data
raw_counts <- as.data.frame(data.table::fread("SRP029880.raw_counts.txt", header=TRUE))
rownames (raw_counts) <- raw_counts[, 1]
raw_counts <- raw_counts[, -1]

# create a matrix out of count data without the 'width' column
countData <- as.matrix(subset(raw_counts, select = c(-width)))

# read the experimental setup data
colData <- as.data.frame(data.table::fread("SRP029880.colData.txt", header=TRUE))
rownames(colData) <- colData [, 1]
colData <- colData [, -1]

# define the design formula
designFormula <- "~ group"

# create a DESeq dataset object from the count matrix and the colData 
dds <- DESeqDataSetFromMatrix(countData = countData, 
                              colData = colData, 
                              design = as.formula(designFormula))

# for each gene, we count the total number of reads for that gene in all samples 
# -and remove those that don't have at least 1 read
dds <- dds[ rowSums(DESeq2::counts(dds)) > 1, ]

# estimate size factors, dispersion values, gene-wise dispersion,
# -mean-dispersion relationship and computer a GLM model based on 
# -experimental design formula
dds <- DESeq(dds)

# dispersion plot
plotDispEsts(dds)

```

--- TASK END
