---
title: "prompt/completion pairs chapter 9"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling |  ML/Stats | visualization, numData: 6, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data in the following files: 
GM12878_hg38_CTCF_r1.chr21.bam, GM12878_hg38_CTCF_r2.chr21.bam, 
GM12878_hg38_H3K4me1.chr21.bam,
GM12878_hg38_SMC3_r1.chr21.bam, GM12878_hg38_Input_r1.chr21.bam 
and GM12878_hg38_Input_r2.chr21.bam. These files are in BAM format.
These are from CTCF_r1, CTCF_r2, H3K4me1, SMC3_r1 and input_r1 and 
input_r2 experiments on human chromosome 21 respectivelly. I would 
like to count the amount of reads in these files when using a genomic
window of 1000bp for chromosome 21. Once I have the reads, I want to
normalize these by calculating the Reads Per Million (RPM). I then want
to remove any windows which do not have any reads for any of the experiments.
I then want to see if my different experiments are correlated in any way since
some are replicates of eachother, aby visualizing this using a heatmap. 
How do I do this?

completion:
To count the amount of reads from the ChIP-seq data of chromosome
 21 in the specified files using a genomic window of 1000bp, we 
first need to determine the length of chromosome 21, to calculate 
appropriate tiling windows. This can be done using the 
`GenomeInfoDb::getChromInfoFromUCSC ("hg38")` which retrieves 
chromosome information from the hg38 genome assembly for human. 
We can then make tiling windows spanning over 1000bp each using 
`GenomicRanges::tileGenome`. 
We then get the readcounts using `GenomicAlignments::summarizeOverlaps`. 
After this, we can normalize by dividing each entry in a column by its 
column sum, and multiplying this by 1e6. We can then calculate the pairwise
 Pearson correlation coefficient using the `cor` function. 
Finally we can visualize the correlations in a heatmap using the 
`pheatmap` function. 

```{r,eval=FALSE}
require(GenomicAlignments)
require(GenomeInfoDb)
require (GenomicRanges)

# fetch the chromosome lengths for the human genome
hg_chrs <- GenomeInfoDb::getChromInfoFromUCSC("hg38")

# find the length of chromosome 21
hg_chrs <- subset(hg_chrs, grepl("chr21$", chrom))[,"size"]
names(hg_chrs) <- "chr21"

# create tiling window to use with files of chromosome 21
tiling_window <- unlist(GenomicRanges::tileGenome(hg_chrs, tilewidth=1000))

# specify files of interest
file_list <- file.path(list("GM12878_hg38_CTCF_r1.chr21.bam",
                            "GM12878_hg38_CTCF_r2.chr21.bam",
                            "GM12878_hg38_H3K4me1.chr21.bam",
                            "GM12878_hg38_Input_r1.chr21.bam",
                            "GM12878_hg38_Input_r2.chr21.bam",
                            "GM12878_hg38_SMC3_r1.chr21.bam"))

# get counts
so <- GenomicAlignments::summarizeOverlaps(tiling_window, file_list)
counts <- assays(so)[[1]]

# fix columnnames
colnames(counts) <- c("CTCF_r1", "CTCF_r2", "H3K4me1",
                      "SMC3_r1", "input_r1", "input_r2")

# normalize by rpm
rpm <- counts
col_count <- colSums (counts)
for (i in 1:ncol(counts)){
  rpm[, i] <- (counts[, i]/col_count[i])*1e6
}

# remove tiles which have 0 reads across all columns
# since these provide no information
rpm <- rpm[!rowSums(rpm)==0, ]

# calculate pairwise pearson correlation
correlation_matrix <- cor(rpm, method='pearson')

# create heatmap
library(viridis)
pheatmap::pheatmap(correlation_matrix, color = viridis(10))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from my experiment in the file called 
"GM12878_hg38_CTCF_r1.chr21.bam". This file is in BAM format
and contains information of ChIP-seq data on chromosome 21. 
I want to extend the small reads that I have so they are 200 bp for 
further analysis. I want the extension to happen at the 3'-end 
of the fragments. I then want to export the results as a 
BigWig file. How do I do this?

completion:
You can extend the reads in your ChIP-seq BAM file to 200
base pairs at the 3'-end of the fragments using the `resize`
function in R. We first read in the data using the 
`GenomicAlignments::readGAlignments` function. Then we convert 
the reads to a GRanges object, and resize the reads. We then compress 
the data and create a BigWig file for writing the extended reads. 
Here are the steps to achieve this:

```{r,eval=FALSE}
require(GenomicAlignments)
require(GenomicRanges)
require(rtracklayer)

# read in data
reads_raw <- GenomicAlignments::readGAlignments("GM12878_hg38_CTCF_r1.chr21.bam")

# convert to GRange object
reads_grange <- GenomicRanges::GRanges(reads_raw)

# resize reads
reads <- resize(reads_grange, width=200, fix="start")

# process for saving
reads <- keepSeqlevels(reads, "chr21", pruning.mode="coarse")
span <- max(end(ranges(reads))) - min(start(ranges(reads)))
names(span) <- "chr21"
cov <- coverage(reads, width = span)

# export as bigWig file
output_file <- "new_file.bigWig"
export.bw(cov, "output_file")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from my experiment in the file called 
"GM12878_hg38_CTCF_r1.chr21.bam". This file is in BAM format
and contains information of ChIP-seq data from a CTCF experiment
on chromosome 21. I want to extend the small reads that I have 
so they are 200 bp for further analysis. I want the extension to happen 
at the 3'-end of the fragments. I then want to visualize peaks in 
my ChIP-seq data along the human chromosome 21. How do I do this?

completion:
You can extend the reads in your ChIP-seq BAM file to be 200 
base pairs at the 3'-end of the fragments using the `resize`
function in R. We first read in the data using the 
`GenomicAlignments::readGAlignments` function. Then we convert
the reads to a GRanges object, and resize the reads. To visualize
the peaks in your ChIP-seq data along human chromosome 21, you can 
use the Gviz package. Below is a basic example to create a visualization.
Here we create a DataTrack using the extended reads, and a genome 
axis track. When plotting we set the genomic range. Adjust the range 
to match your region of interest.

```{r,eval=FALSE}
require (GenomeInfoDb)
require(Gviz)
require(GenomicAlignments)
require(GenomicRanges)

# read in data
reads_raw <- GenomicAlignments::readGAlignments("GM12878_hg38_CTCF_r1.chr21.bam")

# convert to GRange object
reads_grange <- GenomicRanges::GRanges(reads_raw)

# resize reads
reads <- resize(reads_grange, width=200, fix="start")

# process for plotting
reads <- keepSeqlevels(reads, "chr21", pruning.mode="coarse")
span <- max(end(ranges(reads))) - min(start(ranges(reads)))
names(span) <- "chr21"
cov <- coverage(reads, width = span)


# convert reads to GRanges object and define the signal track
gcov   <- as(cov, "GRanges")
dtrack <- Gviz::DataTrack(gcov, name = "CTCF",ylim=c(0,50),col.histogram="steelblue",fill.histogram="black")

# create genomic axis track
AT <- Gviz::GenomeAxisTrack()

# define the track ordering 
track_list <- c(dtrack,AT)

# plot
Gviz::plotTracks(track_list, 
    from = min(start(ranges(reads))),
    to = max(end(ranges(reads))),
    window="auto",
    background.title = "black",
    type="histogram"
)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from my experiment in the file called 
"GM12878_hg38_CTCF_r1.chr21.bam". This file is in BAM format
and contains information of ChIP-seq data from a CTCF experiment
on chromosome 21. I want to assess the quality of my ChIP-seq data, 
by using plus and minus strand cross-correlation using the jaccard 
index. I want to plot the results from the cross-correlation analysis
in a scatterplot. How do I do this?

completion:
To assess the quality of your ChIP-seq data using plus and 
minus strand cross-correlation, you can do the following:
Load your ChIP-seq data from the BAM file and prepare it for
analysis. You can use the `GenomicAlignments::readGAlignments` 
function for this. To perform plus and minus strand cross-correlation 
we keep the starting position of each read. We then split the 
reads into minus and plus strand, and calculate the coverage profile. 
The `GenomicAlignments::coverage` counts the coverage per position. 
This is turned into a Boolean vector by asking which genomic positions 
have a coverage > 0. We then shift the coverage vector from the plus 
strand by 1 to 400 base pairs, and for each pair shift we will calculate 
the Jaccard index between the vectors on the plus and minus strand. 
With this, we can plot the results using the `ggplot2::ggplot` function. 

```{r,eval=FALSE}
require(GenomicAlignments)
require(GenomicRanges)
require(ggplot2)

# read in data and convert to GRanges object
reads_raw <- GenomicAlignments::readGAlignments("GM12878_hg38_CTCF_r1.chr21.bam")
reads <- granges(reads_raw)

# keep only the starting position of each read
reads <- resize(reads, width=1, fix="start")

# if you are only interested in keeping information of chromosome 21
# If you want to assess the entire chromosome, you can skip this step.
reads <- keepSeqlevels(reads, "chr21", pruning.mode='coarse')

#  calculate span and split in plus and minus strand
span <- max(end(ranges(reads))) - min(start(ranges(reads)))
names(span) <- "chr21"
reads <- split(reads, strand(reads))

# calculate coverage profile
cov_plus <- GenomicAlignments::coverage(reads[["+"]], width = span)[[1]] > 0
cov_min <- GenomicAlignments::coverage(reads[["-"]], width = span)[[1]] > 0
cov_plus <- as.vector(cov_plus)
cov_min <- as.vector(cov_min)

# define the shift range
wsize <- 1:400

# define jaccard function
jaccard <- function(x,y)sum((x & y)) / sum((x | y))

# shifts the + vector by 1 - 400 nucleotides and
# calculates the correlation coefficient
cc <- shiftApply(
    SHIFT = wsize, 
    X     = cov_plus,
    Y     = cov_min,
    FUN   = jaccard
)

# convert the results into a data frame
cc <- data.frame(fragment_size = wsize, cross_correlation = cc)

# create plot
ggplot2::ggplot(data = cc, aes(fragment_size, cross_correlation)) +
    geom_point() +
    geom_vline(xintercept = which.max(cc$cross_correlation), 
               size=2, color='red', linetype=2) +
    xlab('Shift in base pairs') +
    ylab('Jaccard similarity') 
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from my experiment in the files called 
"GM12878_hg38_CTCF_r1.chr21.bam" and "GM12878_hg38_CTCF_r2.chr21.bam". 
These files are in BAM format and contain information of ChIP-seq data 
from a CTCF experiment on chromosome 21 of human. These two files are 
replicates of the same experiment. I want to look at the GC bias 
quantification for both of these files after normalization of reads per 
million. For reference, I want to use the sequence information of the UCSC 
hg38 genome chromosome 21. To calculate the average GC content, I want to use
a genomic window of 1000bp across the entire chromosome. How do I do this?

completion:
To perform GC bias quantification for ChIP-seq replicates using a genomic 
window of 1000 bp across human chromosome 21 (UCSC hg38) as a reference, we
first read in the chromsome length of chromosome 21. This is used to create 
tiling windows across the genome each window spanning 1000bp. These tiling windows
are then used to count the number of reads from the ChIP-seq replicates per window. 
Counts are then normalized and transformed to log scale for convenient plotting.
To calculate the GC content of tiling widows, the `BSgenome.Hsapiens.UCSC.hg38` 
library is used. Nucleotide frequency per window is calculated and this
is used for plotting. GC content in one kilobase windows is then plotted against
log(reads per million + 1).To avoid taking log(0), a pseudocount of 1 is added.
 
```{r,eval=FALSE}
require(GenomeInfoDb)
require(GenomicRanges)
require(BSgenome.Hsapiens.UCSC.hg38)
require(ggplot2)
require(gridExtra)

#list files
file_list <- file.path(list("GM12878_hg38_CTCF_r1.chr21.bam",
                            "GM12878_hg38_CTCF_r2.chr21.bam"))

# fetch the chromosome lengths for the human genome
hg_chrs <- GenomeInfoDb::getChromInfoFromUCSC("hg38")

# find the length of chromosome 21
hg_chrs <- subset(hg_chrs, grepl("chr21$", chrom))[,"size"]
names(hg_chrs) <- "chr21"

# create tiling window with width of 1000 bp.
tiling_window = unlist(tileGenome(
    seqlengths = hg_chrs,
    tilewidth  = 1000
))

# get counts from ChIP-seq replicates
so <- GenomicAlignments::summarizeOverlaps(tiling_window, file_list)
counts <- SummarizedExperiment::assays(so)[[1]]

# normalize by rpm
rpm <- counts
col_count <- colSums (counts)
for (i in 1:ncol(counts)){
  rpm[, i] <- (counts[, i]/col_count[i])*1e6
}

# because the counts scale often has a large dynamic range
# we transform it using the log function
rpm_log <- log10(rpm+1)


# fetch sequence of human genome.
# Uses the information of the tiling window to fetch chromosome 21.
seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, tiling_window)

# calculates the frequency of all nucleotides 
# in sequences of 100bp of chromosome 21
nuc <- as.data.frame(oligonucleotideFrequency(seq, width = 1))

# calculate the percentage from absolute counts
nuc <- round(nuc/1000,3)

# make dataframe with log_read count info and GC content per window
gc <- cbind(data.frame(rpm_log), GC = (nuc[,"G"]+nuc[,"C"]))

# plot 
plot1 <- ggplot(
    data = gc, 
    aes(
        x = GC, 
        y = GM12878_hg38_CTCF_r1.chr21.bam
    )) +
  geom_point(size=2, alpha=.3) +
  theme_bw() +
  theme(
    axis.text  = element_text(size=10, face="bold"),
    axis.title = element_text(size=14,face="bold"),
    plot.title = element_text(hjust = 0.5)) +
  xlab("GC content in one kilobase windows") +
  ylab("log10( rpm + 1 )") +
  ggtitle("GM12878_hg38_CTCF_r1.chr21.bam")

plot2 <- ggplot(
    data = gc, 
    aes(
        x = GC, 
        y = GM12878_hg38_CTCF_r2.chr21.bam
    )) +
  geom_point(size=2, alpha=.3) +
  theme_bw() +
  theme(
    axis.text  = element_text(size=10, face="bold"),
    axis.title = element_text(size=14,face="bold"),
    plot.title = element_text(hjust = 0.5)) +
  xlab('GC content in one kilobase windows') +
  ylab('log10( rpm + 1 )') +
  ggtitle("GM12878_hg38_CTCF_r2.chr21.bam") 

gridExtra::grid.arrange(plot1, plot2, ncol = 2)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from my experiment in the file called
"GM12878_hg38_CTCF_r1.chr21.bam". This file is in BAM format
and contains information of ChIP-seq data from a CTCF 
experiment on chromosome 21. I want to annotate my reads to
see which reads fall in TSS, intronic, exonic or intergenic 
regions. If reads have mutliple annotations, I want to do the 
following:
If a read falls partially in a TSS area, I want to annotate it
as TSS. If it falls in an exonic region, but also partially in
an intronic or intergenic region, I want to annotate it as exonic.
If it falls within an intronic region, but also partially in an 
intergenic region, I want to annotate it as intronic. So the ordering
is as follows: TSS -> exon -> intron -> intergenic, with TSS having 
the highest priority and the intergenic regions having the lowest
priority. I want to use the location of these regions corresponding 
to the hg38 genome. I want to use version GRCh38.92. 
I then want to plot the results of the frequency of reads falling into
any of these categories. 
How do I do this?

completion:
To annotate your ChIP-seq reads with respect to their 
locations in TSS (Transcription Start Site), exonic, 
intronic, or intergenic regions, we first read in the 
data using the `GenomicAlignments::readGAlignments` function. 
We then convert the reads into a GRange object for further processing.
For this purpose, in the code snippet below the `AnnotationHub` is used.
Connecting to the annotation hub allows us to retrieve genomic 
information from the GRCh38.92 genome annotation from ENSEMBL. We 
retrieve the information of chromosome 21,and use this to annotate 
our reads using the `GenomicAlignments::findOverlaps` 
function. With this we check for any overlaps between 
the GRCh38.92 genome annotation information and the 
ChIP-seq reads. Any reads for which there is no 
overlap will be annotated as intergenic. To ensure 
the correct hierarchy, we order our data based on 
TSS -> exon -> intron -> intergenic, and subsequently
remove any reads with multiple annotations keeping 
only the annotation with the highest priority. The 
frequency of category is then calculated, and plotted 
using the `ggplot2:ggplot` function. 

```{r,eval=FALSE}
require(AnnotationHub)
require(dplyr)
require(GenomicAlignments)

# read in data and convert to GRanges object:
reads <- GenomicAlignments::readGAlignments("GM12878_hg38_CTCF_r1.chr21.bam")
reads <- granges(reads)

# connect to annotation hub
ah <- AnnotationHub()

# get homo sapiens info. 
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]

# fix chromosome names so they fit UCSC naming and keep chr21
# keep only chromosome 21
seqlevels(hum_inf, pruning.mode="coarse") <- paste0("chr", seqlevels(hum_inf))
hum_inf <- hum_inf[seqnames(hum_inf) == "chr21"]
reads <- keepStandardChromosomes(reads, pruning.mode="coarse")
hum_inf <- keepStandardChromosomes(hum_inf, pruning.mode="coarse")

# construct a GRangesList with human annotation
# the order of the list will determine the hierarchy of annotation
annotation_list <- GRangesList(
    # promoters function extends the hum_inf around the TSS
    # by upstream and downstream amount of 1000bp
    tss    = promoters(
      x = subset(hum_inf, type=='gene'),
      upstream   = 1000, 
      downstream = 1000), 
    exon   = subset(hum_inf, type=='exon'),
    intron = subset(hum_inf, type=='gene')
)

# find overlaps between ChIP-seq data and ENSEMBL human data
overlap <- as.data.frame(GenomicAlignments::findOverlaps(reads, annotation_list))

# annotate with names
overlap$annotation <- names(annotation_list)[overlap$subjectHits]

# order the overlaps based on the hierarchy
overlap <- overlap[order(overlap$subjectHits), ]

# select only first found category per read
overlap <- subset(overlap, !duplicated(queryHits))

# find non-overlaps to annotate as intergenic
non_overlap <- which(!reads %over% annotation_list)
non_overlap <- data.frame (queryHits = non_overlap, 
                           subjectHits = rep("NaN",length(non_overlap)),
                           annotation = rep("intergenic",length(non_overlap)))

res <- rbind(overlap,non_overlap)

# group by annotation:
res <- group_by(.data=res, annotation)

# count the number of reads in each category
res <- summarise(.data = res, counts = length(annotation))

# calculate frequency    
res$frequency  <- with(res, round(counts/sum(counts),2))

#give experiment name
res$experiment <- "CTCF"

# plot results
ggplot2::ggplot(data = res, 
       aes(
           x    = experiment, 
           y    = frequency, 
           fill = annotation
        )) +
    geom_bar(stat='identity') +
    theme_bw() +
    scale_fill_brewer(palette='Set2') +
    theme(
        axis.text = element_text(size=10, face='bold'),
        axis.title = element_text(size=14,face="bold"),
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Sample') +
    ylab('Percentage of reads') +
    ggtitle('Percentage of reads in annotation')
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 3, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from 3 different experiments in 
BigWig files "GM12878_hg38_H3K4me3.chr21.bw", 
"GM12878_hg38_H3K36me3.chr21.bw" and "GM12878_hg38_POLR2A.chr21.bw". 
These are H3K4me3, H3K36me3 and POL2 experiments respectively. 
I want to visualize the signal profiles around the highly expressed 
human gene URB1 from chromosome 21. I want to use the annotation of 
these regions corresponding to the hg38 genome, using version GRCh38.92. 
How do I do this?

completion:
To visualize the signal profiles around the highly 
expressed human gene URB1 on chromosome 21 using 
the ChIP-seq BigWig files for H3K4me3, H3K36me3, and
 POL2 experiments, you can follow these steps below.
We first read in the different files using the 
`rtracklayer::import.bw` function. To be able to 
annotate the tracks that will be plotted, we will 
make use of the  `AnnotationHub` library. We use 
this to obtain annotation information corresponding to 
the hg38 genome, using version GRCh38.92. To visualize 
the ChIP-seq data we make use of the `Gviz` library, 
allowing us to construct several `DataTracks` for the 
different experiments and plot them against the annotated 
human URB1 gene.

```{r,eval=FALSE}
require(rtracklayer)
require(AnnotationHub)
require(GenomicFeatures)
require(Gviz)

# loading in data
file_list <- list(H3K4me3 = "GM12878_hg38_H3K4me3.chr21.bw",
                  H3K36me3 = "GM12878_hg38_H3K36me3.chr21.bw",
                  POL2 = "GM12878_hg38_POLR2A.chr21.bw")

chip_profiles <- lapply(file_list, rtracklayer::import.bw)

# connect to annotation hub
ah <- AnnotationHub()

# get homo sapiens info. 
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]

# fix chromosome names so they fit UCSC naming and keep chr21
# keep only chromosome 21
seqlevels(hum_inf, pruning.mode="coarse") <- paste0("chr", seqlevels(hum_inf))
hum_inf <- hum_inf[seqnames(hum_inf) == "chr21"]

# convert info to gene track object to use for later plotting
txdb <- GenomicFeatures::makeTxDbFromGRanges(hum_inf)
gene_track  <- Gviz::GeneRegionTrack(txdb, chr="chr21", genome="hg38")

# find the length of chromosome 21
hg_chrs <- GenomeInfoDb::getChromInfoFromUCSC("hg38")
hg_chrs <- subset(hg_chrs, grepl("chr21$", chrom))[,"size"]
names(hg_chrs) <- "chr21"

# construct genome track
axis <- GenomeAxisTrack(
    range = GRanges('chr21', IRanges(1, width=hg_chrs))
)

# construct data tracks
exp_name <- names(chip_profiles)

data_tracks <- lapply(exp_name, function(exp_name){
    Gviz::DataTrack(
                    range = chip_profiles[[exp_name]], 
                    name  = exp_name,
                    type  = "h",
                   )
})

# focus on URB1 gene
# select the start and end coordinate for the URB1 gene
start <- min(start(subset(hum_inf, gene_name == 'URB1')))
end   <- max(end(subset(hum_inf, gene_name == 'URB1')))

# plot results
Gviz::plotTracks(
    trackList = c(axis, gene_track, data_tracks),
    # relative track sizes
    sizes     = c(1,1,1,1,1), 
    # background color
    background.title     = "black",
    # controls visualization of gene sets
    collapseTranscripts  = "longest", 
    transcriptAnnotation = "symbol",
    # coordinates to visualize 
    from =  start,
    to   =  end
)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF experiment on which I would like to 
do peak calling. The data is contained in a file called 
"GM12878_hg38_CTCF_r1.chr21.bam". As control for background noise, 
I have ChIP-seq data in a file called "GM12878_hg38_Input_r5.chr21.bam". 
Both files are in BAM format. I want to do peak calling using the 
info of the hg38 genome. I am only interested in the enrichment 
on chromosome 21. Furthermore, I would like to only have 
results which have a q-value of lower then 0.01. I would like to
write the results from this analysis to a txt file for further analysis. 
How do I do this?

completion:
Assuming these BAM files are in the same folder as their respecitve
bam.bai files, peak calling can be performed using the `normr::enrichR` function.
After peak The function will automatically create tiling windows (250bp by default),
count the number of reads in each window, and fit a mixture of binomial 
distributions. After peak calling, we can extract these ranges using the 
`normr::getRanges` function, and annotate them with their corresponding 
enrichment- and q-values. We can then filter for q-values < 0.01, and save 
the results in a tab-seperated txt file using the `write.table` function. 

```{r,eval=FALSE}
require(normr)

# prepare data
chip_data <- "GM12878_hg38_CTCF_r1.chr21.bam"
control_data <- "GM12878_hg38_Input_r5.chr21.bam"

# peak calling
ctcf_fit <- normr::enrichR(
            treatment = chip_data,
            control   = control_data,
            genome    = "hg38",
            verbose   = FALSE)

# extract ranges and annotate with q-values and calculated enrichment
ran <- normr::getRanges(ctcf_fit)
ran$qvalue <- normr::getQvalues(ctcf_fit)
ran$enrichment <- normr::getEnrichment(ctcf_fit)

# select regions which correspond to significantly enriched class
ran <- subset(ran,!is.na(component))
ran <- subset(ran,qvalue<0.01)

# set seqLevels to just chromosome 21
# ensure only results for chromosome 21 are kept
seqlevels(ran, pruning.mode = "coarse") <- "chr21"
ran <- subset(ran,seqnames=="chr21")

# write to txt file for further analysis
write.table(ran, 'peaks.txt',
            row.names = F, col.names = T, quote = F, sep = "\t")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF experiment on which I would 
like to do peak calling. The data is contained in a file called 
"GM12878_hg38_CTCF_r1.chr21.bam". 
As control for background noise, I have ChIP-seq data in a file 
called "GM12878_hg38_Input_r5.chr21.bam". Both files are in BAM 
format.I want to do peak calling using the info of the hg38 genome. 
For the peak calling I want to have bins of 250bp, since I expect sharp 
peaks. I am only interested in the enrichment on chromosome 21. 
Furthermore, I would like to only have results which have a q-value of 
lower then 0.01. To then visualize the results in a genome browser format,  
I want to extend the reads 200bp, retrieve the read counts,
and then normalize the read counts by calculating the reads per million. 
I would then like to visualize the signal around the peak which has the lowest 
q-value in a genome browser format. How do I do this?

completion:
To perform ChIP-seq peak calling and downstream analysis in R, 
we can make use of the `normr` library. Assuming these BAM files 
are in the same folder as their respecitve bam.bai files, 
peak calling can be performed using the `normr::enrichR` function.
The function will count the number of reads in each window, and fit
a mixture of binomial distributions. After peak calling, we can extract 
the ranges using the `normr::getRanges` function, and annotate them with
their corresponding enrichment- and q-values. We can then filter for 
q-values < 0.01. To then visualize the results, we retrieve the reads using
the `GenomicAlignments::readGAlignments` and `GenomicRanges::granges` 
function. We extend the reads with 200bp, and retrieve the coverage vector. 
We then normalize the read counts to read counts per million. 
For visualization in a genome browser format we make use of the `Gviz` library. 

```{r,eval=FALSE}
require(normr)
require(GenomicAlignments)
require(GenomicRanges)
require(Gviz)

# prepare data
chip_data <- "GM12878_hg38_CTCF_r1.chr21.bam"
control_data <- "GM12878_hg38_Input_r5.chr21.bam"
reads_chip <- GenomicRanges::granges(GenomicAlignments::readGAlignments(chip_data))
reads_control <- GenomicRanges::granges(GenomicAlignments::readGAlignments(control_data))

# define the window width for the counting
countConfiguration <- normr::countConfigSingleEnd(binsize = 250)

# peak calling
ctcf_fit <- normr::enrichR(
            treatment = chip_data,
            control   = control_data,
            genome    = "hg38",
            countConfig = countConfiguration,
            verbose   = FALSE)

# extract ranges and annotate with q-values and calculated enrichment
ran <- normr::getRanges(ctcf_fit)
ran$qvalue <- normr::getQvalues(ctcf_fit)
ran$enrichment <- normr::getEnrichment(ctcf_fit)

# select regions which correspond to significantly enriched class
ran <- subset(ran, !is.na(component))
ran <- subset(ran, qvalue<0.01)

# set seqLevels to just chromosome 21
seqlevels(ran, pruning.mode = "coarse") <- "chr21"
ran <- subset(ran,seqnames=="chr21")

# order based on qvalue
ran <- ran[order(ran$qvalue), ]

# resize the reads to 200bp
extend <- 200
reads_chip <- resize(reads_chip, width=extend, fix="start")
reads_control <- resize(reads_control, width=extend, fix="start")

# get the coverage vector
cov_chip <- coverage(reads_chip)
cov_control <- coverage(reads_control)

# normalize the coverage vector to the sequencing depth
cov_chip <- round(cov_chip * (1000000/length(reads_chip)),2)
cov_control <- round(cov_control * (1000000/length(reads_control)), 2)

# convert the coverage go a GRanges object
cov_chip <- as(cov_chip, "GRanges")
cov_control <- as(cov_control, "GRanges")


# keep only chromosome 21
cov_chip <- cov_chip[seqnames(cov_chip)=="chr21", ]
cov_control <- cov_control[seqnames(cov_control)=="chr21",]
seqlevels(cov_chip, pruning.mode="coarse") <- "chr21"
seqlevels(cov_control, pruning.mode="coarse") <- "chr21"

# prepare tracks for plotting
# calculate span of reads
span <- max(end(ranges(cov_chip))) - min(start(ranges(cov_chip)))

#chromosome track
axis <- Gviz::GenomeAxisTrack(
    range = GRanges("chr21", IRanges(1, width=span))
)
# peaks track
peaks_track <- Gviz::AnnotationTrack(ran, name = "Peaks")

#data tracks
# adjust limit as needed
lim <- c(0,3000)
chip_track  <- Gviz::DataTrack(
  range = cov_chip,
  name  = "Exp",
  type  = "h",
  lwd   = 3,
  ylim = lim
)

cont_track <- Gviz::DataTrack(
  range = cov_control,
  name  = "Control",
  type  = "h",
  lwd = 3,
  ylim = lim
)

Gviz::plotTracks(
    trackList = list(axis, peaks_track, chip_track, cont_track),
    sizes     = c(.5, .5, 1, 1),
    background.title = "black",
    # selects the range of the read with the highest qvalue. and adds 1000 basepairs 
    # around beginning and end
    from = start(ran)[1] - 1000,
    to   = end(ran)[1]   + 1000
)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF experiment on which I would
 like to do peak calling. The data is contained in a file called 
 "GM12878_hg38_CTCF_r1.chr21.bam". 
As control for background noise, I have ChIP-seq data in a file 
called "GM12878_hg38_Input_r5.chr21.bam". Both files are in BAM format.
I want to do peak calling using the info of the hg38 genome. 
For the peak calling I want to have bins of 250bp, since I expect 
sharp peaks. I am only interested in the enrichment on chromosome 21. 
I then want to have a look at the percentage of reads in peaks and 
visualize this using a barplot. I consider something a peak, when the 
q-value is smaller then 0.05. 
How do I do this?

completion:
To perform ChIP-seq peak calling and downstream analysis in R,
we can make use of the `normr` library. Assuming these BAM files 
are in the same folder as their respecitve bam.bai files, 
peak calling can be performed using the `normr::enrichR` function.
The function will count the number of reads in each window, and fit
a mixture of binomial distributions. After peak calling, we can 
extract the read counts using the `normr::getCounts` function. We 
extract the q-values and make a new column
called enriched which will hold information if this is considered 
a peak or no peak. We then sum the number of reads in the Peak and 
Not Peak regions for control and experiment, and calculate the 
percentage of reads in peak and no peak regions for control and 
experiment. We then plot the results using the 
`ggplot2::ggplot` function. 

```{r,eval=FALSE}
require(normr)
require(GenomicAlignments)
require(GenomicRanges)
require(Gviz)
require(dplyr)
require(tidyr)
require(ggplot2)

# prepare data
chip_data <- "GM12878_hg38_CTCF_r1.chr21.bam"
control_data <- "GM12878_hg38_Input_r5.chr21.bam"
reads_chip <- GenomicRanges::granges(GenomicAlignments::readGAlignments(chip_data))
reads_control <- GenomicRanges::granges(GenomicAlignments::readGAlignments(control_data))

# define the window width for the counting
countConfiguration <- normr::countConfigSingleEnd(binsize = 250)

# peak calling
ctcf_fit <- normr::enrichR(
            treatment = chip_data,
            control   = control_data,
            genome    = "hg38",
            countConfig = countConfiguration,
            verbose   = FALSE)

# extract counts and annotate with q-values and call peaks
counts <- data.frame(normr::getCounts(ctcf_fit))
counts$qvalue <- normr::getQvalues(ctcf_fit)

counts$enriched[counts$qvalue<0.05] <- "Peak"
counts$enriched[counts$qvalue>=0.05] <- "No peak"
counts$enriched[is.na(counts$qvalue)] <- "No peak"
counts$qvalue <- NULL

# reshape the data.frame into a long format
counts_df <- tidyr::pivot_longer(
    data      =  counts, 
    cols      = -enriched,
    names_to  = "experiment",
    values_to = "counts"
)

# sum the number of reads in the Peak and Not Peak regions
counts_df <- dplyr::group_by(.data = counts_df, experiment, enriched)
counts_df <- dplyr::summarize(.data = counts_df, num_of_reads = sum(counts))
    
# calculate the percentage of reads.
counts_df <- dplyr::group_by(.data = counts_df, experiment)
counts_df <- dplyr::mutate(.data = counts_df, total=sum(num_of_reads))
counts_df$percentage <- with(counts_df, round(num_of_reads/total,2))

# plot results
ggplot2::ggplot(
    data = counts_df, 
    aes(
        x = experiment, 
        y = percentage, 
        fill = enriched
    )) +
    geom_bar(stat='identity', position='dodge') +
    xlab('Experiment') +
    ylab('Percetage of reads in region') +
    ggtitle('Percentage of reads in peaks for experiment') +
    scale_fill_manual(values=c('gray','red'))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment
on which I would like to do peak calling. The data is contained 
in a file called "GM12878_hg38_CTCF_r1.chr21.bam". As control for 
background noise, I have ChIP-seq data in a file called 
"GM12878_hg38_Input_r5.chr21.bam". Both files are in 
BAM format. I want to do peak calling using the info of the hg38
genome. For the peak calling I want to have bins of 250bp, since
I expect sharp peaks. I am only interested in the enrichment on 
chromosome 21.  I consider something a peak, when the q-value is 
smaller then 0.05. For subsequent analysis, I want to calculate the 
percentage of CTCF peaks which contain a known CTCF motif. I wan to 
order the peaks I found based on their q-value and plot a reverse 
cumulative distribution of peak occurrences. How do I do this?

completion:
To perform ChIP-seq peak calling and downstream analysis in 
R, we can make use of the `normr` library. Assuming these 
BAM files are in the same folder as their respecitve bam.bai 
files, peak calling can be performed using the `normr::enrichR` 
function.The function will count the number of reads in each 
window, and fit a mixture of binomial distributions. After 
peak calling, we can extract the ranges using the `normr::getRanges`
function, and annotate them with their corresponding
q-values. We can then filter for q-values < 0.5. Subsequently, 
we order the GRange object based on the q-values and use the BSgenome
package to extract the sequences corresponding to the peak regions.
The DNA binding motifs of the CTCF transcription factor can be extracted
from the MotifDB Bioconductor database. The MotifDB is an agglomeration 
of multiple motif databases. This database we use to extract the position
frequency matrix (PFM). Using the `searchSeq` function we then search the
function for the CTCF-motif. The searchSeq function will set a cutoff to 
determine when a sequence contains the motif, and when it doesn't. The cutoff 
is set by determining the maximal possible score produced by the motif matrix;
a percentage of that score is then taken as the threshold value. The 
default setting for this is 80%. Using the results from this analysis, 
we can then plot the reverse cumulative distribution of peak occurrences.

```{r,eval=FALSE}
require(normr)
require(MotifDb)
require(GenomicAlignments)
require(GenomicRanges)
require(Gviz)
require(BSgenome.Hsapiens.UCSC.hg38)
require(TFBSTools)
require(ggplot2)

# prepare data
chip_data <- "GM12878_hg38_CTCF_r1.chr21.bam"
control_data <- "GM12878_hg38_Input_r5.chr21.bam"
reads_chip <- GenomicRanges::granges(GenomicAlignments::readGAlignments(chip_data))
reads_control <- GenomicRanges::granges(GenomicAlignments::readGAlignments(control_data))

# define the window width for the counting
countConfiguration <- normr::countConfigSingleEnd(binsize = 250)

# peak calling
ctcf_fit <- normr::enrichR(
            treatment = chip_data,
            control   = control_data,
            genome    = "hg38",
            countConfig = countConfiguration,
            verbose   = FALSE)

# extract ranges, qvalues and enrichment
ctcf_peaks <- getRanges(ctcf_fit)
ctcf_peaks$qvalue <- getQvalues(ctcf_fit)
ctcf_peaks$enrichment <- getEnrichment(ctcf_fit)

#extract reads with enrichment and qvalue<0.05
ctcf_peaks <- subset(ctcf_peaks,!is.na(component))
ctcf_peaks <- subset(ctcf_peaks,qvalue<0.05)

#order based on qvalue
ctcf_peaks <- ctcf_peaks[order(ctcf_peaks$qvalue), ]

# get position frequency matrix for CTCF motif
# based on the MotifDB version, the location of the CTCF motif
# might change, if you do not get the expected results please try
# to subset with different indices
pfm_ctcf <- MotifDb::query(MotifDb::query(MotifDb, 'Hsapiens'), 'CTCF')[[1]]

#convert pfm matrix to PWM object for later analysis
ctcf_pwm = TFBSTools::PWMatrix(
    ID = 'CTCF', 
    profileMatrix = pfm_ctcf
)

# extend the peak regions with 400bp
ctcf_peaks <- resize(ctcf_peaks, width = 400, fix = 'center')

# extract the sequences around the peaks
seqs <- getSeq(BSgenome.Hsapiens.UCSC.hg38, ctcf_peaks)

# scan each sequence and determine the probability of CTCF binding.
res <- as.data.frame(TFBSTools::searchSeq(ctcf_pwm,seqs,min.score="80%"))

# create dataframe with length of amount of peaks
motif_hits = data.frame(
  peak_order     = 1:length(ctcf_peaks)
)

# annotate if peaks contained motif or not and sort in desending order
motif_hits$contains_motif <- motif_hits$peak_order %in% res$seqnames
motif_hits <- motif_hits[order(-motif_hits$peak_order), ]

# calculate the percentage of peaks with motif for peaks of descending strength
motif_hits$perc_peaks <- with(motif_hits,
                                round(cumsum(contains_motif) / max(peak_order),2))

# plot the cumulative distribution of motif hit percentages
ggplot2::ggplot(
    motif_hits, 
    aes(
        x = peak_order, 
        y = perc_peaks
    )) +
  geom_line(size=2) +
  theme_bw() +
  theme(
    axis.text = element_text(size=10, face='bold'),
    axis.title = element_text(size=14,face="bold"),
    plot.title = element_text(hjust = 0.5)) +
  xlab('Peak rank') +
  ylab('Percetage of peaks with motif') +
  ggtitle('Percentage of CTCF peaks with the CTCF motif')
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment
on chromome 21 on which I did peak calling. The results are stored in
a txt file called `peaks.txt`. This is a tab separated file, which has the 
following columns:
"seqnames",	"start", "end",	"width",	"strand",	"component",	"qvalue" and	
"enrichment". It holds the sequence names, start of the sequence, end of the 
sequence, width of the sequence, the strand information, component information,
q-values and enrichment values respectively. I want to do motif discovery.
I want to select the top 250 peaks based on their q-value, 
which I then want to use for the motif discovery, based on the genome version hg38. 
I want to plot the results and save the resulting PWM matrix into a txt file.
How do I do this?

completion:
To perform motif discovery on the top 250 peaks based on their q-value from a ChIP-seq experiment,
We first read in the table using `data.table::fread()`, and sort it by q-value. We then convert
our dataframe to a GRanges object, and merge nearby peaks to one region. We then expand
the regions with 50bp, to look for the motif,  and use the `rGADEM::GADEM` function for
motif discovery, based on the `BSgenome.Hsapiens.UCSC.hg38::Hsapiens` genome. 
We can then plot the results using the `plot` function, and save the PWM matrix into a txt file
using the `write.table` function. 

```{r,eval=FALSE}
require(GenomicRanges)
require(rGADEM)
require(BSgenome.Hsapiens.UCSC.hg38)
require(TFBSTools)

# read in data
peaks <- data.table::fread("peaks.txt", header=TRUE)

# order based on q-value
peaks <- peaks[order(peaks$qvalue),]

# take first 250 sequences:
peaks <- peaks[1:250,]

# convert to GRanges object
gr_peaks <- makeGRangesFromDataFrame(peaks, keep.extra.columns = TRUE)

# merge nearby peaks
gr_peaks <- reduce(gr_peaks)

# expand region around the peaks with 50bp
peaks_resized <- resize(gr_peaks, width = 50, fix='center')

# run motif discovery
# needs random seed to begin and number of motifs to find
novel_motif <- rGADEM::GADEM(peaks_resized, seed=42, nmotifs=1, genome=BSgenome.Hsapiens.UCSC.hg38::Hsapiens)

# plot novel motif
plot(novel_motif)

# extract the PWM matrix from the GADEM object
unknown_motif <- getPWM(novel_motif)[[1]]

# Save the PWM matrix to a text file
write.table(unknown_motif, file = "new_motif.txt", sep = "\t", quote = FALSE, col.names = NA)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have performed motif discovery, and saved the resulting PWM matrix into a txt file
called "new_motif.txt". This holds the PWM matrix for my newly found motif.
with rownames in the first column. I want to 
find some of the most similar known transcription factor motifs in human to my motif, 
and extract their names. How do I do this?

completion:
To read in the PWM matrix, we make use of the data.table::fread() function.
We convert this data.table to a PWM matrix using the `TFBSTools::PWMatrix`
function. To see which transcription factor motifs are most similar to
your found motif, we can make use of the `JASPAR2018` library. 
Using the `TFBSTools::getMatrixSet` we extract the PWM matrices
for various transcription factor binding motifs. With this, we scan for
similar motifs using the `TFBSTools::PWMSimilarity` function. We
convert the results into a dataframe, and order based on similarity score.
We then assign the appropriate names to our results, and print out
the 5 motifs with the highest similarity score. 

```{r,eval=FALSE}
require(JASPAR2018)
require(TFBSTools)

# read in data
my_motif <-as.matrix(data.table::fread("new_motif.txt", header=TRUE),rownames=1)

# convert the motif to a PWM matrix
unknown_pwm <- TFBSTools::PWMatrix(
    ID = 'unknown', 
    profileMatrix = my_motif
)

# extract motifs corresponding to human transcription factors
pwm_library <- TFBSTools::getMatrixSet(
  JASPAR2018,
  opts=list(
    collection = "CORE",
    species    = "Homo sapiens",
    matrixtype = "PWM"
  ))

# create dataframe out of library for later annotation
pwm_library_dt <- data.frame (ID = ID(pwm_library), name = name(pwm_library))

# scan for similar motifs
pwm_sim <- TFBSTools::PWMSimilarity(pwm_library,
  unknown_pwm, method = 'Pearson')

# make into dataframe
pwm_sim <- as.data.frame(pwm_sim)

#order based on similarity score:
pwm_sim <- pwm_sim [order(pwm_sim$pwm_sim,decreasing=TRUE),,drop=FALSE]

# replace row names with motif name
rownames(pwm_sim)<-c(pwm_library_dt[match(rownames(pwm_sim), pwm_library_dt$ID),"name"])

# show first 5 hits with best similarity score
pwm_sim[1:5, , drop=FALSE]
```

--- TASK END
