---
title: "prompt/completion pairs chapter 9"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I need to know the length of the human chromosome 21 to use for
further analysis. How do I get this information in R?

completion:
To obtain the length of the human chromosome 21 for further analysis 
in R, you can use the BSgenome.Hsapiens.UCSC.hg38 package, which contains the
reference genome information for human chromosome 21. Here's how you can retrieve the
chromosome length:

```{r,eval=FALSE}
require(BSgenome.Hsapiens.UCSC.hg38)

# Access the genome object
genome_hg38 <- BSgenome.Hsapiens.UCSC.hg38

# Get the length of human chromosome 21
chromosome21_length <- seqlengths(genome_hg38)["chr21"]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have counts of reads from my ChIP-seq experiment saved in 
a dataframe called `counts`. This dataframe has 6 columns 
corresponding to 6 different experiments, and the rows represent
the different tiling windows I used for these counts. 
I want to normalize my reads by reads per million per experiment.
How do I do this?

completion:
To normalize the read counts by reads per million (RPM) per experiment in R, 
you first calculate the total number of reads for each experiment. 
Then, for all columns we divide the read count by the total nr of reads in that experiment
and multiply by 1000. 

```{r,eval=FALSE}

# normalize by rpm
rpm <- counts

# total nr of reads per experiment
col_count <- colSums (counts)

# for all columns
# divide the read count by the total nr of reads in that experiment
# and multiply by 1000. 
for (i in 1:ncol(counts)){
  rpm[, i] <- (counts[, i]/col_count[i])*1e6
}
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have counts of reads from my CHiP-seq experiment saved in 
a dataframe called `counts`. This dataframe has 6 columns 
corresponding to 6 different experiments, and the rows represent
the different tiling windows I used for these counts. I want to 
see if there is any correlation between my experiments,
and plot the results in a heatmap. How do I do this?

completion:
To determine the correlation between the experiments based on the 
read counts and create a heatmap in R, you can make use of the `cor`
function in R. To then create a heatmap of our results, you can make 
use of the `pheatmap::pheatmap` function. 

```{r,eval=FALSE}
require(viridis)
require(pheatmap)

# calculate pairwise parson correlation
correlation_matrix <- cor(counts, method='pearson')

# create heatmap
pheatmap::pheatmap(correlation_matrix, color = viridis(10))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have CHiP-seq data in a variable called `reads`. This is
a GRanges object. I want to extend the reads that I have with 200 bp 
for further analysis. I want the extension to happen at the 3'-end 
of the fragments. How do I do this?

completion:
To extend the reads in a GRanges object at the 3'-end by 200 base pairs in R, 
you can use the resize function from the GenomicRanges package.
The width argument can be used to determine the size of the extension, and 
the fix argument is set to 'start' so that extension happens at the 3'-end. 

```{r,eval=FALSE}
require(GenomicRanges)

#calculate width of new size
new_size <- width(ranges(reads)) + 200

# resize reads
reads <- GenomicRanges::resize(reads, width=new_size, fix="start")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have CHiP-seq data in a variable called `reads`. This is
a GRanges object. I want to extend the reads that I have so that 
my reads are 200 bp in total for further analysis. I want the extension 
to happen at the 3'-end of the fragments. How do I do this?

completion:
To extend the reads in a GRanges object at the 3'-end so they are 200
base pairs in total,you can use the resize function from the GenomicRanges package.
The width argument can be used to determine the new size of the fragments, and 
the fix argument is set to 'start' so that extension happens at the 3'-end. 

```{r,eval=FALSE}
require (GenomicRanges)

# resize reads
reads <- GenomicRanges::resize(reads, width=200, fix="start")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have CHiP-seq data in a variable called `reads`. This is
a GRanges object. I want to extend the reads that I have with 200 bp 
for further analysis. I want the extension to happen at the 5'-end 
of the fragments. How do I do this?

completion:
To extend the reads in a GRanges object at the 3'-end by 200 base pairs in R, 
you can use the resize function from the GenomicRanges package.
The width argument can be used to determine the size of the extension, and 
the fix argument is set to 'end' so that extension happens at the 5'-end. 

```{r,eval=FALSE}
require (GenomicRanges)

new_size <- width(ranges(reads)) + 200

# resize reads
reads <- GenomicRanges::resize(reads, width=new_size, fix="end")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have CHiP-seq data in a variable called `reads`. This is
a GRanges object. I want to extend the reads that I have so that 
my reads are 200 bp in total for further analysis. I want the extension 
to happen at the 5'-end of the fragments. How do I do this?

completion:
To extend the reads in a GRanges object at the 3'-end so they are 200
base pairs in total,you can use the resize function from the GenomicRanges package.
The width argument can be used to determine the new size of the fragments, and 
the fix argument is set to 'end' so that extension happens at the 5'-end. 

```{r,eval=FALSE}
require (GenomicRanges)

# resize reads
reads <- GenomicRanges::resize(reads, width=200, fix="end")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data in a variable called `reads`. This is
a GRanges object. I want to extend these reads so that
in total these are 1000 bp long. I want the extension 
to happen on both sides equally. How do I do this?

completion:
To extend the reads in a GRanges object at the 3'-end so they are 1000
base pairs in total,you can use the resize function from the GenomicRanges package.
The width argument can be used to determine the new size of the fragments, and 
the fix argument is set to 'center' so that extension happens at both sides.

```{r,eval=FALSE}
require (GenomicRanges)

# resize reads
reads <- GenomicRanges::resize(reads, width=1000, fix="center")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data in a variable called `reads`. This is
a GRanges object, and contains information of all the reads
on chromosome 21 of human. I want to visualize peaks in 
my ChIP-seq data along the human chromosome 21. How do I do this?

completion:
To visualize peaks in your ChIP-seq data along human chromosome 21 using 
a GRanges object, you first make sure to only keep the data of `chr21`. 
You then fetch the length of chromosome 21, and 
use this as an argument in the `coverage` function, which creates the 
coverage vector. you can use the Gviz package in R to visualize your results.
Create a Tracks object to represent your data. You can create one or more tracks for 
different types of information you want to display. In your case, you might have one track 
for the peaks. You can also create a GenomeAxis track for further visualization. 
Then, create a Gviz plot using the created tracks.

```{r,eval=FALSE}
require(Gviz)
require(GenomicAlignments)
require(GenomicRanges)
require(GenomeInfoDb)

# ensyre only 'chr21' information is there
reads <- subset(reads,seqnames=="chr21")
reads <- keepSeqlevels(reads, "chr21", pruning.mode="coarse")

# calculate total length of chr 21
hg_chrs <- GenomeInfoDb::getChromInfoFromUCSC("hg38")
hg_chrs <- subset(hg_chrs, grepl("chr21$", chrom))[,"size"]
names(hg_chrs) <- "chr21"

#calculate coverage vector
cov <- coverage(reads, width = hg_chrs)

# convert coverage vector to GRanges object and define the signal track
gcov   <- as(cov, "GRanges")
dtrack <- Gviz::DataTrack(gcov, name = "CTCF",ylim=c(0,50),col.histogram="steelblue",fill.histogram="black")

# create genomic axis track
AT <- Gviz::GenomeAxisTrack()

# define the track ordering and plot
track_list <- c(dtrack,AT)
Gviz::plotTracks(track_list, 
    from = min(start(ranges(reads))),
    to = max(end(ranges(reads))),
    window="auto",
    background.title = "black",
    type="histogram"
)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I want to retrieve ENSEMBL information on the genomic annotation
such as intron, exon, or gene of different locations in the human genome 
using a GRCh38.92 assembly. I want to retrieve a GRanges object so that
I can use this for annotation. How do I do this?

completion:
To retrieve ENSEMBL information on genomic annotation, we can make
use of the `AnnotationHub` library. Here we look for the following keywords:
- ENSEMBL to retrieve the ENSEMBL information
- "Homo" to retrieve the information of homo sapiens
- "GRCh38.92" to retrieve the correct version
- and "chr.gtf" to grab the correct file 
From this we retrieve the correct entry for the annotationhub,
which we can use to extract the human annotation information. 

```{r,eval=FALSE}
require(AnnotationHub)

# connect to annotation hub
ah <- AnnotationHub()

# get homo sapiens info. 
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 GRanges objects. One called `reads` and the
other called `hum_inf`. They contain information about my
ChIP-seq experiment and info about ENSEMBL annotation 
of the human genome respectively. The variable `hum_inf` 
has an extra column called `type`, which holds information
about if a range is a gene, transcript, exon, intron, etc. 
I want to see where these 2 GRanges objects overlap, and annotate
the ranges from my experiment with their respective type. 
How do I do this?

completion:
Assuming appropriate seqnames and seqLevels in these 2 GRanges objects,
we can look for overlaps between these objects using the 
`findOverlaps` function from the GenomicAlignments library. 
To annotate the overlaps based on the `type` column in the hum_inf
object, we extract the names for annotation using the 
`hum_inf[overlaps$subjectHits]$type` indexing, and create a 
new column called annotation for the overlaps object. 

```{r,eval=FALSE}
require(GenomicAlignments)

# find overlaps between GRanges objects
overlaps <- as.data.frame(findOverlaps(reads,hum_inf))

# annotate using the $type column
overlaps$annotation <- c(hum_inf[overlaps$subjectHits]$type)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called 'overlaps' which holds information
about overlaps in genomic location between a query and a subject.
This dataframe has 3 columns, called "queryHits", "subjectHits" and
"annotation" respectively. I want to count the amount of
times certain annotations occur in this dataframe, and subsequently calculate
their relative frequency. How do I do this?

completion:
To group by a certain column in a dataframe and count the number of times each 
annotation appears within that group, you can use the dplyr package in R.
We can use the `group_by` function to group our data by the column `annotation`.
We can then summarize the data, making a column count which is
holds the counts of the different annotation types. Finally, to calculate
the relative frequency we divide each count by the total sum of all counts.

```{r,eval=FALSE}
require (dplyr)

# we first group the data by annotation
res <- group_by(.data=overlaps, annotation)

# we then summarize the data making a column count which is
# holds the counts of the different annotation types. 
res <- summarize(.data=res, count=length(annotation))
res$frequency <- round((res$count / sum(res$count)),2)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I want to retrieve ENSEMBL information on the genomic annotation
such as intron, exon, or gene of the URB1 gene in the human genome 
using a GRCh38.92 assembly. I want to retrieve a GRanges object so that
I can use this for annotation. How do I do this?

completion:
To retrieve ENSEMBL information on genomic annotation, we can make
use of the `AnnotationHub` library. Here we look for the following keywords:
- ENSEMBL to retrieve the ENSEMBL information
- "Homo" to retrieve the information of homo sapiens
- "GRCh38.92" to retrieve the correct version
- and "chr.gtf" to grab the correct file 
From this we retrieve the correct entry for the annotationhub,
which we can use to extract the human annotation information. 
We can then subset this information to only hold the GRanges information
of the "URB1" gene by using the `subset` function. 

```{r,eval=FALSE}
require (AnnotationHub)
# connect to annotation hub
ah <- AnnotationHub()

# get homo sapiens info. 
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]

#Look for gene_name = "URB1"
gr_URB1<-subset(hum_inf, gene_name == "URB1")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 2 GRanges objects called `chip_data` and `control_data`. 
This is from a ChIP-seq experiment on the human genome. 
I want to perform peak calling using the first object as the treatment,
and the second object as the control. How do I do this?

completion:
To perform peak calling using a treatment (ChIP-seq) GRanges object and a 
control GRanges object in R, you can use a peak calling tool such as the
`normr::enrichR`. For this it is important to define the window width for the 
counting. In the example below we use a binsize of 250. However, you can adjust this based 
on the specifics of your experiment. After peak calling, we can extract the ranges, q-values 
and enrichment values using the appropriate `normr` functions. Here is an example of how 
to do it:

```{r,eval=FALSE}
require (normr)

# define the window width for the counting
countConfiguration <- normr::countConfigSingleEnd(binsize = 250)

fit <- normr::enrichR(
            treatment = chip_data,
            control   = control_data,
            genome    = "hg38",
            countConfig = countConfiguration,
            verbose   = FALSE)

# extract ranges and annotate with q-values and calculated enrichment
res <- normr::getRanges(fit)
res$qvalue <- normr::getQvalues(fit)
res$enrichment <- normr::getEnrichment(fit)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment 
on which I did peak calling. The results of this peak calling
are in the variable `ctcf_fit`. This is a NormRFit-class object.
From this, I want to extract the ranges, q-values and enrichment scores.
I then want to extract the reads which have enrichment and a q-value
of lower then 0.05. I then want to order the result based on q-value. 
How do I do this?

completion:
To extract the ranges, q-values, and enrichment scores from a ChIP-seq
peak calling result stored in a NormRFit-class object, filter the peaks 
based on q-value and enrichment score, and order the results based on q-value, 
you can use the `normr` library. This library holds the functions `getRanges`,
`getQvalues` and `getEnrichment` to extract the wanted information. 
The resulting GRanges object called ctcf_peaks we can then manipulate
to retrieve the wanted information, and order based on q-value. 

```{r,eval=FALSE}
require(normr)

# extract ranges, qvalues and enrichment
ctcf_peaks <- getRanges(ctcf_fit)
ctcf_peaks$qvalue <- getQvalues(ctcf_fit)
ctcf_peaks$enrichment <- getEnrichment(ctcf_fit)

#extract reads with enrichment and qvalue<0.05
ctcf_peaks <- subset(ctcf_peaks,!is.na(component))
ctcf_peaks <- subset(ctcf_peaks,qvalue<0.05)

#order based on qvalue
ctcf_peaks <- ctcf_peaks[order(ctcf_peaks$qvalue), ]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment 
on which I did peak calling. The results of this are stored in 
a variabe called `ctcf_peaks`. This is a GRanges object. I 
want to resize the ranges so that the region around the peaks
is +/- 1kb. How do I do this?

completion:
To resize a GRanges object, we can make use of the `resize` function
from the `GenomicRanges` library. To resize the ranges so that the region
around the found peaks is +/- 1kb, we set the `width` argument to 2000, 
and the `fix` argument to 'center'.

```{r,eval=FALSE}
require (GenomicRanges)
ctcf_peaks <- resize(ctcf_peaks, width = 2000, fix='center')
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment 
on which I did peak calling. The results of this are stored in 
a variabe called `ctcf_peaks`. This is a GRanges object. I now
want to retrieve the sequences which belong to these peak regions
using a hg38 version of the human genome. How do I do this?

completion:
To retrieve the sequences that belong to the peak regions in your ChIP-seq 
data (stored in a GRanges object called ctcf_peaks) using the hg38 version 
of the human genome, you can use the Bioconductor package GenomicRanges along 
with the BSgenome package. Here's how you can do it in R:

```{r,eval=FALSE}
require(GenomicRanges)
require(BSgenome.Hsapiens.UCSC.hg38)

seq <- getSeq(BSgenome.Hsapiens.UCSC.hg38, ctcf_peaks)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment 
on which I did peak calling. The results of this are stored in 
a variabe called `ctcf_peaks`. This is a GRanges object. I want
to retrieve the sequences of these peaks, using a hg38 version of the 
human genome.. I then want to perform motif localization for the CTCF motif. 
How do I do this?

completion:
To retrieve the sequences of the ChIP-seq peaks (stored in a GRanges 
object called ctcf_peaks) and perform motif localization for the CTCF 
motif, we can make use of the `MotifDb` library, which allows us to retreive 
the CTCF motif. We can retrieve the sequences associated to the peaks using 
the Bioconductor package GenomicRanges along with the BSgenome package.
We can then use the `searchSeq` function to scan the sequences associated
to the peaks for the CTCF motif. 

```{r,eval=FALSE}
require(BSgenome)
require(MotifDb)
require(TFBSTools)

# get position frequency matrix for CTCF motif
# based on the MotifDB version, the location of the CTCF motif
# might change, if you do not get the expected results please try
# to subset with different indices
pfm_ctcf <- MotifDb::query(MotifDb::query(MotifDb, "Hsapiens"), "CTCF")[[1]]

#convert pfm matrix to PWM object for later analysis
ctcf_pwm <- TFBSTools::PWMatrix(
    ID = "CTCF", 
    profileMatrix = pfm_ctcf
)

# fetch the sequence of the peaks
seq <- BSgenome::getSeq(BSgenome.Hsapiens.UCSC.hg38,ctcf_peaks)

# and scan the peaks
hits <- TFBSTools::searchSeq(ctcf_pwm, seq, min.score = "80%", strand = "*")
hits <- as.data.frame(hits)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I want to retrieve the binding sequence from the transcription
factor called CTCF. I want to have the PWM matrix for this
sequence. How do I do this?

completion:
To retrieve the Position Weight Matrix (PWM) for the CTCF transcription 
factor's binding site in R, you can use a motif database or a motif analysis package. 
One commonly used motif database in R is the MotifDb database. Here's how you can retrieve the 
PWM for the CTCF motif:

```{r,eval=FALSE}
require(MotifDb)
require(TFBSTools)

# Scan the MotifDb database for entries containing
# Hsapiens and CTCF
pfm_ctcf <- MotifDb::query(MotifDb::query(MotifDb, 'Hsapiens'), 'CTCF')[[1]]

#convert pfm matrix to PWM object
ctcf_pwm <- TFBSTools::PWMatrix(
    ID = "CTCF", 
    profileMatrix = pfm_ctcf
)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats,  numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a DNAStringSet object called `seqs` containing sequences that
belong to peaks in my ChIP-seq data. I want to scan these
sequences to see if they contain the motif from 
transcription factor CTCF. I have the motif from CTCF in a
variable called `pwm_ctcf`. This is an object of class PWMatrix.
How do I scan these sequences to see if they contain
the motif, and make sure I only get hits that have 
a score of at least 90%?

completion:
To scan sequences for the presence of a motif from the CTCF transcription 
factor using a position weight matrix (PWM) and filter for hits with a score 
of at least 90%, you can use the Bioconductor package TFBSTools. By making
use of the `searchSeq` function we can search the sequences in variable
`seqs` for the motif in `ctcf_pwm`. By using the argument min.score we can 
determine the minimum score percentage. 
Here's how you can do it:

```{r,eval=FALSE}
require (TFBSTools)

# scan each sequence and determine the probability of CTCF binding.
res <- TFBSTools::searchSeq(ctcf_pwm,seqs,min.score="90%")

# convert to dataframe 
res <- as.data.frame(res)

# print sequence names:
print(res$seqnames)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling , numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I performed motif scanning on my DNAStringSet object `seqs`,
using a variable called `pwm_ctcf`,which is an object of class PWMatrix.
From this I got the results which I stored in a variable called `res`. This
is a SiteSetList object. How do I retrieve the sequences 
that contained the motif, and also print out their percentage scores?

completion:
To retrieve sequences containing the motif from a SiteSetList object, 
you can access the sequences by first converting the `res` SiteSetList object
to a dataframe. After this, we can access which sequences contained the motif
by accessing the `seqnames` column. This we can use to retrieve the sequences
which contained the motif from the DNAStringSet object `seqs`, and also
retrieve the relative scores in percentages. 

```{r,eval=FALSE}
require(TFBSTools)

# Retrieve sequences containing the motif
res <- as.data.frame(res)
seqs_motif_idx <- as.numeric(res$seqnames)
sequences_motif <- seqs[seqs_motif_idx]

# retrieve scores:
scores <- res$relScore
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I performed motif scanning on my DNAStringSet object `seqs`,
using a variable called `pwm_ctcf`,which is an object of class PWMatrix.
From this I got the results which I stored in a variable called `res`. This
is a SiteSetList object. I want to retrieve the subsequence of where the match
is located. How do I do this?

completion:
To retrieve subsequences which matched the motif from a SiteSetList object, 
you can access the sequences by first converting the `res` SiteSetList object
to a dataframe. This dataframe will contain a column called `seqSites` which
holds the subsequences of where the match was made. 

```{r,eval=FALSE}
# Convert SiteSetlist to dataframe
res <- as.data.frame(res)

# retrieve subsequences of match
subseqs <- res$siteSeqs
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I performed motif scanning on my DNAStringSet object `seqs`,
using a variable called `pwm_ctcf`,which is an object of class PWMatrix.
From this I got the results which I stored in a variable called `res`. This
is a SiteSetList object. I want to retrieve the start and end positions of where
the match was found. How do I do this?

completion:
To retrieve the start and end positions of where
the match was found you can access the information of 
the match by first converting the `res` SiteSetList object
to a dataframe. This dataframe will contain columns called
`start` and `end` which holds the starting and ending positions
of the match. 

```{r,eval=FALSE}

# Convert SiteSetlist to dataframe
res <- as.data.frame(res)

# access the start and end columns:
starts <- res$start
ends <- res$end
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a GRanges object called `peaks` which
holds the information about peaks I found in my ChIP-seq experiment.
This experiment was done on chromosome 21 of human. I want to perform motif discovery.
Then I want to plot the newly found motif as a sequence logo. I also 
want to extract the PWM of this newly found motif.
How do I do this?

completion:
To perform motif discovery on your found peaks
which are contained in the GRanges object called `peaks`,
you can make use of the rGADEM library. This is an an unsupervised, 
stochastic motif discovery tools, which uses sampling with subsequent 
enrichment analysis to find over-represented sequence motifs.
You can then plot the sequence logo using the `plot` function. 
To extract the PWM you can make use of the `getPWM` function. 

```{r,eval=fALSE}
require(BSgenome.Hsapiens.UCSC.hg38)
require (rGADEM)

# perform motif discovery using 
# the BSgenome.Hsapiens.UCSC.hg38::Hsapiens genome
# as reference. 
novel_motif <- rGADEM::GADEM(peaks, seed=42, nmotifs=1, genome=BSgenome.Hsapiens.UCSC.hg38::Hsapiens)

# plot sequence logo
plot (novel_motif)

# extract the PWM matrix from the GADEM object
pwm_motif <- getPWM(novel_motif)[[1]]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment 
on which I did peak calling. I then scanned the results of my peaks
to see if the CTCF motif was present in these peaks. The results 
of this are stored in a variable called `hits`. This is a dataframe with the
following columns:"seqnames", "source", "feature",  "start", "end", "absScore",
"relScore", "strand", "ID", "TF", "class" and "siteSeqs". I want to construct a 
plot, where the X-axis represents where the hit starts, and the y-axis the enrichment 
at that position.  Since I extended my reads with 1kb +/- on each side,
I first want to subtract 1000 from the start position and then plot. How do I do this?

completion:
To construct a plot where the X-axis represents the position where the hit starts 
(with a subtraction of 1000) and the Y-axis represents the enrichment at that position 
based on your ChIP-seq data, the `ggplot2` library. Here is how you do it in R:

```{r,eval=FALSE}
require(ggplot2)

#create new column called position
hits$position <- hits$start-1000

#plot results
ggplot(data=hits, aes(position)) +
    geom_density(size=2) +
    xlab("Start position") +
    ylab("Enrichment")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment
on chromome 21 on which I did peak calling. The results are stored in a variable
called `CTCF_peaks`. This is a GRanges object. I want to visualize the 
distribution of peaks in different functional genomic regions, namely
TSS, exon, intron and intergenic regions. 
If a read falls partially in a TSS area, I want to annotate it
as TSS. If it falls in an exonic region, but also partially in
an intronic or intergenic region, I want to annotate it as exonic.
If it falls within an intronic region, but also partially in an 
intergenic region, I want to annotate it as intronic. So the ordering
is as follows: TSS -> exon -> intron -> intergenic, with TSS having 
the highest priority and the intergenic regions having the lowest
priority.I then want to visualize these results in a barplot. 
How do I do this?

completion:
To visualize the distribution of ChIP-seq peaks in different functional genomic 
regions (TSS, exon, intron, and intergenic regions) for chromosome 21 using a GRanges 
object called `CTCF_peaks`, we make use of the `AnnotationHub` library. 
Connecting to the annotation hub allows us to retrieve genomic 
information from the GRCh38.92 genome annotation from ENSEMBL. We 
retrieve the information of chromosome 21,and use this to annotate 
our reads using the `GenomicAlignments::findOverlaps` 
function. With this we check for any overlaps between 
the GRCh38.92 genome annotation information and the 
ChIP-seq reads. Any reads for which there is no 
overlap will be annotated as intergenic. To ensure 
the correct hierarchy, we order our data based on 
TSS -> exon -> intron -> intergenic, and subsequently
remove any reads with multiple annotations keeping 
only the annotation with the highest priority. The 
frequency of category is then calculated, and plotted 
using the `ggplot2:ggplot` function. 

```{r,eval=FALSE}
require(AnnotationHub)
require(dplyr)
require(GenomicAlignments)
require(ggplot2)

#connect to annotationhub
ah <- AnnotationHub()

#query annotation hub for human file
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]

# fix chromosome names so they fit UCSC naming and keep chr21
# keep only chromosome 21
seqlevels(hum_inf, pruning.mode="coarse") <- paste0("chr", seqlevels(hum_inf))
hum_inf <- hum_inf[seqnames(hum_inf) == "chr21"]


# Make annotation list. Any read which does not 
# produce a hit will be labeled as intergenic
annotation_list <- GRangesList(
    # promoters function extends the hum_inf around the TSS
    # by upstream and downstream amount of 1000bp
    tss    = promoters(
      x = subset(hum_inf, type=='gene'),
      upstream   = 1000, 
      downstream = 1000), 
    exon   = subset(hum_inf, type=='exon'),
    intron = subset(hum_inf, type=='gene')
)

# find overlaps between ChIP-seq data and ENSEMBL human data
overlap <- as.data.frame(GenomicAlignments::findOverlaps(ctcf_peaks, annotation_list))

# annotate with names
overlap$annotation <- names(annotation_list)[overlap$subjectHits]

# order the overlaps based on the hierarchy
overlap <- overlap[order(overlap$subjectHits), ]

# select only first found category per read
overlap <- subset(overlap, !duplicated(queryHits))

# find non-overlaps to annotate as intergenic
non_overlap <- which(!ctcf_peaks %over% annotation_list)
non_overlap <- data.frame (queryHits = non_overlap, 
                           subjectHits = rep("NaN",length(non_overlap)),
                           annotation = rep("intergenic",length(non_overlap)))

res <- rbind(overlap,non_overlap)

# group by annotation:
res <- dplyr::group_by(.data=res, annotation)
res <- summarise(.data = res, counts = length(annotation))

# calculate frequency    
res$frequency  <- with(res, round(counts/sum(counts),2))

ggplot(data=res,
        aes(x = 1,
            y = frequency,
            fill= annotation)) +
    geom_bar(stat='identity') +
    labs (x=NULL, y="Frequency") +
    ggtitle('Peak distribution in\ngenomic regions') +
    theme_minimal () +
     theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5))  # Remove x-axis text
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment
on chromome 21 on which I did peak calling. The results are stored in
a dataframe called `peaks`, which has the 
following columns:
"sequence name","starting position","ending position", "strand information", "component",        
"qvalue" and  "enrichment". It holds the sequence names, start of the sequence, 
end of the sequence, the strand information, component information,
q-values and enrichment values respectively. I want to convert this to a GRange object.
How do I do this?

completion:
To convert the information in the dataframe to a GRanges object,
We need to ensure that the correct naming for the columns are used. 
Important are the start and end columns, the sequence names,
and strand information. We will relabel these columns
with the appropriate names. We then use the `makeGRangesFromDataFrame`
function to convert our object to a GRanges object. 

```{r,eval=FALSE}
require (GenomicRanges)
colnames (peaks)[1]<-"seqnames"
colnames (peaks)[2]<-"start"
colnames (peaks)[3]<-"end"
colnames(peaks)[4] <- "strand"

gr_peaks <- makeGRangesFromDataFrame(peaks, keep.extra.columns = TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have ChIP-seq data from a CTCF transcription factor experiment
on chromome 21 of the human genome on which I did peak calling. The results are stored in
a GRanges object called `peaks`, which has an extra column called 'qvalues'
which contains the q-values of the peaks. I want to do motif discovery.
I want to select the top 250 peaks based on their q-value, and extract the DNA sequence
that belongs to these hits. How do I do this?

completion:
To perform motif discovery on the top 250 peaks based on their q-values from your ChIP-seq experiment,
you can first sort the GRanges object based on the column `qvalue`. Then we take the first 250 sequences
in the GRanges object, and use the `getSeq` function from the `BSgenome` library together with the
human sequence information stored in `BSgenome.Hsapiens.UCSC.hg38` to extract the sequences. 

```{r,eval=FALSE}
require(BSgenome.Hsapiens.UCSC.hg38)

# order GRanges object based on q-value:
peaks<-peaks[order(peaks$qvalue)]

# take first 250 sequences:
peaks<-peaks[1:250]

# Extract the sequences
seqs <- BSgenome::getSeq(BSgenome.Hsapiens.UCSC.hg38, peaks)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `my_motif` which holds information of a 
PWM matrix, and has nucleotides as rowname. I want to convert this to a
class PWMatrix. How do I do this?

completion:
To convert a dataframe with PWM (Position Weight Matrix) information to a class 
PWMatrix in R, you can use the TFBSTools package, which provides tools for working 
with transcription factor binding sites. Assuming you have a 
dataframe called my_motif with nucleotides as row names and the PWM values, 
You can convert this dataframe to a PWMatrix, by making use of the `TFBSTools::PWMatrix`
function. 

```{r,eval=FALSE}
require (TFBSTools)

# convert dataframe to matrix
my_motif <- as.matrix(my_motif)

# create PWM matrix
unknown_pwm <- TFBSTools::PWMatrix(
    ID = 'unknown', 
    profileMatrix = my_motif
)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a PWM matrix of type class PWMatrix in a variable called 
`my_motif`. I want to find some of the most similar known transcription factor motifs 
in human to my motif, and extract their names. How do I do this?

completion:
To see which transcription factor motifs are most similar to
your found motif, we can make use of the `JASPAR2018` library. 
Using the `TFBSTools::getMatrixSet` we extract the PWM matrices
for various transcription factor binding motifs. With this, we scan for
similar motifs using the `TFBSTools::PWMSimilarity` function. We
convert the results into a dataframe, and order based on similarity score.
We then assign the appropriate names to our results, and print out
the 5 motifs with the highest similarity score.

```{r,eval=FALSE}
require(JASPAR2018)
require(TFBSTools)

# extract motifs corresponding to human transcription factors
pwm_library <- TFBSTools::getMatrixSet(
  JASPAR2018,
  opts=list(
    collection = "CORE",
    species    = "Homo sapiens",
    matrixtype = "PWM"
  ))

# create dataframe out of library for later annotation
pwm_library_dt <- data.frame (ID = ID(pwm_library), name= name(pwm_library))

# scan for similar motifs
pwm_sim <- TFBSTools::PWMSimilarity(pwm_library,
  my_motif, method = 'Pearson')

# make into dataframe
pwm_sim <- as.data.frame(pwm_sim)

#order based on similarity score:
pwm_sim <- pwm_sim [order(pwm_sim$pwm_sim,decreasing=TRUE),,drop=FALSE]

# replace row names with motif name
rownames(pwm_sim)<-c(pwm_library_dt[match(rownames(pwm_sim), pwm_library_dt$ID),"name"])

# show first 5 hits with best similarity score
pwm_sim[1:5, , drop=FALSE]
```

--- TASK END

