---
title: "prompt/completion pairs chapter 10"
author: "authors"
date: "2023-11-1"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a variable called `met_data` which is a GRanges object that
holds information about methylation score in the extra metadata 
column `score`, from chromosome 21 and 22. 
I want to perform methylation segmentation analysis. 
I want to test partitioning from 1 to 4 segments, and want to 
investigate the methylation value per group and the length of the segment 
as a scatter plot. How do I do this?

completion:
To perform methylation segmentation analysis on a GRanges object 
and investigate the methylation value per group and length of 
segments in a scatterplot, you can make use of the `methylKit` library in R. 
You can partition the GRange object into different methylated segments using the 
`methylKit::methSeg` function. Since we want to test a partitioning from 1 to 4 
segments, we set `G=1:4`. To then investigate the methylation value and length of
segments using a scatterplot, we plot the methylation proportion against the 
log length of the segment. 

```{r,eval=FALSE}
require(methylKit)

#perform segmentation analysis
res <- methylKit::methSeg(met_data,minSeg=10,G=1:4,
                       join.neighbours = TRUE)

#create plot
plot(res$seg.mean,
     log10(width(res)),pch=20,
     col=scales::alpha(rainbow(4)[as.numeric(res$seg.group)], 0.2),
     ylab="log10(length)",
     xlab="methylation proportion")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a variable called `met_data` which is a GRanges object that
holds information about methylation score in the extra metadata 
column `score`, from chromosome 21 and 22. It also has another 
metadata column which is irrelevant for the analysis called
`extra_info`.I want to test partitioning 
from 1 to 4 segments, and want to investigate the methylation value per 
group and the length of the segment as a scatter plot.

completion:
To perform methylation segmentation analysis on a GRanges object 
and investigate the methylation value per group and length of 
segments in a scatterplot, you can make use of the `methylKit` library in R. 
To ensure correct analysis, you first need to remove the metadata column
`extra_info`. You can then partition the GRange object into different 
methylated segments using the `methylKit::methSeg` function. Since we want 
to test a partitioning from 1 to 4 segments, we set `G=1:4`. To then investigate 
the methylation value and length of segments using a scatterplot, we plot the 
methylation proportion against the log length of the segment. 

```{r,eval=FALSE}
require(methylKit)

# remove extra column
met_data$extra_info <- NULL

# perform segmentation analysis
res <- methylKit::methSeg(met_data,minSeg=10,G=1:4,
                       join.neighbours = TRUE)

# create plot
plot(res$seg.mean,
     log10(width(res)),pch=20,
     col=scales::alpha(rainbow(4)[as.numeric(res$seg.group)], 0.2),
     ylab="log10(length)",
     xlab="methylation proportion")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylRaw object called `met_data`. I want to plot the 
methylation status. How do I do this?

completion:
To plot the methylation status of a methylRaw object,
you can make use of the `methylKit` library. The function
`getMethylationStats` allows you to plot the methylation
status in a histogram. 

```{r,eval=FALSE}
require (methylKit)

#Creating methylation status plot
methylKit::getMethylationStats(met_data, plot=TRUE, both.strands=FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylRaw object called `met_data`. I want to plot a 
coverage plot. How do I do this?

completion:
To plot the methylation status of a methylRaw object,
you can make use of the `methylKit` library. The function
`getCoverageStats` allows you to plot the methylation
status in a histogram. 

```{r,eval=FALSE}
require(methylKit)

#Creating coverage plot
methylKit::getCoverageStats(met_data, plot=TRUE, both.strands=FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylRaw object called `met_data`. I want to filter my data, 
so that reads with very low coverage (below 10X),
and reads with abnormally high coverage (that have more than 99.9th percentile 
of coverage) are discarded. How do I do this?

completion:
To filter data in a methylRaw object so that reads
with low coverage (below 10x) and high coverage 
(over 99.9th percentile) are discarded, you can 
make use of the `methylKit` library in R. The function
`filterByCoverage` can be used to filter a methylRaw object
by amount of coverage. 

```{r,eval=FALSE}
require (methylKit)

#filtering the data
filtered_data <- methylKit::filterByCoverage(met_data, lo.count=10, lo.perc=NULL,
                                             hi.count=NULL, hi.perc=99.9)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylRaw object called `met_data`. I want to filter my data, 
so that reads with very low coverage (below 10% of coverage),
and reads with abnormally high coverage (that have more than 200x coverage) are 
discarded. How do I do this?

completion:
To filter data in a methylRaw object so that reads
with low coverage (below 10%) and high coverage 
(more then 200x) are discarded, you can 
make use of the `methylKit` library in R. The function
`filterByCoverage` can be used to filter a methylRaw object
by amount of coverage. 

```{r,eval=FALSE}
require (methylKit)

#filtering the data
filtered_data <- filterByCoverage(met_data, lo.count=NULL, lo.perc=10,
                                  hi.count=200, hi.perc=NULL)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a methylRawList object with 4 methylRaw objects called 
`data`. 2 objects are from control files, and 2 are from treatment 
files. I would like to unite these methylRaw objects to 
one dataframe which only holds bases which are covered in all
all files. How do I do this?

completion:
To unite a methylRawList object with 4 methylRaw objects
you can make use of the methylKit library in R. 
To unite methylRaw objects ensuring that only bases
which are covered in all files are kept, you can
make use of the 'unite' function. 

```{r,eval=FALSE}
require(methylKit)

# unite data
data_united <- methylKit::unite(data, destrand = FALSE, min.per.group = NULL)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a methylRawList object with 4 methylRaw objects called 
`data`. 2 objects are from control files, and 2 are from treatment 
files. I would like to unite these methylRaw objects to 
one dataframe which only holds bases which are in least 1 
file for each replicate.
How do I do this?

completion:
To unite a methylRawList object with 4 methylRaw objects
you can make use of the methylKit library in R. 
To unite methylRaw objects ensuring that only bases
which are covered in all files are kept, you can
make use of the 'unite' function. By setting the
argument `min.per.group = 1L` we ensure that bases 
are kept which are at least in 1 file for 
each replicate. 

```{r,eval=FALSE}
require (methylKit)

# unite data
data_united <- methylKit::unite(data, destrand = FALSE, min.per.group = 1L)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | Ml/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples. 
I want to get the percent methylation matrix 
for this. How do I do this?

completion:
To extract the percent methylation matrix
from a methylBase object, you can make use
of the methylKit library in R. The function 
`percMethylation` allows you to extract
the percent methylation matrix. 

```{r,eval=FALSE}
require (methylKit)

# get percent methylation data
pm <- methylKit::percMethylation(met_data)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples. 
I want to get the standard deviation of CpGs 
for this. I then want to plot a histogram of
this. How do I do this?

completion:
To extract the standard deviation of CpGs
from a methylBase object, you can make use
of the methylKit library in R. The function 
`percMethylation` allows you to extract
the percent methylation matrix. You can then
calculate the standard deviation of CpGs using
the `rowSds` function. To plot
a histogram of the percent methylation 
matrix you can make use of the `hist`
function in R. 

```{r,eval=FALSE}
require(methylKit)
require(metrixStats)

# get percent methylation data
pm <- methylKit::percMethylation(met_data)

# calculate standard deviation of CpGs
mds <- matrixStats::rowSds(pm)

# plot histogram
hist(mds,col="#d98adb",xlab="Std. dev. (%) per CpG")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples. I 
want to convert this object to a dataframe. How
do I do this?

completion:
To convert a methylBase object to a dataframe 
you can make use of the methylKit library in R. 
The function `getData` allows you to convert
the methylBase object to a dataframe. 

```{r,eval=FALSE}
require(methylKit)

# convert to dataframe
df_met_data <- methylKit::getData(met_data)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples. I 
want to filter out positions that lie on chr21 which 
start and end at position 9853296 and 10000000 respectively. 
How do I do this?

completion:
To filter out positions that lie on chr21 which 
start and end at position 9853296 and 10000000 from a 
methylBase object, you can make use of the GenomicRanges
library in R. To filterout the positions that lie on chr21 
which start and end at position  9853296 and 10000000 you can 
make use of a GRanges object, and find the indices which
do not overlap with this object.  

```{r,eval=FALSE}
require (GenomicRanges)

# positions to be deleted as GRanges object
del <- GenomicRanges::GRanges(seqnames=c("chr21"),
                              ranges=IRanges(start=c(9853296),
                                             end=c(10000000)))


# select data that does not overlap with GRanges object
data_filtered <- met_data[!as(met_data,"GRanges") %over% del, ]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples.
I want to see if the results from the test files and the 
control files cluster together, as a way of looking at how 
similar they are. How do I do this?

completion:
To cluster a methylBase object you can make use of the
methylKit library in R. The function `methylkit::ClusterSamples` 
will cluster the samples and draw a dendrogram. It will use 
correlation distance, which is 1-p where p is the correlation 
coefficient between two pairs of samples. The cluster tree 
will be drawn using the “ward” method. This specific variant 
uses a “bottom up” approach: each data point starts in its own 
cluster, and pairs of clusters are merged as one moves up the 
hierarchy. In Ward's method, two clusters are merged if the variance 
is minimized compared to other possible merge operations.

```{r,eval=FALSE}
require(methylKit)

#Cluster samples and show dendogram
methylKit::clusterSamples(met_data, dist="correlation", method="ward", plot=TRUE)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples.
I want to perform PCA and visualize the results 
in a scatterplot. How do I do this?

completion:
To perform PCA on a methylBase object,
you can make use of the methylKit library in R. 
The function `PCASamples`. Alternatively, you could 
extract the percent methylation matrix and perform PCA 
on this transposed matrix using the `prcomp` function. 
Using the `methylKit::PCASamples` function 
however filters low variation regions before PCA analysis 
automatically, which might be beneficial when performing 
further methylation analysis. Below is an example on how to 
use both methods. 

```{r,eval=FALSE}
require (methylKit)

# perform PCA using methylKit function
# automatically will plot scatterplot
pca_mk <- methylKit::PCASamples(met_data,obj.return = TRUE, adj.lim=c(1,1))

# perform PCA using prcomp
# get percent methylation data
pm <- methylKit::percMethylation(met_data)

# perform PCA. Here, rows with NA are removed from the percent metylation matrix. 
pca_pr <- prcomp (na.omit(pm))
mat <- as.matrix(pca_pr$rotation)

# plot scatterplot using first 2 dimensions
# and annotate
plot (mat[, 1:2])
text(x = mat[, 1], y = (mat[,2]+0.01), labels = rownames(mat))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples.
There are 2 test samples and 2 control samples
in this object.
I want to see if there are any differentially methylated 
regions when comparing the results from the test 
samples to the control samples. For this purpose,
I want to perform a fishers exact test. I want the control 
and test samples to have at least a methylation 
difference of 25% before they are called significantly different, 
and I also want the cutoff for the q-value to be 0.01.

completion:
To perform a fishers exact test on a methylBase object
to find differentially methylated regions, you can make
use of the methylKit library in R.
In simple comparisons between pairs of samples (i.e. test and control), 
methods such as Fisher's exact test can be used. 
Since you mentioned there are 2 test samples and 2 control samples
in your methylBase object, replicates should be pooled within groups 
to a single sample per group. This entails adding up the number of 
Cs and Ts per group. You can do this using the 
`methylKit::pool` function. This way even if you have replicated experiments 
you treat them as single experiments, and can apply Fisher's exact test. 
You can then apply the `methylKit::calculateDiffMeth()` function, which 
will return the p-value and q-values for all bases or regions in the input 
methylBase object. Filtering to get differentially methylated CpGs. 
can be done via the `methylKit::getMethlyDiff()` function

```{r,eval=FALSE}
require(methylKit)

#pool data for differential methylation analysis
pooled_data <- pool(met_data,sample.ids=c("test","control"))

#perform fishers exact test for each base:
res <- calculateDiffMeth(pooled_data, test= "fast.fisher")

#get differentially methylated positions:
all_diff <- getMethylDiff(res,difference=25,qvalue=0.01,type="all")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples.
There are 2 test samples and 2 control samples
in this object.
I want to see if there are any differentially methylated 
regions when comparing the results from the test 
samples to the control samples. For this purpose,
I want to use the logistic regression based Chi-square test 
with correction for overdispertion. How do I do this?

completion:
To perform the regression based Chi-square 
test with correction for overdispersion, we can make use of the 
`methylKit::calculateDiffMeth` function. This will report back, 
the p-value, q-value and methylation difference for further analysis.

```{r,eval=FALSE}
require(methylKit)

#perform Chi-square test with overdispersion correction
res <- calculateDiffMeth(met_data,overdispersion = "MN",test = "Chisq")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylDiff object called `res`. This object 
holds the results from using the function
`methylKit::calculateDiffMeth` using
logistic regression based Chi-square test 
with correction for overdispertion.
How do I filter out the positions 
which are differentially methylated?
I consider somehting differentially 
methylated if the q-value is below 0.05 and
the difference in percentage is at least 25%.

completion:
To filter out the positions which are differentially 
methylated from a methylDiff object you can make 
use of the methylKit library in R. The function
`getMethylDiff` allows you to filter out positions
where he q-value is below 0.05 and
the difference in percentage is at least 25% 
by setting the `qvalue` and `difference` argument. 

```{r,eval=FALSE}
require(methylKit)

# get differentially methylated positions:
all_diff <- getMethylDiff(res,difference=25,qvalue=0.05,type="all")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples.
There are 2 test samples and 2 control samples
in this object.
I want to see if there are any differentially methylated 
regions when comparing the results from the test 
samples to the control samples. For this purpose,
I want to use a Betabinomial-distribution-based test 
namely DSS.  How do I do this?

completion:
To perform DSS on a methylBase object, you can make
use of the methylKit library in R. The function
`calculateDiffMethDSS` can be used for this purpose. 

```{r,eval=FALSE}
require (methylKit)

#perform Betabinomial-distribution-based test DSS.
res <- calculateDiffMethDSS(met_data)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylBase object called `met_data`. This holds
methylation information from different samples.
There are 2 test samples and 2 control samples
in this object.
The age of the first sample in the object is 30. The others have an age of 80,
34 and 75 respectively. I want to include this covariate in my test to see if 
there are any differentially methylated regions when comparing the results 
from the test samples to the control samples. For this, I want to use the 
logistic regression based Chi-square test with correction for overdispertion. 
How do I do this?

completion:
To include the covariate age, we make a dataframe out of the age 
information. We then pass this to the `methylKit::calculateDiffMeth` function 
using the `covariates = covariates` argument. 

```{r,eval=FALSE}
require (methylKit)

#define covariate age for all objects
covariates <- data.frame(age=c(30,80,34,75))

#perform Chi-square test with correction for overdispersion
res <- calculateDiffMeth(met_data,
                         covariates=covariates,
                         overdispersion="MN",
                         test="Chisq")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I want to retrieve ENSEMBL information on the genomic annotation
such as intron, exon, or gene in the human genome 
using a GRCh38.92 assembly. I want to retrieve a GRanges object so that
I can use this for annotation. How do I do this?

completion:
To retrieve ENSEMBL information on genomic annotation, we can make
use of the `AnnotationHub` library. Here we look for the following keywords:
- ENSEMBL to retrieve the ENSEMBL information
- "Homo" to retrieve the information of homo sapiens
- "GRCh38.92" to retrieve the correct version
- and "chr.gtf" to grab the correct file 
From this we retrieve the correct entry for the annotationhub,
which we can use to extract the human annotation information. 

```{r,eval=FALSE}
require(AnnotationHub)

# connect to annotation hub
ah <- AnnotationHub()

# get homo sapiens info. 
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylDiff object called `res`. This object 
holds the results from using the function
`methylKit::calculateDiffMeth` using
logistic regression based Chi-square test 
with correction for overdispertion.
How do I filter out the positions 
which are differentially methylated?
I consider somehting differentially 
methylated if the q-value is below 0.05 and
the difference in percentage is at least 25%.
I then want to annotate the differentially methylated
positions to see if they fall within intronic or 
exonic regions etc using info from ENSEMBL using 
the GRCh38.92 version. How do I do this?

completion:
To filter out the positions which are differentially 
methylated from a methylDiff object you can make 
use of the methylKit library in R. The function
`getMethylDiff` allows you to filter out positions
where he q-value is below 0.05 and
the difference in percentage is at least 25% 
by setting the `qvalue` and `difference` argument. 
To annotate the differentially methylated positions, you can 
make use of the`AnnotationHub` library. 
Connecting to the annotation hub allows you to retrieve genomic 
information from the GRCh38.92 genome annotation from ENSEMBL.This 
is used to find overlaps between the ENSEMLS information 
and differentially methylated positions `GenomicAlignments::findOverlaps` 
function. Annotation is done using the `type` column from the human 
information retrieved from ENSEMBL. 

```{r,eval=FALSE}
require(methylKit)
require(AnnotationHub)
require(GenomicAlignments)

# get differentially methylated positions:
all_diff <- getMethylDiff(res,difference=25,qvalue=0.05,type="all")

# convert to GRanges object
diff_gr <- as(all_diff,"GRanges")

# get human genome information with intron exon annotation
# connect to annotation hub
ah <- AnnotationHub()

# get homo sapiens info. 
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]

# fix chromosome names so they fit UCSC naming and
# keep only chromosome 21
seqlevels(hum_inf, pruning.mode="coarse") <- paste0("chr", seqlevels(hum_inf))
seqlevels(hum_inf,pruning.mode="coarse") <- "chr21"
hum_inf <- hum_inf[seqnames(hum_inf) == "chr21"]

# find overlaps
overlap <- as.data.frame(GenomicAlignments::findOverlaps(diff_gr,hum_inf))

# annotate using overlapping info
overlap$annotation <- hum_inf$type[overlap$subjectHits]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have methylation information of chromosome 21
in a methylDiff object called `res`. This object 
holds the results from using the function
`methylKit::calculateDiffMeth` using
logistic regression based Chi-square test 
with correction for overdispertion.
How do I filter out the positions 
which are differentially methylated?
I consider somehting differentially 
methylated if the q-value is below 0.05 and
the difference in percentage is at least 25%.
I then want to annotate the differentially methylated
positions seeing in which genes these regions fall. 
How do I do this?

completion:
To filter out the positions which are differentially 
methylated from a methylDiff object you can make 
use of the methylKit library in R. The function
`getMethylDiff` allows you to filter out positions
where he q-value is below 0.05 and
the difference in percentage is at least 25% 
by setting the `qvalue` and `difference` argument. 
To annotate in which genes the differentially methylated positions 
are, you can make use of the`AnnotationHub` library. 
Connecting to the annotation hub allows you to retrieve genomic 
information from the GRCh38.92 genome annotation from ENSEMBL.This 
is used to find overlaps between the ENSEMLS information 
and differentially methylated positions `GenomicAlignments::findOverlaps` 
function. Annotation is done using the `gene_name` column from the human 
information retrieved from ENSEMBL.

```{r,rval=FALSE}
require(methylKit)
require(AnnotationHub)
require(GenomicAlignments)

# get differentially methylated positions:
all_diff <- getMethylDiff(res,difference=25,qvalue=0.05,type="all")

# convert to GRanges object
diff_gr <- as(all_diff,"GRanges")

# get human genome information with intron exon annotation
# connect to annotation hub
ah <- AnnotationHub()

# get homo sapiens info. 
hs <- query(ah, pattern=c("ENSEMBL", "Homo", "GRCh38.92", "chr.gtf"))
hum_inf <- ah[[rownames(mcols(hs)[1])]]

# fix chromosome names so they fit UCSC naming and
# keep only chromosome 21
seqlevels(hum_inf, pruning.mode="coarse") <- paste0("chr", seqlevels(hum_inf))
seqlevels(hum_inf) <- "chr21"
hum_inf <- hum_inf[seqnames(hum_inf) == "chr21"]

# find overlaps
overlap <- as.data.frame(GenomicAlignments::findOverlaps(diff_gr,hum_inf))

overlap$gene_name <- hum_inf$gene_name[overlap$subjectHits]
```

--- TASK END
