---
title: "prompt/completion pairs chapter 10"
author: "authors"
date: "2023-10-19"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called 'test1.myCpG.txt' which holds information of 
methylation status of different spots in the genome. Column 2 holds the 
information of which chromosome it is, column 3 holds information of where 
the read begins and ends, column 4 holds the information of which strand it 
is from (indicated with F and R), column 5 holds the coverage information and 
column 6 holds the frequency of C's from 0 to 100. I want to see if the methylation 
status has a bimodal distribution as a way of checking the quality. 
Furthermore, I would like to have a plot of the coverage values to see if there is 
any PCR bias problem. How do I do this?

completion:
To check if the methylation status holds a bimodal distribution, we first 
read in the data using the `methylKit::methRead` function. Here we specify 
which columns hold what type of information using the `pipeline` argument. 
Since the frequency column holds values from 0 to 100, we set the option 'fraction' 
to false. To then check the quality of the data and see if it has a bimodal distribution, 
we make use of the `methylKit::getMethylationStats` function. To create a coverage plot, 
we make use of the `methylKit::getCoverageStats` function. 

```{r,eval=TRUE}
require(methylKit)

# reading in data
met_data_test <- methylKit::methRead(location = "test1.myCpG.txt",
                                     sample.id = "test1", assembly="hg18",
                                     pipeline = list(fraction = FALSE, start.col = 3, 
                                                     end.col = 3, chr.col = 2,
                                                     coverage.col = 5, strand.col = 4,
                                                     freqC.col = 6))

#Creating methylation status plot
methylKit::getMethylationStats(met_data_test, plot=TRUE, both.strands=FALSE)

#Creating coverage plot
methylKit::getCoverageStats(met_data_test, plot=TRUE, both.strands=FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a file called 'test1.myCpG.txt' which holds information
of methylation status of different spots in the genome. Column 2 holds
the information of which chromosome it is, column 3 holds information
of where the read begins and also ends, column 4 holds the information
of which strand it is from (indicated with F and R), column 5 holds the
coverage information and column 6 holds the frequency of C's from 0 to 100.
I want to filter my data, so that reads with very low coverage (below 10X),
and reads with abnormally high coverage (that have more than 99.9th percentile 
of coverage) are discarded. How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
Here we specify which columns hold what type of information using the 
`pipeline` argument. Since the frequency column holds values from 0 to 100, 
we set the option 'fraction' to false. To then filter the data and discard 
bases with low coverage and abnormally high coverage, we make use of the 
`methylKit::filterByCoverage` function. Since we want to filter out bases 
with below 10 times coverage, we set the argument `lo.count = 10`. Alternatively, 
if we would want to filter out bases which are below the 10th percentile, we could 
make use of the `lo.percentage`argument. To filter out bases with abnormally high 
coverage (above the 99.9th percentile) we `set hi.perc = 99.9`. Alternatively, 
if we would want to filter on absolute count, we could make use of the `hi.count` argument. 

```{r,eval=FALSE}
# reading in data
require(methylKit)
met_data_test <- methylKit::methRead(location = "test1.myCpG.txt",
                                     sample.id = "test1", assembly="hg18",
                                     pipeline = list(fraction = FALSE, start.col = 3, 
                                                     end.col = 3, chr.col = 2,
                                                     coverage.col = 5, strand.col = 4,
                                                     freqC.col = 6))

#filtering the data
filtered_test <- filterByCoverage(met_data_test,lo.count=10,lo.perc=NULL,
                               hi.count=NULL,hi.perc=99.9)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome.
Column 2 holds the information of which chromosome it is, 
column 3 holds information of where the read begins and ends, 
column 4 holds theinformation of which strand it is from 
(indicated with F and R), column 5 holds the coverage information 
and column 6 holds the frequency of C's from 0 to 100.
The first two files are from test samples, whereas the other 
files are from control samples. I want a way to unite these files 
to one data structure where only bases are present which where covered 
in either both sample groups or both control groups. 
I then want to see how much of a difference there is between the 
methylation percentages across all the different files, to see how much the 
test1, test2, control1 and control2 files agree. I want to do this by plotting 
a histogram with the standard deviation of the percentages across the files. 
How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files we 
want to use. To ensure we mark files as test or control, we make use of 
the `treatment` argument. We also specify which columns hold what type of 
information using the `pipeline` argument. Since the frequency column holds 
values from 0 to 100, we set the option 'fraction' to false. We unite our files 
to one big dataframe where only bases are present which are covered in either 
the two sample groups or both control groups using the `methylkit::unite` function 
using the argument `min.per.group = 2 L` to ensure that only bases are kept which are in 
2/2 files of the sample files and 2/2 files of the control files. To then visualize the 
difference between methylation percentages across the different files, we create a percent 
methylation matrix using the `methylKit::percMethylation` function. The standard deviation 
of the percentage is calculated using `matrixStats::rowSds` and this is used to plot the 
histogram. 

```{r,eval=TRUE}
require(methylKit)
require(matrixStats)

#reading in data:
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt", "control1.myCpG.txt", "control2.myCpG.txt" )

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"), assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

#unite to one big dataframe where there are only bases which are covered in all files:
data_united <- methylKit::unite(data, destrand = FALSE, min.per.group = 2L) #argument 2L ensures only bases are added which are found in at least 2 files per treatment

pm <- methylKit::percMethylation(data_united) # get percent methylation matrix
mds <- matrixStats::rowSds(pm) # calculate standard deviation of CpGs

#plot histogram
hist(mds,col="#d98adb",xlab="Std. dev. (%) per CpG")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, column 3
 holds information of where the read begins and ends, column 4 holds the
information of which strand it is from (indicated with F and R), column 
5 holds the coverage information and column 6 holds the frequency of C's 
from 0 to 100.The first two files are from test samples, whereas the 
other files are from control samples.I want to filter my data, so that 
reads with very low coverage (below 10X), and reads with abnormally high 
coverage (that have more than 99.9th percentile of coverage) are discarded. 
I want then want a way to unite these files to one data structure where only 
bases are present which where convered in either both sample groups or both 
control groups. How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files we 
want to use. To ensure we mark files as test or control, we make use of 
the `treatment` argument. We also specify which columns hold what type 
of information using the `pipeline` argument. Since the frequency column 
holds values from 0 to 100, we set the option 'fraction' to false. To then 
filter the data and discard bases with low coverage and abnormally high 
coverage, we make use of the `methylKit::filterByCoverage` 
function. Since we want to filter out bases with below 10 times coverage, 
we set the argument `lo.count = 10`. Alternatively, if we would want to 
filter out bases which are below the 10th percentile, we could make use of 
the `lo.percentage` argument. To filter out bases with abnormally high 
coverage (above the 99.9th percentile) we `set hi.perc = 99.9`. Alternatively, 
if we would want to filter on absolute count, we could make use of the `hi.count` 
argument. We unite our files to one big dataframe where only bases are present 
which are covered in either the two sample groups or both control groups 
using the `methylkit::unite` function using the argument `min.per.group = 2 L` 
to ensure that only bases are kept which are in 2/2 files of the sample files and
2/2 files of the control files. 

```{r,eval=FALSE}
require(methylKit)

# reading in data:
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3,
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

# filtering the data
filtered_data <- methylKit::filterByCoverage(data, lo.count = 10, lo.perc = NULL,
                                  hi.count = NULL,
                                  hi.perc = 99.9)

# unite files to one datastructure
#argument 2L ensures only bases are added which are found in at least 2 files per treatment
data_united <- methylKit::unite(data, destrand = FALSE, min.per.group = 2L) 
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, 
column 3 holds information of where the read begins and ends, 
column 4 holds theinformation of which strand it is from 
(indicated with F and R), column 5 holds the coverage information 
and column 6 holds the frequency of C's from 0 to 100.The first two 
files are from test samples, whereas the other files are from control 
samples. I want to unite these files to one dataframe and then
filter out positions which I know hold a C -> T mutation which does 
not come from the bisulfite-treatment. The positions I want to filter 
out lie on chromosome 21 (chr21) between start and end at positions 9853296 
and 10000000 respectively. How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make use of 
the `location`argument where we will pass a list with files we want to use. 
To ensure we mark files as test or control, we make use of the `treatment` argument. 
We also specify which columns hold what type of information using the 
`pipeline` argument. Since the frequency column holds values from 
0 to 100, we set the option 'fraction' to false. We unite our files 
to one big dataframe where only bases are present which are covered 
in either the two sample groups or both control groups using the 
`methylkit::unite` function using the argument `min.per.group = 2 L` 
to ensure that only bases are kept which are in 2/2 files of the 
sample files and 2/2 files of the control files. To then filter 
out the positions that lie on chr21 which start and end at position 
9853296 and 10000000, we make use of a Granges object. We use this for 
filtering out our positions of interest. 

```{r,eval=FALSE}
require(GenomicRanges)
require(methylKit)

#reading in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

#unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE) 

# Position to be deleted
mut=GenomicRanges::GRanges(seqnames=c("chr21"),
                           ranges=IRanges(start=c(9853296),
                                          end=c(10000000)))

# select CpGs that do not overlap with mutations
data_filtered <- data_united[! as(data_united,"GRanges") %over% mut, ]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, column 
3 holds information of where the read begins and ends, column 4 
holds theinformation of which strand it is from (indicated with F and R), 
column 5 holds the coverage information and column 6 holds the frequency 
of C's from 0 to 100. The first two files are from test samples, whereas 
the other files are from control samples. I want to see if the results 
from the test files and the control files cluster together, as a way of 
looking at how similar they are. How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files 
we want to use. To ensure we mark files as test or control, we make 
use of the `treatment` argument. We also specify which columns hold 
what type of information using the `pipeline` argument. Since the 
frequency column holds values from 0 to 100, we set the option 
'fraction' to false. We unite our files to one big dataframe where 
only bases are present which are covered in either the two sample 
groups or both control groups using the `methylkit::unite` function 
using the argument `min.per.group = 2 L`. To cluster the information 
given in the different files, we make use of the `methylkit::ClusterSamples` 
function. This function will cluster the samples and draw a dendrogram. 
It will use correlation distance, which is 1-p where p is the correlation 
coefficient between two pairs of samples. The cluster tree will be drawn 
using the “ward” method. This specific variant uses a “bottom up” approach: 
each data point starts in its own cluster, and pairs of clusters are merged 
as one moves up the hierarchy. In Ward's method, two clusters are merged if 
the variance is minimized compared to other possible merge operations.

```{r,eval=FALSE}
require(methylKit)

#reading in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

#unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE)

#Cluster samples and show dendogram
methylKit::clusterSamples(data_united, dist="correlation", method="ward", plot=TRUE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the 
genome. Column 2 holds the information of which chromosome 
it is, column 3 holds information of where the read begins and ends, 
column 4 holds the information of which strand it is from 
(indicated with F and R), column 5 holds the coverage information 
and column 6 holds the frequency of C's from 0 to 100.
The first two files are from test samples, whereas the other 
files are from control samples. I want to perform PCA on the 
variables that are regions/bases, and visualize the results 
in a scatterplot to see if sample and control groups are similar. 
How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files 
we want to use. To ensure we mark files as test or control, we make use 
of the `treatment` argument. We also specify which columns hold what 
type of information using the `pipeline` argument. Since the frequency 
column holds values from 0 to 100, we set the option 'fraction' to false. 
We unite our files to one big dataframe where only bases are present which 
are covered in either the two sample groups or both control groups 
using the `methylkit::unite` function using the argument `min.per.group = 2 L`. 
To perform PCA on the variables that are regions/bases we can make use of the
`methylKit::PCASamples` function. Alternatively, we could extract the 
percent methylation matrix and perform PCA on this transposed matrix 
using the `prcomp` function. Using the `methylKit::PCASamples` function 
however filters low variation regions before PCA analysis automatically, 
which might be beneficial when performing further methylation analysis. 

```{r,eval=FALSE}
require(methylKit)

# reading in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

#unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE)

#perform PCA
pc <- methylKit::PCASamples(data_united,obj.return = TRUE, adj.lim=c(1,1))
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, column 3 
holds information of where the read begins and ends, column 4 holds 
the information of which strand it is from (indicated with F and R), 
column 5 holds the coverage information and column 6 holds the frequency 
of C's from 0 to 100. The first two files are from test samples, whereas 
the other files are from control samples. I want to see if there are any 
differentially methylated regions when comparing the results from the test 
samples to the control samples. For this I want to perform a fishers exact 
test. I want the control and test samples to have at least a methylation 
difference of 25% before they are called significantly different, and I also 
want the cutoff for the q-value to be 0.01. I then also want to investigate 
which regions are hypomethylated and which regions are hypermethylated when 
compared to the control. How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files 
we want to use. To ensure we mark files as test or control, we make 
use of the `treatment` argument. We also specify which columns hold 
what type of information using the `pipeline` argument. Since the 
frequency column holds values from 0 to 100, we set the option 
'fraction' to false. We unite our files to one big dataframe where 
only bases are present which are covered in either the two sample groups 
or both control groups using the `methylkit::unite` function using 
the argument `min.per.group = 2 L`. Differential DNA methylation is 
usually calculated by comparing the proportion of methylated Cs in a 
test sample relative to a control. In simple comparisons 
between such pairs of samples (i.e. test and control), methods such 
as Fisher's exact test can be used. If there are replicates, replicates 
can be pooled within groups to a single sample per group. This entails 
adding up the number of Cs and Ts per group. We can do this using the 
`methylKit::pool` function. This way even if we have replicated experiments 
we treat them as single experiments, and can apply Fisher's exact test. 
We will then apply the `methylKit::calculateDiffMeth()` function, which 
will return the p-value and q-values for all bases or regions in the input 
methylBase object. We need to filter to get differentially methylated CpGs. 
This can be done via the `methylKit::getMethlyDiff()` function

```{r,eval=FALSE}
require(methylKit)

#reading in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

#unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE)

#pool data for differential methylation analysis
pooled_data <- methylKit::pool(data_united, sample.ids = c("test", "control"))

#perform fishers exact test for each base:
dm.pooledf <- methylKit::calculateDiffMeth(pooled_data)

#get differentially methylated positions:
all.diff <- methylKit::getMethylDiff(dm.pooledf,difference=25,qvalue=0.01,type="all")

#get hypomethylated regions:
hypo.diff <- methylKit::getMethylDiff(dm.pooledf,difference=25,qvalue=0.01,type="hypo")

#get hypermethylated regions:
hypo.diff <- methylKit::getMethylDiff(dm.pooledf,difference=25,qvalue=0.01,type="hyper")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, column 3 
holds information of where the read begins and ends, column 4 holds 
theinformation of which strand it is from (indicated with F and R), 
column 5 holds the coverage information and column 6 holds the frequency 
of C's from 0 to 100. The first two files are from test samples, whereas 
the other files are from control samples. I want to see if there are any 
differentially methylated regions when comparing the results from the 
test samples to the control samples. For this, I want to use the logistic 
regression based Chi-square test with correction for 
overdispertion. How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files 
we want to use. To ensure we mark files as test or control, we make use 
of the `treatment` argument. We also specify which columns hold what 
type of information using the `pipeline` argument. Since the frequency 
column holds values from 0 to 100, we set the option 
'fraction' to false. We unite our files to one big dataframe where 
only bases are present which are covered in either the two sample groups 
or both control groups using the `methylkit::unite` function using the 
argument `min.per.group = 2 L`. To perform the regression based Chi-square 
test with correction for overdispersion, we can make use of the 
`methylKit::calculateDiffMeth` function. This will report back, 
the p-value, q-value and methylation difference for further analysis.

```{r,eval=FALSE}
require (methylKit)

# load in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

# unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE)

# perform Chi-square test with overdispersion correction
dm_lr <- methylKit::calculateDiffMeth(data_united,overdispersion = "MN",test ="Chisq")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, column 3 
holds information of where the read begins and ends, column 4 holds 
the information of which strand it is from (indicated with F and R), 
column 5 holds the coverage information and column 6 holds the frequency 
of C's from 0 to 100. The first two files are from test samples, whereas 
the other files are from control samples. I want to see if there are any 
differentially methylated regions when comparing the results from the 
test samples to the control samples. For this, I want to use a 
Betabinomial-distribution-based test namely DSS.  How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files we want 
to use. To ensure we mark files as test or control, we make use of the 
`treatment` argument. We also specify which columns hold what type of 
information using the `pipeline` argument. Since the frequency column 
holds values from 0 to 100, we set the option 'fraction' to false. 
We unite our files to one big dataframe where only bases are present 
which are covered in either the two sample groups or both control groups 
using the `methylkit::unite` function using the argument `min.per.group = 2 L`. 
To perform DDS, we can make use of the `methylKit::calculateDiffMethDSS` 
function. This will report back, the p-value, q-value and methylation 
difference for further analysis.

```{r,eval=FALSE}
require(methylKit)

#load in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

#unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE)

#perform Betabinomial-distribution-based test DSS.
dm_dss <- methylKit::calculateDiffMethDSS(data_united)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 5, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, column 3 
holds information of where the read begins and ends, column 4 holds the
information of which strand it is from (indicated with F and R), column 5 
holds the coverage information and column 6 holds the frequency of C's 
from 0 to 100. The first two files are from test samples, whereas the other 
files are from control samples. The rows represent basepairs. 
I however want to work not with individual bases, but with genomic regions 
of 1000 bp long. I also do not want them to overlap at all. 
I then would like to extract certain regions that I am interested in 
which are held in a file called 'refseq.hg18.bed.txt'. 
This is a BED file which holds the info on which regions are exons, 
introns, promoters or intergenic regions. I want to extract the regions 
which correspond to promotor regions.
How do I process my data so that I can do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make use 
of the `location` argument where we will pass a list with files we want 
to use. To ensure we mark files as test or control, we make use of the 
`treatment` argument. We also specify which columns hold what type of 
information using the `pipeline` argument. Since the frequency column 
holds values from 0 to 100, we set the option 'fraction' to false. To then 
convert the base pair information to genomic regions, we make use of the 
`methylKit::tileMethylCounts` function. This function adds up C and T counts 
from each covered cytosine and returns a total C and T count for each region. 
To read in the BED data file we make use of the `genomation::readTranscriptFeatures` 
function. We then use the output of this function to filter our data for 
the regions of interest. 

```{r,eval=FALSE}
require (methylKit)
require(genomation)

#load in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

bed_data <- genomation::readTranscriptFeatures("refseq.hg18.bed.txt")

#divide into genomic regions of 1000 basepairs
#setting stepsize to 1000 ensures we have no overlapping regions
gen_reg <- methylKit::tileMethylCounts(data,win.size=1000,step.size=1000) 

#extract regions of interest using BED data
reg_int <- regionCounts(gen_reg, bed_data$promoters)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType:data wrangling |  ML/Stats , numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the genome. 
Column 2 holds the information of which chromosome it is, column 3 
holds information of where the read begins and ends, column 4 holds the
information of which strand it is from (indicated with F and R), 
column 5 holds the coverage information and column 6 holds the 
frequency of C's from 0 to 100.The first two files are from test samples, 
whereas the other files are from control samples. The rows represent basepairs. 
The age of the sample in the first file is 30. The others have an age of 80,
34 and 75 respectively. I want to include this covariate in my test to see if 
there are any differentially methylated regions when comparing the results 
from the test samples to the control samples. For this, I want to use the 
logistic regression based Chi-square test with correction for overdispertion. 
How do I do this?

completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files w
e want to use. To ensure we mark files as test or control, we make use 
of the `treatment` argument. We also specify which columns hold what 
type of information using the `pipeline` argument. Since the frequency 
column holds values from 0 to 100, we set the option 'fraction' to false. 
We unite our files to one big dataframe where only bases are present which 
are covered in either the two sample groups or both control groups 
using the `methylkit::unite` function using the argument `min.per.group = 2 L`. 
To perform the regression based Chi-square test with correction for 
overdispersion, we can make use of the `methylKit::calculateDiffMeth` function. 
This will report back, the p-value, q-value and methylation difference for further 
analysis. To include the covariate age, we make a dataframe out of the age 
information. We then pass this to the `methylKit::calculateDiffMeth` function 
using the `covariates = covariates` argument. 

```{r,eval=FALSE}
require(methylKit)

#load in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

#unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE)

#define covariate age for all files
covariates <- data.frame(age=c(30, 80, 34, 75))

#perform Chi-square test with correction for overdispersion
my_diffmet <- methylKit::calculateDiffMeth(data_united,
                                           covariates = covariates,
                                           overdispersion = "MN",
                                           test = "Chisq")
```

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats , numData: 6, seeData: no, openEnded: no, tags: merge

prompt:
I have 4 files called 'test1.myCpG.txt', 'test2.myCpG.txt', 
'control1.myCpG.txt' and 'control2.myCpG.txt' which holds 
information of methylation status of different spots in the
human genome. Column 2 holds the information of which chromosome 
it is, column 3 holds information of where the read begins and ends, 
column 4 holds theinformation of which strand it is from (indicated with F and R), 
column 5 holds the coverage information and column 6 holds the frequency of 
C's from 0 to 100.The first two files are from test samples, whereas the 
other files are from control samples. The rows represent basepairs. 
I would like to test for differentially methylated regions 
between my test and control groups, using fishers exact test. 
I want the control and test samples to have at least a methylation difference 
of 25% before they are called significantly different, and I also want the 
cutoff for the q-value to be 0.01. 
I then would like to annotate these regions that I am interested in. 
For this I have the file 'refseq.hg18.bed.txt'. This is a BED file,  which
holds the info on which regions are exons, introns, promoters or intergenic 
regions. I want to know if the regions that are differentially methylated 
are often part of introns, exons promotor or intergenic regions.

I would also like to investigate the annotation when considering CpG island 
annotation. For this I have a file called 'cpgi.hg18.bed.txt'. This is a BED file.
It holds the information which regions of the genome are  CpGi and which are shores. 

Finally I would like to use the information of the the file 
'refseq.hg18.bed.txt' to not only know if regions are part of 
introns, exons or intergenic regions, but I also want to know 
if exons or introns are differentially methylated, which genes 
are effected by this. How do I do this?


completion:
We first read in the data using the `methylKit::methRead` function. 
To ensure the data from different files is processed as one, we make 
use of the `location` argument where we will pass a list with files 
we want to use. To ensure we mark files as test or control, we make 
use of the `treatment` argument. We also specify which columns hold 
what type of information using the `pipeline` argument. Since the 
frequency column holds values from 0 to 100, we set the option 
'fraction' to false. We unite our files to one big dataframe where 
only bases are present which are covered in either the two sample groups 
or both control groups using the `methylkit::unite` function using 
the argument `min.per.group = 2 L`.Saples are then pooled using 
the `methylKit::pool` function. This way even if we have replicated 
experiments we treat them as single experiments, and can apply Fisher's 
exact test. We will then apply the `methylKit::calculateDiffMeth()` 
function, which will return the results from the Fisher's exact test 
including the p-value and q-values for all bases or regions 
in the input methylBase object. We need to filter to get differentially 
methylated CpGs. This can be done via the `methylKit::getMethlyDiff()` function. 
After filtering, we annotate our differentially methylated basepairs 
using the BED file. This we can do using the `enomation::annotateWithGeneParts` 
function, which will return the percentage of target features overlapping with 
promotor, exon, intron and intergenic regions. 
To find out which genes are differentially methylated, we make use of the 
'refseq.hg18.bed.txt' file. We match our differentially methylated regions to
exons and/or introns in this file. Next we map these introns and exons 
back to genes by making use of the `org.Hs.eg` database. 
This information we can use to extract the refseq names from our 
differentially methylated regions, and match them to their gene names. 

```{r,eval=FALSE}
require(methylKit)
require(genomation)
require(GenomicFeatures)
require(org.Hs.eg.db)
require(RSQLite)

#load in data
file.list <- list("test1.myCpG.txt", "test2.myCpG.txt",
                  "control1.myCpG.txt", "control2.myCpG.txt")

data <- methylKit::methRead(location = file.list,
                            sample.id = list("test1","test2","control1","control2"),
                            assembly="hg18",
                            pipeline = list(fraction = FALSE, start.col = 3, 
                                            end.col = 3, chr.col = 2,
                                            coverage.col = 5, strand.col = 4,
                                            freqC.col = 6),
                            treatment=c(1,1,0,0)) #treatment indicates whether file is a test (1) or control (0) file

anno <- genomation::readTranscriptFeatures ("refseq.hg18.bed.txt")

cpg <- genomation::readFeatureFlank ("cpgi.hg18.bed.txt", feature.flank.name=c("CpGi","shores"))

#unite files to one datastructure
data_united <- methylKit::unite(data, destrand = FALSE)

#pool data for differential methylation analysis
pooled_data <- methylKit::pool(data_united,sample.ids=c("test", "control"))

#perform fishers exact test for each base:
dm.pooledf <- methylKit::calculateDiffMeth(pooled_data)

#get differentially methylated positions:
all.diff <- methylKit::getMethylDiff(dm.pooledf,difference=25,qvalue=0.01,type="all")

#annotate all parts as belonging to introns, exons, etc.
genomation::annotateWithGeneParts(as(all.diff,"GRanges"),anno)

#annotate with CpG islands
genomation::annotateWithFeatureFlank(as(all.diff,"GRanges"), cpg$CpGi,cpg$shores,
                         feature.name="CpGi",flank.name="shores")


#find out on which genes differential methylation took place:
## Get the database connection
dbcon <- org.Hs.eg_dbconn()
refseq <- dbGetQuery(dbcon, "select * from refseq")
gene_name <- dbGetQuery(dbcon, "select * from gene_info")
overlap_exon <- findOverlaps(GRanges(anno$exons),as(all.diff,"GRanges"))
overlap_intron <- findOverlaps(GRanges(anno$introns),as(all.diff,"GRanges"))

#genes which are differentially methylated in introns
intron_nr <- unique(as.data.frame(overlap_intron)[,"queryHits"])
ref_int <- as.data.frame(anno$introns)[intron_nr,"name"]
overlap_int <- which(refseq[,"accession"] %in% intersect(refseq[,"accession"], ref_int))
intron_genes<-unique(gene_name[refseq[overlap_int,"_id"],"gene_name"])

#genes which are differentially methylated in exons
exon_nr <- unique(as.data.frame(overlap_exon)[,"queryHits"])
ref_ex <- as.data.frame(anno$exons)[exon_nr,"name"]
overlap_ex <- which(refseq[,"accession"] %in% intersect(refseq[,"accession"], ref_ex))
ex_genes<-unique(gene_name[refseq[overlap_ex,"_id"],"gene_name"])
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I would like to perform methylation segmentation analysis on my file 
"test1.myCpG.txt". This holds methylation information from chromosome
21. Column 2 holds the information of which chromosome it is, column 
3 holds information of where the read begins and ends, column 4 holds the
information of which strand it is from (indicated with F and R), column 5 
holds the coverage information and column 6 holds the frequency of C's from 0 to 100.
I want to see if there are distinct methylation regions within this sample, and if so 
I want to investigate the methylation value per group and the length of the segment 
as a scatter plot.

completion:
We first read in the data using the `methylKit::methRead` function. 
Here we specify which columns hold what type of information using the 
`pipeline` argument. Since the frequency column holds values from 0 to 100, 
we set the option 'fraction' to false. We convert our data into a GRange object, 
and then we partition the GRange object into different methylated segments using the 
`methylKit::methSeg` function. To then investigate the methylation value and 
length of segments using a scatterplot, we plot the mean methylation proportion against 
the log length of the segment. 

```{r,eval=FALSE}
require(methylKit)
require(GenomicRanges)

#reading in data
met_data_test <- methylKit::methRead(location = "test1.myCpG.txt",
                                     sample.id = "test1", assembly="hg18",
                                     pipeline = list(fraction = FALSE, start.col = 3, 
                                                     end.col = 3, chr.col = 2,
                                                     coverage.col = 5, strand.col = 4,
                                                     freqC.col = 6))

#convert to GRange object
met <- GenomicRanges::GRanges(met_data_test)

#perform methylation segmentation analysis
res <- methylKit::methSeg(met)

#create plot
plot(res$seg.mean,
     log10(width(res)),pch=20,
     col=scales::alpha(rainbow(4)[as.numeric(res$seg.group)], 0.8),
     ylab="log10(length)",
     xlab="methylation proportion")
```

--- TASK END
