---
title: "prompt/completion pairs chapter 11"
author: "authors"
date: "2023-10-30"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 
'gene_exp.txt'. This dataset contains information from 
samples of 2 different cancer subgroups. Columns represent the samples, 
and the rows represent the features. There is a seperate file 
called 'subgroup.txt' which holds extra information of each sample. 
Here, columns represent features and rows represent the different 
samples. The column cms_label holds which sample belongs to which subgroup.
How can I visualize a possible relationship between this gene 
expression data and the 2 different subgroups in the column cms_label?

completion:
To visualize a possible relationship between your gene 
expression data and the two different subgroups, you can 
use various data visualization techniques. 
One common method you could consider is plotting a heatmap 
and looking at the clustering the cancer subgroups. Here is how:

```{r,eval=FALSE}
require (pheatmap)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)

#setting rownames
rownames(gene_exp) <- gene_exp[, 1]
rownames(anno_dat) <- anno_dat[, 1]

#delete column with rownames
gene_exp <- gene_exp[,-1]
anno_dat <- anno_dat [,-1]

#ensure ordering is the same
anno_dat <- anno_dat[colnames(gene_exp),]

# create a dataframe which will be used to annotate later graph
anno_col <- data.frame(cms=as.factor(anno_dat$cms_label))
rownames(anno_col) <- rownames(anno_dat)

#plotting heatmap
pheatmap::pheatmap(as.matrix(gene_exp),
                   annotation_col = anno_col,
                   show_colnames = FALSE,
                   show_rownames = FALSE,
                   main="Gene expression data")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 
'gene_exp.txt'. I also have data containing mutation status 
of certain genes and copy number data. These files are called 
'mut.txt' and 'cnv.txt' respectively. Columns represent the samples, 
and the rows represent the features. There is a seperate file called 
'subgroup.txt' which holds information of each sample. Here, columns 
represent features and rows represent the different samples. 
The column 'cms_label' holds which sample belongs to which subgroup.
This dataset contains information from samples of 2 different cancer subgroups. 
How do I combine these forms of data to perform MFA and visualize the results to 
see if I can find seperation between different subgroups?

completion:
To perform MFA on these datasets and see if seperation can be found, 
we will first read in the datasets. Then MFA is performed using the 
'FactoMineR::MFA' function. MFA is a multivariate analysis method that 
can handle data from different sources and analyze them collectively. 
In your case, you have gene expression data, mutation status data and
copy number data. To visualize the results and see if seperation 
can be found between different subgroups, a heatmap will be plotted.
Here is how to perform MFA and visualize your results:

```{r,eval=FALSE}
require(pheatmap)
require(FactoMineR)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
mut <- data.table::fread("mut.txt", header = TRUE, data.table=FALSE)
cnv <- data.table::fread("cnv.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)

#setting rownames
rownames(gene_exp) <- gene_exp[, 1]
rownames(mut) <- mut[, 1]
rownames(cnv) <- cnv[, 1]
rownames(anno_dat) <- anno_dat[, 1]

#delete column with rownames
gene_exp <- gene_exp[,-1]
mut <- mut [,-1]
cnv <- cnv[,-1]
anno_dat <- anno_dat [,-1]

#ensure ordering is the same
mut <- mut[,colnames(gene_exp)]
cnv <- cnv[,colnames(gene_exp)]
anno_dat <- anno_dat[colnames(gene_exp),]

# create a dataframe which will be used to annotate later graph
anno_col <- data.frame(cms=as.factor(anno_dat$cms_label))
rownames(anno_col) <- rownames(anno_dat)

#perform MFA
r.mfa <- FactoMineR::MFA(
  t(rbind(gene_exp,mut,cnv)), # binding the omics types together
  c(dim(gene_exp)[1], dim(mut)[1], dim(cnv)[1]), # specifying the dimensions of each
  graph=FALSE)

# first, extract the H matrix from the MFA run result
mfa.h <- r.mfa$global.pca$ind$coord

#plot heatmap with annotation
pheatmap::pheatmap(t(mfa.h)[1:2, ], annotation_col = anno_col,
                  show_colnames = FALSE,
                  main="MFA for multi-omics integration")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 'gene_exp.txt'. 
I also have data containing mutation status of certain genes and copy 
number data. These files are called 'mut.txt' and 'cnv.txt' respectively. 
Columns represent the samples, and the rows represent the features.
I also have another file holding information about the different samples
called 'subgroup.txt'. Here, samples are represented by rows, and features
are represented by columns. The column 'cms_label' holds the different cancer 
subgroups each sample belongs to. I want to do joint NMF on these datasets with 
2 components. I then want to visualize the results with a scatterplot to see if 
there is a distinction that can be made between these my different cancer subgroups. 
How do I do this?

completion:
To perform NMF on these datasets, you first need to read in the data using 
`data.table::fread'. It is then ensured that the ordering of all samples 
across the different datasets are the same. NMF is a dimensionality reduction 
technique that can help identify patterns and sources of variation in multi-omics 
datasets. In your case, you have gene expression data, mutation status data and 
copy number data. To perform NMF you first need to  
do a feature normalization on the data. You divide each value of the features
by the mean of that feature across all samples. Then we normalize the data 
by dividing all values using each omics type's frobenius norm. We then perform 
NFM using the `NFM::nfm()` function. To visualize the results
with a scatterplot we then make use of the `ggplot2:ggplot()` function

```{r,eval=TRUE}
require(NMF)
require(ggplot2)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
mut <- data.table::fread("mut.txt", header = TRUE, data.table=FALSE)
cnv <- data.table::fread("cnv.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)

#setting rownames
rownames(gene_exp) <- gene_exp[, 1]
rownames(mut) <- mut[, 1]
rownames(cnv) <- cnv[, 1]
rownames(anno_dat) <- anno_dat[, 1]

#delete column with rownames
gene_exp <- gene_exp[, -1]
mut <- mut [, -1]
cnv <- cnv[, -1]
anno_dat <- anno_dat [, -1]

#ensure ordering is the same
mut <- mut[,colnames(gene_exp)]
cnv <- cnv[,colnames(gene_exp)]
anno_dat <- anno_dat[colnames(gene_exp),]

# create a dataframe which will be used to annotate later graph
anno_col <- data.frame(cms=as.factor(anno_dat$cms_label))
rownames(anno_col) <- rownames(anno_dat)

# Feature-normalize the data
x1.featnorm <- gene_exp / rowSums(gene_exp)
x2.featnorm <- mut / rowSums(mut)
x3.featnorm <- cnv / rowSums(cnv)

# Normalize by each omics type's frobenius norm
x1.featnorm.frobnorm <- x1.featnorm / norm(as.matrix(x1.featnorm), type="F")
x2.featnorm.frobnorm <- x2.featnorm / norm(as.matrix(x2.featnorm), type="F")
x3.featnorm.frobnorm <- x3.featnorm / norm(as.matrix(x3.featnorm), type="F")

#perform NMF
r.nmf <- NMF::nmf(t(rbind(x1.featnorm.frobnorm, x2.featnorm.frobnorm,x3.featnorm.frobnorm)),rank=2, method='Frobenius')

# exctract the H matrix from the nmf run result
nmf.h <- NMF::basis(r.nmf)

# create a dataframe with the H matrix and the CMS label (subtype)
nmf_df <- as.data.frame(nmf.h,row.names=rownames(nmf.h))
colnames(nmf_df) <- c("dim1", "dim2")
nmf_df$subtype <- factor(anno_dat[rownames(nmf_df),]$cms_label)

# create the scatter plot
ggplot2::ggplot(nmf_df, ggplot2::aes(x=dim1, y=dim2, color=subtype)) +
ggplot2::geom_point() +
ggplot2::ggtitle("Scatter plot of 2-component NMF for multi-omics integration")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 'gene_exp.txt'. 
I also have data containing mutation status of certain genes and copy 
number data. These files are called 'mut.txt' and 'cnv.txt' respectively. 
Columns represent the samples, and the rows represent the features. I also 
have another file holding information about the different samples called 
'subgroup.txt'. Here, samples are represented by rows, and features are 
represented by columns. The column 'cms_label' holds the different cancer 
subgroups each sample belongs to. 
I want the mutation file to just hold 1 for when a feature is mutated, 
and 0 if it is not, instead of a continuous value above zero when it is 
mutated like now. Then I want to do the following:
These different omic types are from different distributions, so to analyze 
them I would like to perform iCluster+ analysis. The gene expression, mutation 
status and copy number data are from a gaussian, binomial and multinomial 
distribution respectively. I would like to learn 2 factors. I would then like to 
take these results and visualize them as both a scatterplot and heatmap to see 
if there is a distinction between cancer subgroups. How do I do this?

completion:
To work with these datasets, we first read in the data using 
`data.table::fread'. We then ensure that the ordering of all 
samples across the different datasets are the same.
To change the values of the mutation dataset, so that mutated 
genes are indicated by 1 and non-mutated genes are indicated by 0, 
we overwrite all values in this matrix that are greater then 0, and 
change these to 1. We then do the iClusterPlus analysis using the 
`iClusterPlus::iClusterPlus`, providing the function with the specified 
distributions and desired number of factors. We then extract the H matrix 
from the resulting iClusterPlus object, and use this to plot a scatterplot 
and heatmap.

```{r,eval=FALSE}
require(ggplot2)
require(pheatmap)
require(iClusterPlus)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
mut <- data.table::fread("mut.txt", header = TRUE, data.table=FALSE)
cnv <- data.table::fread("cnv.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)

#setting rownames
rownames(gene_exp) <- gene_exp[,1]
rownames(mut) <- mut[,1]
rownames(cnv) <- cnv[,1]
rownames(anno_dat) <- anno_dat[,1]

#delete column with rownames
gene_exp <- gene_exp[,-1]
mut <- mut [,-1]
cnv <- cnv[,-1]
anno_dat <- anno_dat [,-1]

#ensure ordering is the same
mut <- mut[,colnames(gene_exp)]
cnv <- cnv[,colnames(gene_exp)]
anno_dat <- anno_dat[colnames(gene_exp),]

#change mutation states to 0 or 1 instead of continuous value:
mut[mut>0] <- 1

# run the iClusterPlus function
r.icluster <- iClusterPlus::iClusterPlus(
  t(as.matrix(gene_exp)), # Providing each omics type
  t(as.matrix(mut)),
  t(as.matrix(cnv)),
  type=c("gaussian", "binomial", "multinomial"), # Providing the distributions
  K=2) #providing the amount of factors to learn

#Extract H (z) matrix
icluster.z <- r.icluster$meanZ
rownames(icluster.z) <- rownames(anno_dat) # fix the row names

# construct a dataframe with the H matrix (z) and the cancer subtypes
# for later plotting
icp_df <- as.data.frame(icluster.z)
colnames(icp_df) <- c("dim1", "dim2")
rownames(icp_df) <- colnames(gene_exp)
icp_df$subtype <- factor(anno_dat[rownames(icp_df),]$cms_label)

# create the scatter plot
ggplot2::ggplot(icp_df, ggplot2::aes(x=dim1, y=dim2, color=subtype)) +
ggplot2::geom_point() +
ggplot2::ggtitle("Scatter plot of iCluster+ factors")

# create the heatmap
anno_col <- data.frame(cms=as.factor(anno_dat$cms_label))
rownames(anno_col) <- rownames(anno_dat)

pheatmap::pheatmap(t(icp_df[,1:2]),annotation_col=anno_col,show_colnames=FALSE, main="Heatmap of iCluster+ factors")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 'gene_exp.txt'. 
I also have data containing mutation status of certain genes and copy number 
data. These files are called 'mut.txt' and 'cnv.txt' respectively. Columns 
represent the samples, and the rows represent the features. I also have another 
file holding information about the different samples called 'subgroup.txt'. 
Here, samples are represented by rows, and features are represented by
columns. The column 'cms_label' holds two different cancer subgroups each 
sample can belong to. I want to do joint NMF on these datasets with 2 components. 
Then I want to cluster the samples based on which latent variable influences 
it the most, and compare this to the actual cancer subgroups in the data. How do I do this?

completion:
To perform NMF on these datasets, we first read in the data using 
`data.table::fread'. We then ensure that the ordering of all samples across 
the different datasets are the same. To perform NMF we first do a feature 
normalization on the data. We divide each value of the features by the mean 
of that feature across all samples. Then we normalize the data by dividing all 
values using each omics type's frobenius norm. We then perform NFM using the `NFM::nfm()` 
function. Here we use `rank = 2` to ensure we learn 2 factors. We then extract the 
H matrix from the results, and make our prediction based on which factor influences the 
sample the most. Finally, we plot our heatmap with both annotations of the actual label, 
and the predicted label. Note that here we don't cluster the rows or columns, but plot
a heatmap ordered based on the predicted subgroups. 

```{r,eval=FALSE}
require(NMF)
require(pheatmap)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
mut <- data.table::fread("mut.txt", header = TRUE, data.table=FALSE)
cnv <- data.table::fread("cnv.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)                               

#setting rownames
rownames(gene_exp) <- gene_exp[, 1]
rownames(mut) <- mut[, 1]
rownames(cnv) <- cnv[, 1]
rownames(anno_dat) <- anno_dat[, 1]

#delete column with rownames
gene_exp <- gene_exp[, -1]
mut <- mut [, -1]
cnv <- cnv[, -1]
anno_dat <- anno_dat [, -1]

#ensure ordering is the same
mut <- mut[,colnames(gene_exp)]
cnv <- cnv[,colnames(gene_exp)]
anno_dat <- anno_dat[colnames(gene_exp),]

# Feature-normalize the data
x1.featnorm <- gene_exp / rowSums(gene_exp)
x2.featnorm <- mut / rowSums(mut)
x3.featnorm <- cnv / rowSums(cnv)

# Normalize by each omics type's frobenius norm
x1.featnorm.frobnorm <- x1.featnorm / norm(as.matrix(x1.featnorm), type="F")
x2.featnorm.frobnorm <- x2.featnorm / norm(as.matrix(x2.featnorm), type="F")
x3.featnorm.frobnorm <- x3.featnorm / norm(as.matrix(x3.featnorm), type="F")

#perform NMF with rank of 2 to learn two factors
r.nmf <- NMF::nmf(t(rbind(x1.featnorm.frobnorm, x2.featnorm.frobnorm,x3.featnorm.frobnorm)),rank=2, method='Frobenius')

# exctract the H matrix from the nmf run result
nmf.h <- NMF::basis(r.nmf)

# create a dataframe with the H matrix and the CMS label (subtype)
nmf_df <- as.data.frame(nmf.h,row.names=rownames(nmf.h))
colnames(nmf_df) <- c("dim1", "dim2")
nmf_df$subtype <- factor(anno_dat[rownames(nmf_df),]$cms_label)

#create our predictions for which sample belongs to which group (arbitrarily named group 1 and 2) based on which factor influences it the most
pred_anno <- max.col(nmf.h)

# Create annotation column for later plotting
anno_col <- data.frame(cms=as.factor(anno_dat$cms_label), predicted=as.factor(pred_anno))
rownames(anno_col) <- rownames(anno_dat)

#order dataframes in order of predicted subtype:
nmf_df_ordered <- nmf_df[order(anno_col$predicted),1:2]
anno_col <- anno_col[order(anno_col$predicted),]

#plot heatmap with both predicted and actual annotations
pheatmap::pheatmap(t(nmf_df_ordered),annotation_col=anno_col,show_colnames=FALSE, 
                    main="Heatmap of NMF factors",cluster_cols=FALSE, cluster_rows=FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 
'gene_exp.txt'. I also have data containing mutation status of 
certain genes and copy number data. These files are called 'mut.txt' 
and 'cnv.txt' respectively. Columns represent the samples, and the rows 
represent the features. For the 'mut.txt' file, I would like al features 
(genes) to have only a zero when it is not mutated, and a 1 if it is 
mutated instead of a continuous value. I also have another file holding
information about the different samples called 'subgroup.txt'. 
Here, samples are represented by rows, and features are represented by
columns. The column 'cms_label' holds two different cancer subgroups 
each sample can belong to. I want to perform iClusterPlus analysis. 
Since these different omic types are from different distributions, 
I would like to take this into account when doing my analysis. 
The gene expression, mutation status and copy number data are from a 
gaussian, binomial and multinomial distribution respectively. 
I would like to learn 2 factors, and subsequently use K-means clustering 
to cluster the samples into two groups. 
I then would like to visualize the results in a heatmap, which shows the
predicted group, as well as the actual label. How do I do this?

completion:
To work with these datasets, we first read in the data using 
`data.table::fread'. We then ensure that the ordering of all samples 
across the different datasetsare the same. To change the values of 
the mutation dataset, so that mutated genes are indicated by 1 
and non-mutated genes are indicated by 0, we overwrite all values
in this matrix that are greater then 0, and change these to 1. 
We then do the iClusterPlus analysis using the `iClusterPlus::iClusterPlus`
function. You can provide the function with the specified distributions and desired 
number of factors. We then extract the H matrix from the resulting 
iClusterPlus object, and perform k-means clustering with this using 
the euclidian distance and 2 K. Finally, we plot our heatmap with 
both annotations of the actual label, and the predicted label. 
Note that here we don't cluster the rows or columns, but plot
a heatmap ordered based on the predicted subgroups based on the Kmeans 
clustering.  

```{r,eval=FALSE}
require(iClusterPlus)
require(pheatmap)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
mut <- data.table::fread("mut.txt", header = TRUE, data.table=FALSE)
cnv <- data.table::fread("cnv.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)                               

#setting rownames
rownames(gene_exp) <- gene_exp[, 1]
rownames(mut) <- mut[, 1]
rownames(cnv) <- cnv[, 1]
rownames(anno_dat) <- anno_dat[, 1]

#delete column with rownames
gene_exp <- gene_exp[, -1]
mut <- mut [, -1]
cnv <- cnv[, -1]
anno_dat <- anno_dat [, -1]

#ensure ordering is the same
mut <- mut[, colnames(gene_exp)]
cnv <- cnv[, colnames(gene_exp)]
anno_dat <- anno_dat[colnames(gene_exp), ]

#change mutation states to 0 or 1 instead of continuous value:
mut[mut > 0] <- 1

# run the iClusterPlus function
r.icluster <- iClusterPlus::iClusterPlus(
      t(as.matrix(gene_exp)), # Providing each omics type
      t(as.matrix(mut)),
      t(as.matrix(cnv)),
      type=c("gaussian", "binomial", "multinomial"), # Providing the distributions
      K=2) #providing the amount of factors to learn

#Extract the H (z) matrix
icluster.z <- r.icluster$meanZ
rownames(icluster.z) <- rownames(anno_dat) # fix the row names

#perform K-means clustering on the H (z) matrix with
#2 groups and extract the cluster labels. 
icluster.clusters <- as.data.frame(kmeans(icluster.z, 2)$cluster)
rownames(icluster.clusters) <- rownames(icluster.z)


# construct a dataframe with the H matrix (z) and the
# cancer subtypes and predictions for later plotting
icp_df <- as.data.frame(icluster.z)
colnames(icp_df) <- c("dim1", "dim2")
rownames(icp_df) <- colnames(gene_exp)
icp_df$subtype <- factor(anno_dat[rownames(icp_df), ]$cms_label)
icp_df$predicted <- factor (icluster.clusters[rownames(icp_df), ])

#order dataframe based on predicted label:
#order dataframes in order of predicted subtype:
icp_df <- icp_df [order(icp_df$predicted), ]

#plot heatmap
pheatmap::pheatmap(t(icp_df[, 1:2]), annotation_col = icp_df[, 3:4],
                   show_colnames = FALSE,
                   main = "Heatmap of iCluster+ factors",
                   cluster_rows = FALSE, cluster_cols = FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 
'gene_exp.txt'. I also have data containing mutation status
of certain genes and copy number data. These files are called 
'mut.txt' and 'cnv.txt' respectively. Columns represent the 
samples, and the rows represent the features.
I also have another file holding information about the different
samples called 'subgroup.txt'. Here, samples are represented by rows, 
and features are represented by columns. The column 'cms_label' holds
two different cancer subgroups each sample can belong to. 
I want to do joint NMF on these datasets with 2 factors. I then want 
to visualize the results with a heatmap to see if there is a distinction 
that can be made between these my different cancer subgroups. 
I also want to create a heatmap where there is an annotation of the rows
to see from which source the information came (gene expression, mutation
status or copy number variation). I then 
want to interpret these learnt factors in a biological context by looking at
enrichment in GO terms for the gene expression part,
to see which features influence these factors the most. 
How do I do this?

completion:
To perform NMF on these datasets, we first read in the data using 
`data.table::fread'. We then ensure that the ordering of all samples 
across the different datasets are the same. To perform NMF we first 
do a feature normalization on the data. We divide each value of the 
features by the mean of that feature across all samples. 
Then we normalize the data by dividing all values using each omics 
type's frobenius norm. We then perform NFM using the `NFM::nfm()` function. 
Here we use `rank = 2` to ensure we learn 2 factors. We then visualize the 
results using the `pheatmap::pheatmap` function. To interpret the learnt 
factors in a biological context, we can look at how the different features 
differ from eachother in the different factors. This we do by extracting the 
W matrix and plotting a heatmap of this. 
This however tells us only which features influence the factors, but don't 
tell us anything about possible enrichments in biological processes. For this 
we can use the `enrichR::enrichr` function together with the differentially 
expressed genes. Then we can test if there is any enrichment in factors
in biological processes using a hypergeometric test. We do this by selecting 
those genes that have a higher weight in one factor compared to the other. 
We can visualize these results in a table showing the enriched GO terms. 

```{r,eval=TRUE}
require(NMF)
require(pheatmap)
require(enrichR)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
mut <- data.table::fread("mut.txt", header = TRUE, data.table=FALSE)
cnv <- data.table::fread("cnv.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)                               

#setting rownames
rownames(gene_exp) <- gene_exp[,1]
rownames(mut) <- mut[,1]
rownames(cnv) <- cnv[,1]
rownames(anno_dat) <- anno_dat[,1]

#delete column with rownames
gene_exp <- gene_exp[,-1]
mut <- mut [,-1]
cnv <- cnv[,-1]
anno_dat <- anno_dat [,-1]

#ensure ordering is the same
mut <- mut[,colnames(gene_exp)]
cnv <- cnv[,colnames(gene_exp)]
anno_dat <- anno_dat[colnames(gene_exp),]

# Feature-normalize the data
x1.featnorm <- gene_exp / rowSums(gene_exp)
x2.featnorm <- mut / rowSums(mut)
x3.featnorm <- cnv / rowSums(cnv)

# Normalize by each omics type's frobenius norm
x1.featnorm.frobnorm <- x1.featnorm / norm(as.matrix(x1.featnorm), type="F")
x2.featnorm.frobnorm <- x2.featnorm / norm(as.matrix(x2.featnorm), type="F")
x3.featnorm.frobnorm <- x3.featnorm / norm(as.matrix(x3.featnorm), type="F")

#perform NMF with rank of 2 to learn two factors
r.nmf <- NMF::nmf(t(rbind(x1.featnorm.frobnorm, x2.featnorm.frobnorm,x3.featnorm.frobnorm)),rank=2, method='Frobenius')

# exctract the H and W matrices from the nmf run result
nmf.h <- NMF::basis(r.nmf)
nmf.w <- NMF::coef(r.nmf)
nmfw <- t(nmf.w)

# create a dataframe with the H matrix and the CMS label (subtype)
nmf_df <- as.data.frame(nmf.h, row.names=rownames(nmf.h))
colnames(nmf_df) <- c("dim1", "dim2")
nmf_df$subtype <- factor(anno_dat[rownames(nmf_df),]$cms_label)

# Create annotation column for later plotting
anno_col <- data.frame(cms=as.factor(anno_dat$cms_label))
rownames(anno_col) <- rownames(anno_dat)

#plot heatmap with annotations
pheatmap::pheatmap(t(nmf_df[,1:2]),annotation_col=anno_col,show_colnames=FALSE, 
                    main="Heatmap NMF factors")

#heatmap annotation of biological interpretation
data_anno <- data.frame(
  omics=c(rep('expression',dim(gene_exp)[1]),
          rep('mut',dim(mut)[1]),
          rep('cnv',dim(cnv)[1])))
rownames(data_anno) <- rownames(nmfw)

# generate the heat map
pheatmap::pheatmap(nmfw,
                   cluster_cols = FALSE,
                   annotation_row = data_anno,
                   main="NMF coefficients",
                   clustering_distance_rows = "manhattan",
                   fontsize_row = 1)

# select genes associated preferentially with each factor
# by their relative loading in the W matrix
genes.factor.1 <- names(which(nmfw[1:dim(gene_exp)[1],1] > nmfw[1:dim(gene_exp)[1],2]))
genes.factor.2 <- names(which(nmfw[1:dim(gene_exp)[1],1] < nmfw[1:dim(gene_exp)[1],2]))

# call the enrichr function to find gene sets enriched
# in each latent factor in the GO Biological Processes 2018 library
go.factor.1 <- enrichR::enrichr(genes.factor.1,
                                databases = c("GO_Biological_Process_2018")
                                )$GO_Biological_Process_2018
go.factor.2 <- enrichR::enrichr(genes.factor.2,
                                databases = c("GO_Biological_Process_2018")
                                )$GO_Biological_Process_2018

#Plot enrichment table
plotEnrich(go.factor.1, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value")
plotEnrich(go.factor.2, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: ML/Stats | visualization, numData: 4, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataset containing gene expression data called 'gene_exp.txt'.
I also have data containing mutation status of certain genes and copy 
number data. These files are called 'mut.txt' and 'cnv.txt' respectively. 
Columns represent the samples, and the rows represent the features.
I also have another file holding information about the different samples 
called 'subgroup.txt'. Here, samples are represented by rows, 
and features are represented by columns. The column 'cms_label' holds two 
different cancer subgroups each sample can belong to. There are also columns
called 'gender' and 'msi' which also have two different subgroups.  I want 
to do joint NMF on these datasets with 2 factors. I then want to see if any 
of these factors are correlated to the different
features like 'cancer subgroup', 'gender' or 'msi'. In other words, I want 
to see if these factors are able to provide a split between these groups.

completion:
To perform NMF on these datasets, we first read in the data using 
`data.table::fread'. We then ensure that the ordering of all samples 
across the different datasets are the same. To perform NMF we first do 
a feature normalization on the data. We divide each value of the features 
by the mean of that feature across all samples. Then we normalize the 
data by dividing all values using each omics type's frobenius norm. 
We then perform NFM using the `NFM::nfm()` function. Here we use
`rank = 2` to ensure we learn 2 factors. To see if any of the additional 
features 'gender','msi' and 'cancer subtype' are correlated to the learnt factors,
we can visualize their relationship using a boxplot, scatterplot or heatmap to see 
if these factors are able to split up the groups. To calculate the correlation between
a continuous variable such as our factor and these features, different 
statistical tests such as the Point-Biserial Correlation can be performed. 
Which test is appropriate will however depend on the distribution of your data. 
For now, here is how you visualize the possible relationship using a boxplot, 
heatmap and a scatterplot:

```{r,eval=FALSE}
require(pheatmap)
require(NFM)
require(ggplot2)

#reading in data
gene_exp <- data.table::fread("gene_exp.txt", header = TRUE, data.table=FALSE)
mut <- data.table::fread("mut.txt", header = TRUE, data.table=FALSE)
cnv <- data.table::fread("cnv.txt", header = TRUE, data.table=FALSE)
anno_dat <- data.table::fread("subgroup.txt", header = TRUE, data.table=FALSE)

#setting rownames
rownames(gene_exp) <- gene_exp[,1]
rownames(mut) <- mut[,1]
rownames(cnv) <- cnv[,1]
rownames(anno_dat) <- anno_dat[,1]

#delete column with rownames
gene_exp <- gene_exp[,-1]
mut <- mut [,-1]
cnv <- cnv[,-1]
anno_dat <- anno_dat [,-1]

#ensure ordering is the same
mut <- mut[,colnames(gene_exp)]
cnv <- cnv[,colnames(gene_exp)]
anno_dat <- anno_dat[colnames(gene_exp),]

# Feature-normalize the data
x1.featnorm <- gene_exp / rowSums(gene_exp)
x2.featnorm <- mut / rowSums(mut)
x3.featnorm <- cnv / rowSums(cnv)

# Normalize by each omics type's frobenius norm
x1.featnorm.frobnorm <- x1.featnorm / norm(as.matrix(x1.featnorm), type="F")
x2.featnorm.frobnorm <- x2.featnorm / norm(as.matrix(x2.featnorm), type="F")
x3.featnorm.frobnorm <- x3.featnorm / norm(as.matrix(x3.featnorm), type="F")

#perform NMF with rank of 2 to learn two factors
require(NMF)
r.nmf <- NMF::nmf(t(rbind(x1.featnorm.frobnorm,
                  x2.featnorm.frobnorm,x3.featnorm.frobnorm)),
                  rank=2, method="Frobenius")

# exctract the H matrix from the nmf run result and create dataframe for plotting
nmf_df <- as.data.frame(NMF::basis(r.nmf))
colnames(nmf_df) <- c("dim1", "dim2")
nmf_df$cms <- factor(anno_dat[rownames(nmf_df), ]$cms_label)
nmf_df$gender <- factor(anno_dat[rownames(nmf_df), ]$gender)
nmf_df$msi <- factor(anno_dat[rownames(nmf_df), ]$msi)

#scatterplot annotating points by cms,gender or msi
for (i in c(nmf_df$cms,nmf_df$gender,nmf_df$msi)){
  ggplot2::ggplot(nmf_df, ggplot2::aes(x=dim1, y=dim2, color=i)) +
  ggplot2::geom_point() +
  ggplot2::ggtitle("Scatter plot of 2-component NMF for multi-omics integration")
}

#plot heatmap with annotations
pheatmap::pheatmap(t(nmf_df[,1:2]),annotation_col=nmf_df[, 3:5],show_colnames=FALSE, 
                    main="Heatmap NMF factors")

#plot barplots exploring relationships of factor 1
for (i in c("cms","gender","msi")){
  plot <- ggplot2::ggplot(nmf_df, ggplot2::aes(x=eval(parse(text=i)), y=dim1, group=eval(parse(text=i)))) +
            ggplot2::geom_boxplot() +
            ggplot2::xlab(i) +
            ggplot2::ggtitle(paste0("NMF factor 1 with ",i))
  print (plot)
}

#plot barplots exploring relationships of factor 2
for (i in c("cms","gender","msi")){
  plot <- ggplot2::ggplot(nmf_df, ggplot2::aes(x=eval(parse(text=i)), y=dim2, group=eval(parse(text=i)))) +
            ggplot2::geom_boxplot() +
            ggplot2::xlab(i) +
            ggplot2::ggtitle(paste0("NMF factor 2 with ",i))
  print (plot)
}
```

--- TASK END
