---
title: "prompt/completion pairs chapter 11"
author: "authors"
date: "2023-10-30"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
i <- 1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe holding gene expression data 
called 'gene_exp'. I also have a dataframe with 
annotation data called 'anno_dat'. The dataframe
`gene_exp` has the columnnames which correspond
to the rownames in `anno_dat`. How do I ensure that
they have the same ordering?

completion:
To ensure that the column names in your gene expression dataframe
'gene_exp' correspond to the row names in your annotation dataframe
'anno_dat' and that they have the same ordering, you can use the 
columnames of the 'gene_exp' dataframe to order the 'anno_dat'
dataframe. Here is how:

```{r,eval=FALSE}
# Make ordering the same
anno_dat <- anno_dat[colnames(gene_exp), ]
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe holding gene expression data 
called 'gene_exp'. Columns represent 
the samples, and the rows represent the features.
I also have a dataframe with 
annotation data called 'anno_dat'. The dataframe
`gene_exp` has the columnnames which correspond
to the rownames in `anno_dat`. I want to plot a 
heatmap of my gene expression data, while annotating
the columns with the data from the column `cms_label`
from the `anno_dat` dataframe. How do I do this?

completion:
To create a heatmap of your gene expression data while annotating
the columns with the data from the cms_label column of the anno_dat 
dataframe in R, you can use the pheatmap function from the "pheatmap" 
package. 

```{r,eval=FALSE}
require (pheatmap)

annotation <- anno_dat[, "cms_label", drop = FALSE]

#plotting heatmap
pheatmap::pheatmap(as.matrix(gene_exp),
                   annotation_col = annotation,
                   show_colnames = FALSE,
                   show_rownames = FALSE,
                   main = "Gene expression data")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe holding gene expression data 
called 'gene_exp'. Columns represent 
the samples, and the rows represent the features.
I also have a dataframe with annotation data called 
'anno_dat'. The dataframe `gene_exp` has the columnnames 
which correspond to the rownames in `anno_dat`. 
I want to plot a heatmap of my gene expression data, 
while annotating the columns with the data from the 
columns `cms_label` and 'gender' from the `anno_dat` 
dataframe. How do I do this?

completion:
To create a heatmap of your gene expression data while annotating
the columns with the data from the cms_label column of the anno_dat 
dataframe in R, you can use the pheatmap function from the "pheatmap" 
package. 

```{r,eval=FALSE}
require (pheatmap)

annotation <- anno_dat[, c("cms_label", "gender"), drop = FALSE]

#plotting heatmap
pheatmap::pheatmap(as.matrix(gene_exp),
                   annotation_col = annotation,
                   show_colnames = FALSE,
                   show_rownames = FALSE,
                   main="Gene expression data")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 3 dataframes called `mut`,`gene_exp` and `cnv`.
These store omics data with columns representing samples,
and rows representing different features. The ordering
of the samples is the same across all dataframes. 
How do I combine these forms of data to perform MFA?

completion:
To combine the data in dataframes `mut`, `gene_exp` and `cnv`
which have the same columns that represent samples, we can make use
of the `rbind` function to bind these dataframes together. 
In MFA, you typically analyze the data with observations (samples) 
in rows and variables (features) in columns. Therefore, 
you can transpose the combined dataframe. You can then perform
MFA using the `FactoMineR::MFA` function, specifying the
dimensions of each omics type. The results are stored in 
the variable `r_mfa`

```{r,eval=FALSE}
require (FactoMineR)

# binding omics types together:
omics <- rbind(gene_exp, mut, cnv)
t_omics <- t(omics)

# perform MFA
r_mfa <- FactoMineR::MFA(
  t_omics, #transpose matrix so features are columns
  c(dim(gene_exp)[1], dim(mut)[1], dim(cnv)[1]), # specifying the dimensions of each omics type
  graph = FALSE)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I performed MFA using the `FactoMineR::MFA` function.
The results are stored in a variable called `r_mfa`.
How do I extract the H-matrix from this analysis

completion:
To extract the H-matrix from the results of
Multiple Factor Analysis (MFA) using the
`FactoMineR::MFA` function, you can use 
the `r_mfa$global.pca$ind$coord` component
of `r_mfa`. Here is how you do it:

```{r,eval=FALSE}
# extract H-matrix from
# Results of the Multiple Factor Analysis (MFA)
h_matrix <- r_mfa$global.pca$ind$coord
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a H-matrix from my MFA analysis
which has samples as rows and dimensions as columns.
I want to plot the samples and the first 2 dimensions 
in a heatmap, and annotate them using another dataframe I have
called `anno_dat`, which has samples as rows and contains a column
called 'cms_label' which I want to use for annotation. How 
do I do this?

completion:
To plot the samples and the first two dimensions from your H-matrix 
in a heatmap and annotate them using the 'cms_label' column from the 
'anno_dat' dataframe, you can use the pheatmap package in R. 
For annotating the samples we can use the `annotation_row`
function. 

```{r,eval=FALSE}
pheatmap::pheatmap(h_matrix[ ,1:2],
                   annotation_row = anno_dat[, "cms_label", drop=FALSE])

```

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a H-matrix from my MFA analysis
which has samples as rows (and rownames correspond to sample name)
and dimensions as columns. This is called `h_matrix`
I want to plot the samples and the first 2 dimensions in a heatmap, and 
annotate them using another dataframe I have called `anno_dat`, 
which has samples as rows (and rownames correspond to 
sample name) and contains a column called 'cms_label' which I want to use 
for annotation. I want to present the heatmap in a way where samples which 
have the same cms_label are next to eachother. How do I do this?

completion:
To plot the samples and the first two dimensions from your H-matrix 
in a heatmap and annotate them using the 'cms_label' column from the 
'anno_dat' dataframe, you can use the pheatmap package in R. 
To ensure that samples with the same label get plotted next
to eachoter, you can create a new dataframe called cms_label
which holds the information of the `cms_label` column in `anno_dat`.
This dataframe you can order based on label name, and use this 
new ordering to order the H-matrix `h_matrix` as well.
We can then plot the heatmap using the function `pheatmap`.
For annotating the samples we can use the `annotation_row`
function. To keep the same ordering you set in the dataframe, 
you can set the argument `cluster_rows = FALSE`.

```{r,eval=FALSE}
require (pheatmap)

# extract column of interest for labeling
cms_label <- anno_dat[ , "cms_label",drop = FALSE]

# order based on labeling
cms_label <- cms_label[order(cms_label$cms_label), , drop = FALSE]

# order h_matrix based on ordering of
# cms_label
h_matrix <- h_matrix [rownames(cms_label), , drop = FALSE]

# plot heatmap
pheatmap::pheatmap(h_matrix[ ,1:2],
                   annotation_row = cms_label,
                   cluster_rows = FALSE)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `mut`
This stores omics data with columns representing samples,
and rows representing different features. I 
want to feature-normalize the data. How do I do this?

completion:
Feature normalization,is a process of rescaling each 
feature of your data so that they have similar scales. 
This is often done to ensure that features contribute 
equally to the analysis and to prevent features with 
larger scales from dominating the results. Since in 
your dataframe features are represented as rows, 
we can divide the dataframe `mut` by the rowsums of `mut`. 
This divides each value in `mut` by its corresponding rowsum. 
Here is how you do it:

```{r,eval=FALSE}

# Feature normalize data
mut_featnorm <- mut / rowSums(mut)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have a dataframe called `mut`
This stores omics data with columns representing samples,
and rows representing different features. I want to 
normalize my data by its frobenius norm. How do I do this?

completion:
To normalize your omics data by its Frobenius norm,
you can make use of the `norm` function. Using the 
argument `type=F`, this will calculate the 
Frobenius norm of the matrix `mut`. 

```{r,eval=FALSE}
library(matrixStats)

# calculate Frobenius norm
mut_frobnorm <- mut / norm(as.matrix(mut), type = "F")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 3 dataframes called 'x1','x2' and 'x3'.
These hold information about different omics data. 
Columns represent samples,and rows represent different 
features. columnames are sample names and rownames are
featurenames. The ordering of the samples is the same across 
all dataframes. These dataframes are feature normalized and
contain the Frobenius norm matrix. I now want to perform
NMF. How do I do this?

completion:
NMF is a dimensionality reduction and factorization technique 
that's commonly used in data analysis, particularly in omics data 
analysis. To perform NMF, you need the `NMF` library. You mentioned 
that 'x1', 'x2', and 'x3' are already feature-normalized and 
contain the Frobenius norm matrix. Assuming that these dataframes 
are already properly formatted, you can proceed to create an NMF 
model using these data. Combine the three dataframes into a single 
data matrix, as NMF typically works with a single data matrix. 
Make sure that the column order (sample order) is consistent among the 
dataframes. Assuming you want to analyze the data with observations 
(samples) in rows and variables (features) in columns, we transpose the 
dataframes. You need to specify the number of factors (also known as rank) 
that you want the NMF algorithm to extract. This depends on your data and 
objectives. The nmf function requires specifying the rank using the rank 
parameter. You may need to experiment with different rank values to determine 
the most informative factorization for your omics data.

```{r,eval=FALSE}
require(NMF)

# perform NMF
data <- rbind (x1,x2,x3)
t_data <- t(data)

# adjust as needed
rank <- 2

# perform  NMF
r_nmf <- NMF::nmf(t_data,rank=rank, method="Frobenius")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have performed NMF using the `NMF::nmf` function.
The results of this are stored in an object of class: NMFfit
called `r_nmf`. I want to extract the H- and W-matrices
from this. How do I do this?

completion:
To extract the H- and W-matrices from a 
NMFfit object, you can make use of the 
`basis` and `coef` functions from the NMF
library. Here is how you do it:

```{r,eval=FALSE}
require(NMF)

# exctract the H and W matrices from the nmf run result
nmf_h <- NMF::basis(r_nmf)
nmf_w <- NMF::coef(r_nmf)
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have performed NMF using the `NMF::nmf` function with a rank of 2.
I extracted the H-matrix from this analysis called `h_matrix` 
and now I want to create a scatterplot based on the 2
dimensions or ranks which are held in the two columns 
of this matrix. The rows of this matrix represent my samples.
The rownames correspond to the sample names. I also have a dataframe 
called `anno_dat`. This has samples represented as 
rows (samplenames are rownames). There is a column in this dataframe 
called `cms_label` which I want to use to annotate the points in the 
scatterplot. How do I do this?

completion:
To create a scatterplot based on the two dimensions (ranks) from 
the H-matrix obtained from NMF analysis and annotate the points using the 
'cms_label' column from the 'anno_dat' dataframe, you can use the ggplot2 
package. Here is how you do it:

```{r,eval=FALSE}
require(ggplot2)

# extract the desired dims for plotting
data_plot <- as.data.frame(h_matrix[, 1:2, drop = FALSE])

# change colnames
colnames(data_plot) <- c("dim1", "dim2")

# extract labeling data
labels <- anno_dat [,"cms_label",drop=FALSE]

# ensure ordering of samples is the same:
labels <- labels[rownames(data_plot), ,drop=FALSE]

# plot scatterplot
ggplot2::ggplot(data_plot,aes(x=dim1, y=dim2, color=labels[,"cms_label"])) +
ggplot2::geom_point() +
ggplot2::ggtitle("Scatter plot of 2 dim NMF")
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have 3 dataframes called 'x1','x2' and 'x3'.
These hold information about different omics data. 
Columns represent samples,and rows represent different 
features. columnames are sample names and rownames are
featurenames. The ordering of the samples is the same across 
all dataframes. I want to perform iCluster+ analysis. The `x1`, 'x2'
and 'x3' are from a gaussian, binomial and multinomial 
distribution respectively. I would like to learn 2 factors.
How do I do this?

completion:
iCluster+ is a statistical method used for integrative clustering of 
multi-omics data. It can handle different types of data, including 
Gaussian, binomial, and multinomial distributions. To perform iCluster+ 
analysis in R with your 'x1,' 'x2,' and 'x3' dataframes, you can use the 
iClusterPlus package. Make sure that the column order (sample order) is 
consistent among the dataframes. Assuming you want to analyze the data 
with observations (samples) in rows and variables (features) in columns, 
the dataframes are transposed. Use the iClusterPlus function to perform the 
analysis. Specify the data, data types, and the number of factors you want to learn 
(in this case, 2 factors). 

```{r,eval=FALSE}
require(iClusterPlus)

# transpose dataframe
t_x1 <- as.matrix(t(x1))
t_x2 <- as.matrix(t(x2))
t_x3 <- as.matrix(t(x3))

# run the iClusterPlus function
r_icluster <- iClusterPlus::iClusterPlus(
    t_x1,
    t_x2,
    t_x3,
    type = c("gaussian", "binomial", "multinomial"), # Providing the distributions
    K = 2) #providing the amount of factors to learn
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I performed iCluster+ analysis using the `iClusterPlus::iClusterPlus`
function. The results I stored in a variable called `r_cluster`. I now
want to extract the different cluster labels. How do I do this?

completion:
The iCluster_result contains various information about the clustering results, 
including the factor loadings, sample cluster assignments, and more. You can explore and 
interpret these results based on your research objectives. To access the cluster label 
information, you can make use of the `$clusters` operator. Here is how you do it:

```{r,eval=FALSE}
# get cluster labels
clus_lab <- r_icluster$clusters
```

--- TASK END

## Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I performed iCluster+ analysis using the `iClusterPlus::iClusterPlus`
function. The results I stored in a variable called `r_cluster`. I now
want to extract the H- and W-matrices from this analysis. How do I 
do this?

completion:
The iCluster_result contains various information about the clustering results, 
including the factor loadings, sample cluster assignments, and more. You can explore and 
interpret these results based on your research objectives. To access the H-matrix 
you can make use of the `$meanZ` operator. To access the W-matrix, you can make
use of the `$beta` operator. Here is how you do it:

```{r,eval=FALSE}
# Extract the H (z) matrix
icluster_h <- r_icluster$meanZ

# Extract the W matrix
icluster_w <- r_icluster$beta
```

--- TASK END

## Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I performed iCluster+ analysis using the `iClusterPlus::iClusterPlus`
function. The results I stored in a variable called `r_cluster`. I want
to make a scatterplot which plots the first two dimensions agains eachother,
and annotate the points based on their clusterlabel. How do I do this?

completion:
The iCluster_result contains various information about the clustering results, 
including the factor loadings, sample cluster assignments, and more. You can explore and 
interpret these results based on your research objectives. To access the H-matrix 
you can make use of the `$meanZ` operator. This matrix you can use to create the 
scatterplot. To access the cluster label information, you can make use of the `$clusters` 
operator. You can then use this information to annotate the scatterplot. 

```{r,eval=FALSE}
require (ggplot2)

# Extract the H (z) matrix
icluster_h <- as.data.frame(r_icluster$meanZ)

# get cluster labels
clus_lab <- as.factor(r_icluster$clusters)

# plot first two dimensions in scatterplot
# and annotate with cluster label
ggplot2::ggplot(icluster_h, ggplot2::aes(x=V1, y=V2, color=clus_lab)) +
ggplot2::geom_point() +
ggplot2::ggtitle("Scatter plot of iCluster+ factors")
```

--- TASK END

## Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have performed NMF using the `NMF::nmf` function.
The results of this are stored in a, object of class: NMFfit
called `r_nmf`. I want to make a label for the samples based on 
which latent variable influences it the most. How do I do this?

completion:
To create sample labels based on which latent variable (factor) influences 
each sample the most in the NMF analysis, you first need to extract the
H matrix. The H matrix contains the loadings for each sample across
the different dimensions. You can extract it from the NMFfit object using the 
`basis` function. You can then create a list containing the labels by seeing which
column holds the highest value for each sample. 

```{r,eval=FALSE}
# exctract the H matrix from the nmf run result
nmf_h <- NMF::basis(r_nmf)

# label based on which latent
# variable influences it the most

l_label <- c()
for (i in 1:nrow(nmf_h)){
    l_label <- c(l_label, which(max(nmf_h[i,])==nmf_h[i,]))
}

# provide sample name to label
names(l_label) <- rownames(nmf_h)
```

--- TASK END

## Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have performed NMF using the `NMF::nmf` function.
The results of this are stored in a, object of class: NMFfit
called `r_nmf`. I have extracted the H-matrix from this, which is
in a variable called nmf_h. This has the samples as rows, and the dimensions
as columns. I also have a list called 'l_label'. This has the samplenames
as `names`. It holds the predicted label based on which latent 
variable influences it the most. Furthermore, I also have a dataframe 
called `anno_dat`. This has samples represented as 
rows (samplenames are rownames). There is a column in this dataframe 
called `cms_label`, which holds the actual labels.
I would like to visualize the H-matrix in a heatmap, which shows the
predicted group, as well as the actual label. How do I do this?

completion:
To create a heatmap of the H-matrix showing both the predicted group labels 
(from the l_label list) and the actual labels (from the anno_dat dataframe), 
you can use the pheatmap function. You first need to ensure that the order
of samples is the same across all data. You can then create a new dataframe 
for annotation of the rows when using the `pheatmap` function. 
Here is how you do it:

```{r,eval=FALSE}
require (pheatmap)

# ensure ordering of samples is the same
# as annotation dataframe
nmf_h <- nmf_h [rownames(anno_dat), ,drop=FALSE]
l_label <- l_label [rownames(anno_dat)]


# create annotation dataframe
annotation <- anno_dat [, "cms_label", drop=FALSE]
colnames(annotation) <- "true_label"
annotation$predicted_label <- as.factor(l_label)

# plot heatmap:
pheatmap::pheatmap (nmf_h,
                    annotation_row = annotation,
                    main = "Heatmap NMF analysis and annotation")
```

--- TASK END

## Task `r i`,`r i<-i+1` taskType: data wrangling | ML/Stats | visualization, numData: 0, seeData: no, openEnded: no, tags: merge

prompt:
I have performed NMF using the `NMF::nmf` function.
The results of this are stored in a, object of class: NMFfit
called `r_nmf`. I have extracted the W-matrix from this, which is
in a variable called nmf_w. This has the dimensions as rows (of which
there are 2), and the features as columns. Column 1 to 500 holds features 
which are genes. I want to select genes associated preferentially with 
each dimension by their relative loading in the W matrix. I then
want to find GO terms enriched in each dimension, and plot the results. 
How do I do this?

completion:
To select genes associated preferentially with each dimension based on their 
relative loading in the W matrix, find gene sets enriched in each dimension, 
and plot the results, you first need to select the top genes for each 
dimension (in this case 2). Performing gene set enrichment analysis can 
be done using the `enrichR` tool. After finding enriched GO terms 
you can plot the results using the `plotEnrich` function. 

```{r,eval=FALSE}
require(enrichR)

# prepare lists for annotation
gene_label <- c()

# check which dimension holds max value
for (i in 1:500){
    gene_label <- c(gene_label,which (max(nmf_w[, i])==nmf_w[, i]))
}

# annotate with gene names
names(gene_label) <- colnames(nmf_w)[1:500]

# get names per factor
factor_1 <- names(gene_label)[gene_label==1]
factor_2 <- names(gene_label)[gene_label==2]

# find enriched GO terms per factor
go.factor.1 <- enrichR::enrichr(factor_1,
                                databases = c("GO_Biological_Process_2018")
                                )$GO_Biological_Process_2018

go.factor.2 <- enrichR::enrichr(factor_2,
                                databases = c("GO_Biological_Process_2018")
                                )$GO_Biological_Process_2018                                


#Plot enrichment table
plotEnrich(go.factor.1, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value")
plotEnrich(go.factor.2, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value")
```

--- TASK END
