---
title: "prompt/completion pairs single-cell_intro_chapter 3"
author: "authors"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: read, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have single cell RNA sequence (scRNA-seq) data set in file 
"GSE85241_cellsystems_dataset_4donors_updated.csv.gz" in .csv.gz format,
which I would like to read.
How can I do this?

completion:
Single cell RNA sequence (scRNA-seq) data set can be read by using read.delim()
function and stored as a matrix as follow.
 
```{r,eval=FALSE}

# read the scRNA-Seq dataset
mat <- as.matrix(read.delim("GSE85241_cellsystems_dataset_4donors_updated.csv.gz"))

# check dimensions (number of rows, number of columns) of the created matrix
dim(mat) 

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have single cell RNA sequence (scRNA-seq) data set in file 
"GSE85241_cellsystems_dataset_4donors_updated.csv.gz" in .csv.gz format,
which I would like to read but store only non-zero values.
How can I do this?

completion:
For reading single cell RNA sequence (scRNA-seq) data set and storing only 
non-zero values one can make use of readSparseCounts() function 
from the scuttle package as follow. The data are stored as a matrix.
 
```{r,eval=FALSE}

library(scuttle)

# read the scRNA-Seq dataset
sparse.mat <- readSparseCounts("GSE85241_cellsystems_dataset_4donors_updated.csv.gz")

# check dimensions (number of rows, number of columns) of the created matrix
dim(sparse.mat)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have single cell RNA sequence (scRNA-seq) data set in file 
"GSE61533_HTSEQ_count_results.xls" in Excel format. I would like to extract
the data into a matrix.
How can I do this?

completion:
Single cell RNA sequence (scRNA-seq) data set file in Excel format can be
read using read_excel() function from readxl package. After which the first
row containing the gene IDs can be extracted, data baring the first row 
converted to matrix and then the extracted gene IDs added to the matrix as 
row names, to have the gene names as row names and counts in each cell in each
column. 
 
```{r,eval=FALSE}

library(readxl)

# read the scRNA-Seq dataset
all.counts <- read_excel("GSE61533_HTSEQ_count_results.xls")

# extract IDs
gene.names <- all.counts$ID

# convert to matrix
all.counts <- as.matrix(all.counts[,-1])

# add extracted IDs as row names to the above created matrix
rownames(all.counts) <- gene.names

# check dimensions (number of rows, number of columns) of the created matrix
dim(all.counts)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have 10X genomics data in a directory which I would like to read. 
How can I do this?

completion:
10X genomics data can be read by supplying the directory path to 
read10xCounts() function from the DropletUtils package. The output is a
SingleCellExperiment object containing the matrix.

```{r,eval=FALSE}

library(DropletUtils)

# read the scRNA-Seq dataset
sce <- read10xCounts("tenx-2.1.0-pbmc4k/filtered_gene_bc_matrices/GRCh38")

# inspect the singlecellexperiment object
sce

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have multiple 10X genomics data in multiple directories. All the datasets 
have same annotations. I would like to read the datasets and combine them 
into a single object. 
How can I do this?

completion:
10X genomics data can be read by supplying the directory path to 
read10xCounts() function from the DropletUtils package. To combine multiple 
data set into a single object, their directory paths are supplied directly
to the read10xCounts() function. 

```{r,eval=FALSE}

library(DropletUtils)

# store scRNA-Seq dataset A path
dirA <- "tenx-2.1.0-pbmc4k/filtered_gene_bc_matrices/GRCh38"

# store scRNA-Seq dataset B path
dirB <- "tenx-2.1.0-pbmc4k-2/filtered_gene_bc_matrices/GRCh38"

# read the scRNA-Seq dataset A and B
sce <- read10xCounts(c(dirA, dirB))

# inspect the object (combined singlecellexperiment object)
sce

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
What is Hierarchical Data Format version 5 (HDF5) and how can data in H5AD 
format be read?

completion:
A family of scRNA-seq storage formats is based around Hierarchical Data Format 
version 5 (HDF5). These formats offer the ability to store, in the same file, 
both the expression values and associated gene and cell annotations. 
It has an additional advantage in that Bioconductor-based analyses can be 
performed without reading all of the data into R. This allows us to analyze 
very large datasets in the presence of limited computer memory.
One flavor of this approach is the H5AD format, which can be read as a 
SingleCellExperiment using the zellkonverter package, as demonstrated here 
with the example dataset from zellkonverter package using readH5AD() function.

```{r,eval=FALSE}

library(zellkonverter)

# get the data set
#demo <- system.file("extdata", "krumsiek11.h5ad", package = "zellkonverter")

# read the datset
sce <- readH5AD(demo)

# inspect the singlecellexperiment object
sce

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
What is Hierarchical Data Format version 5 (HDF5) and how can data in Loom file 
format be read?

completion:
A family of scRNA-seq storage formats is based around Hierarchical Data Format 
version 5 (HDF5). These formats offer the ability to store, in the same file, 
both the expression values and associated gene and cell annotations. 
It has an additional advantage in that Bioconductor-based analyses can be 
performed without reading all of the data into R. This allows us to analyze 
very large datasets in the presence of limited computer memory.
One flavor of this approach is the Loom file format, which can be read using 
the LoomExperiment package as a SingleCellLoomExperiment, which is effectively 
a plug-and-play equivalent to the SingleCellExperiment. Demonstrated here 
with the example dataset from LoopExperiment package using import() function.


```{r,eval=FALSE}

library(LoomExperiment)

# get the data set
#demo <- system.file("extdata", "L1_DRG_20_example.loom", package = "LoomExperiment")

# read the datset
scle <- import(demo, type="SingleCellLoomExperiment")

# inspect the singlecellloomexperiment object
scle

```