---
title: "prompt/completion pairs single-cell_intro_chapter 4"
author: "authors"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA sequence (scRNA-seq) count matrix data in file 
"2261a12c52_counts_Calero_20160113.tsv" in .tsv format. Using this data
I would like to create a rudimentary SingleCellExperiment object that contains
information about the genes and the samples.
How can I do this?

completion:
To create a rudimentary SingleCellExperiment object, read the count matrix
data using the read.delim() function. Remove the gene length column from
the data frame and then convert it to a matrix. Using the created matrix 
construct the SingleCellExperiment object using SingleCellExperiment() function 
from SingleCellExperiment package. Note that we provide our data as a named 
list where each entry of the list is a matrix - in this case, named "counts".
 
```{r,eval=FALSE}

library(SingleCellExperiment)

# read the scRNA-Seq dataset
mat <- read.delim("raw_data/2261a12c52_counts_Calero_20160113.tsv", header=TRUE, row.names=1, check.names=FALSE)

# only considering endogenous genes for now
spike.mat <- mat[grepl("^ERCC-", rownames(mat)),] 
mat <- mat[grepl("^ENSMUSG", rownames(mat)),] 

# splitting off the gene length column
gene.length <- mat[,1]
mat <- as.matrix(mat[,-1]) 

# construct the SingleCellExperiment object
sce <- SingleCellExperiment(assays = list(counts = mat))

# inspect the singlecellexperiment object
sce

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA sequence (scRNA-seq) count matrix data in file 
"2261a12c52_counts_Calero_20160113.tsv" in .tsv format. Using this data
I would like to create a rudimentary SingleCellExperiment object that contains
information about the genes and the samples. I would then like to access the
count data from the created SingleCellExperiment object.
How can I do this?

completion:
To create a rudimentary SingleCellExperiment object, read the count matrix
data using the read.delim() function. Remove the gene length column from
the data frame and then convert it to a matrix. Using the created matrix 
construct the SingleCellExperiment object using SingleCellExperiment() function 
from SingleCellExperiment package. Note that we provide our data as a named 
list where each entry of the list is a matrix - in this case, named "counts".
To access the count data from the SingleCellExperiment object we can use 
assay() function from SingleCellExperiment package, which takes first argument 
as the SingleCellExperiment object and second argument as the name of the 
assay - here "counts". Similarly, it can be used to access any assay from 
the SingleCellExperiment object.
 
```{r,eval=FALSE}

library(SingleCellExperiment)

# read the scRNA-Seq dataset
mat <- read.delim("raw_data/2261a12c52_counts_Calero_20160113.tsv", header=TRUE, row.names=1, check.names=FALSE)

# only considering endogenous genes for now
spike.mat <- mat[grepl("^ERCC-", rownames(mat)),] 
mat <- mat[grepl("^ENSMUSG", rownames(mat)),] 

# splitting off the gene length column
gene.length <- mat[,1]
mat <- as.matrix(mat[,-1]) 

# construct the SingleCellExperiment object
sce <- SingleCellExperiment(assays = list(counts = mat))

# inspect the singlecellexperiment object
sce

# access count data
assay(sce, "counts")

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA sequence (scRNA-seq) count matrix data in
"2261a12c52_counts_Calero_20160113.tsv" file in .tsv format. Using this data
I would like to create a rudimentary SingleCellExperiment object that contains
information about the genes and the samples. In addition to the raw count data 
I would also like to have normalized data in the same SingleCellExperiment 
object.
How can I do this?

completion:
To create a rudimentary SingleCellExperiment object, read the count matrix
data using the read.delim() function. Remove the gene length column from
the data frame and then convert it to a matrix. Using the created matrix 
construct the SingleCellExperiment object using SingleCellExperiment() function 
from SingleCellExperiment package. Note that we provide our data as a named 
list where each entry of the list is a matrix - in this case, named "counts".
To compute a log-transformed normalized expression matrix use logNormCounts()
function from scuttle package, and store it as another assay. By reassigning 
the results of logNormCounts() function back to "sce" we overwrite the previous
"sce" and store both original and normalised data in the same 
SingleCellExpression object "sce".
 
```{r,eval=FALSE}

library(SingleCellExperiment)
library(scuttle)

# read the scRNA-Seq dataset
mat <- read.delim("raw_data/2261a12c52_counts_Calero_20160113.tsv", header=TRUE, row.names=1, check.names=FALSE)

# only considering endogenous genes for now
spike.mat <- mat[grepl("^ERCC-", rownames(mat)),] 
mat <- mat[grepl("^ENSMUSG", rownames(mat)),] 

# splitting off the gene length column
gene.length <- mat[,1]
mat <- as.matrix(mat[,-1]) 

# construct the SingleCellExperiment object
sce <- SingleCellExperiment(assays = list(counts = mat))

# compute a log-transformed normalized expression matrix and store it as another assay
sce <- logNormCounts(sce)

# inspect the singlecellexperiment object
sce

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA sequence (scRNA-seq) count matrix data in 
"2261a12c52_counts_Calero_20160113.tsv" file in .tsv format. Using this data
I would like to create a rudimentary SingleCellExperiment object that contains
information about the genes and the samples. In addition to the raw count data 
I would also like to have normalized data in the same SingleCellExperiment 
object. Moreover I would like to create a matrix which adds 100 to all the 
counts of the created SingleCellExperiment object, which I would like to use
as a new assay and add to the SingleCellExperiment object.
How can I do this?

completion:
To create a rudimentary SingleCellExperiment object, read the count matrix
data using the read.delim() function. Remove the gene length column from
the data frame and then convert it to a matrix. Using the created matrix 
construct the SingleCellExperiment object using SingleCellExperiment() function 
from SingleCellExperiment package. Note that we provide our data as a named 
list where each entry of the list is a matrix - in this case, named "counts".
To compute a log-transformed normalized expression matrix use logNormCounts()
function from scuttle package, and store it as another assay. By reassigning 
the results of logNormCounts() function back to "sce" we overwrite the previous
"sce" and store both original and normalised data in the same 
SingleCellExpression object "sce". To create a matrix which adds 100 to all 
the counts of the "sce" object use counts() function from SingleCellExperiment 
package. The created matrix can be added to the "sce" object using assay() 
function, which can be confirmed by retrieving all the available assays in the 
"sce" object by assays() getter, from the SingleCellExperiment package. 
 
```{r,eval=FALSE}

library(SingleCellExperiment)
library(scuttle)

# read the scRNA-Seq dataset
mat <- read.delim("raw_data/2261a12c52_counts_Calero_20160113.tsv", header=TRUE, row.names=1, check.names=FALSE)

# only considering endogenous genes for now
spike.mat <- mat[grepl("^ERCC-", rownames(mat)),] 
mat <- mat[grepl("^ENSMUSG", rownames(mat)),] 

# splitting off the gene length column
gene.length <- mat[,1]
mat <- as.matrix(mat[,-1]) 

# construct the SingleCellExperiment object
sce <- SingleCellExperiment(assays = list(counts = mat))

# compute a log-transformed normalized expression matrix and store it as another assay
sce <- logNormCounts(sce)

# create a matrix from the SingleCellExperiment object
counts_100 <- counts(sce) + 100

# assign a new entry to assays slot
assay(sce, "counts_100") <- counts_100 

# inspect the singlecellexperiment object for new assay addition
assays(sce) 

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object with three assays. I would
like to only keep first two assays.
How can I do this?

completion:
SingleCellExperiment package can be used for creating SingleCellExperiment 
objects and for working with them. Here using assays() getter, we first 
retrieve all the available assays, then we subset "sce" object with first
two assays. Here setter of the same name can be used to modify the set of 
available assays.

```{r,eval=FALSE}

library(SingleCellExperiment)

# inspect the singlecellexperiment object for number of assays 
assays(sce) 

# only keeping the first two assays
assays(sce) <- assays(sce)[1:2]

# inspect the singlecellexperiment object
sce

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object with number of assays. I would
like to get the names of the assays. 
How can I do this?

completion:
SingleCellExperiment package can be used for creating SingleCellExperiment 
objects and for working with them. Here using assayNames() function we can 
get and set the names of the assays. We can also use names() function, which
is less efficient. 

```{r,eval=FALSE}

library(SingleCellExperiment)

# get the names of the assays
assayNames(sce)

# same result, but slightly less efficient
names(assays(sce)) 

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I am metadata of my single cell RNA sequence (scRNA-seq) experiment in
"37064fb2008_E-MTAB-5522.sdrf.txt" file in .txt format. How can I read it?

completion:
Data in .txt format can be read using read.delim() function.

```{r,eval=FALSE}

# read metadata in .txt file format
coldata <- read.delim("raw_data/37064fb2008_E-MTAB-5522.sdrf.txt", check.names=FALSE)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I am metadata of my single cell RNA sequence (scRNA-seq) experiment in 
"37064fb2008_E-MTAB-5522.sdrf.txt" file in .txt format. I would like to 
read it and add it to my already created SingleCellExperiment object
of the single cell data.
How can I do this?

completion:
We can read the metadata in .txt format using read.delim() function. Metadata
describes the columns of our primary data, e.g., the samples or cells of our 
experiment. This is stored in the colData slot, a DataFrame object where 
rows correspond to cells and columns correspond to metadata fields, 
e.g., batch of origin, treatment condition. From the read data only keep
the cells involved in count matrix of already created SingleCellExperiment 
object. Filter the columns of interest and create a data frame using 
Dataframe() function. Then add the data frame to an existing 
SingleCellExperiment object using colData() function from 
SingleCellExperiment package.

```{r,eval=FALSE}

library(SingleCellExperiment)

# read metadata in .txt file format
coldata <- read.delim("raw_data/37064fb2008_E-MTAB-5522.sdrf.txt", check.names=FALSE)

# only keeping the cells involved in the count matrix in 'mat', 
# -used to create SingleCellExperiment object
coldata <- coldata[coldata[,"Derived Array Data File"]=="counts_Calero_20160113.tsv",]

# only keeping interesting columns, and setting the library names as the row names
coldata <- DataFrame(
    genotype=coldata[,"Characteristics[genotype]"],
    phenotype=coldata[,"Characteristics[phenotype]"],
    spike_in=coldata[,"Factor Value[spike-in addition]"],
    row.names=coldata[,"Source Name"]
)

# inspect the coldata
coldata

# create/previously created SingleCellExperiment object
sce <- SingleCellExperiment(list(counts=mat))

# add coldata to an existing SingleCellExperiment object
colData(sce) <- coldata

# inspect the SingleCellExperiment object
sce

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I am metadata of my single cell RNA sequence (scRNA-seq) experiment in 
"37064fb2008_E-MTAB-5522.sdrf.txt" file in .txt format. I would like to 
read it, and use it and the count matrix of my single cell data, to create 
a SingleCellExperiment object.
How can I do this?

completion:
We can read the metadata in .txt format using read.delim() function. Metadata
describes the columns of our primary data, e.g., the samples or cells of our 
experiment. This is stored in the colData slot, a DataFrame object where 
rows correspond to cells and columns correspond to metadata fields, 
e.g., batch of origin, treatment condition. From the read data only keep
the cells involved in count matrix of already created SingleCellExperiment 
object. Filter the columns of interest and using the Dataframe() function 
create a data frame that we will use along with the count matrix of  
the single cell data to create a SingleCellExperiment object using 
SingleCellExperiment() constructor from the SingleCellExperiment package.

```{r,eval=FALSE}

library(SingleCellExperiment)

# read metadata in .txt file format
coldata <- read.delim("raw_data/37064fb2008_E-MTAB-5522.sdrf.txt", check.names=FALSE)

# only keeping the cells involved in the count matrix in 'mat', 
# -used to create SingleCellExperiment object
coldata <- coldata[coldata[,"Derived Array Data File"]=="counts_Calero_20160113.tsv",]

# only keeping interesting columns, and setting the library names as the row names
coldata <- DataFrame(
    genotype=coldata[,"Characteristics[genotype]"],
    phenotype=coldata[,"Characteristics[phenotype]"],
    spike_in=coldata[,"Factor Value[spike-in addition]"],
    row.names=coldata[,"Source Name"]
)

# inspect coldata
coldata

# create a SingleCellExperiment object
sce <- SingleCellExperiment(assays = list(counts=mat), colData=coldata)

# inspect the SingleCellExperiment object
sce

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data and I would like to 
access the column data of my object.
How can I do this?

completion:
Column data of a SingleCellExperiment object can be accessed using 
colData() function from the SingleCellExperiment package.

```{r,eval=FALSE}

library(SingleCellExperiment)

# access column data
colData(sce)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data and I would like to 
access a specific column of the colData slot of my object.
How can I do this?

completion:
We can extract a single field/column of a SingleCellExperiment
object's colData by using the $ shortcut.

```{r,eval=FALSE}

library(SingleCellExperiment)

# access genotype column of colData
head(sce$genotype)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
add quality control metrics to the colData slot of my object.
How can I do this?

completion:
Some functions automatically add column metadata by returning a 
SingleCellExperiment with extra fields in the colData slot. For example, 
the scuttle package contains the addPerCellQC() function that appends a number 
of quality control metrics to the colData.

```{r,eval=FALSE}

library(SingleCellExperiment)
library(scuttle)

# append quality control metrics to the colData
sce <- addPerCellQC(sce)

# inspect the singlecellexperiment object's column data
colData(sce)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
store feature-level annotation in the rowData slot of my object.
How can I do this?

completion:
We store feature-level annotation in the rowData slot, a DataFrame where 
each row corresponds to a gene and contains annotations like the transcript 
length or gene symbol. We can manually get and set this with the rowData() 
function from SingleCellExperiment package.

```{r,eval=FALSE}

library(SingleCellExperiment)

# store gene length in the rowData slot
rowData(sce)$Length <- gene.length

# inspect the singlecellexperiment object's row data
rowData(sce)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
extract feature-level data from my object and store them in the rowData slot.
How can I do this?

completion:
addPerFeatureQC() function from the scuttle package returns a 
SingleCellExperiment object with the rowData populated with relevant bits of 
information.

```{r,eval=FALSE}

library(SingleCellExperiment)
library(scuttle)

# store feature-level data in the rowData slot
sce <- addPerFeatureQC(sce)

# inspect the singlecellexperiment object's row data
rowData(sce)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
add genomic coordinates in it in the form of GRanges or GRangesList.
How can I do this?

completion:
In a SingleCellExperiment object, there is a special rowRanges slot to hold 
genomic coordinates in the form of a GRanges or GRangesList. This stores 
describes the chromosome, start, and end coordinates of the features 
(genes, genomic regions) that can be easily queryed and manipulated via the 
GenomicRanges framework. The manner with which we populate the rowRanges 
depends on the organism and annotation used during alignment and quantification. 
For example, when we have Ensembl identifiers, we might use rtracklayer 
package to load in a GRanges from an GTF file containing the Ensembl annotation 
used in this dataset using import() function.

```{r,eval=FALSE}

library(SingleCellExperiment)
library(rtracklayer)

# load GRanges from the file 
gene.data <- rtracklayer::import("raw_data/22a336d3609_Mus_musculus.GRCm38.82.gtf.gz")

# cleaning up the object
gene.data <- gene.data[gene.data$type=="gene"]
names(gene.data) <- gene.data$gene_id
is.gene.related <- grep("gene_", colnames(mcols(gene.data)))
mcols(gene.data) <- mcols(gene.data)[,is.gene.related]

# creating rowRanges
rowRanges(sce) <- gene.data[rownames(sce)]

# inspecting rowRanges
rowRanges(sce)[1:10,]

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I have 
results/annotations that do not fit in the any of the "colData, rowData, 
rowRanges" slots. Nevertheless, I would like to add them in my object.
How can I do this?

completion:
Some analyses contain results or annotations that do not fit into the 
aforementioned, "colData, rowData, rowRanges", slots, e.g., study metadata. 
This can be stored in the metadata slot, a named list of arbitrary objects, 
by simply appending them to the metadata slot using metadata() function from
SingleCellExperiment package, for our analysis.
The main disadvantage of storing content in the metadata() is that it will 
not be synchronized with the rows or columns when subsetting or combining. 
Thus, if an annotation field is related to the rows or columns, it should 
be stored in the rowData() or colData() instead.

```{r,eval=FALSE}

library(SingleCellExperiment)

# vector containing gene of interest
my_genes <- c("gene_1", "gene_5")

# adding them to the metadata slot
metadata(sce) <- list(favorite_genes = my_genes)

# inspecting metadata slot
metadata(sce)

## or simply appending them using $
# vector containing gene of interest
your_genes <- c("gene_4", "gene_8")

# adding them to the metadata slot using $
metadata(sce)$your_genes <- your_genes

# inspecting metadata slot
metadata(sce)

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
subset my object to contain the first 10 cells along with the 
associated annotation.
How can I do this?

completion:
One of the major advantages of using the SingleCellExperiment is that 
operations on the rows or columns of the expression data are synchronized 
with the associated annotation. So when we subset our SingleCellExperiment 
object to the first 10 cells, both the count matrix and the colData() are 
automatically subsetted.

```{r,eval=FALSE}

library(SingleCellExperiment)

# subset the object to the firsts 10 cells
first.10 <- sce[,1:10]

# count the columns in the subsetted object (only 10 columns)
ncol(counts(first.10)) 

# inspect coldata of the subsetted object (only 10 rows)
colData(first.10) 

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
subset my object to contain only wild-type cells along with the 
associated annotation.
How can I do this?

completion:
One of the major advantages of using the SingleCellExperiment is that 
operations on the rows or columns of the expression data are synchronized 
with the associated annotation. We can subset our SingleCellExperiment object 
to contain only wild-type cells based on its column data entries using colData()
function from SingleCellExperiment package.

```{r,eval=FALSE}

library(SingleCellExperiment)

# subset the object to contain only wild-type cells
wt.only <- sce[, sce$phenotype == "wild type phenotype"]

# count the columns in the subsetted object (wild-type only)
ncol(counts(wt.only))

# inspect coldata of the subsetted object
colData(wt.only)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
subset my object to contain only protein-coding genes along with the 
associated annotation.
How can I do this?

completion:
One of the major advantages of using the SingleCellExperiment is that 
operations on the rows or columns of the expression data are synchronized 
with the associated annotation. We can subset our SingleCellExperiment object 
to contain only protein-coding genes based on its cell data entries using 
rowData() function from SingleCellExperiment package.

```{r,eval=FALSE}

library(SingleCellExperiment)

# subset the object to contain only protein-coding genes
coding.only <- sce[rowData(sce)$gene_biotype == "protein_coding",]

# count the rows in the subsetted object (protein-coding genes only)
nrow(counts(coding.only))

# inspect rowdata of the subsetted object
rowData(coding.only)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have multiple SingleCellExperiment objects of my single cell data. They have 
the same row annotation values, using which I would like to combine them.
How can I do this?

completion:
One of the major advantages of using the SingleCellExperiment is that 
operations on the rows or columns of the expression data are synchronized 
with the associated annotation. If we were to combine multiple 
SingleCellExperiment objects, the class would take care of combining both 
the expression values and the associated annotation in a coherent manner. 
Assuming that all objects involved have the same row annotation values 
and compatible column annotation fields, we can use cbind() function 
to combine objects by column.

```{r,eval=FALSE}

library(SingleCellExperiment)

# combining multiple objects with same row annotation
sce2 <- cbind(sce, sce)

# count the columns (twice as many)
ncol(counts(sce2)) 

# inspecting column data (twice as many rows)
colData(sce2) 

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have multiple SingleCellExperiment objects of my single cell data. They have 
the same column annotation values, using which I would like to combine them.
How can I do this?

completion:
One of the major advantages of using the SingleCellExperiment is that 
operations on the rows or columns of the expression data are synchronized 
with the associated annotation. If we were to combine multiple 
SingleCellExperiment objects, the class would take care of combining both 
the expression values and the associated annotation in a coherent manner. 
Assuming that all objects involved have the same column annotation values 
and compatible row annotation fields, we can use rbind() function 
to combine objects by row.

```{r,eval=FALSE}

library(SingleCellExperiment)

# combining multiple objects with same column annotation
sce2 <- rbind(sce, sce)

# count the rows (twice as many)
nrow(counts(sce2))

# inspecting row data (twice as many rows)
rowData(sce2) 

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like
to perform PCA on it, store the results, and access the reduced dimensions. 
How can I do this?

completion:
The reducedDims slot is specially designed to store reduced dimensionality 
representations of the primary data obtained by methods such as PCA and  
t-SNE. This slot contains a list of numeric matrices of low-reduced 
representations of the primary data, where the rows represent the columns 
of the primary data (i.e., cells), and columns represent the dimensions. 
As this slot holds a list, we can store multiple PCA/t-SNE/etc results 
for the same dataset. To calculate a PCA representation of the data we first
normalize the data and store it in the object, if not there already, using 
logNormCounts() function, and then perform PCA on the data using runPCA() 
function from scater package. The reduced dimensions can be accessed from 
the reducedDim slot using dim() function, and the names of all the entries in 
the reducedDim slot can be accessed via reducedDims() accessor.

```{r,eval=FALSE}

library(SingleCellExperiment)
library(scater)

# normalized the data
sce <- scater::logNormCounts(sce)

# run PCA
sce <- scater::runPCA(sce)

# count the dimensions
dim(reducedDim(sce, "PCA"))

# get the names of all entries in reducedDims slot
reducedDims(sce)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: read | data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have data for a distinct set of features but the same set of samples/cells
as my SingleCellExperiment object of my single cell data. I would like to 
store them in my SingleCellExperiment object as an alternative experiment.
How can I do this?

completion:
The SingleCellExperiment class provides the concept of “alternative experiments” 
where we have data for a distinct set of features but the same set of 
samples/cells. To store distinct feature sets as an alternative
experiment, we first create a separate SummarizedExperiment object using 
SummarizedExperiment() function from SummarizedExperiment package. Then we 
store this SummarizedExperiment object in our SingleCellExperiment object 
via the altExp() setter. 

```{r,eval=FALSE}

library(SingleCellExperiment)
library(SummarizedExperiment)

# read the scRNA-Seq dataset
mat <- read.delim("raw_data/2261a12c52_counts_Calero_20160113.tsv", header=TRUE, row.names=1, check.names=FALSE)

# only considering endogenous genes for now
spike.mat <- mat[grepl("^ERCC-", rownames(mat)),] 

# create SummarizedExperiment object (-1 to get rid of the first gene length column)
spike_se <- SummarizedExperiment(list(counts=spike.mat[,-1]))

# store SummarizedExperiment object in SingleCellExperiment object
altExp(sce, "spike") <- spike_se

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data, that may also
contain alternative experiments. I would like to retrieve all of the 
available alternative experiments. 
How can I do this?

completion:
The SingleCellExperiment class provides the concept of “alternative experiments” 
where we have data for a distinct set of features but the same set of 
samples/cells. Alternative experiments can be stored in a SingleCellExperiment 
object via the altExp() setter. And all the available alternative experiments 
can be retrieve with altExps() getter from SingleCellExperiment package.

```{r,eval=FALSE}

library(SingleCellExperiment)

# retrieve all of the available alternative Experiments
altExps(sce)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
get a numeric vector of per-cell scaling factors used for normalization. 
How can I do this?

completion:
Scaling factors used for normalization are typically automatically added by 
normalization functions. An example is deconvolution-based size factors 
added by computeSumFactors() function from scran package. Or per-cell 
size factors from the library sizes by librarySizeFactors() function 
from scater package. Per-cell scaling factors used for normalization
can be obtained by sizeFactors() function.

```{r,eval=FALSE}

library(SingleCellExperiment)
library(scran)
library(scater)

# add scaling factors
sce <- scran::computeSumFactors(sce)

# get scaling factors
summary(sizeFactors(sce))

# SECOND METHOD; add scaling factors 
sizeFactors(sce) <- scater::librarySizeFactors(sce)

# get scaling factors
summary(sizeFactors(sce))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also 
contains PCA representation. I would like to set a factor of per-cell labels 
corresponding to groupings assigned by unsupervised clustering. 
How can I do this?

completion:
The colLabels() function from SingleCellExperiment package allows us to get or 
set a factor of per-cell labels corresponding to groupings assigned by 
unsupervised clustering, or predicted cell type identities from classification 
algorithms.

```{r,eval=FALSE}

library(SingleCellExperiment)
library(scran)

# assigned labels based on clusters
colLabels(sce) <- scran::clusterCells(sce, use.dimred="PCA")

# table of labels
table(colLabels(sce))

```
