---
title: "prompt/completion pairs single-cell_basic_chapter 1"
author: "authors"
date: "2023-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells based on their 
expression profiles.
How do I do this?

completion:
To calculate quality control metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and calculate
the quality control metrics on them. The created quality control metrics
data frame contains a number of columns. The sum column contains 
the total count for each cell and the detected column contains the number of 
detected genes. The subsets_Mito_percent column contains the percentage of 
reads mapped to mitochondrial transcripts. Finally, the altexps_ERCC_percent 
column contains the percentage of reads mapped to ERCC transcripts.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# total count per cell
summary(df$sum)

# number of detected genes
summary(df$detected)

# percentage of reads mapped to mitochondrial transcripts
summary(df$subsets_Mito_percent)

# percentage of reads mapped to ERCC transcripts
summary(df$altexps_ERCC_percent)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells based on their 
expression profiles and add them to the colData slot of my SingleCellExperiment 
object.
How do I do this?

completion:
To calculate quality control metrics for each cell and append them to the 
colData slot of our SingleCellExperiment object, we can use the
addPerCellQCMetrics() function from scuttle package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the quality control metrics on them and append the calculated
metrics at the same time using addPerCellQCMetrics() function. The sum column 
contains the total count for each cell and the detected column contains the 
number of detected genes. The subsets_Mito_percent column contains the 
percentage of reads mapped to mitochondrial transcripts. Finally, the 
altexps_ERCC_percent column contains the percentage of reads mapped to ERCC 
transcripts.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
sce.416b <- addPerCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# inspect column names to confirm new column addition
colnames(colData(sce.416b))

```
--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells based on fixed 
thresholds. The thresholds for low-quality cells being library sizes below 
100,000 reads; express fewer than 5,000 genes; have spike-in proportions above 
10%; or have mitochondrial proportions above 10%.
How do I do this?

completion:
To calculate quality control metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the quality control metrics on them. The created quality control 
metrics data frame contains a number of columns. The sum column contains 
the total count for each cell and the detected column contains the number of 
detected genes. The subsets_Mito_percent column contains the percentage of 
reads mapped to mitochondrial transcripts. Finally, the altexps_ERCC_percent 
column contains the percentage of reads mapped to ERCC transcripts. By applying
thresholds on the respective data columns of the quality control metrics data
frame we can identify the low-quality cells.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# total count per cell
summary(df$sum)

# number of detected genes
summary(df$detected)

# percentage of reads mapped to mitochondrial transcripts
summary(df$subsets_Mito_percent)

# percentage of reads mapped to ERCC transcripts
summary(df$altexps_ERCC_percent)

# library size threshold
qc.lib <- df$sum < 1e5

# number of expressed genes threshold
qc.nexprs <- df$detected < 5e3

# spike-in proportions threshold
qc.spike <- df$altexps_ERCC_percent > 10

# mitochondrial proportions threshold
qc.mito <- df$subsets_Mito_percent > 10

# remove the cells based on the threshold
discard <- qc.lib | qc.nexprs | qc.spike | qc.mito

# summarize the number of cells removed for each reason
DataFrame(LibSize=sum(qc.lib), NExprs=sum(qc.nexprs),
    SpikeProp=sum(qc.spike), MitoProp=sum(qc.mito), Total=sum(discard))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells assuming that 
most of my dataset consists of high-quality cells. 
How do I do this?

completion:
To calculate quality control (QC) metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the QC metrics on them. The created QC metrics data frame contains 
a number of columns. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. Assuming that most of the 
dataset consists of high-quality cells, we identify cells that are outliers for 
the various QC metrics, based on the median absolute deviation (MAD) from the 
median value of each metric across all cells. By default, we consider a value 
to be an outlier if it is more than 3 MADs from the median in the “problematic” 
direction. We use perCellQCFilters() function from the scater package for this.
This function identifies cells with log-transformed library sizes that are more 
than 3 MADs below the median as well as the proportion-based metrics specified 
in the sub.fields = arguments. A cell that is an outlier for any of these 
metrics is considered to be of low quality.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# total count per cell
summary(df$sum)

# number of detected genes
summary(df$detected)

# percentage of reads mapped to mitochondrial transcripts
summary(df$subsets_Mito_percent)

# percentage of reads mapped to ERCC transcripts
summary(df$altexps_ERCC_percent)

# identifying outliers for QC metrics
reasons <- perCellQCFilters(df, 
    sub.fields=c("subsets_Mito_percent", "altexps_ERCC_percent"))

# low-quality cells based on various QC metrics
colSums(as.matrix(reasons))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells assuming that 
most of my dataset consists of high-quality cells. And at the end I would like
to get the number of cells that are not and are of low-quality respectively.
How do I do this?

completion:
To calculate quality control (QC) metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the QC metrics on them. The created QC metrics data frame contains 
a number of columns. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. Assuming that most of the 
dataset consists of high-quality cells, we identify cells that are outliers for 
the various QC metrics, based on the median absolute deviation (MAD) from the 
median value of each metric across all cells. By default, we consider a value 
to be an outlier if it is more than 3 MADs from the median in the “problematic” 
direction. We use perCellQCFilters() function from the scater package for this.
This function identifies cells with log-transformed library sizes that are more 
than 3 MADs below the median as well as the proportion-based metrics specified 
in the sub.fields = arguments. A cell that is an outlier for any of these 
metrics is considered to be of low quality and discarded. This is captured in 
the discard column.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# total count per cell
summary(df$sum)

# number of detected genes
summary(df$detected)

# percentage of reads mapped to mitochondrial transcripts
summary(df$subsets_Mito_percent)

# percentage of reads mapped to ERCC transcripts
summary(df$altexps_ERCC_percent)

# identifying outliers for QC metrics
reasons <- perCellQCFilters(df, 
    sub.fields=c("subsets_Mito_percent", "altexps_ERCC_percent"))

# summary showing number of low-quality cells (as TRUE)
summary(reasons$discard)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells assuming that 
most of my dataset consists of high-quality cells. Moreover, I would like to 
extract the automatically selected thresholds for the quality control metrics.
How do I do this?

completion:
To calculate quality control (QC) metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the QC metrics on them. The created QC metrics data frame contains 
a number of columns. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. Assuming that most of the 
dataset consists of high-quality cells, we identify cells that are outliers for 
the various QC metrics, based on the median absolute deviation (MAD) from the 
median value of each metric across all cells. By default, we consider a value 
to be an outlier if it is more than 3 MADs from the median in the “problematic” 
direction. We use perCellQCFilters() function from the scuttle package for this.
This function identifies cells with log-transformed library sizes that are more 
than 3 MADs below the median as well as the proportion-based metrics specified 
in the sub.fields = arguments. A cell that is an outlier for any of these 
metrics is considered to be of low quality. The filter thresholds used for the 
QC metrics can be extracted from the attributes of each of the logical vectors
using attr() function.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# total count per cell
summary(df$sum)

# number of detected genes
summary(df$detected)

# percentage of reads mapped to mitochondrial transcripts
summary(df$subsets_Mito_percent)

# percentage of reads mapped to ERCC transcripts
summary(df$altexps_ERCC_percent)

# identifying outliers for various QC metrics
reasons <- perCellQCFilters(df, 
    sub.fields=c("subsets_Mito_percent", "altexps_ERCC_percent"))

# extract filter thresholds
attr(reasons$low_lib_size, "thresholds")

# extract filter thresholds
attr(reasons$low_n_features, "thresholds")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control (QC) metrics to identify outliers in high-dimensional 
space based on the QC metrics for each cell. 
How do I do this?

completion:
To calculate quality control (QC) metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the QC metrics on them. The created QC metrics data frame contains 
a number of columns. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. Assuming that most of the 
dataset consists of high-quality cells, we identify cells that are outliers for 
the various QC metrics, based on the median absolute deviation (MAD) from the 
median value of each metric across all cells. By default, we consider a value 
to be an outlier if it is more than 3 MADs from the median in the “problematic” 
direction. A cell that is an outlier for any of these metrics is considered to 
be of low quality. So to identify them we proceed with log-transforming the sum 
and the detected columns, and creating a matrix with the four data columns; 
"sum", "detected", "subsets_Mito_percent" and "altexps_ERCC_percent". We then 
use methods from robustbase package on the created matrix to quantify the 
“outlyingness” of each cells based on their QC metrics, and then use 
isOutlier() function from scuttle package to identify low-quality cells that 
exhibit unusually high levels of outlyingness, which can be checked by 
summary() function. This and related approaches like PCA-based outlier detection 
and support vector machines can provide more power to distinguish low-quality 
cells from high-quality counterparts as they can exploit patterns across many
QC metrics. However, this comes at some cost to interpretability, as the reason 
for removing a given cell may not always be obvious.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)
library(robustbase)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# log-transformation and matrix creation
stats <- cbind(log10(df$sum), log10(df$detected),
    df$subsets_Mito_percent, df$altexps_ERCC_percent)

# quantifying outlyingness 
outlying <- adjOutlyingness(stats, only.outlyingness = TRUE)

# outliers identification
multi.outlier <- isOutlier(outlying, type = "higher")

# summary showing number of low-quality cells (as TRUE)
summary(multi.outlier)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control (QC) metrics to identify outliers and to create 
diagnostic plots for inspecting the QC metric distribution.
How can I do this?

completion:
To calculate quality control (QC) metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them to 
identify interesting feature subsets, here mitochondrial genes, and calculate 
the QC metrics on them. The created QC metrics data frame "df" contains a 
number of columns. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. Assuming that most of the 
dataset consists of high-quality cells, we identify cells that are outliers for 
the various QC metrics, based on the median absolute deviation (MAD) from the 
median value of each metric across all cells. By default, we consider a value 
to be an outlier if it is more than 3 MADs from the median in the “problematic” 
direction. We use perCellQCFilters() function from the scuttle package for this.
This function identifies cells with log-transformed library sizes that are more 
than 3 MADs below the median as well as the proportion-based metrics specified 
in the sub.fields = arguments. A cell that is an outlier for any of these 
metrics is considered to be of low quality and discarded. This is captured in 
the discard column of the created "reasons" data frame consisting of identified
outliers for various QC metrics. Using cbind() function we add the previously 
created QC metrics data frame "df" to the colData slot of the 
SingleCellExperiment object. Rename elements in phenotype column of the object. 
Create a column "discard" in the colData slot of the object that contains 
discarded cell info from the "reasons" data frame. Now using the updated
object "sce.416b" we plot the distribution of QC metrics for each batch and 
phenotype by making use of plotColData() function from scater package.
It is good practice to inspect the distributions of QC metrics to identify 
possible problems. In the most ideal case, we would see normal distributions 
that would justify the 3 MAD threshold used in outlier detection. A large 
proportion of cells in another mode suggests that the QC metrics might be 
correlated with some biological state, potentially leading to the loss of 
distinct cell types during filtering; or that there were inconsistencies with 
library preparation for a subset of cells, a not-uncommon phenomenon in 
plate-based protocols.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)
library(scater)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# identify outliers for various QC metrics
reasons <- perCellQCFilters(df, 
    sub.fields=c("subsets_Mito_percent", "altexps_ERCC_percent"))

# add "df" to the colData slot of the SingleCellExperiment object
colData(sce.416b) <- cbind(colData(sce.416b), df)

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# rename elements in phenotype column
sce.416b$phenotype <- ifelse(grepl("induced", sce.416b$phenotype),
    "induced", "wild type")

# add discarded cells info to the colData slot of the object
sce.416b$discard <- reasons$discard

# plot the distribution of QC metrics
gridExtra::grid.arrange(
    plotColData(sce.416b, x="block", y="sum", colour_by="discard",
        other_fields="phenotype") + facet_wrap(~phenotype) + 
        scale_y_log10() + ggtitle("Total count"),
    plotColData(sce.416b, x="block", y="detected", colour_by="discard", 
        other_fields="phenotype") + facet_wrap(~phenotype) + 
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(sce.416b, x="block", y="subsets_Mito_percent", 
        colour_by="discard", other_fields="phenotype") + 
        facet_wrap(~phenotype) + ggtitle("Mito percent"),
    plotColData(sce.416b, x="block", y="altexps_ERCC_percent", 
        colour_by="discard", other_fields="phenotype") + 
        facet_wrap(~phenotype) + ggtitle("ERCC percent"),
    ncol=1
)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells assuming that 
most of my dataset consists of high-quality cells. Then I would like to subset 
a SingleCellExperiment object without low-quality cells.
How do I do this?

completion:
To calculate quality control (QC) metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the QC metrics on them. The created QC metrics data frame contains 
a number of columns. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. Assuming that most of the 
dataset consists of high-quality cells, we identify cells that are outliers for 
the various QC metrics, based on the median absolute deviation (MAD) from the 
median value of each metric across all cells. By default, we consider a value 
to be an outlier if it is more than 3 MADs from the median in the “problematic” 
direction. We use perCellQCFilters() function from the scuttle package for this.
This function identifies cells with log-transformed library sizes that are more 
than 3 MADs below the median as well as the proportion-based metrics specified 
in the sub.fields = arguments. A cell that is an outlier for any of these 
metrics is considered to be of low quality and discarded. This is captured in 
the discard column of the created "reasons" data frame consisting of identified
outliers for various QC metrics. Using this information we can subset the
SingleCellExperiment by column to remove low-quality cells.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# total count per cell
summary(df$sum)

# number of detected genes
summary(df$detected)

# percentage of reads mapped to mitochondrial transcripts
summary(df$subsets_Mito_percent)

# percentage of reads mapped to ERCC transcripts
summary(df$altexps_ERCC_percent)

# identifying outliers for various QC metrics
reasons <- perCellQCFilters(df, 
    sub.fields=c("subsets_Mito_percent", "altexps_ERCC_percent"))

# keeping the columns we don't want to discard
filtered <- sce.416b[,!reasons$discard]

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells assuming that 
most of my dataset consists of high-quality cells. Then I would like to mark
the low-quality cells and keep them in the SingleCellExperiment object.
How do I do this?

completion:
To calculate quality control (QC) metrics for each cell, we can use the
perCellQCMetrics() function from the scater package. In the absence of 
spike-in transcripts, the proportion of reads mapped to genes in the 
mitochondrial genome can be used. To do this, we first identify the 
mitochondrial transcripts in our SingleCellExperiment and then use them 
to identify interesting feature subsets, here mitochondrial genes, and 
calculate the QC metrics on them. The created QC metrics data frame contains 
a number of columns. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. Assuming that most of the 
dataset consists of high-quality cells, we identify cells that are outliers for 
the various QC metrics, based on the median absolute deviation (MAD) from the 
median value of each metric across all cells. By default, we consider a value 
to be an outlier if it is more than 3 MADs from the median in the “problematic” 
direction. We use perCellQCFilters() function from the scuttle package for this.
This function identifies cells with log-transformed library sizes that are more 
than 3 MADs below the median as well as the proportion-based metrics specified 
in the sub.fields = arguments. A cell that is an outlier for any of these 
metrics is considered to be of low quality and discarded. This is captured in 
the discard column of the created "reasons" data frame consisting of identified
outliers for various QC metrics. Using this information about the low-quality 
cells we can mark them in the SingleCellExperiment object and retain them in 
the downstream analysis. However, for routine analyses it is advised to remove
them to avoid complications.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load SingleCellExperiment
sce.416b <- LunSpikeInData(which="416b") 

# convert data column to factor
sce.416b$block <- factor(sce.416b$block)

# identifying the mitochondrial transcripts in our SingleCellExperiment
location <- rowRanges(sce.416b)

# mitochondial transcripts
is.mito <- any(seqnames(location)=="MT")

# calculate quality control metrics 
df <- perCellQCMetrics(sce.416b, subsets=list(Mito=is.mito))

# total count per cell
summary(df$sum)

# number of detected genes
summary(df$detected)

# percentage of reads mapped to mitochondrial transcripts
summary(df$subsets_Mito_percent)

# percentage of reads mapped to ERCC transcripts
summary(df$altexps_ERCC_percent)

# identifying outliers for various QC metrics
reasons <- perCellQCFilters(df, 
    sub.fields=c("subsets_Mito_percent", "altexps_ERCC_percent"))

# copy SingleCellExperiment
marked <- sce.416b

# add low-quality cell info to colData "discard"
marked$discard <- reasons$discard

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells based on their 
expression profiles and add them to the colData slot of my SingleCellExperiment 
object. Then I would like to perform diagnostic by plotting the proportion of 
mitochondrial counts against the total count for each cell.
How can I do this?

completion:
To start off, we load the SingleCellExperiment object. Sum the expression 
values (by default, counts) for each feature set in each cell of the object by 
using aggregateAcrossFeatures() function from scuttle package. Annotate genes 
using mabIDs() function from org.Mm.eg.db package (for mice). Then calculate 
quality control (QC) metrics for each cell and append them to the colData slot 
of the SingleCellExperiment object using addPerCellQCMetrics() function from 
scuttle package. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. We then use quickPerCellQC() 
function from scuttle package on the SingleCellExperiment object to identifies 
low-quality cells and store them in a data frame "qc". A cell that is an outlier 
for any of QC metrics is considered to be of low quality and discarded. This is 
captured in the discard column of the created "qc" data frame. We include this 
information in the SingleCellExperiment object by creating a column "discard" 
in the colData slot of the object and assigning the discard information to it 
from the "qc" data frame. Using the updated SingleCellExperiment object we plot 
the proportion of mitochondrial counts against total count for each cell for 
diagnostics by making use of plotColData() function from scater package. Here, 
the aim is to confirm that there are no cells with both large total counts and 
large mitochondrial counts, to ensure that we are not inadvertently removing 
high-quality cells that happen to be highly metabolically active. 

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)
library(scater)
library(org.Mm.eg.db)

# load SingleCellExperiment object
sce.zeisel <- ZeiselBrainData()

# sum the expression values
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

# gene annotation
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

# calculate quality control metrics 
sce.zeisel <- addPerCellQCMetrics(sce.zeisel, 
    subsets=list(Mt=rowData(sce.zeisel)$featureType=="mito"))

# identifies low-quality cells
qc <- quickPerCellQC(colData(sce.zeisel), 
    sub.fields=c("altexps_ERCC_percent", "subsets_Mt_percent"))

# add discarded cells info to the colData slot of the object
sce.zeisel$discard <- qc$discard

# plot percent of mitochondrial counts against total count for each cell
scater::plotColData(sce.zeisel, x="sum", y="subsets_Mt_percent", colour_by="discard")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
calculate quality control metrics to identify low-quality cells based on their 
expression profiles and add them to the colData slot of my SingleCellExperiment 
object. Then I would like to perform diagnostic by plotting the proportion of 
mitochondrial counts against the proportion of ERCC counts.
How can I do this?

completion:
To start off, we load the SingleCellExperiment object. Sum the expression 
values (by default, counts) for each feature set in each cell of the object by 
using aggregateAcrossFeatures() function from scuttle package. Annotate genes 
using mabIDs() function from org.Mm.eg.db package (for mice). Then calculate 
quality control (QC) metrics for each cell and append them to the colData slot 
of the SingleCellExperiment object using addPerCellQCMetrics() function from 
scuttle package. The sum column contains the total count for each cell and 
the detected column contains the number of detected genes. The 
subsets_Mito_percent column contains the percentage of reads mapped to 
mitochondrial transcripts. Finally, the altexps_ERCC_percent column contains 
the percentage of reads mapped to ERCC transcripts. We then use quickPerCellQC() 
function from scuttle package on the SingleCellExperiment object to identifies 
low-quality cells and store them in a data frame "qc". A cell that is an outlier 
for any of QC metrics is considered to be of low quality and discarded. This is 
captured in the discard column of the created "qc" data frame. We include this 
information in the SingleCellExperiment object by creating a column "discard" 
in the colData slot of the object and assigning the discard information to it 
from the "qc" data frame. Using the updated SingleCellExperiment object we plot 
the proportion of mitochondrial counts against proportion of ERCC counts for 
diagnostics by making use of plotColData() function from scater package. 
Comparison of the ERCC and mitochondrial percentages can also be informative. 
Low-quality cells with small mitochondrial percentages, large spike-in 
percentages and small library sizes are likely to be stripped nuclei, i.e., 
they have been so extensively damaged that they have lost all cytoplasmic 
content. On the other hand, cells with high mitochondrial percentages and low 
ERCC percentages may represent undamaged cells that are metabolically active. 
This interpretation also applies for single-nuclei studies but with a switch 
of focus: the stripped nuclei become the libraries of interest while the 
undamaged cells are considered to be low quality.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)
library(scater)
library(org.Mm.eg.db)

# load SingleCellExperiment object
sce.zeisel <- ZeiselBrainData()

# sum the expression values
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

# gene annotation
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

# calculate quality control metrics 
sce.zeisel <- addPerCellQCMetrics(sce.zeisel, 
    subsets=list(Mt=rowData(sce.zeisel)$featureType=="mito"))

# identifies low-quality cells
qc <- quickPerCellQC(colData(sce.zeisel), 
    sub.fields=c("altexps_ERCC_percent", "subsets_Mt_percent"))

# add discarded cells info to the colData slot of the object
sce.zeisel$discard <- qc$discard

# plot percent of mitochondrial counts against percent of ERCC
scater::plotColData(sce.zeisel, x="altexps_ERCC_percent", y="subsets_Mt_percent",
    colour_by="discard")

```

--- TASK END
