---
title: "prompt/completion pairs single-cell_basic_chapter 2"
author: "authors"
date: "2023-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data and I would like to 
perform scaling normalization on it. 
How can I do this?

completion:
Library size normalization is the simplest strategy for performing scaling 
normalization. Library size is defined as the total sum of counts across all 
genes for each cell. Library sizes are converted into size factors by scaling 
them so that their mean across cells is 1. This ensures that the normalized 
values are still on the same scale as the raw counts. With library size-derived 
size factors, we implicitly assume that sequencing coverage is the only 
difference between cells. To perform scaling normalization we can make use of
librarySizeFactors() function from the scuttle package.

```{r,eval=FALSE}

library(scuttle)

# load SingleCellExperiment 
sce.zeisel <- ZeiselBrainData()

# perform scaling normalization
lib.sf.zeisel <- librarySizeFactors(sce.zeisel)

# get summary of the normalized data
summary(lib.sf.zeisel)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. My data contains
unbalanced differential expression between samples/cell types. Taking this into
account how can I perform normalization on my data?

completion:
Composition biases will be present when any unbalanced differential expression 
exists between samples. Normalization can be performed with the 
estimateSizeFactorsFromMatrix() function from the DESeq2 package or with the 
calcNormFactors() function from the edgeR package. However, single-cell data 
can be problematic for these bulk normalization methods due to the dominance 
of low and zero counts. To overcome this, we pool counts from many cells to 
increase the size of the counts for accurate size factor estimation. Pool-based 
size factors are then “deconvolved” into cell-based factors for normalization 
of each cell’s expression profile. This is performed using the 
calculateSumFactors() function from scran package. We use a pre-clustering step 
with quickCluster() function from scran package first, where cells in each 
cluster are normalized separately and the size factors are rescaled to be 
comparable across clusters. This avoids the assumption that most genes are 
non-DE across the entire population - only a non-DE(differentially expressed) 
majority is required between pairs of clusters, which is a weaker assumption 
for highly heterogeneous populations. By default, quickCluster() will use an 
approximate algorithm for PCA based on methods from the irlba package. The 
approximation relies on stochastic initialization so we need to set the 
random seed (via set.seed()) for reproducibility.

```{r,eval=FALSE}

library(scran)

# load SingleCellExperiment 
sce.zeisel <- ZeiselBrainData()

# set seed
set.seed(100)

# perform clustering  
clust.zeisel <- quickCluster(sce.zeisel) 

# visualize clusters
table(clust.zeisel)

# scaling normalization by deconvolution
deconv.sf.zeisel <- calculateSumFactors(sce.zeisel, cluster=clust.zeisel)

# get summary of the normalized data
summary(deconv.sf.zeisel)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
perform normalization assuming, first, that my data does not contain unbalanced 
differential expression between samples/cell types, and second, that it does. 
Moreover, I would like to visualize and compare the two normalization results. 
How can I do this?

completion:
Library size normalization is the simplest strategy for performing scaling 
normalization. Library size is defined as the total sum of counts across all 
genes for each cell. Library sizes are converted into size factors by scaling 
them so that their mean across cells is 1. This ensures that the normalized 
values are still on the same scale as the raw counts. With library size-derived 
size factors, we implicitly assume that sequencing coverage is the only 
difference between cells. To perform scaling normalization we can make use of
librarySizeFactors() function from the scuttle package.
Composition biases will be present when any unbalanced differential expression 
exists between samples. Normalization can be performed with the 
estimateSizeFactorsFromMatrix() function from the DESeq2 package or with the 
calcNormFactors() function from the edgeR package. However, single-cell data 
can be problematic for these bulk normalization methods due to the dominance 
of low and zero counts. To overcome this, we pool counts from many cells to 
increase the size of the counts for accurate size factor estimation. Pool-based 
size factors are then “deconvolved” into cell-based factors for normalization 
of each cell’s expression profile. This is performed using the 
calculateSumFactors() function from scran package. We use a pre-clustering step 
with quickCluster() function from scran package first, where cells in each 
cluster are normalized separately and the size factors are rescaled to be 
comparable across clusters. This avoids the assumption that most genes are 
non-DE (differentially expressed) across the entire population - only a non-DE 
majority is required between pairs of clusters, which is a weaker assumption 
for highly heterogeneous populations. By default, quickCluster() will use an 
approximate algorithm for PCA based on methods from the irlba package. The 
approximation relies on stochastic initialization so we need to set the 
random seed (via set.seed()) for reproducibility. 
The two normalization results can be visualize using plot() function.
Cell type-specific deviations between the two (deconvolution and library) size 
factors is consistent with the presence of composition biases that are 
introduced by strong differential expression between cell types. Use of the 
deconvolution size factors adjusts for these biases to improve normalization 
accuracy for downstream applications.

```{r,eval=FALSE}

library(scuttle)
library(scran)

# load SingleCellExperiment 
sce.zeisel <- ZeiselBrainData()

# perform scaling normalization
lib.sf.zeisel <- librarySizeFactors(sce.zeisel)

# get summary of the normalized data
summary(lib.sf.zeisel)

# set seed
set.seed(100)

# perform clustering  
clust.zeisel <- quickCluster(sce.zeisel) 

# visualize clusters
table(clust.zeisel)

# scaling normalization by deconvolution
deconv.sf.zeisel <- calculateSumFactors(sce.zeisel, cluster=clust.zeisel)

# get summary of the normalized data
summary(deconv.sf.zeisel)

# plot the two normalization results
plot(lib.sf.zeisel, deconv.sf.zeisel, xlab="Library size factor",
    ylab="Deconvolution size factor", log='xy', pch=16,
    col=as.integer(factor(sce.zeisel$level1class))) +
  abline(a=0, b=1, col="red")
  
```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
preserve the differences in the total RNA content of individual cells in 
downstream analysis of my data. Which normalization method should I use and 
how do I go about it?

completion:
Spike-in normalization should be used if differences in the total RNA content 
of individual cells are of interest and must be preserved in downstream 
analyses. Spike-in normalization is based on the assumption that the same 
amount of spike-in RNA was added to each cell. Systematic differences in the 
coverage of the spike-in transcripts can only be due to cell-specific biases.
To remove these biases, we equalize spike-in coverage across cells by scaling 
with “spike-in size factors”. Spike-in normalization assumes that the spike-in 
transcripts were (i) added at a constant level to each cell, and (ii) respond 
to biases in the same relative manner as endogenous genes. For a given cell, 
an increase in its overall amount of endogenous RNA will not increase its 
spike-in size factor. This ensures that the effects of total RNA content on 
expression across the population will not be removed upon scaling. 
We apply the computeSpikeFactors() method from the scuttle package to estimate 
spike-in size factors for all cells. This is defined by converting the total 
spike-in count per cell into a size factor, using the same reasoning* as in 
librarySizeFactors(). Scaling will subsequently remove any differences in 
spike-in coverage across cells.
*The reasoning while performing scaling normalization using librarySizeFactors()
is that the library sizes are converted into size factors by scaling 
them so that their mean across cells is 1. This ensures that the normalized 
values are still on the same scale as the raw counts. 

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)

# load the SingleCellExperiment 
sce.richard <- RichardTCellData()

# subset sce with OK quality data
sce.richard <- sce.richard[,sce.richard$`single cell quality`=="OK"]

# get summary of the data
sce.richard

# perform spike-in normalization on alternative experiment "ERCC"
sce.richard <- computeSpikeFactors(sce.richard, "ERCC")

# get summary of the normalized data
summary(sizeFactors(sce.richard))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
preserve the differences in the total RNA content of individual cells in 
downstream analysis of my data. Therefore, I would like to perform spike-in as 
well as deconvolution normalization and visualize the results to examine any 
correlation between size factors within each treatment condition.
How can I do this?

completion:
Spike-in normalization should be used if differences in the total RNA content 
of individual cells are of interest and must be preserved in downstream 
analyses. Spike-in normalization is based on the assumption that the same 
amount of spike-in RNA was added to each cell. Systematic differences in the 
coverage of the spike-in transcripts can only be due to cell-specific biases.
To remove these biases, we equalize spike-in coverage across cells by scaling 
with “spike-in size factors”. Spike-in normalization assumes that the spike-in 
transcripts were (i) added at a constant level to each cell, and (ii) respond 
to biases in the same relative manner as endogenous genes. For a given cell, 
an increase in its overall amount of endogenous RNA will not increase its 
spike-in size factor. This ensures that the effects of total RNA content on 
expression across the population will not be removed upon scaling. 
We apply the computeSpikeFactors() method from the scuttle package to estimate 
spike-in size factors for all cells. This is defined by converting the total 
spike-in count per cell into a size factor, using the same reasoning* as in 
librarySizeFactors(). Scaling will subsequently remove any differences in 
spike-in coverage across cells.
*The reasoning while performing scaling normalization using librarySizeFactors()
is that the library sizes are converted into size factors by scaling 
them so that their mean across cells is 1. This ensures that the normalized 
values are still on the same scale as the raw counts.
In normalization by deconvolution, we pool counts from many cells to 
increase the size of the counts for accurate size factor estimation. Pool-based 
size factors are then “deconvolved” into cell-based factors for normalization 
of each cell’s expression profile. This is performed using the 
calculateSumFactors() function from scran package.
We can create a data frame containing the normalization results and use it
to visualize the results using ggplot() function to examine any correlation 
between size factors within each treatment condition.
A positive correlation between the spike-in size factors and deconvolution 
size factors within each treatment condition, would indicate that they are 
capturing similar technical biases in sequencing depth and capture efficiency.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)
library(scran)

# load the SingleCellExperiment 
sce.richard <- RichardTCellData()

# subset sce with OK quality data
sce.richard <- sce.richard[,sce.richard$`single cell quality`=="OK"]

# get summary of the data
sce.richard

# perform spike-in normalization on alternative experiment "ERCC"
sce.richard <- computeSpikeFactors(sce.richard, "ERCC")

# get summary of the normalized data
summary(sizeFactors(sce.richard))

# create a data frame with normalization data
to.plot <- data.frame(
    # perform normalization by deconvolution and save the results in DeconvFactor 
    DeconvFactor=calculateSumFactors(sce.richard),
    # save the spike-in normalizaiton results in SpikeFactor 
    SpikeFactor=sizeFactors(sce.richard),
    # save stimulus data in Stimulus
    Stimulus=sce.richard$stimulus, 
    # save time data in Time
    Time=sce.richard$time
)

# create and visualize the plot
ggplot(to.plot, aes(x=DeconvFactor, y=SpikeFactor, color=Time)) +
    geom_point() + facet_wrap(~Stimulus) + scale_x_log10() + 
    scale_y_log10() + geom_abline(intercept=0, slope=1, color="red")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. I would like to 
preserve the differences in the total RNA content of individual cells in 
downstream analysis of my data. Therefore, I would like to perform spike-in
normalization. In addition, I would like to perform deconvolution normalization 
and use the two size factors for downstream analysis.
How can I do this?

completion:
Spike-in normalization should be used if differences in the total RNA content 
of individual cells are of interest and must be preserved in downstream 
analyses. Spike-in normalization is based on the assumption that the same 
amount of spike-in RNA was added to each cell. Systematic differences in the 
coverage of the spike-in transcripts can only be due to cell-specific biases.
To remove these biases, we equalize spike-in coverage across cells by scaling 
with “spike-in size factors”. Spike-in normalization assumes that the spike-in 
transcripts were (i) added at a constant level to each cell, and (ii) respond 
to biases in the same relative manner as endogenous genes. For a given cell, 
an increase in its overall amount of endogenous RNA will not increase its 
spike-in size factor. This ensures that the effects of total RNA content on 
expression across the population will not be removed upon scaling. 
We apply the computeSpikeFactors() method from the scuttle package to estimate 
spike-in size factors for all cells. This is defined by converting the total 
spike-in count per cell into a size factor, using the same reasoning* as in 
librarySizeFactors(). Scaling will subsequently remove any differences in 
spike-in coverage across cells.
*The reasoning while performing scaling normalization using librarySizeFactors()
is that the library sizes are converted into size factors by scaling 
them so that their mean across cells is 1. This ensures that the normalized 
values are still on the same scale as the raw counts.
In normalization by deconvolution, we pool counts from many cells to 
increase the size of the counts for accurate size factor estimation. Pool-based 
size factors are then “deconvolved” into cell-based factors for normalization 
of each cell’s expression profile. This is performed using the 
calculateSumFactors() function from scran package.
We store the normalization results in a data frame "to.plot". We then compute 
log-transformed normalized expression values using "to.plot" by making use of 
logNormalCounts() function from scuttle package, and store them as assays in 
our SingleCellExperiment (sce) objects. These log-values that are stored in 
logcounts assays of our sce, are “log-transformed normalized expression values”,
that will be the basis of our downstream analyses. These counts can be visualize
by using plotExpression() function from scater package for our analysis.

```{r,eval=FALSE}

library(scRNAseq)
library(scuttle)
library(scran)
library(scater)

# load the SingleCellExperiment 
sce.richard <- RichardTCellData()

# subset sce with OK quality data
sce.richard <- sce.richard[,sce.richard$`single cell quality`=="OK"]

# get summary of the data
sce.richard

# perform spike-in normalization on alternative experiment "ERCC"
sce.richard <- computeSpikeFactors(sce.richard, "ERCC")

# get summary of the normalized data
summary(sizeFactors(sce.richard))

# create a data frame with normalization data 
to.plot <- data.frame(
    # perform normalization by deconvolution and save the results in DeconvFactor 
    DeconvFactor=calculateSumFactors(sce.richard),
    # save the spike-in normalizaiton results in SpikeFactor 
    SpikeFactor=sizeFactors(sce.richard),
    # save stimulus data in Stimulus
    Stimulus=sce.richard$stimulus, 
    # save time data in Time
    Time=sce.richard$time
)

# create sce with logcounts assays
sce.richard.deconv <- logNormCounts(sce.richard, size_factors=to.plot$DeconvFactor)
sce.richard.spike <- logNormCounts(sce.richard, size_factors=to.plot$SpikeFactor)

# create and visualize the plot
gridExtra::grid.arrange(
    plotExpression(sce.richard.deconv, x="stimulus", 
        colour_by="time", features="ENSMUSG00000092341") + 
        theme(axis.text.x = element_text(angle = 90)) + 
        ggtitle("After deconvolution"),
    plotExpression(sce.richard.spike, x="stimulus", 
        colour_by="time", features="ENSMUSG00000092341") + 
        theme(axis.text.x = element_text(angle = 90)) +
        ggtitle("After spike-in normalization"),
    ncol=2
)

```

--- TASK END
