---
title: "prompt/completion pairs single-cell_basic_chapter 3"
author: "authors"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. I would like to calculate per-gene
variance.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. We then extract the meta data from the sce using metadata()
function from scran package and use it to visualize the fit. The demo example
shows variance in the PBMC data set as a function of the mean. Each point 
represents a gene while the blue line represents the trend fitted to all genes.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# variance with respect to abundance across all genes
dec.pbmc <- modelGeneVar(sce.pbmc)

# visualizing the fit
fit.pbmc <- metadata(dec.pbmc)
plot(fit.pbmc$mean, fit.pbmc$var, xlab="Mean of log-expression",
    ylab="Variance of log-expression")
curve(fit.pbmc$trend(x), col="dodgerblue", add=TRUE, lwd=2)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select an appropriate 
set of highly variable genes (HVGs).
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. By applying order to the biological 
component of the fitted trend to the variance we can order the genes by most 
interesting (i.e., highly variable genes) to least interesting.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# variance with respect to abundance across all genes
dec.pbmc <- modelGeneVar(sce.pbmc)

# ordering by most interesting genes
dec.pbmc[order(dec.pbmc$bio, decreasing=TRUE),] 

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
spike-in transcripts data. Using the spike-in data I would like to quantify the
variation per gene and select an appropriate set of highly variable genes (HVGs).
How can I do this?

completion:
Under the assumption that the variation in expression for most genes is driven 
by uninteresting processes like sampling noise, the fitted value of the trend 
at any given gene’s abundance represents an estimate of its uninteresting 
variation, which we call the technical component. The biological component for 
each gene is then defined as the difference between its total variance and the 
technical component. This biological component represents the “interesting” 
variation for each gene and can be used as the metric for HVG selection.
The above stated assumption may be problematic in rare scenarios where many 
genes at a particular abundance are affected by a biological process. For 
example, strong upregulation of cell type-specific genes may result in an 
enrichment of HVGs at high abundances. This would inflate the fitted trend in 
that abundance interval and compromise the detection of the relevant genes. We 
can avoid this problem by fitting a mean-dependent trend to the variance of the 
spike-in transcripts, if they are available. The premise here is that spike-ins 
should not be affected by biological variation, so the fitted value of the 
spike-in trend should represent a better estimate of the technical component 
for each gene. 
To calculate per-gene variance we start of by loading the SingleCellExperiment 
(sce) objects. Then using modelGeneVarWithSpikes() from the scran package we 
fit a mean-dependent trend to the variance of the spike-in transcripts. By 
applying order to the biological component of the fitted trend to the variance 
we can order the genes by most interesting (i.e., highly variable genes) to 
least interesting.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS ABOVE

library(scran)

# variance of the spike-in transcripts
dec.spike.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC")

# ordering by most interesting genes
dec.spike.416b[order(dec.spike.416b$bio, decreasing=TRUE),]

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
spike-in transcripts data. Using the spike-in data I would like to fit a 
mean-dependent trend to the variance of the spike-in transcripts and visualize 
it. How can I do this?

completion:
Under the assumption that the variation in expression for most genes is driven 
by uninteresting processes like sampling noise, the fitted value of the trend 
at any given gene’s abundance represents an estimate of its uninteresting 
variation, which we call the technical component. The biological component for 
each gene is then defined as the difference between its total variance and the 
technical component. This biological component represents the “interesting” 
variation for each gene and can be used as the metric for HVG selection.
The above stated assumption may be problematic in rare scenarios where many 
genes at a particular abundance are affected by a biological process. For 
example, strong upregulation of cell type-specific genes may result in an 
enrichment of HVGs at high abundances. This would inflate the fitted trend in 
that abundance interval and compromise the detection of the relevant genes. We 
can avoid this problem by fitting a mean-dependent trend to the variance of the 
spike-in transcripts, if they are available. The premise here is that spike-ins 
should not be affected by biological variation, so the fitted value of the 
spike-in trend should represent a better estimate of the technical component 
for each gene. 
To calculate per-gene variance we start of by loading the SingleCellExperiment 
(sce) objects. Then using modelGeneVarWithSpikes() from scran package we 
fit a mean-dependent trend to the variance of the spike-in transcripts. 
To visualize the fit we can plot the mean-dependent trend to the variance and 
additional information can be added to the plot by first extracting them from 
the metadata using metadata() function from scran package and then adding them.
The demo example shows variance in the 416B data set as a function of the mean. 
Each point represents a gene (black) or spike-in transcript (red) and the blue 
line represents the trend fitted to all spike-ins.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS ABOVE

library(scran)

# variance of the spike-in transcripts
dec.spike.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC")

# ordering by most interesting genes
dec.spike.416b[order(dec.spike.416b$bio, decreasing=TRUE),]

# extracting meta data
fit.spike.416b <- metadata(dec.spike.416b)

# visualizing the fit
plot(dec.spike.416b$mean, dec.spike.416b$total, xlab="Mean of log-expression",
    ylab="Variance of log-expression")
#fit.spike.416b <- metadata(dec.spike.416b)
points(fit.spike.416b$mean, fit.spike.416b$var, col="red", pch=16)
curve(fit.spike.416b$trend(x), col="dodgerblue", add=TRUE, lwd=2)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. My data does not
contain spike-in transcripts data. What approach should I follow to quantify
the variation per gene and select an appropriate set of highly variable genes 
(HVGs)?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. Using the modelGeneVar() function from the scran package 
we fit a trend to the variance with respect to abundance across all genes. 
Under the assumption that the variation in expression for most genes is driven 
by uninteresting processes like sampling noise, the fitted value of the trend 
at any given gene’s abundance represents an estimate of its uninteresting 
variation, which we call the technical component. The biological component for 
each gene is then defined as the difference between its total variance and the 
technical component. This biological component represents the “interesting” 
variation for each gene and can be used as the metric for HVG selection.
The above stated assumption may be problematic in rare scenarios where many 
genes at a particular abundance are affected by a biological process. For 
example, strong upregulation of cell type-specific genes may result in an 
enrichment of HVGs at high abundances. This would inflate the fitted trend in 
that abundance interval and compromise the detection of the relevant genes. We 
can avoid this problem by fitting a mean-dependent trend to the variance of the 
spike-in transcripts, if they are available. The premise here is that spike-ins 
should not be affected by biological variation, so the fitted value of the 
spike-in trend should represent a better estimate of the technical component 
for each gene.
In the absence of spike-in data, one can attempt to create a trend by making 
some distributional assumptions about the noise. For example, UMI counts 
typically exhibit near-Poisson variation if we only consider technical noise 
from library preparation and sequencing. For this, we start of by loading the 
SingleCellExperiment (sce) objects. Then using modelGeneVarByPoisson() function 
from scran package we fit a mean-variance trend to the log-counts. Applying 
order to the biological component of the fitted trend to the log-counts we can 
order the genes by most interesting (i.e., highly variable genes) to least 
interesting.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# set seed
set.seed(0010101)

# variance trend in the log-counts
dec.pois.pbmc <- modelGeneVarByPoisson(sce.pbmc)

# ordering by most interesting genes
dec.pois.pbmc <- dec.pois.pbmc[order(dec.pois.pbmc$bio, decreasing=TRUE),]

# check the order
head(dec.pois.pbmc)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. My data does not
contain spike-in transcripts data. What approach should I follow to quantify
the variation per gene and then visualize the fit?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. Using the modelGeneVar() function from the scran package 
we fit a trend to the variance with respect to abundance across all genes. 
Under the assumption that the variation in expression for most genes is driven 
by uninteresting processes like sampling noise, the fitted value of the trend 
at any given gene’s abundance represents an estimate of its uninteresting 
variation, which we call the technical component. The biological component for 
each gene is then defined as the difference between its total variance and the 
technical component. This biological component represents the “interesting” 
variation for each gene and can be used as the metric for HVG selection.
The above stated assumption may be problematic in rare scenarios where many 
genes at a particular abundance are affected by a biological process. For 
example, strong upregulation of cell type-specific genes may result in an 
enrichment of HVGs at high abundances. This would inflate the fitted trend in 
that abundance interval and compromise the detection of the relevant genes. We 
can avoid this problem by fitting a mean-dependent trend to the variance of the 
spike-in transcripts, if they are available. The premise here is that spike-ins 
should not be affected by biological variation, so the fitted value of the 
spike-in trend should represent a better estimate of the technical component 
for each gene.
In the absence of spike-in data, one can attempt to create a trend by making 
some distributional assumptions about the noise. For example, UMI counts 
typically exhibit near-Poisson variation if we only consider technical noise 
from library preparation and sequencing. This can be used to construct a 
mean-variance trend in the log-counts with the modelGeneVarByPoisson() function.
For this, we start of by loading the SingleCellExperiment (sce) objects. Then 
using modelGeneVarByPoisson() function from scran package we fit a mean-variance 
trend to the log-counts. Applying order to the biological component of the 
fitted trend to the log-counts we can order the genes by most interesting 
(i.e., highly variable genes) to least interesting.
To visualize the fit we can plot the variance against the mean of log-expression.
The demo example shows variance of normalized log-expression values for each 
gene in the PBMC dataset, plotted against the mean log-expression. The blue 
line represents represents the mean-variance relationship corresponding to 
Poisson noise. Note the increased residuals of the high-abundance genes, which 
can be interpreted as the amount of biological variation that was assumed to be 
“uninteresting” when fitting the gene-based trend.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# set seed
set.seed(0010101)

# variance trend in the log-counts
dec.pois.pbmc <- modelGeneVarByPoisson(sce.pbmc)

# ordering by most interesting genes
dec.pois.pbmc <- dec.pois.pbmc[order(dec.pois.pbmc$bio, decreasing=TRUE),]

# check the order
head(dec.pois.pbmc)

# visualize the fit
plot(dec.pois.pbmc$mean, dec.pois.pbmc$total, pch=16, xlab="Mean of log-expression",
    ylab="Variance of log-expression")
curve(metadata(dec.pois.pbmc)$trend(x), col="dodgerblue", add=TRUE)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. Keeping in mind 
that my data contains multiple batches what approach should I follow to select 
an appropriate set of highly variable genes (HVGs)?

completion:
Data containing multiple batches will often exhibit batch effects. We are 
usually not interested in HVGs that are driven by batch effects; instead, 
we want to focus on genes that are highly variable within each batch. This is 
naturally achieved by performing trend fitting and variance decomposition 
separately for each batch. Example below demonstrate this approach by treating 
each plate (block) in the 416B dataset as a different batch, using the 
modelGeneVarWithSpikes() function from scran package. 
We start of by loading the SingleCellExperiment (sce) objects. Then using 
modelGeneVarWithSpikes() function we fit the trend and variance batch wise. 
The biological component represents the “interesting” variation for each gene 
and can be used as the metric for HVG selection. So by applying order to the 
biological component of the fitted trend we can order the genes by most 
interesting (i.e., highly variable genes) to least interesting.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS ABOVE

library(scran)

# trend fitting and variance decomposition separately for each batch
dec.block.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC", block=sce.416b$block)

# ordering by most interesting genes
head(dec.block.416b[order(dec.block.416b$bio, decreasing=TRUE),1:6])

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data. Keeping in mind 
that my data contains multiple batches I would like to select an appropriate 
set of highly variable genes (HVGs). In addition, I would like to visualize the 
differences between batches. How can I do this?

completion:
Data containing multiple batches will often exhibit batch effects. We are 
usually not interested in HVGs that are driven by batch effects; instead, 
we want to focus on genes that are highly variable within each batch. This is 
naturally achieved by performing trend fitting and variance decomposition 
separately for each batch. Example below demonstrate this approach by treating 
each plate (block) in the 416B dataset as a different batch, using the 
modelGeneVarWithSpikes() function from scran package. 
We start of by loading the SingleCellExperiment (sce) objects. Then using 
modelGeneVarWithSpikes() function we fit the trend and variance batch wise. 
The biological component represents the “interesting” variation for each gene 
and can be used as the metric for HVG selection. So by applying order to the 
biological component of the fitted trend we can order the genes by most 
interesting (i.e., highly variable genes) to least interesting.
The use of a batch-specific trend fit is useful as it accommodates differences 
in the mean-variance trends between batches. This is especially important if 
batches exhibit systematic technical differences, e.g., differences in coverage 
or in the amount of spike-in RNA added. 
In our example, there are only minor differences between the trends, which 
indicates that the experiment was tightly replicated across plates. The analysis 
of each plate yields estimates of the biological and technical components for 
each gene, which are averaged across plates to take advantage of information 
from multiple batches.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

## FOR DEMO PURPOSE WE'LL USE 416B DATASET CREATED AS ABOVE

library(scran)

# trend fitting and variance decomposition separately for each batch
dec.block.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC", block=sce.416b$block)

# ordering by most interesting genes
head(dec.block.416b[order(dec.block.416b$bio, decreasing=TRUE),1:6])

# batch differnce visualization
par(mfrow=c(1,2))
blocked.stats <- dec.block.416b$per.block
for (i in colnames(blocked.stats)) {
    current <- blocked.stats[[i]]
    plot(current$mean, current$total, main=i, pch=16, cex=0.5,
        xlab="Mean of log-expression", ylab="Variance of log-expression")
    curfit <- metadata(current)
    points(curfit$mean, curfit$var, col="red", pch=16)
    curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2) 
}

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select an appropriate 
set of highly variable genes (HVGs). For the selection of HVGs is there a 
strategy that I can follow?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
While making a selection a larger subset will reduce the risk of discarding 
interesting biological signal by retaining more potentially relevant genes, at 
the cost of increasing noise from irrelevant genes that might obscure said 
signal. It is difficult to determine the optimal trade-off for any given 
application as noise in one context may be useful signal in another. For 
example, heterogeneity in T cell activation responses is an interesting 
phenomena but may be irrelevant noise in studies that only care about 
distinguishing the major immunophenotypes.
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 1000.
The choice of n also has a fairly straightforward biological interpretation. 
It is based on trend-fitting assumption that most genes do not exhibit 
biological heterogeneity; this implies that they are not differentially 
expressed (DE) between cell types or states in our population. If we quantify 
this assumption into a statement that, e.g., no more than 5% of genes are 
differentially expressed, we can naturally set n to 5% of the number of genes. 
In practice, we usually do not know the proportion of DE genes beforehand so 
this interpretation just exchanges one unknown for another. Nonetheless, it is 
still useful as it implies that we should lower n for less heterogeneous 
datasets, retaining most of the biological signal without unnecessary noise 
from irrelevant genes. Conversely, more heterogeneous datasets should use 
larger values of n to preserve secondary factors of variation beyond those 
driving the most obvious HVGs.
The main disadvantage of this approach is that it turns HVG selection into a 
competition between genes, whereby a subset of very highly variable genes can 
push other informative genes out of the top set. This can be problematic for 
analyses of highly heterogeneous populations if the loss of important markers 
prevents the resolution of certain subpopulations. In the most extreme example, 
consider a situation where a single subpopulation is very different from the 
others. In such cases, the top set will be dominated by differentially expressed 
genes involving that distinct subpopulation, compromising resolution of 
heterogeneity between the other populations. (This can be recovered with a 
nested analysis, but we would prefer to avoid the problem in the first place.)
Another potential concern with this approach is the fact that the choice of n
is fairly arbitrary, with any value from 500 to 5000 considered “reasonable”. 
We have chosen n = 1000 here, though there is no particular a priori reason for 
doing so. Our recommendation is to simply pick an arbitrary n and proceed with 
the rest of the analysis, with the intention of testing other choices later, 
rather than spending much time worrying about obtaining the “optimal” value.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# variance with respect to abundance across all genes
dec.pbmc <- modelGeneVar(sce.pbmc)

# taking the top 1000 genes
hvg.pbmc.var <- getTopHVGs(dec.pbmc, n=1000)

# get the HVgenes
str(hvg.pbmc.var)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select the top 10% of highly 
variable genes (HVGs).
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. Similarly, selection of top 10% of HVGs can be performed. For 
modelGeneVar() and modelGeneVarWithSpikes() functions, we would select the 
genes with the largest biological components. This is conveniently done by 
using getTopHVgs() function from scran package.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# variance with respect to abundance across all genes
dec.pbmc <- modelGeneVar(sce.pbmc)

# select top 10% of HVGs
chosen <- getTopHVGs(dec.pbmc, prop=0.1)

# get the HVgenes
str(chosen)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select the top 10% of highly 
variable genes (HVGs). Following the selection I would like to subset my 
SingleCellExperiment to only retain the selected genes (HVGs).
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. Similarly, selection of top 10% of HVGs can be performed. For 
modelGeneVar() and modelGeneVarWithSpikes() functions, we would select the 
genes with the largest biological components. This is conveniently done by 
using getTopHVgs() function from scran package.
Once we have the selection of interest we can subset the SingleCellExperiment 
to only retain our selection of HVGs. This ensures that downstream methods will 
only use these genes for their calculations. The downside is that the non-HVGs 
are discarded from the new SingleCellExperiment, making it slightly more 
inconvenient to interrogate the full dataset for interesting genes that are not 
HVGs.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# variance with respect to abundance across all genes
dec.pbmc <- modelGeneVar(sce.pbmc)

# select top 10% of HVGs
chosen <- getTopHVGs(dec.pbmc, prop=0.1)

# get the HVgenes
str(chosen)

# subset sce to retain only HVGs
sce.pbmc.hvg <- sce.pbmc[chosen,]

# dimensions of new sce
dim(sce.pbmc.hvg)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select different sets of 
highly variable genes (HVGs). I would then like to save my selections in my 
SingleCellExperiment for downstream analysis. 
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. Similarly, selection of top 10%, 20% & 30% of HVGs can be performed. 
For modelGeneVar() and modelGeneVarWithSpikes() functions, we would select the 
genes with the largest biological components. This is conveniently done by 
using getTopHVgs() function from scran package.
We can keep the original SingleCellExperiment object and specify the genes to 
use for downstream functions via an extra argument like subset.row = . This is 
useful if the analysis uses multiple sets of HVGs at different steps, whereby 
one set of HVGs can be easily swapped for another in specific steps.
This approach is facilitated by the rowSubset() utility, which allows us to 
easily store one or more sets of interest in our SingleCellExperiment. By doing 
so, we avoid the need to keep track of a separate chosen variable and ensure 
that our HVG set is synchronized with any downstream row subsetting of
SingleCellExperiment object.
It can be inconvenient to repeatedly specify the desired feature set across 
steps, so some downstream functions will automatically subset to the default 
rowSubset() if present in the SingleCellExperiment. However, we find that it is 
generally safest to be explicit about which set is being used for a particular 
step.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# variance with respect to abundance across all genes
dec.pbmc <- modelGeneVar(sce.pbmc)

# select top 10% of HVGs
chosen <- getTopHVGs(dec.pbmc, prop=0.1)

# get the HVgenes
str(chosen)

# performing PCA only on the chosen HVGs
library(scater)
sce.pbmc <- runPCA(sce.pbmc, subset_row=chosen)
reducedDimNames(sce.pbmc)

# top 10% of HVGs stored in the default 'subset'
rowSubset(sce.pbmc) <- chosen 

# top 20% of HVGs stored in 'HVGs.more'
rowSubset(sce.pbmc, "HVGs.more") <- getTopHVGs(dec.pbmc, prop=0.2)

# top 30% of HVGs stored in 'HVGs.less'
rowSubset(sce.pbmc, "HVGs.less") <- getTopHVGs(dec.pbmc, prop=0.3)

# confirm the newly added data
colnames(rowData(sce.pbmc))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select the top 10% of highly 
variable genes (HVGs). Using the selection I would like to create a 
SingleCellExperiment to only retain the selected genes (HVGs), and have it and 
my original SingleCellExperiment in the same SingleCellExperiment object to be 
able to access them both conveniently.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. Similarly, selection of top 10% of HVGs can be performed. For 
modelGeneVar() and modelGeneVarWithSpikes() functions, we would select the 
genes with the largest biological components. This is conveniently done by 
using getTopHVgs() function from scran package.
Once we have the selection of interest we can subset the SingleCellExperiment 
to only retain our selection of HVGs. This ensures that downstream methods will 
only use these genes for their calculations. The downside is that the non-HVGs 
are discarded from the new SingleCellExperiment, making it slightly more 
inconvenient to interrogate the full dataset for interesting genes that are not 
HVGs. This can be tackled by having the original SingleCellExperiment as well
in the same SingleCellExperiment object.
“alternative Experiment” system in the SingleCellExperiment class designed for 
storing alternative features like spike-ins or antibody tags, can instead be 
used to hold our full dataset while we perform our downstream operations 
conveniently on the HVG subset. This avoids book-keeping problems in long 
analyses when the original dataset is not synchronized with the HVG subsetted 
data.
We do this by using the subsetted SingleCellExperiment object and storing in 
its alternative experiment our original/full dataset. This way we can perform 
our downstream analysis directly on the HVG subsetted dataset without 
explicitly specifying it (HVG).

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)    # Version=1.31.0
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## FOR DEMO PURPOSE WE'LL USE 10X PBMC DATASET CREATED AS ABOVE

library(scran)

# variance with respect to abundance across all genes
dec.pbmc <- modelGeneVar(sce.pbmc)

# select top 10% of HVGs
chosen <- getTopHVGs(dec.pbmc, prop=0.1)

# get the HVgenes
str(chosen)

# subset sce to retain only HVGs
sce.pbmc.hvg <- sce.pbmc[chosen,]

# dimensions of new sce
dim(sce.pbmc.hvg)

# storing full dataset in the above subsetted class
altExp(sce.pbmc.hvg, "original") <- sce.pbmc

# confirming the addition
altExpNames(sce.pbmc.hvg)

# no need for explicit subset_row= specification in downstream operations
sce.pbmc.hvg <- runPCA(sce.pbmc.hvg)

# recover original data
sce.pbmc.original <- altExp(sce.pbmc.hvg, "original", withColData=TRUE)

```

--- TASK END