---
title: "prompt/completion pairs single-cell_basic_chapter 4"
author: "authors"
date: "2023-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). Moreover, I would like to reduce random noise and 
dimensions in my data by performing PCA. 
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields.
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

# set seed
set.seed(100) 

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

# confirm storage of first 50 PCs
dim(reducedDim(sce.zeisel, "PCA"))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). Since I am working with large data set, how can I 
perform PCA to reduce random noise and dimensions in my data?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields. 
For large data sets, greater efficiency is obtained by using approximate SVD 
algorithms that only compute the top PCs. By default, most PCA-related functions
in scater and scran will use methods from the irlba or rsvd packages to perform
the SVD. We can explicitly specify the SVD algorithm to use by passing an
BiocSingularParam object to the BSPARAM= argument. Many of these approximate 
algorithms are based on randomization and thus require set.seed() to obtain 
reproducible results.
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

library(BiocSingular)
# set seed
set.seed(1000)

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel, 
    BSPARAM=RandomParam(), name="randomized")

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). As I am working with large data set I would like to 
perform PCA to reduce random noise and dimensions in my data. Following which 
how can I decide on the number of principle components to retain?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields. 
For large data sets, greater efficiency is obtained by using approximate SVD 
algorithms that only compute the top PCs. By default, most PCA-related functions
in scater and scran will use methods from the irlba or rsvd packages to perform
the SVD. We can explicitly specify the SVD algorithm to use by passing an
BiocSingularParam object to the BSPARAM= argument. Many of these approximate 
algorithms are based on randomization and thus require set.seed() to obtain 
reproducible results.
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.
Now, the choice of the number of PCs d is a decision that is analogous to the 
choice of the number of HVGs to use. Using more PCs will retain more biological
signal at the cost of including more noise that might mask said signal. On the
other hand, using fewer PCs will introduce competition between different factors
of variation, where weaker (but still interesting) factors may be pushed down 
into lower PCs and inadvertently discarded from downtream analyses.
Much like the choice of the number of HVGs, it is hard to determine whether an 
“optimal” choice exists for the number of PCs. Certainly, we could attempt to 
remove the technical variation that is almost always uninteresting. However, 
even if we were only left with biological variation, there is no straightforward
way to automatically determine which aspects of this variation are relevant. 
One analyst’s biological signal may be irrelevant noise to another analyst with 
a different scientific question. Most practitioners will simply set d to a 
“reasonable” but arbitrary value, typically ranging from 10 to 50. This is often
satisfactory as the later PCs explain so little variance that their inclusion 
or omission has no major effect.
For example, in the Zeisel dataset used for demo here, few PCs explain more than
1% of the variance in the entire dataset (Figure) and choosing between, say, 20 
and 40 PCs would not even amount to four percentage points’ worth of difference 
in variance. In fact, the main consequence of using more PCs is simply that 
downstream calculations take longer as they need to compute over more 
dimensions, but most PC-related calculations are fast enough that this is not 
a practical concern.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

library(BiocSingular)
# set seed
set.seed(1000)

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel, 
    BSPARAM=RandomParam(), name="randomized")

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

# extract percentage of variance explained by successive PCs for visualization
percent.var <- attr(reducedDim(sce.zeisel), "percentVar")

# visualize percentage variance on log scale
plot(percent.var, log="y", xlab="PC", ylab="Variance explained (%)")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). Using the selected HVGs I would like to perform PCA to 
reduce random noise and dimensions in my data. Moreover, I would like to use the 
first two PCs to visualize the resolution of difference between subpopulations. 
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields. 
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.
To visualize the first two PCs we can make use of plotReducedDim() function 
from scater package. In demo plot each point represents a cell, colored 
according to the published annotation.
The problem is that PCA is a linear technique, i.e., only variation along a line
in high-dimensional space is captured by each PC. As such, it cannot efficiently 
pack differences in d dimensions into the first 2 PCs. This is demonstrated in 
demo Figure where the top two PCs fail to resolve some subpopulations identified 
by Zeisel et al. (2015). If the first PC is devoted to resolving the biggest 
difference between subpopulations, and the second PC is devoted to resolving the 
next biggest difference, then the remaining differences will not be visible in 
the plot.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

# set seed
set.seed(100) 

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

library(scater)
# visualize the first two PCs
plotReducedDim(sce.zeisel, dimred="PCA", colour_by="level1class")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). Using the selected HVGs I would like to perform PCA to 
reduce random noise and dimensions in my data. Moreover, I would like to use the 
first few PCs to visualize the resolution of difference between subpopulations. 
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields. 
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.
To visualize the PCs we can make use of plotReducedDim() function from scater 
package. In demo plot each point represents a cell, colored according to the 
published annotation.
However, the problem is that PCA is a linear technique, i.e., only variation 
along a line in high-dimensional space is captured by each PC. As such, it 
cannot efficiently pack differences in d dimensions into the first 2 PCs. This 
is demonstrated in demo Figure where the top two PCs fail to resolve some 
subpopulations identified by Zeisel et al. (2015). If the first PC is devoted 
to resolving the biggest difference between subpopulations, and the second PC 
is devoted to resolving the next biggest difference, then the remaining 
differences will not be visible in the plot.
One workaround is to plot several of the top PCs against each other in pairwise 
plots. However, it is difficult to interpret multiple plots simultaneously, and 
even this approach is not sufficient to separate some of the annotated 
subpopulations in the example shown.
Thus, plotting the top few PCs is not satisfactory for visualization of complex 
populations. That said, the PCA itself is still of great value in visualization 
as it compacts and denoises the data prior to downstream steps. The top PCs are 
often used as input to more sophisticated (and computationally intensive) 
algorithms for dimensionality reduction.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

# set seed
set.seed(100) 

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

library(scater)
# pairwise visualization of the first four PCs 
plotReducedDim(sce.zeisel, dimred="PCA", ncomponents=4,
    colour_by="level1class")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). Moreover, I would like to visualize my scRNA-seq data
using t-stochastic neighbour embedding method.  
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000. 
The de facto standard for visualization of scRNA-seq data is the t-stochastic 
neighbor embedding (t-SNE) method. This attempts to find a low-dimensional 
representation of the data that preserves the distances between each point and 
its neighbors in the high-dimensional space. Unlike PCA, it is not restricted 
to linear transformations, nor is it obliged to accurately represent distances 
between distant populations. This means that it has much more freedom in how it 
arranges cells in low-dimensional space, enabling it to separate many distinct 
clusters in a complex population.
We can use runTSNE from scater package to compute t-SNE coordinates.  
One of the main disadvantages of t-SNE is that it is much more computationally 
intensive than other visualization methods. We mitigate this effect by setting 
dimred="PCA" in runTSNE(), which instructs the function to perform the t-SNE 
calculations on the top PCs of sce. This exploits the data compaction and noise 
removal of the PCA for faster and cleaner results in the t-SNE. It is also 
possible to run t-SNE on the original expression matrix but this is less 
efficient.
Another issue with t-SNE is that it requires the user to be aware of additional 
parameters. It involves a random initialization so we need to set the seed to 
ensure that the chosen results are reproducible. We may also wish to repeat the 
visualization several times to ensure that the results are representative.
To be able to insturuct runTSNE() to perform the t-SNE calcualtion on the top 
PCs, we first need to perform principal components analysis (PCA) on the data. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields. 
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.
To visualize the t-SNE plots constructed from the top PCs we can use 
plotReducedDim() function from scater package. In demo plot each point 
represents a cell, colored according to the published annotation.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

# set seed
set.seed(100) 

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

set.seed(00101001101)

# store the t-SNE coordinates in the reducedDims
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA")

library(scater)
# t-SNE plots from the top PCs 
plotReducedDim(sce.zeisel, dimred="TSNE", colour_by="level1class")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). Moreover, I would like to visualize my scRNA-seq data
using t-stochastic neighbour embedding method. In my visualizaiton I would like
to resolve finer details, but also take measures not to include random noise.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000. 
The de facto standard for visualization of scRNA-seq data is the t-stochastic 
neighbor embedding (t-SNE) method. This attempts to find a low-dimensional 
representation of the data that preserves the distances between each point and 
its neighbors in the high-dimensional space. Unlike PCA, it is not restricted 
to linear transformations, nor is it obliged to accurately represent distances 
between distant populations. This means that it has much more freedom in how it 
arranges cells in low-dimensional space, enabling it to separate many distinct 
clusters in a complex population.
One of the main disadvantages of t-SNE is that it is much more computationally 
intensive than other visualization methods. We mitigate this effect by setting 
dimred="PCA" in runTSNE(), which instructs the function to perform the t-SNE 
calculations on the top PCs of sce. This exploits the data compaction and noise 
removal of the PCA for faster and cleaner results in the t-SNE. It is also 
possible to run t-SNE on the original expression matrix but this is less 
efficient.
Another issue with t-SNE is that it requires the user to be aware of additional 
parameters. It involves a random initialization so we need to set the seed to 
ensure that the chosen results are reproducible. We may also wish to repeat the 
visualization several times to ensure that the results are representative.
The “perplexity” is another important parameter that determines the granularity 
of the visualization. Low perplexities will favor resolution of finer structure, 
possibly to the point that the visualization is compromised by random noise. 
Thus, it is advisable to test different perplexity values to ensure that the 
choice of perplexity does not drive the interpretation of the plot.
We can use runTSNE from scater package to compute t-SNE coordinates using a 
range of perplexity values.  
To be able to insturuct runTSNE() to perform the t-SNE calcualtion on the top 
PCs, we first need to perform principal components analysis (PCA) on the data. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields. 
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.
To visualize the t-SNE plots constructed from the top PCs using the range of 
perplexity values, we can use plotReducedDim() function from scater package. In 
demo plot each point represents a cell, colored according to the published 
annotation.
Finally, it is unwise to read too much into the relative sizes and positions of 
the visual clusters. t-SNE will inflate dense clusters and compress sparse ones, 
such that we cannot use the size as a measure of subpopulation heterogeneity. 
In addition, t-SNE is not obliged to preserve the relative locations of 
non-neighboring clusters, such that we cannot use their positions to determine 
relationships between distant clusters. Despite its shortcomings, t-SNE is 
proven tool for general-purpose visualization of scRNA-seq data and remains a 
popular choice in many analysis pipelines. 

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

# set seed
set.seed(100) 

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

set.seed(100)
# stores the t-SNE coordinates using perplexity=5 
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA", perplexity=5)

# t-SNE plots from the top PCs using perplexity=5
out5 <- plotReducedDim(sce.zeisel, dimred="TSNE",
    colour_by="level1class") + ggtitle("perplexity = 5")

set.seed(100)
# stores the t-SNE coordinates using perplexity=20
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA", perplexity=20)

# t-SNE plots from the top PCs using perplexity=20
out20 <- plotReducedDim(sce.zeisel, dimred="TSNE",
    colour_by="level1class") + ggtitle("perplexity = 20")

set.seed(100)
# stores the t-SNE coordinates using perplexity=80
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA", perplexity=80)

# t-SNE plots from the top PCs using perplexity=80
out80 <- plotReducedDim(sce.zeisel, dimred="TSNE", 
    colour_by="level1class") + ggtitle("perplexity = 80")

# create a grid of plots
gridExtra::grid.arrange(out5, out20, out80, ncol=3)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain
log transformed normalized expression values. Using the log transformed data
I would like to quantify the variation per gene and select top 2000 highly 
variable genes (HVGs). Moreover, I would like to visualize my scRNA-seq data
using the unifrom manifold approximation and projection (UMAP) method, an 
alternative to the t-stochastic neighbour embedding (t-SNE) method.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells. To calculate per-gene variance we start of by loading 
the SingleCellExperiment (sce) objects. Using the modelGeneVar() function from
the scran package we fit a trend to the variance with respect to abundance 
across all genes. At any given abundance, we assume that the variation in 
expression for most genes is driven by uninteresting processes like sampling 
noise. Under this assumption, the fitted value of the trend at any given gene’s 
abundance represents an estimate of its uninteresting variation, which we call 
the technical component. We then define the biological component for each gene 
as the difference between its total variance and the technical component. This 
biological component represents the “interesting” variation for each gene and 
can be used as the metric for HVG selection. 
The most obvious selection strategy is to take the top n genes with the largest 
values for the relevant variance metric. The main advantage of this approach is 
that the user can directly control the number of genes retained, which ensures 
that the computational complexity of downstream calculations is easily 
predicted. For modelGeneVar() and modelGeneVarWithSpikes() functions, we would 
select the genes with the largest biological components. This is conveniently 
done by using getTopHVgs() function from scran package, as shown below with 
n = 2000.
The de facto standard for visualization of scRNA-seq data is the t-stochastic 
neighbor embedding (t-SNE) method. This attempts to find a low-dimensional 
representation of the data that preserves the distances between each point and 
its neighbors in the high-dimensional space. Unlike PCA, it is not restricted 
to linear transformations, nor is it obliged to accurately represent distances 
between distant populations. This means that it has much more freedom in how it 
arranges cells in low-dimensional space, enabling it to separate many distinct 
clusters in a complex population.
The uniform manifold approximation and projection (UMAP) method is an 
alternative to t-SNE for non-linear dimensionality reduction. It is roughly 
similar to t-SNE in that it also tries to find a low-dimensional representation 
that preserves relationships between neighbors in high-dimensional space. 
However, the two methods are based on different theory, represented by 
differences in the various graph weighting equations. This manifests as a 
different visualization.
Compared to t-SNE, the UMAP visualization tends to have more compact visual 
clusters with more empty space between them. It also attempts to preserve more 
of the global structure than t-SNE. From a practical perspective, UMAP is much 
faster than t-SNE, which may be an important consideration for large datasets. 
Here we run UMAP on the top PCs. UMAP also involves a series of randomization 
steps so setting the seed is critical.
We can use runUMAP from scater package to compute UMAP coordinates.
To be able to insturuct runUMAP() to perform the UMAP calcualtion on the top 
PCs, we first need to perform principal components analysis (PCA) on the data. 
Principal components analysis (PCA) discovers axes in high-dimensional space 
that capture the largest amount of variation. The variance captured by each 
axis is defined as the variance in the positions of cells along that axis. 
In PCA, the first axis (or “principal component”, PC) is chosen such that it 
maximizes this variance. The next PC is chosen such that it is orthogonal to 
the first and captures the greatest remaining amount of variation, and so on.
By definition, the top PCs capture the dominant factors of heterogeneity in the 
data set. In the context of scRNA-seq, our assumption is that biological 
processes affect multiple genes in a coordinated manner. This means that the 
earlier PCs are likely to represent biological structure as more variation can 
be captured by considering the correlated behavior of many genes. By comparison, 
random technical or biological noise is expected to affect each gene 
independently. There is unlikely to be an axis that can capture random variation 
across many genes, meaning that noise should mostly be concentrated in the later 
PCs. This motivates the use of the earlier PCs in our downstream analyses, which 
concentrates the biological signal to simultaneously reduce computational work 
and remove noise. The use of the earlier PCs for denoising and data compaction 
is a strategy that is simple, highly effective and widely used in a variety of 
fields. 
We perform a PCA on the log-normalized expression values using the fixedPCA() 
function from scran. By default, fixedPCA() will compute the first 50 PCs and 
store them in the reducedDims() of the output SingleCellExperiment object. Here,
we use only the top 2000 genes with the largest biological components to reduce 
both computational work and high-dimensional random noise. In particular, while 
PCA is robust to random noise, an excess of it may cause the earlier PCs to 
capture noise instead of biological structure. This effect can be mitigated by 
restricting the PCA to a subset of HVGs.
To visualize the UMAP plots constructed from the top PCs, we can use 
plotReducedDim() function from scater package. In demo plot each point 
represents a cell, colored according to the published annotation.
It is arguable whether the UMAP or t-SNE visualizations are more useful or 
aesthetically pleasing. UMAP aims to preserve more global structure but this 
necessarily reduces resolution within each visual cluster. However, UMAP is 
unarguably much faster, and for that reason alone, it is increasingly displacing
t-SNE as the method of choice for visualizing large scRNA-seq data sets.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

## FOR DEMO PURPOSE WE'LL USE ZEISEL ET AL.(2015) DATASET CREATED AS ABOVE

library(scran)
# taking the top 2000 genes
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

# set seed
set.seed(100) 

# perform PCA on the selected genes (n=2000)
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

# confirm storage of PCA in sce
reducedDimNames(sce.zeisel)

# set seed
set.seed(1100101001)

# stores the UMAP coordinates
sce.zeisel <- runUMAP(sce.zeisel, dimred="PCA")

# UMAP plots from the top PCs
plotReducedDim(sce.zeisel, dimred="UMAP", colour_by="level1class")

```

--- TASK END