---
title: "prompt/completion pairs single-cell_basic_chapter 5"
author: "authors"
date: "2024-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to get the cluster assignment of each cell in lower dimensional space.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. The cluster assignment of each cell in lower dimensional space 
can be obtained with clusterCells() function from scran package. It returns a 
vector containing cluster assignments for each cell in the SingleCellExperiment 
object.
By default, clusterCells() uses the 10 nearest neighbors of each cell to 
construct a shared nearest neighbor graph. Two cells are connected by an edge 
if any of their nearest neighbors are shared, with the edge weight defined from 
the highest average rank of the shared neighbors. The Walktrap method from the 
igraph package is then used to identify communities.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# get cluster assignments
nn.clusters <- clusterCells(sce.pbmc, use.dimred="PCA")

# check cluster assignments
table(nn.clusters)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to get the cluster assignment of each cell in lower dimensional space and
visualize the distribution of clusters in a t-SNE plot.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. The cluster assignment of each cell in lower dimensional space 
can be obtained with clusterCells() function from scran package. It returns a 
vector containing cluster assignments for each cell in the SingleCellExperiment 
object. We then assign the cluster assignments back into our 
SingleCellExperiment object as a factor in the column metadata. This allows us 
to conveniently visualize the distribution of clusters in a t-SNE plot using
plotReducedDim() function from scater package. In the demo plot, each point 
represents a cell and is coloured according to the identity of the assigned 
cluster from graph-based clustering.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# get cluster assignments
nn.clusters <- clusterCells(sce.pbmc, use.dimred="PCA")

# check cluster assignments
table(nn.clusters)

library(scater)
# cluster assignments into sce object
colLabels(sce.pbmc) <- nn.clusters

# visualization of clusters 
plotReducedDim(sce.pbmc, "TSNE", colour_by="label")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. Using 
the lower dimension information I would like to create a graph.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. The cluster assignment of each cell in lower dimensional space 
can be obtained with clusterCells() function from scran package. The graph can 
be obtained by specifying full=TRUE in the clusterCells() call. Doing so will 
return all intermediate structures that are used during clustering, including 
a graph object from the igraph package.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# store a graph
nn.clust.info <- clusterCells(sce.pbmc, use.dimred="PCA", full=TRUE)

# output graph
nn.clust.info$objects$graph

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. Using 
the lower dimension information I would like to create and visualize a graph.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. The cluster assignment of each cell in lower dimensional space 
can be obtained with clusterCells() function from scran package. The graph can 
be obtained by specifying full=TRUE in the clusterCells() call. Doing so will 
return all intermediate structures that are used during clustering, including 
a graph object from the igraph package. This graph can be visualized using a 
force-directed layout, closely related to t-SNE and UMAP. The cells are labeled
as per their cluster assignments.
As mentioned earlier cluster assignments can be obtained with clusterCells() 
function, which returns a vector containing cluster assignments for each cell 
in the SingleCellExperiment object. We then assign the cluster assignments back 
into our SingleCellExperiment object as a factor in the column metadata. This 
allows us to visualize the distribution of clusters in a plot using
plotReducedDim() function from scater package. The demo plot is the 
force-directed layout for the shared nearest-neighbor graph of the PBMC dataset. 
Each point represents a cell and is coloured according to its assigned cluster 
identity.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# get cluster assignments
nn.clusters <- clusterCells(sce.pbmc, use.dimred="PCA")

# check cluster assignments
table(nn.clusters)

library(scater)
# cluster assignments into sce object
colLabels(sce.pbmc) <- nn.clusters

library(scran)
# store a graph
nn.clust.info <- clusterCells(sce.pbmc, use.dimred="PCA", full=TRUE)

# output graph
nn.clust.info$objects$graph

# set seed
set.seed(11000)

# store graph object
reducedDim(sce.pbmc, "force") <- igraph::layout_with_fr(nn.clust.info$objects$graph)

library(scater)
# nearest-neighbor graph visualization
plotReducedDim(sce.pbmc, colour_by="label", dimred="force")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to get the cluster assignment of each cell in lower dimensional space.
Moreover, what are the different parameters I could use for my clustering?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. The cluster assignment of each cell in lower dimensional space 
can be obtained with clusterCells() function from scran package. It returns a 
vector containing cluster assignments for each cell in the SingleCellExperiment 
object.
By default, clusterCells() uses the 10 nearest neighbors of each cell to 
construct a shared nearest neighbor graph. Two cells are connected by an edge 
if any of their nearest neighbors are shared, with the edge weight defined from 
the highest average rank of the shared neighbors. The Walktrap method from the 
igraph package is then used to identify communities.
If we wanted to explicitly specify all of these parameters, we would use the 
more verbose call below. This uses a SNNGraphParam object from the bluster 
package to instruct clusterCells() to detect communities from a shared 
nearest-neighbor graph with the specified parameters. The appeal of this 
interface is that it allows us to easily switch to a different clustering 
algorithm by simply changing the BLUSPARAM= argument.
One of the most important parameters is k, the number of nearest neighbors used 
to construct the graph. This controls the resolution of the clustering where 
higher k yields a more inter-connected graph and broader clusters. Users can 
exploit this by experimenting with different values of k to obtain a 
satisfactory resolution.
Further tweaking can be performed by changing the edge weighting scheme during 
graph construction. Setting type="number" will weight edges based on the number 
of nearest neighbors that are shared between two cells. Similarly, 
type="jaccard" will weight edges according to the Jaccard index of the two sets 
of neighbors. We can also disable weighting altogether by using a simple 
k-nearest neighbor graph, which is occasionally useful for downstream graph 
operations that do not support weights.
The community detection can be performed by using any of the algorithms 
provided by igraph, such as Walktrap, but many others are available to choose 
from.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# get cluster assignments
nn.clusters <- clusterCells(sce.pbmc, use.dimred="PCA")

# check cluster assignments
table(nn.clusters)

### parameter: number of neighbours

library(bluster)
# more resolved
clust.5 <- clusterCells(sce.pbmc, use.dimred="PCA", BLUSPARAM=NNGraphParam(k=5))
table(clust.5)

# less resolved
clust.50 <- clusterCells(sce.pbmc, use.dimred="PCA", BLUSPARAM=NNGraphParam(k=50))
table(clust.50)

### parameter: edges weight
clust.num <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(type="number"))
table(clust.num)

clust.jaccard <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(type="jaccard"))
table(clust.jaccard)

clust.none <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=KNNGraphParam())
table(clust.none)

### parameter: community detection
clust.walktrap <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="walktrap"))

clust.louvain <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="louvain"))

clust.infomap <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="infomap"))

clust.fast <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="fast_greedy"))

clust.labprop <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="label_prop"))

clust.eigen <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="leading_eigen"))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to get the cluster assignment of each cell in lower dimensional space.
Moreover, I would like to use different parameters for my clustering and 
compare the different clustering strategies.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. The cluster assignment of each cell in lower dimensional space 
can be obtained with clusterCells() function from scran package. It returns a 
vector containing cluster assignments for each cell in the SingleCellExperiment 
object.
By default, clusterCells() uses the 10 nearest neighbors of each cell to 
construct a shared nearest neighbor graph. Two cells are connected by an edge 
if any of their nearest neighbors are shared, with the edge weight defined from 
the highest average rank of the shared neighbors. The Walktrap method from the 
igraph package is then used to identify communities.
If we wanted to explicitly specify all of these parameters, we would use the 
more verbose call below. This uses a SNNGraphParam object from the bluster 
package to instruct clusterCells() to detect communities from a shared 
nearest-neighbor graph with the specified parameters. The appeal of this 
interface is that it allows us to easily switch to a different clustering 
algorithm by simply changing the BLUSPARAM= argument.
One of the most important parameters is k, the number of nearest neighbors used 
to construct the graph. This controls the resolution of the clustering where 
higher k yields a more inter-connected graph and broader clusters. Users can 
exploit this by experimenting with different values of k to obtain a 
satisfactory resolution.
Further tweaking can be performed by changing the edge weighting scheme during 
graph construction. Setting type="number" will weight edges based on the number 
of nearest neighbors that are shared between two cells. Similarly, 
type="jaccard" will weight edges according to the Jaccard index of the two sets 
of neighbors. We can also disable weighting altogether by using a simple 
k-nearest neighbor graph, which is occasionally useful for downstream graph 
operations that do not support weights.
The community detection can be performed by using any of the algorithms 
provided by igraph, such as Walktrap, but many others are available to choose 
from.
Two clustering strategies can be compared with the help of heatmap plots by 
using heatmap() function from heatmap package. The demo example suggests that 
Infomap yields finer clusters than Walktrap while fast-greedy yields coarser 
clusters. In the plot each entry of each heatmap represents a pair of labels, 
coloured proportionally to the log-number of cells with those labels.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# get cluster assignments
nn.clusters <- clusterCells(sce.pbmc, use.dimred="PCA")

# check cluster assignments
table(nn.clusters)

### parameter: number of neighbours

library(bluster)
# more resolved
clust.5 <- clusterCells(sce.pbmc, use.dimred="PCA", BLUSPARAM=NNGraphParam(k=5))
table(clust.5)

# less resolved
clust.50 <- clusterCells(sce.pbmc, use.dimred="PCA", BLUSPARAM=NNGraphParam(k=50))
table(clust.50)

### parameter: edges weight
clust.num <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(type="number"))
table(clust.num)

clust.jaccard <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(type="jaccard"))
table(clust.jaccard)

clust.none <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=KNNGraphParam())
table(clust.none)

### parameter: community detection
clust.walktrap <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="walktrap"))

clust.louvain <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="louvain"))

clust.infomap <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="infomap"))

clust.fast <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="fast_greedy"))

clust.labprop <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="label_prop"))

clust.eigen <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=NNGraphParam(cluster.fun="leading_eigen"))

library(pheatmap)

# Using a large pseudo-count for a smoother color transition
# between 0 and 1 cell in each 'tab'.
tab <- table(paste("Infomap", clust.infomap), 
    paste("Walktrap", clust.walktrap))
ivw <- pheatmap(log10(tab+10), main="Infomap vs Walktrap",
    color=viridis::viridis(100), silent=TRUE)

tab <- table(paste("Fast", clust.fast), 
    paste("Walktrap", clust.walktrap))
fvw <- pheatmap(log10(tab+10), main="Fast-greedy vs Walktrap",
    color=viridis::viridis(100), silent=TRUE)

gridExtra::grid.arrange(ivw[[4]], fvw[[4]])

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform k-means clustering on my data. 
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
k-means clustering is a classic vector quantization technique that divides cells 
into k clusters. Each cell is assigned to the cluster with the closest centroid, 
which is done by minimizing the within-cluster sum of squares using a random 
starting configuration for the k centroids. We usually set k to a large value 
such as the square root of the number of cells to obtain fine-grained clusters. 
These are not meant to be interpreted directly, but rather, the centroids are 
used in downstream steps for faster computation. The main advantage of this 
approach lies in its speed, given the simplicity and ease of implementation of 
the algorithm.
To perform k-means clustering with the specified number of clusters in centers=, 
we supply a KmeansParam object in clusterCells(). We again use our top PCs after 
setting the random seed to ensure that the results are reproducible. 

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

# set seed
set.seed(100)

# get k-mean clusters
clust.kmeans <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=KmeansParam(centers=10))

# check clusters
table(clust.kmeans)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform k-means clustering on my data and visualize the clusters. 
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
k-means clustering is a classic vector quantization technique that divides cells 
into k clusters. Each cell is assigned to the cluster with the closest centroid, 
which is done by minimizing the within-cluster sum of squares using a random 
starting configuration for the k centroids. We usually set k to a large value 
such as the square root of the number of cells to obtain fine-grained clusters. 
These are not meant to be interpreted directly, but rather, the centroids are 
used in downstream steps for faster computation. The main advantage of this 
approach lies in its speed, given the simplicity and ease of implementation of 
the algorithm.
To perform k-means clustering with the specified number of clusters in centers=, 
we supply a KmeansParam object in clusterCells(). We again use our top PCs after 
setting the random seed to ensure that the results are reproducible. 
We then assign the cluster assignments back into our SingleCellExperiment object 
as a factor in the column metadata. This allows us to conveniently visualize the 
distribution of clusters in a t-SNE plot using plotReducedDim() function from 
scater package. In the demo t-SNE plot of the 10X PBMC dataset, each point 
represents a cell and is coloured according to the identity of the assigned 
cluster from k-mean clustering.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

# set seed
set.seed(100)

# get k-mean clusters
clust.kmeans <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=KmeansParam(centers=10))

# check clusters
table(clust.kmeans)

library(scater)
# cluster assignments into sce object
colLabels(sce.pbmc) <- clust.kmeans

# visualization of clusters 
plotReducedDim(sce.pbmc, "TSNE", colour_by="label")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform clustering on my data using a method that gives me large number 
of clusters.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
k-means clustering is a classic vector quantization technique that divides cells 
into k clusters. Each cell is assigned to the cluster with the closest centroid, 
which is done by minimizing the within-cluster sum of squares using a random 
starting configuration for the k centroids. We usually set k to a large value 
such as the square root of the number of cells to obtain fine-grained clusters. 
These are not meant to be interpreted directly, but rather, the centroids are 
used in downstream steps for faster computation. The main advantage of this 
approach lies in its speed, given the simplicity and ease of implementation of 
the algorithm.
A more practical use of k-means is to deliberately set k to a large value to 
achieve overclustering. This will forcibly partition cells inside broad clusters 
that do not have well-defined internal structure. For example, we might be 
interested in the change in expression from one “side” of a cluster to the 
other, but the lack of any clear separation within the cluster makes it 
difficult to separate with graph-based methods, even at the highest resolution.
k-means has no such problems and will readily split these broad clusters for 
greater resolution.
To perform k-means clustering with the specified number of clusters in centers=, 
we supply a KmeansParam object in clusterCells(). We again use our top PCs after 
setting the random seed to ensure that the results are reproducible. 

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

# set seed
set.seed(100)

# get k-mean clusters
clust.kmeans2 <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=KmeansParam(centers=20))

# check clusters
table(clust.kmeans2)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform clustering on my data using a method that gives me large number 
of clusters and visualize the clusters as well.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
k-means clustering is a classic vector quantization technique that divides cells 
into k clusters. Each cell is assigned to the cluster with the closest centroid, 
which is done by minimizing the within-cluster sum of squares using a random 
starting configuration for the k centroids. We usually set k to a large value 
such as the square root of the number of cells to obtain fine-grained clusters. 
These are not meant to be interpreted directly, but rather, the centroids are 
used in downstream steps for faster computation. The main advantage of this 
approach lies in its speed, given the simplicity and ease of implementation of 
the algorithm.
A more practical use of k-means is to deliberately set k to a large value to 
achieve overclustering. This will forcibly partition cells inside broad clusters 
that do not have well-defined internal structure. For example, we might be 
interested in the change in expression from one “side” of a cluster to the 
other, but the lack of any clear separation within the cluster makes it 
difficult to separate with graph-based methods, even at the highest resolution.
k-means has no such problems and will readily split these broad clusters for 
greater resolution.
To perform k-means clustering with the specified number of clusters in centers=, 
we supply a KmeansParam object in clusterCells(). We again use our top PCs after 
setting the random seed to ensure that the results are reproducible. 
For convenience of visualizing the distribution of clusters we can assign the 
cluster assignments back into our SingleCellExperiment object as a factor in the 
column metadata. Visualization of t-SNE plot can be done using plotReducedDim() 
function from scater package. In the demo t-SNE plot of the 10X PBMC dataset, 
each point represents a cell and is coloured according to the identity of the 
assigned cluster from k-mean clustering with k=20.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

# set seed
set.seed(100)

# get k-mean clusters
clust.kmeans2 <- clusterCells(sce.pbmc, use.dimred="PCA", 
    BLUSPARAM=KmeansParam(centers=20))

# check clusters
table(clust.kmeans2)

library(scater)
# cluster assignments into sce object
colLabels(sce.pbmc) <- clust.kmeans2

# visualization of clusters 
plotTSNE(sce.pbmc, colour_by="label", text_by="label")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. What 
approach can I follow to cluster my large dataset using k-means clustering 
method?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
k-means clustering is a classic vector quantization technique that divides cells 
into k clusters. Each cell is assigned to the cluster with the closest centroid, 
which is done by minimizing the within-cluster sum of squares using a random 
starting configuration for the k centroids. We usually set k to a large value 
such as the square root of the number of cells to obtain fine-grained clusters. 
These are not meant to be interpreted directly, but rather, the centroids are 
used in downstream steps for faster computation. The main advantage of this 
approach lies in its speed, given the simplicity and ease of implementation of 
the algorithm.
To perform k-means clustering with the specified number of clusters in centers=, 
we supply a KmeansParam object in clusterCells(). We again use our top PCs after 
setting the random seed to ensure that the results are reproducible. 
For larger datasets, we can use a variant of this approach named mini-batch 
k-means from the mbkmeans package. At each iteration, we only update the cluster 
assignments and centroid positions for a small subset of the observations. This 
reduces memory usage and computational time - especially when not all of the 
observations are informative for convergence - and supports parallelization via 
BiocParallel. Using this variant is as simple as switching to a MbkmeansParam() 
object in our clusterCells() call.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

# set seed
set.seed(100)

library(mbkmeans)
# get k-mean clusters
clust.mbkmeans <- clusterCells(sce.pbmc, use.dimred="PCA",
    BLUSPARAM=bluster::MbkmeansParam(centers=10))

# check clusters
table(clust.mbkmeans)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform combined k-means/graph-based clustering on my data and visualize 
the clusters.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
k-means clustering is a classic vector quantization technique that divides cells 
into k clusters. Each cell is assigned to the cluster with the closest centroid, 
which is done by minimizing the within-cluster sum of squares using a random 
starting configuration for the k centroids. We usually set k to a large value 
such as the square root of the number of cells to obtain fine-grained clusters. 
These are not meant to be interpreted directly, but rather, the centroids are 
used in downstream steps for faster computation. The main advantage of this 
approach lies in its speed, given the simplicity and ease of implementation of 
the algorithm.
By itself, k-means suffers from several shortcomings that reduce its appeal for 
obtaining interpretable clusters.
These concerns however are less relevant when k-means is being used for vector 
quantization. In this application, k-means is used as a prelude to more 
sophisticated and interpretable - but computationally expensive - clustering 
algorithms. The clusterCells() function supports a “two-step” mode where k-means 
is initially used to obtain representative centroids that are subjected to 
graph-based clustering. Each cell is then placed in the same graph-based cluster 
that its k-means centroid was assigned to.
To perform the combined clustering with the specified number of clusters in 
centers=, and number of neighbours in graph construction in k=, we supply a
KmeansParam object and a NNGraphParam object respectively in clusterCells().
We again use our top PCs after setting the random seed to ensure that the 
results are reproducible. 
The obvious benefit of this approach over direct graph-based clustering is the 
speed improvement. We avoid the need to identifying nearest neighbors for each 
cell and the construction of a large intermediate graph, while benefiting from 
the relative interpretability of graph-based clusters compared to those from 
k-means. This approach also mitigates the “inflation” effect. Each centroid 
serves as a representative of a region of space that is roughly similar in 
volume, ameliorating differences in cell density that can cause (potentially 
undesirable) differences in resolution.
The choice of the number of k-means clusters determines the trade-off between 
speed and fidelity. Larger values provide a more faithful representation of the 
underlying distribution of cells, at the cost of requiring more computational 
work by the second-step clustering procedure. Note that the second step operates 
on the centroids, so increasing clusters= may have further implications if the 
second-stage procedure is sensitive to the total number of input observations. 
For example, increasing the number of centroids would require an concomitant 
increase in k= (the number of neighbors in graph construction) to maintain the 
same level of resolution in the final output
The clusters can be visualize using plotTSNE() function from scater package. 
In the demo t-SNE plot of the 10X PBMC dataset, each point represents a cell 
and is coloured according to the identity of the assigned cluster from combined 
k-mean/graph-based clustering.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

# setting the seed due to the randomness of k-means
set.seed(0101010)

library(bluster)
# get combined k-means and graph-based clusters
kgraph.clusters <- clusterCells(sce.pbmc, use.dimred="PCA",
    BLUSPARAM=TwoStepParam(
        first=KmeansParam(centers=1000),
        second=NNGraphParam(k=5)
    )
)

# check clusters
table(kgraph.clusters)

library(scater)
# visualization of clusters 
plotTSNE(sce.pbmc, colour_by=I(kgraph.clusters))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform hierarchical clustering on the top principle components of my 
data and visualize the clusters.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
Hierarchical clustering is an old technique that arranges samples into a 
hierarchy based on their relative similarity to each other. Most implementations 
do so by joining the most similar samples into a new cluster, then joining 
similar clusters into larger clusters, and so on, until all samples belong to a 
single cluster. This process yields a dendrogram that defines clusters with 
progressively increasing granularity. Variants of hierarchical clustering 
methods primarily differ in how they choose to perform the agglomerations. For 
example, complete linkage aims to merge clusters with the smallest maximum 
distance between their elements, while Ward’s method aims to minimize the 
increase in within-cluster variance.
In the context of scRNA-seq, the main advantage of hierarchical clustering lies 
in the production of the dendrogram. This is a rich summary that quantitatively 
captures the relationships between subpopulations at various resolutions. 
Cutting the dendrogram at high resolution is also guaranteed to yield clusters 
that are nested within those obtained at a low-resolution cut; this can be 
helpful for interpretation. The dendrogram is also a natural representation of 
the data in situations where cells have descended from a relatively recent 
common ancestor.
In practice, hierarchical clustering is too slow to be used for anything but 
the smallest scRNA-seq datasets. Most implementations require a cell-cell 
distance matrix that is prohibitively expensive to compute for a large number 
of cells.
We use a HclustParam object to instruct clusterCells() to perform hierarchical 
clustering on the top PCs. Specifically, it computes a cell-cell distance matrix 
using the top PCs and then applies Ward’s minimum variance method to obtain a 
dendrogram. The dendrogram can them be visualized using plot() function.
The demo plot shows hierarchy of cells in the 416B data set after hierarchical 
clustering, where each leaf node is a cell that is coloured according to its 
oncogene induction status (red is induced, blue is control) and plate of origin 
(light or dark).

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 416B DATASET AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

#--- variance-modelling ---#
dec.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC", block=sce.416b$block)
chosen.hvgs <- getTopHVGs(dec.416b, prop=0.1)

#--- batch-correction ---#
library(limma)
assay(sce.416b, "corrected") <- removeBatchEffect(logcounts(sce.416b), 
    design=model.matrix(~sce.416b$phenotype), batch=sce.416b$block)

#--- dimensionality-reduction ---#
sce.416b <- runPCA(sce.416b, ncomponents=10, subset_row=chosen.hvgs,
    exprs_values="corrected", BSPARAM=BiocSingular::ExactParam())

set.seed(1010)
sce.416b <- runTSNE(sce.416b, dimred="PCA", perplexity=10)

## FOR DEMO PURPOSE WE'LL USE THE 416B DATASET AS ABOVE

library(bluster)
# perform hierarchial clustering
hclust.416b <- clusterCells(sce.416b, use.dimred="PCA",
    BLUSPARAM=HclustParam(method="ward.D2"), full=TRUE)
tree.416b <- hclust.416b$objects$hclust

# making a prettier dendrogram
library(dendextend)
tree.416b$labels <- seq_along(tree.416b$labels)
dend <- as.dendrogram(tree.416b, hang=0.1)

combined.fac <- paste0(sce.416b$block, ".", 
    sub(" .*", "", sce.416b$phenotype))
labels_colors(dend) <- c(
    "20160113.wild"="blue",
    "20160113.induced"="red",
    "20160325.wild"="dodgerblue",
    "20160325.induced"="salmon"
)[combined.fac][order.dendrogram(dend)]

# visualizing the dendrogram
plot(dend)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform hierarchical clustering on the top principle components of my 
data to obtain explicit clusters and then visualize the clusters.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
Hierarchical clustering is an old technique that arranges samples into a 
hierarchy based on their relative similarity to each other. Most implementations 
do so by joining the most similar samples into a new cluster, then joining 
similar clusters into larger clusters, and so on, until all samples belong to a 
single cluster. This process yields a dendrogram that defines clusters with 
progressively increasing granularity. Variants of hierarchical clustering 
methods primarily differ in how they choose to perform the agglomerations. For 
example, complete linkage aims to merge clusters with the smallest maximum 
distance between their elements, while Ward’s method aims to minimize the 
increase in within-cluster variance.
In the context of scRNA-seq, the main advantage of hierarchical clustering lies 
in the production of the dendrogram. This is a rich summary that quantitatively 
captures the relationships between subpopulations at various resolutions. 
Cutting the dendrogram at high resolution is also guaranteed to yield clusters 
that are nested within those obtained at a low-resolution cut; this can be 
helpful for interpretation. The dendrogram is also a natural representation of 
the data in situations where cells have descended from a relatively recent 
common ancestor.
In practice, hierarchical clustering is too slow to be used for anything but 
the smallest scRNA-seq datasets. Most implementations require a cell-cell 
distance matrix that is prohibitively expensive to compute for a large number 
of cells.
We use a HclustParam object to instruct clusterCells() to perform hierarchical 
clustering on the top PCs. Specifically, it computes a cell-cell distance matrix 
using the top PCs and then applies Ward’s minimum variance method to obtain a 
dendrogram. 
To obtain explicit clusters, we “cut” the tree by removing internal branches 
such that every subtree represents a distinct cluster. This is most simply done 
by removing internal branches above a certain height of the tree, as performed 
by the cutree() function. A more sophisticated variant of this approach is 
implemented in the dynamicTreeCut package, which uses the shape of the branches 
to obtain a better partitioning for complex dendrograms. We enable this option 
by setting cut.dynamic=TRUE, with additional tweaking of the deepSplit= 
parameter to control the resolution of the resulting clusters.
The dendrogram can them be visualized using plot() function.
The demo plot shows hierarchy of cells in the 416B data set after hierarchical 
clustering, where each leaf node is a cell that is coloured according to its 
assigned cluster identity from a dynamic tree cut.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 416B DATASET AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

#--- variance-modelling ---#
dec.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC", block=sce.416b$block)
chosen.hvgs <- getTopHVGs(dec.416b, prop=0.1)

#--- batch-correction ---#
library(limma)
assay(sce.416b, "corrected") <- removeBatchEffect(logcounts(sce.416b), 
    design=model.matrix(~sce.416b$phenotype), batch=sce.416b$block)

#--- dimensionality-reduction ---#
sce.416b <- runPCA(sce.416b, ncomponents=10, subset_row=chosen.hvgs,
    exprs_values="corrected", BSPARAM=BiocSingular::ExactParam())

set.seed(1010)
sce.416b <- runTSNE(sce.416b, dimred="PCA", perplexity=10)

## FOR DEMO PURPOSE WE'LL USE THE 416B DATASET AS ABOVE

library(bluster)
# perform hierarchial clustering
hclust.416b <- clusterCells(sce.416b, use.dimred="PCA",
    BLUSPARAM=HclustParam(method="ward.D2"), full=TRUE)
tree.416b <- hclust.416b$objects$hclust

# making a prettier dendrogram
library(dendextend)
tree.416b$labels <- seq_along(tree.416b$labels)
dend <- as.dendrogram(tree.416b, hang=0.1)

library(bluster)
library(dynamicTreeCut)
# perform hierarchial clustering for explicit clusters
hclust.dyn <- clusterCells(sce.416b, use.dimred="PCA",
    BLUSPARAM=HclustParam(method="ward.D2", cut.dynamic=TRUE,
        cut.params=list(minClusterSize=10, deepSplit=1)))
table(hclust.dyn)

# adding labels
labels_colors(dend) <- as.integer(hclust.dyn)[order.dendrogram(dend)]

# visualizing the dendrogram
plot(dend)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to perform hierarchical clustering on the top principle components of my 
data to obtain explicit clusters for grouping cells on a t-SNE plot.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
Hierarchical clustering is an old technique that arranges samples into a 
hierarchy based on their relative similarity to each other. Most implementations 
do so by joining the most similar samples into a new cluster, then joining 
similar clusters into larger clusters, and so on, until all samples belong to a 
single cluster. This process yields a dendrogram that defines clusters with 
progressively increasing granularity. Variants of hierarchical clustering 
methods primarily differ in how they choose to perform the agglomerations. For 
example, complete linkage aims to merge clusters with the smallest maximum 
distance between their elements, while Ward’s method aims to minimize the 
increase in within-cluster variance.
In the context of scRNA-seq, the main advantage of hierarchical clustering lies 
in the production of the dendrogram. This is a rich summary that quantitatively 
captures the relationships between subpopulations at various resolutions. 
Cutting the dendrogram at high resolution is also guaranteed to yield clusters 
that are nested within those obtained at a low-resolution cut; this can be 
helpful for interpretation. The dendrogram is also a natural representation of 
the data in situations where cells have descended from a relatively recent 
common ancestor.
In practice, hierarchical clustering is too slow to be used for anything but 
the smallest scRNA-seq datasets. Most implementations require a cell-cell 
distance matrix that is prohibitively expensive to compute for a large number 
of cells.
We use a HclustParam object to instruct clusterCells() to perform hierarchical 
clustering on the top PCs. Specifically, it computes a cell-cell distance matrix 
using the top PCs and then applies Ward’s minimum variance method to obtain a 
dendrogram. 
To obtain explicit clusters, we “cut” the tree by removing internal branches 
such that every subtree represents a distinct cluster. This is most simply done 
by removing internal branches above a certain height of the tree, as performed 
by the cutree() function. A more sophisticated variant of this approach is 
implemented in the dynamicTreeCut package, which uses the shape of the branches 
to obtain a better partitioning for complex dendrograms. We enable this option 
by setting cut.dynamic=TRUE, with additional tweaking of the deepSplit= 
parameter to control the resolution of the resulting clusters.
We then assign the cluster assignments back into our SingleCellExperiment object 
as a factor in the column metadata. This allows us to conveniently visualize the 
distribution of clusters in a t-SNE plot using plotReducedDim() function from 
scater package. 
In the demo t-SNE plot of the 416b dataset, each point represents a cell and is 
coloured according to the identity of the assigned cluster from hierarchical 
clustering. Here, cluster 2 is split across two visual clusters in the plot but 
we attribute this to a distortion introduced by t-SNE, given that this cluster 
actually has the highest average silhouette width.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 416B DATASET AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

#--- variance-modelling ---#
dec.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC", block=sce.416b$block)
chosen.hvgs <- getTopHVGs(dec.416b, prop=0.1)

#--- batch-correction ---#
library(limma)
assay(sce.416b, "corrected") <- removeBatchEffect(logcounts(sce.416b), 
    design=model.matrix(~sce.416b$phenotype), batch=sce.416b$block)

#--- dimensionality-reduction ---#
sce.416b <- runPCA(sce.416b, ncomponents=10, subset_row=chosen.hvgs,
    exprs_values="corrected", BSPARAM=BiocSingular::ExactParam())

set.seed(1010)
sce.416b <- runTSNE(sce.416b, dimred="PCA", perplexity=10)

## FOR DEMO PURPOSE WE'LL USE THE 416B DATASET AS ABOVE

library(bluster)
# perform hierarchial clustering
hclust.416b <- clusterCells(sce.416b, use.dimred="PCA",
    BLUSPARAM=HclustParam(method="ward.D2"), full=TRUE)
tree.416b <- hclust.416b$objects$hclust

# making a prettier dendrogram
library(dendextend)
tree.416b$labels <- seq_along(tree.416b$labels)
dend <- as.dendrogram(tree.416b, hang=0.1)

library(bluster)
library(dynamicTreeCut)
# perform hierarchial clustering for explicit clusters
hclust.dyn <- clusterCells(sce.416b, use.dimred="PCA",
    BLUSPARAM=HclustParam(method="ward.D2", cut.dynamic=TRUE,
        cut.params=list(minClusterSize=10, deepSplit=1)))
table(hclust.dyn)


library(scater)
# cluster assignments into sce object
colLabels(sce.416b) <- factor(hclust.dyn)

# visualization of clusters 
plotReducedDim(sce.416b, "TSNE", colour_by="label")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. Since I 
am working with a large data set, how can I perform hierarchical clustering on 
my data and then visualize the clusters?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
Hierarchical clustering is an old technique that arranges samples into a 
hierarchy based on their relative similarity to each other. Most implementations 
do so by joining the most similar samples into a new cluster, then joining 
similar clusters into larger clusters, and so on, until all samples belong to a 
single cluster. This process yields a dendrogram that defines clusters with 
progressively increasing granularity. Variants of hierarchical clustering 
methods primarily differ in how they choose to perform the agglomerations. For 
example, complete linkage aims to merge clusters with the smallest maximum 
distance between their elements, while Ward’s method aims to minimize the 
increase in within-cluster variance.
In the context of scRNA-seq, the main advantage of hierarchical clustering lies 
in the production of the dendrogram. This is a rich summary that quantitatively 
captures the relationships between subpopulations at various resolutions. 
Cutting the dendrogram at high resolution is also guaranteed to yield clusters 
that are nested within those obtained at a low-resolution cut; this can be 
helpful for interpretation. The dendrogram is also a natural representation of 
the data in situations where cells have descended from a relatively recent 
common ancestor.
In practice, hierarchical clustering is too slow to be used for anything but 
the smallest scRNA-seq datasets. Most implementations require a cell-cell 
distance matrix that is prohibitively expensive to compute for a large number 
of cells.
However, we can make use of two-step approach to perform hierarchical clustering 
on larger data set. In this approach hierarchical clustering to perform on the 
representative centroids. This avoids the construction of a distance matrix 
across all cells for faster computation.
The clusterCells() function supports a “two-step” mode where k-means is 
initially used to obtain representative centroids that are then used for 
hierarchical clustering. The parameters for this are supplied in a KmeansParam 
object and a HclustParam object respectively in clusterCells(). We use our top 
PCs after setting the random seed to ensure that the results are reproducible. 
The clusters can be visualize using plotTSNE() function from scater package. 
In the demo t-SNE plot of the 10X PBMC dataset, each point represents a cell 
and is coloured according to the identity of the assigned cluster from combined 
k-mean/hierarchical clustering.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

# setting the seed due to the randomness of k-means
set.seed(1111)

library(bluster)
# get clusters
khclust.info <- clusterCells(sce.pbmc, use.dimred="PCA",
    BLUSPARAM=TwoStepParam(
        first=KmeansParam(centers=1000),
        second=HclustParam(method="ward.D2", cut.dynamic=TRUE,
            cut.param=list(deepSplit=3)) # for higher resolution.
    ),
    full=TRUE
)

# check clusters
table(khclust.info$clusters)

library(scater)
# visualization of clusters 
plotTSNE(sce.pbmc, colour_by=I(khclust.info$clusters), 
    text_by=I(khclust.info$clusters))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. Since I 
am working with a large data set is there another approach to improve resolution 
of my data set?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. 
The cluster assignment of each cell in lower dimensional space can be obtained 
with clusterCells() function from scran package. It returns a vector containing 
cluster assignments for each cell in the SingleCellExperiment object.
Two-step approach can be used to perform for example, combined 
k-means/hierarchical clustering, and combined k-means/affinity propagation 
clustering on large data set, where we could subject our k-means centroids to 
clustering by hierarchy and affinity propagation respectively.
Another simple approach to improving resolution is to repeat the feature 
selection and clustering within a single cluster. This aims to select HVGs and 
PCs that are more relevant to internal structure, improving resolution by 
avoiding noise from unnecessary features. Subsetting also encourages clustering 
methods to separate cells according to more modest heterogeneity in the absence 
of distinct subpopulations. 
Here we demonstrate this with a cluster of putative memory T cells from the 
PBMC dataset, identified according to several markers.
Visualization can be done by plotExpression() function from scater package that 
plots expression values for a set of featues in a SingleExperiment object. The 
demo plot shows the distribution of log-normalized expression values for several 
T cell markers within each cluster in the 10X PBMC dataset. Each cluster is 
color-coded for convenience.
Following repeated modelling and PCA on the subset, we apply graph-based 
clustering within this memory subset to obtain CD4+ and CD8+ subclusters. 
Admittedly, the expression of CD4 is so low that the change is rather modest, 
but the interpretation is clear enough.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# get cluster assignments
clust.full <- clusterCells(sce.pbmc, use.dimred="PCA")

library(scater)
# visualization of clusters 
plotExpression(sce.pbmc, features=c("CD3E", "CCR7", "CD69", "CD44"),
    x=I(clust.full), colour_by=I(clust.full))

# repeating modelling and PCA on the subset
memory <- 10L
sce.memory <- sce.pbmc[,clust.full==memory]
dec.memory <- modelGeneVar(sce.memory)
sce.memory <- denoisePCA(sce.memory, technical=dec.memory,
    subset.row=getTopHVGs(dec.memory, n=5000))

# graph-based clustering
g.memory <- buildSNNGraph(sce.memory, use.dimred="PCA")
clust.memory <- igraph::cluster_walktrap(g.memory)$membership

library(scater)
# visualization of clusters
plotExpression(sce.memory, features=c("CD8A", "CD4"),
    x=I(factor(clust.memory)))

```

--- TASK END