---
title: "prompt/completion pairs single-cell_basic_chapter 6"
author: "authors"
date: "2024-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to identify the genes that drive separation between clusters in my data.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. Clustering can be then performed using buildSNNGraph() function 
from scran package.
The most straightforward approach to marker gene detection involves testing for 
differential expression (DE) between clusters. If a gene is strongly DE between 
clusters, it is likely to have driven the separation of cells in the clustering 
algorithm. Several methods are available to quantify the differences in 
expression profiles between clusters and obtain a single ranking of genes for 
each cluster. 
A general strategy is to compare each pair of clusters and compute scores 
quantifying the differences in the expression distributions between clusters. 
The scores for all pairwise comparisons involving a particular cluster are then 
consolidated into a single DataFrame for that cluster. The scoreMarkers() 
function from scran returns a list of DataFrames where each DataFrame 
corresponds to a cluster and each row of the DataFrame corresponds to a gene. 
In the DataFrame for cluster X, the columns contain the self.average, the mean 
log-expression in X; other.average, the grand mean across all other clusters; 
self.detected, the proportion of cells with detected expression in X; 
other.detected, the mean detected proportion across all other clusters; and 
finally, a variety of effect size summaries generated from all pairwise 
comparisons involving X.
For each cluster, we can then rank candidate markers based on one of these 
effect size summaries. We demonstrate below with the mean AUC for cluster 1.
We can then visualize the distribution of expression values across clusters for
the top potential marker genes (as determined by the mean AUC) using 
plotExpression() function from scater package.
In addition to pairwise comparisons one can compare each cluster to the average 
of all other cells. The latter approach however is sensitive to the population 
composition, which introduces an element of unpredictability to the marker sets 
due to variation in cell type abundances. (In the worst case, the presence of 
one subpopulation containing a majority of the cells will drive the selection 
of top markers for every other cluster, pushing out useful genes that can 
distinguish between the smaller subpopulations.) Moreover, pairwise comparisons 
naturally provide more information to interpret the utility of a marker, e.g., 
by providing log-fold changes to indicate which clusters are distinguished by 
each gene.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

#--- clustering ---#
g <- buildSNNGraph(sce.pbmc, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
colLabels(sce.pbmc) <- factor(clust)

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# pairwise comparison
marker.info <- scoreMarkers(sce.pbmc, colLabels(sce.pbmc))

# show compute scores data frames
marker.info

# statistics for cluster 1
colnames(marker.info[["1"]]) 

# select cluster 1
chosen <- marker.info[["1"]]

# rank candidate markers using mean AUC 
ordered <- chosen[order(chosen$mean.AUC, decreasing=TRUE),]

# showing basic stats only, for brevity
head(ordered[,1:4]) 

library(scater)
# visualization of expression values across clusters
plotExpression(sce.pbmc, features=head(rownames(ordered)), 
    x="label", colour_by="label")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to identify the genes that drive separation between clusters in my data.
Furthermore, I would like to know the ways I could interpret the results?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. Clustering can be then performed using buildSNNGraph() function 
from scran package.
The most straightforward approach to marker gene detection involves testing for 
differential expression (DE) between clusters. If a gene is strongly DE between 
clusters, it is likely to have driven the separation of cells in the clustering 
algorithm. Several methods are available to quantify the differences in 
expression profiles between clusters and obtain a single ranking of genes for 
each cluster. 
A general strategy is to compare each pair of clusters and compute scores 
quantifying the differences in the expression distributions between clusters. 
The scores for all pairwise comparisons involving a particular cluster are then 
consolidated into a single DataFrame for that cluster. The scoreMarkers() 
function from scran returns a list of DataFrames where each DataFrame 
corresponds to a cluster and each row of the DataFrame corresponds to a gene. 
In the DataFrame for cluster X, the columns contain the self.average, the mean 
log-expression in X; other.average, the grand mean across all other clusters; 
self.detected, the proportion of cells with detected expression in X; 
other.detected, the mean detected proportion across all other clusters; and 
finally, a variety of effect size summaries generated from all pairwise 
comparisons involving X.
For each cluster, we can then rank candidate markers based on one of these 
effect size summaries. 
In addition to pairwise comparisons one can compare each cluster to the average 
of all other cells. The latter approach however is sensitive to the population 
composition, which introduces an element of unpredictability to the marker sets 
due to variation in cell type abundances. (In the worst case, the presence of 
one subpopulation containing a majority of the cells will drive the selection 
of top markers for every other cluster, pushing out useful genes that can 
distinguish between the smaller subpopulations.) Moreover, pairwise comparisons 
naturally provide more information to interpret the utility of a marker, e.g., 
by providing log-fold changes to indicate which clusters are distinguished by 
each gene.
In the context of marker detection, the area under the curve (AUC) quantifies 
our ability to distinguish between two distributions in a pairwise comparison. 
The AUC represents the probability that a randomly chosen observation from our 
cluster of interest is greater than a randomly chosen observation from the other 
cluster. A value of 1 corresponds to upregulation, where all values of our 
cluster of interest are greater than any value from the other cluster; a value 
of 0.5 means that there is no net difference in the location of the 
distributions; and a value of 0 corresponds to downregulation. The AUC is 
closely related to the U statistic in the Wilcoxon ranked sum test (a.k.a., 
Mann-Whitney U-test).
Cohen’s d is a standardized log-fold change where the difference in the mean 
log-expression between groups is scaled by the average standard deviation across 
groups. In other words, it is the number of standard deviations that separate 
the means of the two groups. The interpretation is similar to the log-fold 
change; positive values indicate that the gene is upregulated in our cluster of 
interest, negative values indicate downregulation and values close to zero 
indicate that there is little difference. Cohen’s d is roughly analogous to the 
t-statistic in various two-sample t-tests.
Finally, we also compute the log-fold change in the proportion of cells with 
detected expression between clusters. This ignores any information about the 
magnitude of expression, only considering whether any expression is detected 
at all. Again, positive values indicate that a greater proportion of cells 
express the gene in our cluster of interest compared to the other cluster. Note 
that a pseudo-count is added to avoid undefined log-fold changes when no cells 
express the gene in either group. 
The AUC or Cohen’s d is usually the best choice for general purpose marker 
detection, as they are effective regardless of the magnitude of the expression 
values. The log-fold change in the detected proportion is specifically useful 
for identifying binary changes in expression.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

#--- clustering ---#
g <- buildSNNGraph(sce.pbmc, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
colLabels(sce.pbmc) <- factor(clust)

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# pairwise comparison
marker.info <- scoreMarkers(sce.pbmc, colLabels(sce.pbmc))

# show compute scores data frames
marker.info

# statistics for cluster 1
colnames(marker.info[["1"]]) 

# select cluster 1
chosen <- marker.info[["1"]]

# extract all AUCs 
auc.only <- chosen[,grepl("AUC", colnames(chosen))]

# show mean AUC
auc.only[order(auc.only$mean.AUC,decreasing=TRUE),]

# extract all log FC cohen 
cohen.only <- chosen[,grepl("logFC.cohen", colnames(chosen))]

# show mean logFC cohen
cohen.only[order(cohen.only$mean.logFC.cohen,decreasing=TRUE),]

# extract all log FC detected 
detect.only <- chosen[,grepl("logFC.detected", colnames(chosen))]

# show mean logFC detected
detect.only[order(detect.only$mean.logFC.detected,decreasing=TRUE),]

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to identify the genes that drive separation between clusters in my data.
In particular I would like to obtain the effect sizes for all pairwise 
comparisons involving a particular cluster.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. Clustering can be then performed using buildSNNGraph() function 
from scran package.
The most straightforward approach to marker gene detection involves testing for 
differential expression (DE) between clusters. If a gene is strongly DE between 
clusters, it is likely to have driven the separation of cells in the clustering 
algorithm. Several methods are available to quantify the differences in 
expression profiles between clusters and obtain a single ranking of genes for 
each cluster. 
A general strategy is to compare each pair of clusters and compute scores 
quantifying the differences in the expression distributions between clusters. 
The scores for all pairwise comparisons involving a particular cluster are then 
consolidated into a single DataFrame for that cluster. The scoreMarkers() 
function from scran returns a list of DataFrames where each DataFrame 
corresponds to a cluster and each row of the DataFrame corresponds to a gene. 
In the DataFrame for cluster X, the columns contain the self.average, the mean 
log-expression in X; other.average, the grand mean across all other clusters; 
self.detected, the proportion of cells with detected expression in X; 
other.detected, the mean detected proportion across all other clusters; and 
finally, a variety of effect size summaries generated from all pairwise 
comparisons involving X.
For each cluster, we can then rank candidate markers based on one of these 
effect size summaries. 
In addition to pairwise comparisons one can compare each cluster to the average 
of all other cells. The latter approach however is sensitive to the population 
composition, which introduces an element of unpredictability to the marker sets 
due to variation in cell type abundances. (In the worst case, the presence of 
one subpopulation containing a majority of the cells will drive the selection 
of top markers for every other cluster, pushing out useful genes that can 
distinguish between the smaller subpopulations.) Moreover, pairwise comparisons 
naturally provide more information to interpret the utility of a marker, e.g., 
by providing log-fold changes to indicate which clusters are distinguished by 
each gene.
In the context of marker detection, the area under the curve (AUC) quantifies 
our ability to distinguish between two distributions in a pairwise comparison. 
The AUC represents the probability that a randomly chosen observation from our 
cluster of interest is greater than a randomly chosen observation from the other 
cluster. A value of 1 corresponds to upregulation, where all values of our 
cluster of interest are greater than any value from the other cluster; a value 
of 0.5 means that there is no net difference in the location of the 
distributions; and a value of 0 corresponds to downregulation. The AUC is 
closely related to the U statistic in the Wilcoxon ranked sum test (a.k.a., 
Mann-Whitney U-test).
Cohen’s d is a standardized log-fold change where the difference in the mean 
log-expression between groups is scaled by the average standard deviation across 
groups. In other words, it is the number of standard deviations that separate 
the means of the two groups. The interpretation is similar to the log-fold 
change; positive values indicate that the gene is upregulated in our cluster of 
interest, negative values indicate downregulation and values close to zero 
indicate that there is little difference. Cohen’s d is roughly analogous to the 
t-statistic in various two-sample t-tests.
Finally, we also compute the log-fold change in the proportion of cells with 
detected expression between clusters. This ignores any information about the 
magnitude of expression, only considering whether any expression is detected 
at all. Again, positive values indicate that a greater proportion of cells 
express the gene in our cluster of interest compared to the other cluster. Note 
that a pseudo-count is added to avoid undefined log-fold changes when no cells 
express the gene in either group. 
The AUC or Cohen’s d is usually the best choice for general purpose marker 
detection, as they are effective regardless of the magnitude of the expression 
values. The log-fold change in the detected proportion is specifically useful 
for identifying binary changes in expression.
For more complex questions, we may need to interrogate effect sizes from 
specific comparisons of interest. To do so, we set full.stats=TRUE to obtain 
the effect sizes for all pairwise comparisons involving a particular cluster. 
This is returned in the form of a nested DataFrame for each effect size type - 
in the example below, full.AUC contains the AUCs for the comparisons between 
cluster 4 and every other cluster.
Say we want to identify the genes that distinguish cluster 4 from other clusters 
with high LYZ expression. We subset full.AUC to the relevant comparisons and 
sort on our summary statistic of choice to obtain a ranking of markers within 
this subset. This allows us to easily characterize subtle differences between 
closely related clusters. To illustrate, we use the smallest rank from 
computeMinRank() to identify the top DE genes in cluster 4 compared to the 
other LYZ-high clusters (Figure). The demo figure shoes the heatmap of the 
centered average log-expression values for the top potential marker genes for 
cluster 4 relative to other LYZ-high clusters in the PBMC dataset. The set of 
markers was selected as those genes with AUC-derived min-ranks less than or 
equal to 10.
Similarly, we can use the full set of effect sizes to define our own summary 
statistic if the precomputed measures are too coarse. For example, we may be 
interested in markers that are upregulated against some percentage - say, 80% - 
of other clusters. This improves the cluster specificity of the ranking by 
being more stringent than the median yet not as strigent as the minimum. We 
achieve this by computing and sorting on the 20th percentile of effect sizes.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

#--- clustering ---#
g <- buildSNNGraph(sce.pbmc, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
colLabels(sce.pbmc) <- factor(clust)

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# full effect for all pairwise comparisons
marker.info <- scoreMarkers(sce.pbmc, colLabels(sce.pbmc), full.stats=TRUE)

# select cluster 4
chosen <- marker.info[["4"]]

# show compute scores data frames
chosen$full.AUC

# lyz.high clusters based on inspection
lyz.high <- c("4", "6", "8", "9", "14") 

# subset lyz.high data
subset <- chosen$full.AUC[,colnames(chosen$full.AUC) %in% lyz.high]

# difference between clusters
to.show <- subset[computeMinRank(subset) <= 10,]

# show differences
to.show

# compute 20th percentile of effect sizes
stat <- rowQuantiles(as.matrix(chosen$full.AUC), p=0.2)

# just showing the basic stats for brevity
chosen[order(stat, decreasing=TRUE), 1:4] 

# visualize cluster 4 relative to other lyz-high clusters
plotGroupedHeatmap(sce.pbmc[,colLabels(sce.pbmc) %in% lyz.high],
    features=rownames(to.show), group="label", center=TRUE, zlim=c(-3, 3))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to identify the genes that drive separation between clusters in my data, 
specifically when variability is low.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. Clustering can be then performed using buildSNNGraph() function 
from scran package.
The most straightforward approach to marker gene detection involves testing for 
differential expression (DE) between clusters. If a gene is strongly DE between 
clusters, it is likely to have driven the separation of cells in the clustering 
algorithm. Several methods are available to quantify the differences in 
expression profiles between clusters and obtain a single ranking of genes for 
each cluster. 
A general strategy is to compare each pair of clusters and compute scores 
quantifying the differences in the expression distributions between clusters. 
The scores for all pairwise comparisons involving a particular cluster are then 
consolidated into a single DataFrame for that cluster. The scoreMarkers() 
function from scran returns a list of DataFrames where each DataFrame 
corresponds to a cluster and each row of the DataFrame corresponds to a gene. 
In the DataFrame for cluster X, the columns contain the self.average, the mean 
log-expression in X; other.average, the grand mean across all other clusters; 
self.detected, the proportion of cells with detected expression in X; 
other.detected, the mean detected proportion across all other clusters; and 
finally, a variety of effect size summaries generated from all pairwise 
comparisons involving X.
For each cluster, we can then rank candidate markers based on one of these 
effect size summaries. 
In addition to pairwise comparisons one can compare each cluster to the average 
of all other cells. The latter approach however is sensitive to the population 
composition, which introduces an element of unpredictability to the marker sets 
due to variation in cell type abundances. (In the worst case, the presence of 
one subpopulation containing a majority of the cells will drive the selection 
of top markers for every other cluster, pushing out useful genes that can 
distinguish between the smaller subpopulations.) Moreover, pairwise comparisons 
naturally provide more information to interpret the utility of a marker, e.g., 
by providing log-fold changes to indicate which clusters are distinguished by 
each gene.
In the context of marker detection, the area under the curve (AUC) quantifies 
our ability to distinguish between two distributions in a pairwise comparison. 
The AUC represents the probability that a randomly chosen observation from our 
cluster of interest is greater than a randomly chosen observation from the other 
cluster. A value of 1 corresponds to upregulation, where all values of our 
cluster of interest are greater than any value from the other cluster; a value 
of 0.5 means that there is no net difference in the location of the 
distributions; and a value of 0 corresponds to downregulation. The AUC is 
closely related to the U statistic in the Wilcoxon ranked sum test (a.k.a., 
Mann-Whitney U-test).
Cohen’s d is a standardized log-fold change where the difference in the mean 
log-expression between groups is scaled by the average standard deviation across 
groups. In other words, it is the number of standard deviations that separate 
the means of the two groups. The interpretation is similar to the log-fold 
change; positive values indicate that the gene is upregulated in our cluster of 
interest, negative values indicate downregulation and values close to zero 
indicate that there is little difference. Cohen’s d is roughly analogous to the 
t-statistic in various two-sample t-tests.
Finally, we also compute the log-fold change in the proportion of cells with 
detected expression between clusters. This ignores any information about the 
magnitude of expression, only considering whether any expression is detected 
at all. Again, positive values indicate that a greater proportion of cells 
express the gene in our cluster of interest compared to the other cluster. Note 
that a pseudo-count is added to avoid undefined log-fold changes when no cells 
express the gene in either group. 
The AUC or Cohen’s d is usually the best choice for general purpose marker 
detection, as they are effective regardless of the magnitude of the expression 
values. The log-fold change in the detected proportion is specifically useful 
for identifying binary changes in expression.
As mentioned, the Cohen’s d and AUC calculations consider both the magnitude of 
the difference between clusters as well as the variability within each cluster. 
If the variability is lower, it is possible for a gene to have a large effect 
size even if the magnitude of the difference is small. These genes tend to be 
somewhat uninformative for cell type identification despite their strong 
differential expression (e.g., ribosomal protein genes). We would prefer genes 
with larger log-fold changes between clusters, even if they have higher 
variability.
To favor the detection of such genes, we can compute the effect sizes relative 
to a log-fold change threshold by setting lfc= in scoreMarkers(). The definition 
of Cohen’s d is generalized to the standardized difference between the observed 
log-fold change and the specified lfc threshold. Similarly, the AUC is redefined 
as the probability of randomly picking an expression value from one cluster that 
is greater than a random value from the other cluster plus lfc. A large positive 
Cohen’s d and an AUC above 0.5 can only be obtained if the observed log-fold 
change between clusters is significantly greater than lfc. We demonstrate below 
by obtaining the top markers for cluster 5 in the PBMC dataset with lfc=2 
(Figure). The figure shows dot plot of the top potential marker genes (as 
determined by the mean AUC) for cluster 5 in the PBMC dataset. Each row 
corrresponds to a marker gene and each column corresponds to a cluster. The 
size of each dot represents the proportion of cells with detected expression of 
the gene in the cluster, while the color is proportional to the average 
expression across all cells in that cluster.
Note that the interpretation of the AUC and Cohen’s d becomes slightly more 
complicated when lfc is non-zero. If lfc is positive, a positive Cohen’s d and 
an AUC above 0.5 represents upregulation. However, a negative Cohen’s d or AUC 
below 0.5 may not represent downregulation; it may just indicate that the 
observed log-fold change is less than the specified lfc. The converse applies 
when lfc is negative, where the only conclusive interpretation occurs for 
downregulated genes. For the most part, this complication is not too problematic 
for routine marker detection, as we are mostly interested in upregulated genes 
with large positive Cohen’s d and AUCs above 0.5.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS BELOW 

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

#--- dimensionality-reduction ---#
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

#--- clustering ---#
g <- buildSNNGraph(sce.pbmc, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
colLabels(sce.pbmc) <- factor(clust)

## FOR DEMO PURPOSE WE'LL USE THE 10X PBMC DATASET CREATED AS ABOVE

library(scran)
# compute effect sizes relative to lfc threshold
marker.info.lfc <- scoreMarkers(sce.pbmc, colLabels(sce.pbmc), lfc=2)

# select cluster 5
chosen2 <- marker.info.lfc[["5"]] 

# extract mean AUC
chosen2 <- chosen2[order(chosen2$mean.AUC, decreasing=TRUE),]

# show mean AUC
chosen2[,c("self.average", "other.average", "mean.AUC")]

# visualize top potential marker genes for cluster 5
plotDots(sce.pbmc, rownames(chosen2)[1:10], group="label")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a SingleCellExperiment object of my single cell data that also contain 
log transformed normalized expression values, variance per gene, and 
representation of top highly variable genes (HVGs) in lower dimension. I would 
like to identify the genes that drive separation between clusters in my data,
taking in account known and uninteresting factors of variation.
How can I do this?

completion:
The simplest approach to quantifying per-gene variation is to compute the 
variance of the log-normalized expression values (i.e., “log-counts”) for each 
gene across all cells using modelGeneVar() or modelGeneVarByPoisson() function 
from the scran package. Highly variable genes(HVGs) on which to perform 
dimensionality reduction can be selected with getTopHVgs() function from scran 
package and then dimensions reduced by runTSNE() and runUMAP() functions from 
scater package. Clustering can be then performed using hclust() function 
from stats package.
The most straightforward approach to marker gene detection involves testing for 
differential expression (DE) between clusters. If a gene is strongly DE between 
clusters, it is likely to have driven the separation of cells in the clustering 
algorithm. Several methods are available to quantify the differences in 
expression profiles between clusters and obtain a single ranking of genes for 
each cluster. 
A general strategy is to compare each pair of clusters and compute scores 
quantifying the differences in the expression distributions between clusters. 
The scores for all pairwise comparisons involving a particular cluster are then 
consolidated into a single DataFrame for that cluster. The scoreMarkers() 
function from scran returns a list of DataFrames where each DataFrame 
corresponds to a cluster and each row of the DataFrame corresponds to a gene. 
In the DataFrame for cluster X, the columns contain the self.average, the mean 
log-expression in X; other.average, the grand mean across all other clusters; 
self.detected, the proportion of cells with detected expression in X; 
other.detected, the mean detected proportion across all other clusters; and 
finally, a variety of effect size summaries generated from all pairwise 
comparisons involving X.
For each cluster, we can then rank candidate markers based on one of these 
effect size summaries. 
In addition to pairwise comparisons one can compare each cluster to the average 
of all other cells. The latter approach however is sensitive to the population 
composition, which introduces an element of unpredictability to the marker sets 
due to variation in cell type abundances. (In the worst case, the presence of 
one subpopulation containing a majority of the cells will drive the selection 
of top markers for every other cluster, pushing out useful genes that can 
distinguish between the smaller subpopulations.) Moreover, pairwise comparisons 
naturally provide more information to interpret the utility of a marker, e.g., 
by providing log-fold changes to indicate which clusters are distinguished by 
each gene.
In the context of marker detection, the area under the curve (AUC) quantifies 
our ability to distinguish between two distributions in a pairwise comparison. 
The AUC represents the probability that a randomly chosen observation from our 
cluster of interest is greater than a randomly chosen observation from the other 
cluster. A value of 1 corresponds to upregulation, where all values of our 
cluster of interest are greater than any value from the other cluster; a value 
of 0.5 means that there is no net difference in the location of the 
distributions; and a value of 0 corresponds to downregulation. The AUC is 
closely related to the U statistic in the Wilcoxon ranked sum test (a.k.a., 
Mann-Whitney U-test).
Cohen’s d is a standardized log-fold change where the difference in the mean 
log-expression between groups is scaled by the average standard deviation across 
groups. In other words, it is the number of standard deviations that separate 
the means of the two groups. The interpretation is similar to the log-fold 
change; positive values indicate that the gene is upregulated in our cluster of 
interest, negative values indicate downregulation and values close to zero 
indicate that there is little difference. Cohen’s d is roughly analogous to the 
t-statistic in various two-sample t-tests.
Finally, we also compute the log-fold change in the proportion of cells with 
detected expression between clusters. This ignores any information about the 
magnitude of expression, only considering whether any expression is detected 
at all. Again, positive values indicate that a greater proportion of cells 
express the gene in our cluster of interest compared to the other cluster. Note 
that a pseudo-count is added to avoid undefined log-fold changes when no cells 
express the gene in either group. 
The AUC or Cohen’s d is usually the best choice for general purpose marker 
detection, as they are effective regardless of the magnitude of the expression 
values. The log-fold change in the detected proportion is specifically useful 
for identifying binary changes in expression.
Large studies may contain factors of variation that are known and not 
interesting (e.g., batch effects, sex differences). If these are not modelled, 
they can interfere with marker gene detection - most obviously by inflating the 
variance within each cluster, but also by distorting the log-fold changes if 
the cluster composition varies across levels of the blocking factor. To avoid 
these issues, we specify the blocking factor via the block= argument.
For each gene, each pairwise comparison between clusters is performed separately 
in each level of the blocking factor - in the demo case, the plate of origin. By 
comparing within each batch, we cancel out any batch effects so that they are 
not conflated with the biological differences between subpopulations. The effect 
sizes are then averaged across batches to obtain a single value per comparison, 
using a weighted mean that accounts for the number of cells involved in the 
comparison in each batch. A similar correction is applied to the mean 
log-expression and proportion of detected cells inside and outside each cluster.
We can then visualize the distribution of expression values across clusters for
the top potential marker genes (as shown, from cluster 1 in the 416b dataset) 
using plotExpression() function from scater package. In the demo figure each 
point represents a cell and is colored by the batch of origin.
The block= argument works for all effect sizes shown above and is robust to 
differences in the log-fold changes or variance between batches. However, it 
assumes that each pair of clusters is present in at least one batch. In 
scenarios where cells from two clusters never co-occur in the same batch, the 
associated pairwise comparison will be impossible and is ignored during 
calculation of summary statistics.

```{r,eval=FALSE}

## FOR DEMO PURPOSE WE'LL USE THE 416b DATASET CREATED AS BELOW 

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

#--- variance-modelling ---#
dec.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC", block=sce.416b$block)
chosen.hvgs <- getTopHVGs(dec.416b, prop=0.1)

#--- batch-correction ---#
library(limma)
assay(sce.416b, "corrected") <- removeBatchEffect(logcounts(sce.416b), 
    design=model.matrix(~sce.416b$phenotype), batch=sce.416b$block)

#--- dimensionality-reduction ---#
sce.416b <- runPCA(sce.416b, ncomponents=10, subset_row=chosen.hvgs,
    exprs_values="corrected", BSPARAM=BiocSingular::ExactParam())

set.seed(1010)
sce.416b <- runTSNE(sce.416b, dimred="PCA", perplexity=10)

#--- clustering ---#
my.dist <- dist(reducedDim(sce.416b, "PCA"))
my.tree <- hclust(my.dist, method="ward.D2")

library(dynamicTreeCut)
my.clusters <- unname(cutreeDynamic(my.tree, distM=as.matrix(my.dist),
    minClusterSize=10, verbose=0))
colLabels(sce.416b) <- factor(my.clusters)

## FOR DEMO PURPOSE WE'LL USE THE 416b DATASET CREATED AS ABOVE

library(scran)
# pairwise comparison with blocking factor
m.out <- scoreMarkers(sce.416b, colLabels(sce.416b), block=sce.416b$block)

# select cluster 1
demo <- m.out[["1"]] 

# order and extract median log FC cohen 
ordered <- demo[order(demo$median.logFC.cohen, decreasing=TRUE),]

# show data
ordered[,1:4]

library(scater)
# visualization of expression values across clusters
plotExpression(sce.416b, features=rownames(ordered)[1:6],
    x="label", colour_by="block")

```

--- TASK END
