---
title: "prompt/completion pairs single-cell_advanced_chapter 2"
author: "authors"
date: "2024-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I would like to apply quality 
control metrics. While scaling normalization of scRNA-seq data what are the 
reasons for adding pseudo-count? 

completion:
When log-transforming, logNormCounts() will add a pseudo-count to avoid 
undefined values at zero. Larger pseudo-counts will shrink the log-fold changes 
between cells towards zero for low-abundance genes, meaning that downstream 
high-dimensional analyses will be driven more by differences in expression for 
high-abundance genes. Conversely, smaller pseudo-counts will increase the 
relative contribution of low-abundance genes. Common practice is to use a 
pseudo-count of 1, for the simple pragmatic reason that it preserves sparsity 
in the original matrix (i.e., zeroes in the input remain zeroes after 
transformation). This works well in all but the most pathological scenarios.
An interesting subtlety of logNormCounts() is that it will center the size 
factors at unity, if they were not already. This puts the normalized 
expression values on roughly the same scale as the original counts for easier 
interpretation. For example, Figure1 shows that interneurons have a median 
Snap25 log-expression from 5-6; this roughly translates to an original count 
of 30-60 UMIs in each cell, which gives us some confidence that it is actually 
expressed. This relationship to the original data would be less obvious - or 
indeed, lost altogether - if the centering were not performed. "Figure1: 
Distribution of log-expression values for Snap25 in each cell type of the 
Zeisel brain dataset."
Centering also allows us to interpret a pseudo-count of 1 as an extra read or 
UMI for each gene. In practical terms, this means that the shrinkage effect of 
the pseudo-count diminishes as read/UMI counts increase. As a result, any 
estimates of log-fold changes in expression (e.g., from differences in the 
log-values between groups of cells) become increasingly accurate with deeper 
coverage. Conversely, at lower counts, stronger shrinkage avoids inflated 
differences due to sampling noise, which might otherwise mask interesting 
features in downstream analyses like clustering. In some sense, the 
pseudo-count aims to protect later analyses from the lack of information at 
low counts while trying to miminize its own effect at high counts.
For comparison, consider the situation where we applied a constant pseudo-count 
to some count-per-million-like measure. It is easy to see that the accuracy of 
the subsequent log-fold changes would never improve regardless of how much 
additional sequencing was performed; scaling to a constant library size of a 
million means that the pseudo-count will have the same effect for all datasets. 
This is ironic given that the whole intention of sequencing more deeply is to 
improve quantification of these differences between cell subpopulations. The 
same criticism applies to popular metrics like the “counts-per-10K”.

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

library(scuttle)
library(scater)
sce.zeisel <- logNormCounts(sce.zeisel)
plotExpression(sce.zeisel, x="level1class", features="Snap25", colour="level1class")

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I would like to apply quality 
control metrics. I would like to know when is scaling normalization of 
scRNA-seq data not appropriate and what can I do instead in such case?

completion:
In rare cases, direct scaling of the counts is not appropriate due to the 
effect described by A. Lun (2018). Briefly, this is caused by the fact that 
the mean of the log-normalized counts is not the same as the log-transformed 
mean of the normalized counts. The difference between them depends on the mean 
and variance of the original counts, such that there is a systematic trend in 
the mean of the log-counts with respect to the count size. This typically 
manifests as trajectories correlated strongly with library size even after 
library size normalization, as shown in Figure1 for synthetic scRNA-seq data 
generated with a pool-and-split approach (Tian et al. 2019). "Figure1: PCA plot 
of all pool-and-split libraries in the SORT-seq CellBench data, computed from 
the log-normalized expression values with library size-derived size factors. 
Each point represents a library and is colored by the mixing ratio used to 
construct it (left) or by the size factor (right)."
As the problem arises from differences in the sizes of the counts, the most 
straightforward solution is to downsample the counts of the high-coverage cells 
to match those of low-coverage cells. This uses the size factors to determine 
the amount of downsampling for each cell required to reach the 1st percentile 
of size factors. (The small minority of cells with smaller size factors are 
simply scaled up. We do not attempt to downsample to the smallest size factor, 
as this would result in excessive loss of information for one aberrant cell 
with very low size factors.) We can see that this eliminates the library size 
factor-associated trajectories from the first two PCs, improving resolution of 
the known differences based on mixing ratios (Figure2). The log-transformation 
is still necessary but no longer introduces a shift in the means when the sizes 
of the counts are similar across cells. "Figure2: PCA plot of pool-and-split 
libraries in the SORT-seq CellBench data, computed from the log-transformed 
counts after downsampling in proportion to the library size factors. Each point 
represents a library and is colored by the mixing ratio used to construct it 
(left) or by the size factor (right)."
While downsampling is an expedient solution, it is statistically inefficient as 
it needs to increase the noise of high-coverage cells in order to avoid 
differences with low-coverage cells. It is also slower than simple scaling. 
Thus, we would only recommend using this approach after an initial analysis 
with scaled counts reveals suspicious trajectories that are strongly correlated 
with the size factors. In such cases, it is a simple matter to re-normalize by 
downsampling to determine whether the trajectory is an artifact of the 
log-transformation.

```{r,eval=FALSE}

# TODO: move to scRNAseq.
library(BiocFileCache)
bfc <- BiocFileCache(ask=FALSE)
qcdata <- bfcrpath(bfc, "https://github.com/LuyiTian/CellBench_data/blob/master/data/mRNAmix_qc.RData?raw=true")

env <- new.env()
load(qcdata, envir=env)
sce.8qc <- env$sce8_qc

# Library size normalization and log-transformation.
sce.8qc <- logNormCounts(sce.8qc)
sce.8qc <- runPCA(sce.8qc)
gridExtra::grid.arrange(
    plotPCA(sce.8qc, colour_by=I(factor(sce.8qc$mix))),
    plotPCA(sce.8qc, colour_by=I(librarySizeFactors(sce.8qc))),
    ncol=2
)

# downsampling
sce.8qc2 <- logNormCounts(sce.8qc, downsample=TRUE)
sce.8qc2 <- runPCA(sce.8qc2)
gridExtra::grid.arrange(
    plotPCA(sce.8qc2, colour_by=I(factor(sce.8qc2$mix))),
    plotPCA(sce.8qc2, colour_by=I(librarySizeFactors(sce.8qc2))),
    ncol=2
)

```

--- TASK END