---
title: "prompt/completion pairs single-cell_advanced_chapter 3"
author: "authors"
date: "2024-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset in which I would like to perform feature
selection. While selecting features, how can I fine-tune the fitted trend?

completion:
The trend fit has several useful parameters that can be tuned for a more 
appropriate fit. For example, the defaults can occasionally yield an overfitted 
trend when the few high-abundance genes are also highly variable. In such cases, 
users can reduce the contribution of those high-abundance genes by turning off 
density weights, as demonstrated in Figure with a single donor from the 
Segerstolpe et al. (2016) dataset.
"Figure: Variance in the Segerstolpe pancreas data set as a function of the 
mean. Each point represents a gene while the lines represent the trend fitted 
to all genes with default parameters (blue) or without weights (red)."

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.seger <- SegerstolpePancreasData()

#--- gene-annotation ---#
library(AnnotationHub)
edb <- AnnotationHub()[["AH73881"]]
symbols <- rowData(sce.seger)$symbol
ens.id <- mapIds(edb, keys=symbols, keytype="SYMBOL", column="GENEID")
ens.id <- ifelse(is.na(ens.id), symbols, ens.id)

# Removing duplicated rows.
keep <- !duplicated(ens.id)
sce.seger <- sce.seger[keep,]
rownames(sce.seger) <- ens.id[keep]

#--- sample-annotation ---#
emtab.meta <- colData(sce.seger)[,c("cell type", "disease",
    "individual", "single cell well quality")]
colnames(emtab.meta) <- c("CellType", "Disease", "Donor", "Quality")
colData(sce.seger) <- emtab.meta

sce.seger$CellType <- gsub(" cell", "", sce.seger$CellType)
sce.seger$CellType <- paste0(
    toupper(substr(sce.seger$CellType, 1, 1)),
    substring(sce.seger$CellType, 2))

#--- quality-control ---#
low.qual <- sce.seger$Quality == "low quality cell"

library(scater)
stats <- perCellQCMetrics(sce.seger)
qc <- quickPerCellQC(stats, percent_subsets="altexps_ERCC_percent",
    batch=sce.seger$Donor,
    subset=!sce.seger$Donor %in% c("HP1504901", "HP1509101"))

sce.seger <- sce.seger[,!(qc$discard | low.qual)]

#--- normalization ---#
library(scran)
clusters <- quickCluster(sce.seger)
sce.seger <- computeSumFactors(sce.seger, clusters=clusters)
sce.seger <- logNormCounts(sce.seger) 

library(scran)
sce.seger <- sce.seger[,sce.seger$Donor=="HP1507101"]
dec.default <- modelGeneVar(sce.seger)
dec.noweight <- modelGeneVar(sce.seger, density.weights=FALSE)

fit.default <- metadata(dec.default)
plot(fit.default$mean, fit.default$var, xlab="Mean of log-expression",
    ylab="Variance of log-expression") 
curve(fit.default$trend(x), col="dodgerblue", add=TRUE, lwd=2)
fit.noweight <- metadata(dec.noweight)
curve(fit.noweight$trend(x), col="red", add=TRUE, lwd=2)
legend("topleft", col=c("dodgerblue", "red"), legend=c("Default", "No weight"), lwd=2)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset in which I would like to perform feature
selection. While selecting features, how do I handle covariates?

completion:
For experiments with multiple batches, the use of block-specific trends with 
block= in modelGeneVar() is the recommended approach for avoiding unwanted 
variation. However, this is not possible for experimental designs involving 
multiple unwanted factors of variation and/or continuous covariates. In such 
cases, we can use the design= argument to specify a design matrix with 
uninteresting factors of variation. This fits a linear model to the expression 
values for each gene to obtain the residual variance. We illustrate this with 
the 416B data set (The Lun et al. (2017)), blocking on the plate of origin and 
oncogene induction. (The same argument is available in modelGeneVar() when 
spike-ins are not available.)
This strategy is simple but somewhat inaccurate as it does not consider the 
mean expression in each blocking level. Imagine that we have two genes with 
variances lying on this trend; the first gene has an average expression of 0 
in one batch and 6 in the other batch, while the second gene with an average 
expression of 3 in both batches. Both genes would have the same mean across 
all cells but quite different variances, making it difficult to fit a single 
mean-variance trend - despite both genes following the mean-variance trend in 
each of their respective batches! The block= approach is safer as it handles 
the trend fitting and decomposition within each batch, and should be preferred 
in all situations where it is applicable.

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.416b <- LunSpikeInData(which="416b") 
sce.416b$block <- factor(sce.416b$block)

#--- gene-annotation ---#
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.416b)$ENSEMBL <- rownames(sce.416b)
rowData(sce.416b)$SYMBOL <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SYMBOL")
rowData(sce.416b)$SEQNAME <- mapIds(ens.mm.v97, keys=rownames(sce.416b),
    keytype="GENEID", column="SEQNAME")

library(scater)
rownames(sce.416b) <- uniquifyFeatureNames(rowData(sce.416b)$ENSEMBL, 
    rowData(sce.416b)$SYMBOL)

#--- quality-control ---#
mito <- which(rowData(sce.416b)$SEQNAME=="MT")
stats <- perCellQCMetrics(sce.416b, subsets=list(Mt=mito))
qc <- quickPerCellQC(stats, percent_subsets=c("subsets_Mt_percent",
    "altexps_ERCC_percent"), batch=sce.416b$block)
sce.416b <- sce.416b[,!qc$discard]

#--- normalization ---#
library(scran)
sce.416b <- computeSumFactors(sce.416b)
sce.416b <- logNormCounts(sce.416b)

design <- model.matrix(~factor(block) + phenotype, colData(sce.416b))
dec.design.416b <- modelGeneVarWithSpikes(sce.416b, "ERCC", design=design)
dec.design.416b[order(dec.design.416b$bio, decreasing=TRUE),]

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset in which I would like to perform feature
selection. In addition to using linear models, are their other ways to handling 
covariates while selecting features?

completion:
An alternative approach to quantification uses the squared coefficient of 
variation (CV2) of the normalized expression values prior to log-transformation. 
The CV2 is a widely used metric for describing variation in non-negative data 
and is closely related to the dispersion parameter of the negative binomial 
distribution in packages like edgeR and DESeq2. We compute the CV2 for each 
gene in the human PBMC 10X genomics dataset using the modelGeneCV2() function, 
which provides a robust implementation of the approach described by 
Brennecke et al. (2013).
This allows us to model the mean-variance relationship when considering the 
relevance of each gene (Figure). Again, our assumption is that most genes 
contain random noise and that the trend captures mostly technical variation. 
Large CV2 values that deviate strongly from the trend are likely to represent 
genes affected by biological structure. If spike-ins are available, we can also 
fit the trend to the spike-ins via the modelGeneCV2WithSpikes() function.
"Figure: CV2 in the PBMC data set as a function of the mean. Each point 
represents a gene while the blue line represents the fitted trend."
For each gene, we quantify the deviation from the trend in terms of the ratio 
of its CV2 to the fitted value of trend at its abundance. This is more 
appropriate than directly subtracting the trend from the CV2, as the magnitude 
of the ratio is not affected by the mean.
We can then select HVGs based on the largest ratios using getTopHVGs().
Both the CV2 and the variance of log-counts are effective metrics for 
quantifying variation in gene expression. The CV2 tends to give higher rank to 
low-abundance HVGs driven by upregulation in rare subpopulations, for which the 
increase in variance on the raw scale is stronger than that on the log-scale. 
However, the variation described by the CV2 is less directly relevant to 
downstream procedures operating on the log-counts, and the reliance on the 
ratio can assign high rank to uninteresting genes with low absolute variance. 
As such, we prefer to use the variance of log-counts for feature selection, 
though many of the same principles apply to procedures based on the CV2.

```{r,eval=FALSE}

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

dec.cv2.pbmc <- modelGeneCV2(sce.pbmc)

fit.cv2.pbmc <- metadata(dec.cv2.pbmc)
plot(fit.cv2.pbmc$mean, fit.cv2.pbmc$cv2, log="xy")
curve(fit.cv2.pbmc$trend(x), col="dodgerblue", add=TRUE, lwd=2)

dec.cv2.pbmc[order(dec.cv2.pbmc$ratio, decreasing=TRUE),]

hvg.pbmc.cv2 <- getTopHVGs(dec.cv2.pbmc, var.field="ratio", n=1000)
str(hvg.pbmc.cv2)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset in which I would like to perform feature
selection. While selecting features (i.e., HVG), I would like to keep all genes
above the trend. How can I do this?

completion:
Here, the aim is to only remove the obviously uninteresting genes with 
variances below the trend. By doing so, we avoid the need to make any judgement 
calls regarding what level of variation is interesting enough to retain. This 
approach represents one extreme of the bias-variance trade-off where bias is 
minimized at the cost of maximizing noise. For modelGeneVar(), it equates to 
keeping all positive biological components:
For modelGeneCV2(), this involves keeping all ratios above 1:
By retaining all potential biological signal, we give secondary population 
structure the chance to manifest. This is most useful for rare subpopulations 
where the relevant markers will not exhibit strong overdispersion owing to the 
small number of affected cells. It will also preserve a weak but consistent 
effect across many genes with small biological components; admittedly, though, 
this is not of major interest in most scRNA-seq studies given the difficulty of 
experimentally validating population structure in the absence of strong marker 
genes.
The obvious cost is that more noise is also captured, which can reduce the 
resolution of otherwise well-separated populations and mask the secondary 
signal that we were trying to preserve. The use of more genes also introduces 
more computational work in each downstream step. This strategy is thus best 
suited to very heterogeneous populations containing many different cell types 
(possibly across many datasets that are to be merged) where there is a 
justified fear of ignoring marker genes for low-abundance subpopulations under 
a competitive top X approach.

```{r,eval=FALSE}

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

dec.pbmc <- modelGeneVar(sce.pbmc)
hvg.pbmc.var.3 <- getTopHVGs(dec.pbmc, var.threshold=0)
length(hvg.pbmc.var.3)

dec.cv2.pbmc <- modelGeneCV2(sce.pbmc)
hvg.pbmc.cv2.3 <- getTopHVGs(dec.cv2.pbmc, var.field="ratio", var.threshold=1)
length(hvg.pbmc.cv2.3)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset in which I would like to perform feature
selection. In addition to using linear models and coefficient of variance, are 
their other ways to handling covariates while selecting features?

completion:
Another approach to feature selection is to set a fixed threshold of one of the 
metrics. This is most commonly done with the (adjusted) p-value reported by 
each of the linear model, coefficient of covariance and top X methods. The 
p-value for each gene is generated by testing against the null hypothesis that 
the variance is equal to the trend. For example, we might define our HVGs as 
all genes that have adjusted p-values below 0.05.
This approach is simple to implement and - if the test holds its size - it 
controls the false discovery rate (FDR). That is, it returns a subset of genes 
where the proportion of false positives is expected to be below the specified 
threshold. This can occasionally be useful in applications where the HVGs 
themselves are of interest. For example, if we were to use the list of HVGs in 
further experiments to verify the existence of heterogeneous expression for 
some of the genes, we would want to control the FDR in that list.
The downside of this approach is that it is less predictable than the top X 
strategy. The number of genes returned depends on the type II error rate of the 
test and the severity of the multiple testing correction. One might obtain no 
genes or every gene at a given FDR threshold, depending on the circumstances. 
Moreover, control of the FDR is usually not helpful at this stage of the 
analysis. We are not interpreting the individual HVGs themselves but are only 
using them for feature selection prior to downstream steps. There is no reason 
to think that a 5% threshold on the FDR yields a more suitable compromise 
between bias and noise compared to the top X selection.
As an aside, we might consider ranking genes by the p-value instead of the 
biological component for use in a top X approach. This results in some 
counterintuitive behavior due to the nature of the underlying hypothesis test, 
which is based on the ratio of the total variance to the expected technical 
variance. Ranking based on p-value tends to prioritize HVGs that are more 
likely to be true positives but, at the same time, less likely to be 
biologically interesting. Many of the largest ratios are observed in 
high-abundance genes and are driven by very low technical variance; the total 
variance is typically modest for such genes, and they do not contribute much 
to population heterogeneity in absolute terms. (Note that the same can be said 
of the ratio of CV2 values.)

```{r,eval=FALSE}

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

dec.pbmc <- modelGeneVar(sce.pbmc)
hvg.pbmc.var.2 <- getTopHVGs(dec.pbmc, fdr.threshold=0.05)
length(hvg.pbmc.var.2)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset in which I would like to perform feature
selection. While selecting features, I would like to select a priori genes
of interest. How can I do this?

completion:
A blunt yet effective feature selection strategy is to use pre-defined sets of 
interesting genes. The aim is to focus on specific aspects of biological 
heterogeneity that may be masked by other factors when using unsupervised 
methods for HVG selection. One example application lies in the dissection of 
transcriptional changes during the earliest stages of cell fate commitment, 
which may be modest relative to activity in other pathways (e.g., cell cycle, 
metabolism). Indeed, if our aim is to show that there is no meaningful 
heterogeneity in a given pathway, we would - at the very least - be obliged to 
repeat our analysis using only the genes in that pathway to maximize power for 
detecting such heterogeneity.
Using scRNA-seq data in this manner is conceptually equivalent to a fluorescence 
activated cell sorting (FACS) experiment, with the convenience of being able to 
(re)define the features of interest at any time. For example, in the human PBMC 
10X genomics dataset, we might use some of the C7 immunologic signatures from 
MSigDB to improve resolution of the various T cell subtypes. We stress that 
there is no shame in leveraging prior biological knowledge to address specific 
hypotheses in this manner. We say this because a common refrain in genomics is 
that the data analysis should be “unbiased”, i.e., free from any biological 
preconceptions. This is admirable but such “biases” are already present at 
every stage, starting with experimental design and ending with the 
interpretation of the data.
Of course, the downside of focusing on pre-defined genes is that it will limit 
our capacity to detect novel or unexpected aspects of variation. Thus, this 
kind of focused analysis should be complementary to (rather than a replacement 
for) the unsupervised feature selection strategies.
Alternatively, we can invert this reasoning to remove genes that are unlikely 
to be of interest prior to downstream analyses. This eliminates unwanted 
variation that could mask relevant biology and interfere with interpretation 
of the results. Ribosomal protein genes or mitochondrial genes are common 
candidates for removal, especially in situations with varying levels of cell 
damage within a population. For immune cell subsets, we might also be inclined 
to remove immunoglobulin genes and T cell receptor genes for which clonal 
expression introduces (possibly irrelevant) population structure.
In practice, we tend to err on the side of caution and abstain from preemptive 
filtering on biological function until these genes are demonstrably problematic 
in downstream analyses.

```{r,eval=FALSE}

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

library(msigdbr)
c7.sets <- msigdbr(species = "Homo sapiens", category = "C7")
head(unique(c7.sets$gs_name))

# Using the Goldrath sets to distinguish CD8 subtypes
cd8.sets <- c7.sets[grep("GOLDRATH", c7.sets$gs_name),]
cd8.genes <- rowData(sce.pbmc)$Symbol %in% cd8.sets$human_gene_symbol
summary(cd8.genes)

# Using GSE11924 to distinguish between T helper subtypes
th.sets <- c7.sets[grep("GSE11924", c7.sets$gs_name),]
th.genes <- rowData(sce.pbmc)$Symbol %in% th.sets$human_gene_symbol
summary(th.genes)

# Using GSE11961 to distinguish between B cell subtypes
b.sets <- c7.sets[grep("GSE11961", c7.sets$gs_name),]
b.genes <- rowData(sce.pbmc)$Symbol %in% b.sets$human_gene_symbol
summary(b.genes)

# Identifying ribosomal proteins:
ribo.discard <- grepl("^RP[SL]\\d+", rownames(sce.pbmc))
sum(ribo.discard)

# A more curated approach for identifying ribosomal protein genes:
c2.sets <- msigdbr(species = "Homo sapiens", category = "C2")
ribo.set <- c2.sets[c2.sets$gs_name=="KEGG_RIBOSOME",]$human_gene_symbol
ribo.discard <- rownames(sce.pbmc) %in% ribo.set
sum(ribo.discard)

library(AnnotationHub)
edb <- AnnotationHub()[["AH73881"]]
anno <- select(edb, keys=rowData(sce.pbmc)$ID, keytype="GENEID", 
    columns="TXBIOTYPE")

# Removing immunoglobulin variable chains:
igv.set <- anno$GENEID[anno$TXBIOTYPE %in% c("IG_V_gene", "IG_V_pseudogene")]
igv.discard <- rowData(sce.pbmc)$ID %in% igv.set
sum(igv.discard)

# Removing TCR variable chains:
tcr.set <- anno$GENEID[anno$TXBIOTYPE %in% c("TR_V_gene", "TR_V_pseudogene")]
tcr.discard <- rowData(sce.pbmc)$ID %in% tcr.set
sum(tcr.discard)

```

--- TASK END