---
title: "prompt/completion pairs single-cell_advanced_chapter 4"
author: "authors"
date: "2024-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
i<-1
```

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I have already applied quality 
control, normalization and computed variance. While performing principal 
component analysis (PCA) to reduce the dimensions of the data how do I choose 
the number of principal components (PCs)?

completion:
A simple heuristic for choosing the suitable number of principal components 
(PCs) d involves identifying the elbow point in the percentage of variance 
explained by successive PCs. This refers to the “elbow” in the curve of a 
scree plot as shown in Figure. "Figure: Percentage of variance explained by 
successive PCs in the Zeisel et al. (2015) brain dataset. The identified elbow 
point is marked with a red line."
Our assumption is that each of the top PCs capturing biological signal should 
explain much more variance than the remaining PCs. Thus, there should be a 
sharp drop in the percentage of variance explained when we move past the last 
“biological” PC. This manifests as an elbow in the scree plot, the location of 
which serves as a natural choice for d. Once this is identified, we can subset 
the reducedDims() entry to only retain the first d PCs of interest.
From a practical perspective, the use of the elbow point tends to retain fewer 
PCs compared to other methods. The definition of “much more variance” is 
relative so, in order to be retained, later PCs must explain a amount of 
variance that is comparable to that explained by the first few PCs. Strong 
biological variation in the early PCs will shift the elbow to the left, 
potentially excluding weaker (but still interesting) variation in the next 
PCs immediately following the elbow.

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)
set.seed(100) 
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

# Percentage of variance explained is tucked away in the attributes.
percent.var <- attr(reducedDim(sce.zeisel), "percentVar")

library(PCAtools)
chosen.elbow <- findElbowPoint(percent.var)
chosen.elbow

plot(percent.var, xlab="PC", ylab="Variance explained (%)") 
abline(v=chosen.elbow, col="red")

# Creating a new entry with only the first 20 PCs, 
# which is useful if we still need the full set of PCs later. 
reducedDim(sce.zeisel, "PCA.elbow") <- reducedDim(sce.zeisel)[,1:chosen.elbow]
reducedDimNames(sce.zeisel)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I have already applied quality 
control, normalization and computed variance. While performing principal 
component analysis (PCA) to reduce the dimensions of the data, is their another 
strategy to choose the number of principal components (PCs), other than the 
elbow method?

completion:
Another strategy is to retain all PCs until the percentage of total variation 
explained reaches some threshold T. For example, we might retain the top set 
of PCs that explains 80% of the total variation in the data. Of course, it 
would be pointless to swap one arbitrary parameter d for another T. Instead, 
we derive a suitable value for T by calculating the proportion of variance in 
the data that is attributed to the biological component. This is done using 
the denoisePCA() function with the variance modelling results from 
modelGeneVarWithSpikes() or related functions, where T is defined as the ratio 
of the sum of the biological components to the sum of total variances. To 
illustrate, we use this strategy to pick the number of PCs in the 10X PBMC 
dataset.
The dimensionality of the output represents the lower bound on the number of 
PCs required to retain all biological variation. This choice of d is motivated 
by the fact that any fewer PCs will definitely discard some aspect of biological 
signal. (Of course, the converse is not true; there is no guarantee that the 
retained PCs capture all of the signal, which is only generally possible if no 
dimensionality reduction is performed at all.) From a practical perspective, 
the denoisePCA() approach usually retains more PCs than the elbow point method 
as the former does not compare PCs to each other and is less likely to discard 
PCs corresponding to secondary factors of variation. The downside is that many 
minor aspects of variation may not be interesting (e.g., transcriptional 
bursting) and their retention would only add irrelevant noise.
Note that denoisePCA() imposes internal caps on the number of PCs that can be 
chosen in this manner. By default, the number is bounded within the “reasonable” 
limits of 5 and 50 to avoid selection of too few PCs (when technical noise is 
high relative to biological variation) or too many PCs (when technical noise is 
very low). For example, applying this function to the Zeisel et al. (2015) 
brain dataset hits the upper limit.
This method also tends to perform best when the mean-variance trend reflects 
the actual technical noise, i.e., estimated by modelGeneVarByPoisson() or 
modelGeneVarWithSpikes() instead of modelGeneVar(). Variance modelling results 
from modelGeneVar() tend to understate the actual biological variation, 
especially in highly heterogeneous datasets where secondary factors of 
variation inflate the fitted values of the trend. Fewer PCs are subsequently 
retained because T is artificially lowered, as evidenced by denoisePCA() 
returning the lower limit of 5 PCs for the PBMC dataset.

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

####

#--- loading ---#
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

#--- gene-annotation ---#
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

#--- cell-detection ---#
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

#--- quality-control ---#
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

#--- variance-modelling ---#
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

library(scran)
set.seed(111001001)
denoised.pbmc <- denoisePCA(sce.pbmc, technical=dec.pbmc, subset.row=top.pbmc)
ncol(reducedDim(denoised.pbmc))

set.seed(001001001)
denoised.zeisel <- denoisePCA(sce.zeisel, technical=dec.zeisel, 
    subset.row=top.zeisel)
ncol(reducedDim(denoised.zeisel))

dec.pbmc2 <- modelGeneVar(sce.pbmc)
denoised.pbmc2 <- denoisePCA(sce.pbmc, technical=dec.pbmc2, subset.row=top.pbmc)
ncol(reducedDim(denoised.pbmc2))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I have already applied quality 
control, normalization and computed variance. While performing principal 
component analysis (PCA) to reduce the dimensions of the data, is their another 
strategy to choose the number of principal components (PCs), other than the 
elbow and technical noise method?

completion:
Yet another method to choose d uses information about the number of 
subpopulations in the data. Consider a situation where each subpopulation 
differs from the others along a different axis in the high-dimensional space 
(e.g., because it is defined by a unique set of marker genes). This suggests 
that we should set d to the number of unique subpopulations minus 1, which 
guarantees separation of all subpopulations while retaining as few dimensions 
(and noise) as possible. We can use this reasoning to loosely motivate an a 
priori choice for d - for example, if we expect around 10 different cell types 
in our population, we would set d≈10.
In practice, the number of subpopulations is usually not known in advance. 
Rather, we use a heuristic approach that uses the number of clusters as a proxy 
for the number of subpopulations. We perform clustering (graph-based by default) 
on the first d∗ PCs and only consider the values of d∗ that yield no more than 
d∗ + 1 clusters. If we detect more clusters with fewer dimensions, we consider 
this to represent overclustering rather than distinct subpopulations, assuming 
that multiple subpopulations should not be distinguishable on the same axes. We 
test a range of d∗ and set d to the value that maximizes the number of clusters 
while satisfying the above condition. This attempts to capture as many distinct 
(putative) subpopulations as possible by retaining biological signal in later 
PCs, up until the point that the additional noise reduces resolution.
"Figure: Number of clusters detected in the Zeisel brain dataset as a function 
of the number of PCs. The red unbroken line represents the theoretical upper 
constraint on the number of clusters, while the grey dashed line is the number 
of PCs suggested by getClusteredPCs()."
We subset the PC matrix by column to retain the first d PCs and assign the 
subsetted matrix back into our SingleCellExperiment object. Downstream 
applications that use the "PCA.clust" results in sce.zeisel will subsequently 
operate on the chosen PCs only.
This strategy is pragmatic as it directly addresses the role of the 
bias-variance trade-off in downstream analyses, specifically clustering. There 
is no need to preserve biological signal beyond what is distinguishable in 
later steps. However, it involves strong assumptions about the nature of the 
biological differences between subpopulations - and indeed, discrete 
subpopulations may not even exist in studies of continuous processes like 
differentiation. It also requires repeated applications of the clustering 
procedure on increasing number of PCs, which may be computational expensive.

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

####

library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

set.seed(100) 
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

pcs <- reducedDim(sce.zeisel)
choices <- getClusteredPCs(pcs)
val <- metadata(choices)$chosen

plot(choices$n.pcs, choices$n.clusters,
    xlab="Number of PCs", ylab="Number of clusters")
abline(a=1, b=1, col="red")
abline(v=val, col="grey80", lty=2)

reducedDim(sce.zeisel, "PCA.clust") <- pcs[,1:val]

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I have already applied quality 
control, normalization and computed variance. While performing principal 
component analysis (PCA) to reduce the dimensions of the data, is their another 
strategy to choose the number of principal components (PCs), other than the 
elbow, technical noise and population structure method?

completion:
We consider the observed (log-)expression matrix to be the sum of (i) a 
low-rank matrix containing the true biological signal for each cell and (ii) a 
random matrix representing the technical noise in the data. Under this 
interpretation, we can use random matrix theory to guide the choice of the 
number of PCs based on the properties of the noise matrix.
The Marchenko-Pastur (MP) distribution defines an upper bound on the singular 
values of a matrix with random i.i.d. entries. Thus, all PCs associated with 
larger singular values are likely to contain real biological structure - or at 
least, signal beyond that expected by noise - and should be retained. We can 
implement this scheme using the chooseMarchenkoPastur() function from the 
PCAtools package, given the dimensionality of the matrix used for the PCA 
(noting that we only used the HVG subset); the variance explained by each PC 
(not the percentage); and the variance of the noise matrix derived from our 
previous variance decomposition results.
We can then subset the PC coordinate matrix by the first mp.choice columns. It 
is best to treat this as a guideline only; PCs below the MP limit are not 
necessarily uninteresting, especially in noisy datasets where the higher noise 
drives a more aggressive choice of d. Conversely, many PCs above the limit may 
not be relevant if they are driven by uninteresting biological processes like 
transcriptional bursting, cell cycle or metabolic variation. Morever, the use 
of the MP distribution is not entirely justified here as the noise distribution 
differs by abundance for each gene and by sequencing depth for each cell.
In a similar vein, Horn’s parallel analysis is commonly used to pick the number 
of PCs to retain in factor analysis. This involves randomizing the input matrix, 
repeating the PCA and creating a scree plot of the PCs of the randomized matrix.
The desired number of PCs is then chosen based on the intersection of the 
randomized scree plot with that of the original matrix (Figure). Here, the 
reasoning is that PCs are unlikely to be interesting if they explain less 
variance than that of the corresponding PC of a random matrix. Note that this 
differs from the MP approach as we are not using the upper bound of randomized 
singular values to threshold the original PCs. "Figure: Percentage of variance 
explained by each PC in the original matrix (black) and the PCs in the 
randomized matrix (grey) across several randomization iterations. The red line 
marks the chosen number of PCs."
The parallelPCA() function helpfully emits the PC coordinates in 
horn$original$rotated, which we can subset by horn$n and add to the 
reducedDims() of our SingleCellExperiment. Parallel analysis is reasonably 
intuitive (as random matrix methods go) and avoids any i.i.d. assumption across 
genes. However, its obvious disadvantage is the not-insignificant computational 
cost of randomizing and repeating the PCA. One can also debate whether the 
scree plot of the randomized matrix is even comparable to that of the original, 
given that the former includes biological variation and thus cannot be 
interpreted as purely technical noise. This manifests in Figure as a 
consistently higher curve for the randomized matrix due to the redistribution 
of biological variation to the later PCs.
Another approach is based on optimizing the reconstruction error of the 
low-rank representation (Gavish and Donoho 2014). Recall that PCA produces both 
the matrix of per-cell coordinates and a rotation matrix of per-gene loadings, 
the product of which recovers the original log-expression matrix. If we subset 
these two matrices to the first d dimensions, the product of the resulting 
submatrices serves as an approximation of the original matrix. Under certain 
conditions, the difference between this approximation and the true low-rank 
signal (i.e., sans the noise matrix) has a defined mininum at a certain number 
of dimensions. This minimum can be defined using the chooseGavishDonoho() 
function from PCAtools as shown.
The Gavish-Donoho method is appealing as, unlike the other approaches for 
choosing d, the concept of the optimum is rigorously defined. By minimizing the 
reconstruction error, we can most accurately represent the true biological 
variation in terms of the distances between cells in PC space. However, there 
remains some room for difference between “optimal” and “useful”; for example, 
noisy datasets may find themselves with very low d as including more PCs will 
only ever increase reconstruction error, regardless of whether they contain 
relevant biological variation. This approach is also dependent on some strong 
i.i.d. assumptions about the noise matrix.
Note: Zeisel et al. (2015) brain dataset used for demo.

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

####

library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

# Generating more PCs for demonstration purposes:
set.seed(10100101)
sce.zeisel2 <- fixedPCA(sce.zeisel, subset.row=top.zeisel, rank=200)

library(PCAtools)
library(RMTstat)
# Actual variance explained is also provided in the attributes:
mp.choice <- chooseMarchenkoPastur(
    .dim=c(length(top.zeisel), ncol(sce.zeisel2)),
    var.explained=attr(reducedDim(sce.zeisel2), "varExplained"),
    noise=median(dec.zeisel[top.zeisel,"tech"]))

mp.choice

set.seed(100010)
horn <- parallelPCA(logcounts(sce.zeisel)[top.zeisel,],
    BSPARAM=BiocSingular::IrlbaParam(), niters=10)
horn$n

plot(horn$original$variance, type="b", log="y", pch=16)
permuted <- horn$permuted
for (i in seq_len(ncol(permuted))) {
    points(permuted[,i], col="grey80", pch=16)
    lines(permuted[,i], col="grey80", pch=16)
}
abline(v=horn$n, col="red")

gv.choice <- chooseGavishDonoho(
    .dim=c(length(top.zeisel), ncol(sce.zeisel2)),
    var.explained=attr(reducedDim(sce.zeisel2), "varExplained"),
    noise=median(dec.zeisel[top.zeisel,"tech"]))

gv.choice

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 2, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I have already applied quality 
control, normalization and computed variance. To reduce the dimensions of the 
data, how can I perform count-based dimensionality reduction?

completion:
For count matrices, correspondence analysis (CA) is a natural approach to 
dimensionality reduction. In this procedure, we compute an expected value for 
each entry in the matrix based on the per-gene abundance and size factors. Each 
count is converted into a standardized residual in a manner analogous to the 
calculation of the statistic in Pearson’s chi-squared tests, i.e., subtraction 
of the expected value and division by its square root. An SVD is then applied 
on this matrix of residuals to obtain the necessary low-dimensional coordinates 
for each cell. To demonstrate, we use the corral package to compute CA factors 
for the Zeisel et al. (2015) brain dataset.
The major advantage of CA is that it avoids difficulties with the mean-variance 
relationship upon transformation. If two cells have the same expression profile 
but differences in their total counts, CA will return the same expected 
location for both cells; this avoids artifacts observed in PCA on 
log-transformed counts (Figure). However, CA is more sensitive to overdispersion 
in the random noise due to the nature of its standardization. This may cause 
some problems in some datasets where the CA factors may be driven by a few 
genes with random expression rather than the underlying biological structure. 
"Figure: Dimensionality reduction results of all pool-and-split libraries in 
the SORT-seq CellBench data, computed by a PCA on the log-normalized expression 
values (left) or using the corral package (right). Each point represents a 
library and is colored by the mixing ratio used to construct it."

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

####

library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)

library(corral)
sce.corral <- corral_sce(sce.zeisel, subset_row=top.zeisel,
   col.w=sizeFactors(sce.zeisel))
dim(reducedDim(sce.corral, "corral"))

# TODO: move to scRNAseq. The rm(env) avoids problems with knitr caching inside
# rebook's use of callr::r() during compilation.
library(BiocFileCache)
bfc <- BiocFileCache(ask=FALSE)
qcdata <- bfcrpath(bfc, "https://github.com/LuyiTian/CellBench_data/blob/master/data/mRNAmix_qc.RData?raw=true")

env <- new.env()
load(qcdata, envir=env)
sce.8qc <- env$sce8_qc
rm(env)

sce.8qc$mix <- factor(sce.8qc$mix)
sce.8qc

# Choosing some HVGs for PCA:
sce.8qc <- logNormCounts(sce.8qc)
dec.8qc <- modelGeneVar(sce.8qc)
hvgs.8qc <- getTopHVGs(dec.8qc, n=1000)
sce.8qc <- fixedPCA(sce.8qc, subset.row=hvgs.8qc)

# By comparison, corral operates on the raw counts:
sce.8qc <- corral_sce(sce.8qc, subset_row=hvgs.8qc, col.w=sizeFactors(sce.8qc))

library(scater)
gridExtra::grid.arrange(
    plotPCA(sce.8qc, colour_by="mix") + ggtitle("PCA"),
    plotReducedDim(sce.8qc, "corral", colour_by="mix") + ggtitle("corral"),
    ncol=2
)

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I have already applied quality 
control, normalization and computed variance. In order to reduce the dimensions 
of the data, is their a faster method?

completion:
Conventional t-SNE algorithms scale poorly with the number of cells. Fast 
interpolation-based t-SNE (FIt-SNE) (Linderman et al. 2019) is an alternative 
algorithm that reduces the computational complexity of the calculations from  
NlogN to ∼ 2pN. This is achieved by using interpolation nodes in the 
high-dimensional space; the bulk of the calculations are performed on the nodes 
and the embedding of individual cells around each node is determined by 
interpolation. To use this method, we can simply set use_fitsne=TRUE when 
calling runTSNE() with scater - this calls the snifter package, which in turn 
wraps the Python library openTSNE using basilisk. As Figure1 shows, the 
embeddings produced by this method are qualitatively similar to those produced 
by other algorithms, supported by some theoretical results from Linderman et al. 
(2019) showing that any difference from conventional t-SNE implementations is 
low and bounded. "Figure1: FI-tSNE embedding and Barnes-Hut t-SNE embeddings 
for the Zeisel et al. (2015) brain dataset."
By using snifter directly, we can also take advantage of openTSNE’s ability to 
project new points into an existing embedding. In this process, the existing 
points remain static while new points are inserted based on their affinities 
with each other and the points in the existing embedding. For example, cells 
are generally projected near to cells of a similar type in Figure2. This may be 
useful as an exploratory step when combining datasets, though the projection 
may not be sensible for cell types that are not present in the existing 
embedding. "Figure2: t-SNE embedding created with snifter, using 80% of the 
cells in the Zeisel brain data. The remaining 20% of the cells were projected 
into this pre-existing embedding."

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

####

library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)
set.seed(100) 
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

set.seed(9000)
sce.zeisel <- runTSNE(sce.zeisel)
sce.zeisel <- runTSNE(sce.zeisel, use_fitsne = TRUE, name="FIt-SNE")

gridExtra::grid.arrange(
    plotReducedDim(sce.zeisel, "TSNE", colour_by="level1class") + ggtitle("t-SNE"),
    plotReducedDim(sce.zeisel, "FIt-SNE", colour_by="level1class") + ggtitle("FIt-SNE"),
    ncol=2
)

set.seed(1000)
ind_test <- as.logical(rbinom(ncol(sce.zeisel), 1, 0.2))
ind_train <- !ind_test

library(snifter)
olddata <- reducedDim(sce.zeisel[, ind_train], "PCA")
embedding <- fitsne(olddata)
newdata <- reducedDim(sce.zeisel[, ind_test], "PCA")
projected <- project(embedding, new = newdata, old = olddata)

all <- rbind(embedding, projected)
label <- c(sce.zeisel$level1class[ind_train], sce.zeisel$level1class[ind_test])

ggplot() +
    aes(all[, 1], all[, 2], col = factor(label), shape = ind_test) +
    labs(x = "t-SNE 1", y = "t-SNE 2") +
    geom_point(alpha = 0.5) +
    scale_colour_brewer(palette = "Set2", name="level1class") +
    theme_bw() +
    scale_shape_manual(values = c(8, 19), name = "Set", labels = c("Training", "Test"))

```

--- TASK END

### Task `r i`,`r i<-i+1` taskType: data wrangling | visualization, numData: 1, seeData: no, openEnded: no, tags: merge

prompt:
I have a single cell RNA-seq dataset to which I have already applied quality 
control, normalization and computed variance. While using t-SNE algorithms to 
reduce the dimensions of the data, is there a way one can preserve the density 
of t-SNE?

completion:
One downside of t−SNE and UMAP is that they preserve the neighbourhood 
structure of the data while neglecting the local density of the data. This can 
result in seemingly compact clusters on a t-SNE or UMAP plot that correspond to 
very heterogeneous groups in the original data. The dens-SNE and densMAP 
algorithms mitigate this effect by incorporating information about the average 
distance to the nearest neighbours when creating the embedding. We demonstrate 
below by applying these approaches on the PCs of the Zeisel et al. (2015) brain 
dataset using the densviz wrapper package.
These methods provide more information about transcriptional heterogeneity 
within clusters (Figure), with the astrocyte cluster being less compact in the 
density-preserving versions. This excessive compactness can imply a lower level 
of within-population heterogeneity. "Figure: t-SNE, UMAP, dens-SNE and densMAP 
embeddings for the Zeisel brain data."

```{r,eval=FALSE}

#--- loading ---#
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

#--- gene-annotation ---#
library(org.Mm.eg.db)
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")

#--- quality-control ---#
stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

#--- normalization ---#
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)

#--- variance-modelling ---#
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop=0.1)

####

library(scran)
top.zeisel <- getTopHVGs(dec.zeisel, n=2000)
set.seed(100) 
sce.zeisel <- fixedPCA(sce.zeisel, subset.row=top.zeisel)

set.seed(9000)
sce.zeisel <- runTSNE(sce.zeisel)
sce.zeisel <- runTSNE(sce.zeisel, use_fitsne = TRUE, name="FIt-SNE")

library(densvis)
dt <- densne(reducedDim(sce.zeisel, "PCA"), dens_frac = 0.4, dens_lambda = 0.2)
reducedDim(sce.zeisel, "dens-SNE") <- dt
dm <- densmap(reducedDim(sce.zeisel, "PCA"), dens_frac = 0.4, dens_lambda = 0.2)
reducedDim(sce.zeisel, "densMAP") <- dm
sce.zeisel <- runUMAP(sce.zeisel) # for comparison

gridExtra::grid.arrange(
    plotReducedDim(sce.zeisel, "TSNE", colour_by="level1class") + ggtitle("t-SNE"),
    plotReducedDim(sce.zeisel, "dens-SNE", colour_by="level1class") + ggtitle("dens-SNE"),
    plotReducedDim(sce.zeisel, "UMAP", colour_by="level1class") + ggtitle("UMAP"),
    plotReducedDim(sce.zeisel, "densMAP", colour_by="level1class") + ggtitle("densMAP"),
    ncol=2
)

```

--- TASK END